### 2023-11-28 paper/main.tex
bibliography

% Copyright 2023 Andy Casey (Monash) and friends
% TeX magic by David Hogg (NYU)

\documentclass[modern]{aastex631}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}

\renewcommand{\twocolumngrid}{}
\addtolength{\topmargin}{-0.35in}
\addtolength{\textheight}{0.6in}
\setlength{\parindent}{3.5ex}
\renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}

% figure setup
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[framemethod=tikz]{mdframed}
\usetikzlibrary{shadows}
\definecolor{captiongray}{HTML}{555555}
\mdfsetup{%
innertopmargin=2ex,
innerbottommargin=1.8ex,
linecolor=captiongray,
linewidth=0.5pt,
roundcorner=1pt,
shadow=false,
}
\newlength{\figurewidth}
\setlength{\figurewidth}{0.75\textwidth}

% text macros
\shorttitle{Stellar continuum modelling}
\shortauthors{Casey}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}

% math macros
\newcommand{\unit}[1]{\mathrm{#1}}
\newcommand{\mps}{\unit{m\,s^{-1}}}
\newcommand{\kmps}{\unit{km\,s^{-1}}}

% notes

\sloppy\sloppypar\raggedbottom\frenchspacing
\begin{document}

\title{\Huge Stellar continuum modelling}

\author[0000-0003-0174-0564]{Andrew R. Casey}
\affiliation{School of Physics \& Astronomy, Monash University}
\affiliation{Centre of Excellence for Astrophysics in Three Dimensions (ASTRO-3D)}

%\author[0000-0003-2866-9403]{David W. Hogg}
%\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
%\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
%\affiliation{Flatiron Institute, a division of the Simons Foundation}

\begin{abstract}\noindent
Continuum normalization is a common step when analyzing stellar spectra.
- various bespoke methods
- best fit simultaneously with the stellar spectrum
- that requires a model of the stellar spectrum, which can be expensive to compute
- 
\end{abstract}

\keywords{Some --- keywords --- here}

\section*{}\clearpage
\section{Introduction}\label{sec:intro}

Code is available at \url{https://github.com/andycasey/continuum} and is registered in the Python Package Index as \texttt{stellar-continuum}.

\paragraph{Software}
\texttt{numpy} \citep{numpy} ---
\texttt{matplotlib} \citep{matplotlib} ---
\texttt{scipy} \citep{scipy}.

\paragraph{Acknowledgements}
It is a pleasure to thank
% All these people are possible co-authors
    Adam Wheeler (Ohio State University),
    David W. Hogg (New York University),
    Megan Bedell (Flatiron Institute),
.
% include bibliography
\bibliographystyle{aasjournal}
\bibliography{bibliography}

\end{document}

### 2023-11-28 setup.py

@@ -1,0 +1,9 @@
+from setuptools import setup, find_packages
+
+setup(
+    name="stellar-continuum",
+    version="0.0.1",
+    package_dir={"": "python"},
+    packages=find_packages(where="python"),
+    install_requires=["numpy", "scipy"]
+)

### 2023-11-29 paper/main.tex
some methods

@@
 
+\newcommand{\project}[1]{\textit{#1}}
+\renewcommand{\vec}[1]{\mathbf{#1}}
+\newcommand{\vectheta}{\boldsymbol{\theta}}
+\newcommand{\vecpsi}{\boldsymbol{\psi}}
+\newcommand{\vecW}{\mathbf{W}}
+\newcommand{\vecH}{\mathbf{H}}
+\newcommand{\vecV}{\mathbf{V}}
+
 % math macros
@@
 \begin{abstract}\noindent
-Continuum normalization is a common step when analyzing stellar spectra.
-- various bespoke methods
-- best fit simultaneously with the stellar spectrum
-- that requires a model of the stellar spectrum, which can be expensive to compute
-- 
+Continuum normalization is often a necessary step when analyzing stellar spectra.
+The best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but this can be computationally expensive.
+%Methods that estimate the continuum before stellar parameters tend to be biased as a function of stellar parameters or signal-to-noise ratio.
+Here we describe a set of constrained linear models that simultaneously fit stellar absorption, and the joint continuum-instrument response.
+The linear basis ensures that inference is stable and fast.
+We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of all evolutionary stages: from M-dwarfs to white dwarfs.
+%The stellar absorption model is built from a non-negative matrix factorization of a theoretical continuum-normalized spectra, which ensures it cannot predict spectra more than 1.
+%The constraints restrict the stellar absorption to 
+We show that we obtain unbiased estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
 \end{abstract}
@@
 \section{Introduction}\label{sec:intro}
+
+\section{Methods}\label{sec:methods}
+
+We will assume a forward model for the data that includes two components: a component to represent continuum-normalized absorption (e.g., line absorption), and a component to represent the smooth continuum.\footnote{Most stellar spectra are not flux-calibrated; the continuum and instrument response enter multiplicatively, but throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.}
+We will assume both components to be linear models.
+This is not a strict requirement, but keeping linearity ensures that  inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation.
+
+% other assumptions
+% - NMF is sufficient for representing continuum-normalized spectra
+% - that the theoretical models are OK enough.
+there are ways they can be wrong (eg hydrogen lines) and still work in practice, or there are ways they can be wrong entirely (eg missing molecular band) and not work at all
+% - that sines and cosines are sufficiently flexible for modelling the large scale continuum
+The data are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$).
+The forward model for these data can be expressed as the element-wise multiplication of the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
+\begin{align}
+    y_i &= f(\lambda_i;\vecpsi)\cdot{}g(\lambda_i;\vectheta) + \mbox{noise}
+\end{align}
+where the components $f(...)$ and $g(...)$ are described below.
+
+% NMF first
+The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a grid of $N$ theoretical spectra each with $D$  continuum-normalized fluxes, which we denote as $\vec{M}$.
+If the data always have the same wavelength sampling and a constant line spread function, then a good choice is to use a grid where $D = P$.
+Other situations are described later in this section.
+With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecV$ to be
+\begin{align}
+    \vec{V} = 1 - \vec{M}
+\end{align}
+such that $\vec{V} \in \left[0, 1\right)$ and is zero when no line absorption exists.
+This is a `trick' in order to construct a highly constrained and sparse approximation to the matrix $\vecV$ using non-negative matrix factorization (NMF) such that
+\begin{align}
+    \vec{V} \approx \vec{W}\vec{H}
+\end{align}
+where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
+%We could construct an approximation directly to $\vec{M}$, but since elements in $\vec{M}$ are non-zero whenever there is line absorption, this would lead to denser matrices $\vecV$ and $\vecH$. 
+For this approximation we must chose the number of components to use $C$, which is significantly smaller than both $N$ and $P$.
+Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times P$ matrix of $C$ corresponding eigenspectra.
+Figure~\ref{fig:nmf} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$.
+
+% Figure NMF
+
+With our approximating matrix $\vec{H}$ we can now define the line absorption function $f(\lambda_i;\vecpsi)$ 
+\begin{align}
+    f(\lambda_i;\vecpsi) = 1 - \vecpsi\vecH \label{eq:f}
+\end{align}
+where $\vecpsi \in [0, \infty)$ is a row vector of $C$ elements.
+$\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ amplitudes needed to reconstruct the stellar absorption from the $C$ eigenspectra in $\vecH$.
+Note that because $\vecpsi$ and $\vecH$ are both restricted to have non-negative elements, Equation~\ref{eq:f} shows that the maximum value that can be predicted by $f(...)$ is 1.
+This restricts the flexibility of $f(...)$ to only be able to model continuum-normalized flux values, leaving $g(...)$ to represent the joint continuum-instrument response.
+While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), such solutions are not favoured in practice.
+
+% put this in discussion
+%As we discuss in Section~\ref{sec:discussion}, this is not true of all linear models: it is a design choice that leads to this strict constraint.
+Other linear models (e.g., PCA) allow for summation of large positive and negative amplitudes, leading to  of positive and negative eigenspectra,
+
+\section{Results}\label{sec:results}
+
+\section{Conclusions}\label{sec:conclusions}
 
@@
 \bibliographystyle{aasjournal}
-\bibliography{bibliography}
+%\bibliography{bibliography}
 

### 2023-11-29 paper/main.tex
meh

@@
 The best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but this can be computationally expensive.
-%Methods that estimate the continuum before stellar parameters tend to be biased as a function of stellar parameters or signal-to-noise ratio.
-Here we describe a set of constrained linear models that simultaneously fit stellar absorption, and the joint continuum-instrument response.
-The linear basis ensures that inference is stable and fast.
+We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption in theoretical (normalized) spectra, and a sines-and-cosines basis for the joint continuum-instrument response.
+The non-negative matrix factorization ensures that eigenspectra are strictly additive, and restricts the predicted normalized stellar flux to be at most 1, with the joint continuum-instrument response component to model the rest.
+The linearity of both components ensures that inference is stable and fast.
 We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of all evolutionary stages: from M-dwarfs to white dwarfs.
+We find good fits to observed spectra, even in situations where the theoretical grid used to approximate line absorption is provably incorrect for some features.
 %The stellar absorption model is built from a non-negative matrix factorization of a theoretical continuum-normalized spectra, which ensures it cannot predict spectra more than 1.
@@
 
+Continuum normalisation is a necessary step
+
+Often needed to be done before stellar parameter determination, as a step towards stellar parameter determination
+
+For measuring equivalent widths, or for full-spectrum fitting, or even for performing radial velocity determination
+
+Many methods. Some done by eye.
+
+Fitting the continuum well requires you to know where the line absorption is. But knowing where the line absorption is requires you to know the stellar parameters.
+
+Best is to fit with the stellar parameters, but is intractable.
+
+For some stars there is NO continuum (e.g., M-dwarfs).
+
+cite cases of NMF use in astronomy (not that many)
+
 \section{Methods}\label{sec:methods}
@@
 We will assume a forward model for the data that includes two components: a component to represent continuum-normalized absorption (e.g., line absorption), and a component to represent the smooth continuum.\footnote{Most stellar spectra are not flux-calibrated; the continuum and instrument response enter multiplicatively, but throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.}
+
 We will assume both components to be linear models.
-This is not a strict requirement, but keeping linearity ensures that  inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation.
+This is not a strict requirement, but keeping linearity ensures that  inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation.\\
 
@@
 % - that sines and cosines are sufficiently flexible for modelling the large scale continuum
-The data are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$).
+The problem set up is where the data are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$).
 The forward model for these data can be expressed as the element-wise multiplication of the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
@@
 \end{align}
-where the components $f(...)$ and $g(...)$ are described below.
+where the components $f(...)$ and $g(...)$ are described below.\\
 
@@
 where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
-%We could construct an approximation directly to $\vec{M}$, but since elements in $\vec{M}$ are non-zero whenever there is line absorption, this would lead to denser matrices $\vecV$ and $\vecH$. 
+We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0, but there are \emph{many} elements in $\vec{V}$ that are 0 (e.g., when there is no line absorption).
 For this approximation we must chose the number of components to use $C$, which is significantly smaller than both $N$ and $P$.
@@
 Figure~\ref{fig:nmf} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$.
+\\
 
@@
 
-With our approximating matrix $\vec{H}$ we can now define the line absorption function $f(\lambda_i;\vecpsi)$ 
+\noindent{}With the matrix $\vec{H}$ we can now define the line absorption function $f(\lambda_i;\vecpsi)$ 
 \begin{align}
@@
 This restricts the flexibility of $f(...)$ to only be able to model continuum-normalized flux values, leaving $g(...)$ to represent the joint continuum-instrument response.
-While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), such solutions are not favoured in practice.
+While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this could be readily compensated by a negative continuum model $g(...) < 0$ and such solutions are not favoured in practice.\\
+
+There are many suitable choices for the continuum-instrument response model $g(\lambda_i;\vectheta)$.
+Here we chose a sine-and-cosine basis because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
+The component $g(\lambda_i;\vectheta)$ is expressed as compactly as
+\begin{align}
+    g(\lambda_i;\vectheta) = \vec{A}(\lambda_i)\vectheta
+\end{align}
+where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column
+\begin{align}
+    \vec{A}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
+                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
+\end{align}
+
+The design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
+
+%In later sections we describe applications of our method to real data.
+For now we will describe a few options that we found to work reasonably well across all settings, which we have since established as default behaviour in the accompanying software implementation.
+When faced with the choice of how many spectra to include when performing non-negative matrix factorization, we found good results by including everything that could fit into memory.
+Using limited precision floats (e.g., float-8) helped.
+Initialising with non-negative double singular value decomposition (with zeros filled with small random values) seemed to work very well.
+Multiplicative updates.
 
@@
 
+We experimented with choices of initialisation and inference.
+Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ seemed to work well in many scenarios.
+Since both model components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously.
+
+\subsection{Application to APOGEE spectra}
+
+- choice of grid
+- regions for continuum modelling
+
+The continuum-instrument response of the APOGEE detector was modelled using a 3 degree sines-and-cosines with a length-scale $L = 1300$, but allowing for different coefficients per chip.
+
+- choices of iterations, initialisation, etc, 
+- eigenspectra examples already shown
+
+\subsection{Application to BOSS spectra}
+
+- bosz grid convolved to what resolution + sampled at X
+
 \section{Results}\label{sec:results}
+
+- figure showing the
 

### 2023-12-06 paper/main.tex
some methods

@@
 \newcommand{\vecV}{\mathbf{V}}
+\newcommand{\hadamard}{\odot}
 
@@
 \newcommand{\kmps}{\unit{km\,s^{-1}}}
+\newcommand{\transpose}{^\top}
 
 % notes
+\definecolor{tab:blue}{HTML}{1170aa}
+\newcommand{\todo}[1]{\textcolor{tab:blue}{#1}}
 
@@
 Continuum normalization is often a necessary step when analyzing stellar spectra.
-The best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but this can be computationally expensive.
+The best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but computing stellar spectra can be expensive.
 We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption in theoretical (normalized) spectra, and a sines-and-cosines basis for the joint continuum-instrument response.
-The non-negative matrix factorization ensures that eigenspectra are strictly additive, and restricts the predicted normalized stellar flux to be at most 1, with the joint continuum-instrument response component to model the rest.
+The non-negative matrix factorization ensures that eigenspectra are strictly additive, thereby restricting the predicted normalized stellar flux to be at most 1, with the joint continuum-instrument response component to describe the rest.
 The linearity of both components ensures that inference is stable and fast.
@@
 We show that we obtain unbiased estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
+This approach largely eliminates subjectivity in stellar continuum normalization.
 \end{abstract}
@@
 
-We will assume a forward model for the data that includes two components: a component to represent continuum-normalized absorption (e.g., line absorption), and a component to represent the smooth continuum.\footnote{Most stellar spectra are not flux-calibrated; the continuum and instrument response enter multiplicatively, but throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.}
-
-We will assume both components to be linear models.
+We will assume a forward model for the data that includes two components: a component to represent continuum-normalized absorption (e.g., line absorption), and a component to represent the smooth continuum.\footnote{Most stellar spectra are not flux-calibrated; the continuum and instrument response enter multiplicatively and require some work to disentangle the two.
+Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.} We will assume both components to be linear models.
 This is not a strict requirement, but keeping linearity ensures that  inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation.\\
@@
 % - that sines and cosines are sufficiently flexible for modelling the large scale continuum
-The problem set up is where the data are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$).
+The problem can be described as having data that are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$).
 The forward model for these data can be expressed as the element-wise multiplication of the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
 \begin{align}
-    y_i &= f(\lambda_i;\vecpsi)\cdot{}g(\lambda_i;\vectheta) + \mbox{noise}
+    y_i &= f(\lambda_i;\vecpsi)\hadamard{}g(\lambda_i;\vectheta) + \mbox{noise}
 \end{align}
-where the components $f(...)$ and $g(...)$ are described below.\\
+where $\hadamard$ represents the element-wise (Hadamard) product and the components $f(...)$ and $g(...)$ are described below.\\
 
@@
 The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a grid of $N$ theoretical spectra each with $D$  continuum-normalized fluxes, which we denote as $\vec{M}$.
-If the data always have the same wavelength sampling and a constant line spread function, then a good choice is to use a grid where $D = P$.
-Other situations are described later in this section.
+In the applications that we present, the data are assumed to have the same wavelength sampling and a consistent line spread function, such that we use a grid where $D = P$.
+Other situations are permitted, but at inference time there is a need to interpolate the line absorption to the $P$ observed pixels.
 With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecV$ to be
@@
 such that $\vec{V} \in \left[0, 1\right)$ and is zero when no line absorption exists.
-This is a `trick' in order to construct a highly constrained and sparse approximation to the matrix $\vecV$ using non-negative matrix factorization (NMF) such that
+This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecV$ using non-negative matrix factorization (NMF) such that
 \begin{align}
-    \vec{V} \approx \vec{W}\vec{H}
+    \vec{V} \approx \vec{W}\vec{H} \label{eq:nmf}
 \end{align}
 where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
-We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0, but there are \emph{many} elements in $\vec{V}$ that are 0 (e.g., when there is no line absorption).
-For this approximation we must chose the number of components to use $C$, which is significantly smaller than both $N$ and $P$.
-Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times P$ matrix of $C$ corresponding eigenspectra.
-Figure~\ref{fig:nmf} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$.
+We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vec{V}$ that are 0 (i.e., no line absorption).
+For this approximation we must select the number of components $C$ to use, which is significantly smaller than both the number of theoretical spectra $N$ and the number of pixels $D$ per theoretical spectrum.
+Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding eigenspectra.
+Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$.
 \\
 
-% Figure NMF
+\begin{figure*}
+    \caption{A schematic illustrating the non-negative matrix factorization procedure, with some example infrared eigenspectra. \label{fig:schematic}}
+\end{figure*}
 
@@
 This restricts the flexibility of $f(...)$ to only be able to model continuum-normalized flux values, leaving $g(...)$ to represent the joint continuum-instrument response.
-While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this could be readily compensated by a negative continuum model $g(...) < 0$ and such solutions are not favoured in practice.\\
+While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this would be readily compensated by a negative continuum model $g(...) < 0$ and in practice such local minima are unfavoured.\\
 
@@
 \end{align}
-where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column
+where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column are,
 \begin{align}
@@
 \end{align}
-
-The design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
+\noindent{}and the design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.\\
 
@@
 
+\subsection{Constructing the line absorption model}
+
+The line absorption model $f(\lambda_i;\vecpsi)$ can be easily constructed from a grid of theoretical continuum-normalized spectra.
+In later sub-sections we discuss the specific choices made for the \emph{APOGEE} and \emph{BOSS} applications, but for now we discuss some general principles and practices that we found successful.
+The objective function to minimize
+\begin{align}
+    \chi^2 = ||\vecV - \vecW\vecH||^2
+\end{align}
+\noindent{}is the Frobenius norm (the squared approximation error). This can be solved through coordinate descent, or through the multiplicative update rules developed by \citet{Lee:2000}:
+\begin{align}
+    \vecW_{ij}^{\mathrm{(new)}} \leftarrow \vecW_{ij} \frac{\left[\vecV\vecH\transpose\right]_{ij}}{\left[\vecW\vecH\vecH\transpose\right]_{ij}} \quad \phantom{.}\\
+    \vecH_{ij}^{\mathrm{(new)}} \leftarrow \vecH_{ij} \frac{\left[\vecW\transpose\vecV\right]_{ij}}{\left[\vecW\transpose\vecW\vecH\right]_{ij}} \quad .
+\end{align}
+
+We used the \texttt{scipy.decomposition.NMF} implementation \citep{scipy} with some minor adjustments to minimize the memory requirements (e.g., allowing for lower precision float types).
+We used no regularization on $\vecW$ or $\vecH$.
+We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition with zeros filled-in with the average of $\vecV$ (the default behavior in the \texttt{scipy} implementation).
+Initializing with random or small non-negative values produced comparable results but required many more iterations.
+We stopped the approximation of $\vecW$ and $\vecH$ after 1,000 iterations of multiplicative updates.
+This took minutes to hours to complete, depending on the size of the theoretical grid $\vecV$.\\
+
+There are no requirements on the number of grid dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on the spacing between grid points.
+The only implicit requirement is that the grid should approximately span the range of stars that you intent to apply the method.\footnote{This is more of a recommendation than a requirement.
+For example, in practice we found that a grid trained on OBA-type stars was also sufficient to model many white dwarf spectra.} For this reason, we chose to include as many spectra as our memory constraints would allow.
+If the number of spectra exceeded memory constraints, we would opt to skip every $n$th theoretical spectrum.\\
+
+%Note that the stellar parameters used to compute the theoretical spectra are ignored when solving for $\vecW$ and $\vecH$ in Equation~\ref{eq:nmf}: only the spectra are used.
+There are correlations between the stellar parameters and the amplitudes $\vecW$, but as we discuss in Section~\ref{sec:discussion}, we avoid trying to use the amplitudes $\vecW$ (or $\vecpsi$) to interpreting stellar parameters.
+
+%\subsection{Inference}
+
+The method can be readily applied to real data once the eigenspectra $\vecH$ have been computed.
 We experimented with choices of initialisation and inference.
-Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ seemed to work well in many scenarios.
-Since both model components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously.
+Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously.
+The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecpsi$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip).
+Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecpsi$) seemed to work well in many scenarios.
+\\
 
 \subsection{Application to APOGEE spectra}
+\label{sec:apogee}
 
@@
 \section{Results}\label{sec:results}
+
+\begin{figure*}
+    \caption{The median pixel $\chi^2$ value as a function of \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence stars observed with the BOSS spectrograph.
+\todo{Should expect to see increasing residuals due to TiO bands not captured by the model, and emission lines.}}
+\end{figure*}
 

### 2023-12-06 paper/main.tex
beer oclock

@@
 \newcommand{\hadamard}{\odot}
+\newcommand{\apogee}{\project{APOGEE}}
+\newcommand{\boss}{\project{BOSS}}
 
@@
 
-- choice of grid
-- regions for continuum modelling
+The \apogee\ spectrograph spans approximately 1.5 to 1.7\,microns and has a resolving power of about 22,500 \citep{wilson}.
+There are three chips, which we chose to model as separate continuum regions (each with their own continuum coefficients $\vecpsi$).
+We used a 3rd order sine-and-cosine basis with length scale $L = 1300$, corresponding to 7 continuum coefficients per chip.
 
-The continuum-instrument response of the APOGEE detector was modelled using a 3 degree sines-and-cosines with a length-scale $L = 1300$, but allowing for different coefficients per chip.
+We use the \todo{DR17 grids computed for FERRE, some in non-LTE}.
+The data are assumed to be resampled onto a rest-frame wavelength sampling that is uniform in log-spacing.
+We took the high-resolution theoretical grids and convolved them to a spectral resolution of 22,500 and sampled it to match the data.
+In total there are nearly \todo{1 M} spectra each with 8,575 pixels.
+This proved too large to fit in memory when computing the eigenspectra $\vecH$, forcing us to skip every 10th spectra.
 
-- choices of iterations, initialisation, etc, 
-- eigenspectra examples already shown
+We chose to use 32 eigenspectra to construct the line absorption model, some of which are shown in Figure~\ref{fig:schematic}.
+During training, the hydrogen lines separate out immediately before other absorption features are recognizable.
+The final eigenspectra include multiple hydro \todo{BEER OCLOCK}
 
@@
 
-Code is available at \url{https://github.com/andycasey/continuum} and is registered in the Python Package Index as \texttt{stellar-continuum}.
+%Code is available at \url{https://github.com/andycasey/continuum} and is registered in the Python Package Index as \texttt{stellar-continuum}.
 

### 2023-12-07 paper/main.tex
m

@@
 \newcommand{\vecH}{\mathbf{H}}
-\newcommand{\vecV}{\mathbf{V}}
+\newcommand{\vecX}{\mathbf{X}}
 \newcommand{\hadamard}{\odot}
@@
 Continuum normalization is often a necessary step when analyzing stellar spectra.
-The best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but computing stellar spectra can be expensive.
+Best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but computing stellar spectra can be expensive.
 We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption in theoretical (normalized) spectra, and a sines-and-cosines basis for the joint continuum-instrument response.
@@
 Other situations are permitted, but at inference time there is a need to interpolate the line absorption to the $P$ observed pixels.
-With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecV$ to be
+With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecX$ to be
 \begin{align}
-    \vec{V} = 1 - \vec{M}
+    \vecX = 1 - \vec{M}
 \end{align}
-such that $\vec{V} \in \left[0, 1\right)$ and is zero when no line absorption exists.
-This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecV$ using non-negative matrix factorization (NMF) such that
+such that $\vecX \in \left[0, 1\right)$ and is zero when no line absorption exists.
+This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecX$ using non-negative matrix factorization (NMF) such that
 \begin{align}
-    \vec{V} \approx \vec{W}\vec{H} \label{eq:nmf}
+    \vecX \approx \vec{W}\vec{H} \label{eq:nmf}
 \end{align}
 where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
-We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vec{V}$ that are 0 (i.e., no line absorption).
+We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX$ that are 0 (i.e., no line absorption).
 For this approximation we must select the number of components $C$ to use, which is significantly smaller than both the number of theoretical spectra $N$ and the number of pixels $D$ per theoretical spectrum.
@@
 \begin{align}
-    \chi^2 = ||\vecV - \vecW\vecH||^2
+    \chi^2 = ||\vecX - \vecW\vecH||^2
 \end{align}
@@
 \begin{align}
-    \vecW_{ij}^{\mathrm{(new)}} \leftarrow \vecW_{ij} \frac{\left[\vecV\vecH\transpose\right]_{ij}}{\left[\vecW\vecH\vecH\transpose\right]_{ij}} \quad \phantom{.}\\
-    \vecH_{ij}^{\mathrm{(new)}} \leftarrow \vecH_{ij} \frac{\left[\vecW\transpose\vecV\right]_{ij}}{\left[\vecW\transpose\vecW\vecH\right]_{ij}} \quad .
+    \vecW_{ij}^{\mathrm{(new)}} \leftarrow \vecW_{ij} \frac{\left[\vecX\vecH\transpose\right]_{ij}}{\left[\vecW\vecH\vecH\transpose\right]_{ij}} \quad \phantom{.}\\
+    \vecH_{ij}^{\mathrm{(new)}} \leftarrow \vecH_{ij} \frac{\left[\vecW\transpose\vecX\right]_{ij}}{\left[\vecW\transpose\vecW\vecH\right]_{ij}} \quad .
 \end{align}
@@
 We used no regularization on $\vecW$ or $\vecH$.
-We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition with zeros filled-in with the average of $\vecV$ (the default behavior in the \texttt{scipy} implementation).
+We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition with zeros filled-in with the average of $\vecX$ (the default behavior in the \texttt{scipy} implementation).
 Initializing with random or small non-negative values produced comparable results but required many more iterations.
 We stopped the approximation of $\vecW$ and $\vecH$ after 1,000 iterations of multiplicative updates.
-This took minutes to hours to complete, depending on the size of the theoretical grid $\vecV$.\\
+This took minutes to hours to complete, depending on the size of the theoretical grid $\vecX$.\\
 
@@
 
-\subsection{Application to APOGEE spectra}
+\subsection{APOGEE spectra}
 \label{sec:apogee}
@@
 There are three chips, which we chose to model as separate continuum regions (each with their own continuum coefficients $\vecpsi$).
-We used a 3rd order sine-and-cosine basis with length scale $L = 1300$, corresponding to 7 continuum coefficients per chip.
+We used a 3rd order sine-and-cosine basis with length scale $L = 1300$, corresponding to 7 continuum coefficients per chip.\\
 
@@
 In total there are nearly \todo{1 M} spectra each with 8,575 pixels.
-This proved too large to fit in memory when computing the eigenspectra $\vecH$, forcing us to skip every 10th spectra.
+This proved too large to fit in memory when computing the eigenspectra $\vecH$, forcing us to skip every 10th spectra.\\
 
@@
 During training, the hydrogen lines separate out immediately before other absorption features are recognizable.
-The final eigenspectra include multiple hydro \todo{BEER OCLOCK}
+After training is complete, there are multiple eigenspectra that only have hydrogen lines: some narrow, and some broad.
+Eigenspectra with only recognizable patterns of molecular features are also present.
+The remaining eigenspectra have line absorption at many different wavelengths, representing something akin to overall metallicity.
+\\
+
+With these eigenspectra and continuum settings, we simultaneously fit line absorption and the joint continuum-instrument response for all \apogee\ spectra in SDSS-IV DR17 \citep{dr17} and \apogee\ spectra acquired by the Milky Way Mapper (MWM) in SDSS-V \citep{kollmeier}.
+In total there are \todo{X} \apogee\ spectra fit by this process: \todo{X} combined spectra and \todo{X} individual visits.
+We fit combined (co-added) spectra first, and for individual visit spectra we analytically computed the continuum coefficients $\vectheta$ for that visit, conditioned on the $\vecpsi$ found from the combined fit.
+In practice this implies that the combined and visit spectra for the same star share the same amplitudes $\vecpsi$ for computing rectified flux, but may have different continuum coefficients per visit.
+Figure~\ref{fig:example-apogee-spectra} show example \apogee\ rectified spectra for different spectral types.
+
+\begin{figure}
+    \caption{Example rectified \apogee\ spectra (black) and best-fitting rectified line absorption (red) for different spectral types. \label{fig:example-apogee-spectra}}
+\end{figure}
 
 \subsection{Application to BOSS spectra}
+\label{sec:}
 
-- bosz grid convolved to what resolution + sampled at X
+SDSS-V is acquiring optical spectra for millions of Milky Way stars using the \boss\ 
+low-resolution ($\mathcal{R} \sim 4{,}000$) spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO).
+Although the highest priority targets observed by \boss\ tend to be Black Hole Mapper (BHM) targets, many Milky Way stars are included per field, either because they are specifically targeted as part of a Milky Way Mapper carton\footnote{`Cartons' are how targets are assigned in SDSS-V.}, or because they are standard stars.\\
+
+We took all \boss\ spectra where the source was allocated to a Milky Way Mapper carton, or spectra where the source was allocated to a stellar-like carton.
+\todo{Some description of stellar-like.} For these spectra we fit the line absorption and joint continuum-instrument response.\\
+
+We used the BOSZ spectral grid to construct eigenspectra for a line absorption model for \boss\ spectra.
+This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a sufficiently large range in stellar parameters.
+We convolved the spectra to the nominal mean resolving power of the \boss\ spectrograph ($\mathcal{R} \sim 4{,}000$) and resampled it to the same pixels that the \boss\ spectra are resampled to: a uniform-in-log wavelength sampling from approximately 360\,nm until \todo{1000\,nm}.
+We clipped any BOSZ flux values with emission lines exceeding normalized flux values of 1, as they would violate our NMF requirements.
+As we discuss in later sections, any emission lines are sufficiently rare in stellar spectra that they do not contribute significantly to the $\chi^2$ at inference time.\\
+
+We used all \todo{X} BOSZ spectra to construct a line absorption matrix $\vecX$ and find the approximating matrices $\vecW$ and $\vecH$.
+For fitting \boss\ spectra, we separated the continuum fits into two chunks to represent the blue and red chip of the detector.
+Most of the time the blue and red chips are correctly merged together to produce a seamless contiguous spectrum from \todo{300\,nm to 1000,\nm}, but we noticed there could some times be jumps in the spectrum due to an imperfect merge.
+For these reasons, we used two sections from \todo{3750\,\AA\ to 6000\,\AA and 6050\,\AA to 10000\,\AA}.
+We chose a \todo{3rd} order sines-and-cosines basis with a length scale $L = X$\,\AA.
+Following what we did with \apogee\ spectra, we fit the combined \boss\ spectrum first for a given star, and then analytically computed $\vectheta{}|\vecpsi$: the continuum coefficients $\vectheta$ for the given visit, conditioned on the amplitudes optimized from the combined spectrum $\vecpsi$.
 

### 2023-12-07 paper/main.tex
some stuff

@@
 
+\begin{figure}
+    \caption{Median observed rectified flux values as a function of S/N ratio showing that there is no bias as a function of S/N ratio.}
+\end{figure}
+
+\begin{figure}
+    \caption{Median model continuum-rectified flux values as a function of effective temperature, compared to the median observed rectified flux values as a function of Gaia colours, showing there is no bias as a function of spectral type?}
+\end{figure}
+
+\begin{figure}
+    \caption{Gaia H-R diagram with each bin coloured by the median $\vecpsi_i$ value, picking some $i$th eigenspectra that looks like metallicity. \label{fig:gaia_hrd_metallicity}}
+\end{figure}
+
 \begin{figure*}
@@
 
-%Code is available at \url{https://github.com/andycasey/continuum} and is registered in the Python Package Index as \texttt{stellar-continuum}.
+We solved continuum normalization.
+
+Code is available at \url{https://github.com/andycasey/continuum} and is registered in the Python Package Index as \texttt{stellar-continuum}.
 

### 2023-12-18 applications/gaia_rvs/20231218_gaia_rvs_nmf.py
add gaia rvs example

@@ -1,0 +1,44 @@
+import numpy as np
+from sklearn.decomposition import NMF
+
+import h5py as h5
+
+fp = h5.File("/Users/andycasey/Downloads/dr3-rvs-all.hdf5", "r")
+
+np.random.seed(1234)
+
+N = 100_000
+
+try:
+    snr
+except NameError:
+    snr = np.nanmean(fp["flux"][:] / fp["flux_error"][:], axis=1)
+    indices = np.sort(np.random.choice(np.where(snr >= 50)[0], N, replace=False))
+
+    flux = fp["flux"][indices]
+    source_id = fp["source_id"][indices]
+
+
+absorption = np.clip(1 - flux, 0, 1)
+absorption[~np.isfinite(absorption)] = 0
+
+
+model = NMF(n_components=8, solver="cd", max_iter=1_000, verbose=True)
+W = model.fit_transform(absorption)
+H = model.components_
+
+import matplotlib.pyplot as plt
+fig, ax = plt.subplots()
+for i in range(model.n_components):
+    ax.plot(1 - H[i] / H[i].max() + i, label=i)
+    
+    
+from astropy.table import Table
+
+names, data = (["source_id", "gaia_rvs_snr"], [source_id, snr[indices]])
+for i in range(W.shape[1]):
+    data.append(W[:, i])
+    names.append(f"W{i}")
+    
+t = Table(data=data, names=names)
+t.write("20231218_gaia_rvs_nmf.csv")

### 2023-12-18 applications/gaia_rvs/20231219_gaia_rvs_nmf.py

@@ -1,0 +1,229 @@
+import numpy as np
+from sklearn.decomposition import NMF
+from astropy.io import fits
+import h5py as h5
+from astropy.table import Table
+import matplotlib.pyplot as plt
+from sklearn.ensemble import RandomForestRegressor
+
+
+EPSILON = np.finfo(np.float32).eps
+
+galah = Table.read("GALAH_DR3_main_allstar_v2.fits")
+keep = (
+    np.isfinite(galah["teff"]) \
+&   np.isfinite(galah["logg"]) \
+&   np.isfinite(galah["fe_h"]) \
+&   np.isfinite(galah["alpha_fe"])
+)
+galah = galah[keep]
+
+fp = h5.File("/Users/andycasey/Downloads/dr3-rvs-all.hdf5", "r")
+
+
+_, A_idx, B_idx = np.intersect1d(fp["source_id"][:], np.array(galah["dr3_source_id"]), assume_unique=False, return_indices=True)
+
+N = 10_000
+
+np.random.seed(1234)
+
+
+try:
+    flux
+except NameError:
+    A_indices = np.sort(np.random.choice(A_idx, N, replace=False))
+
+    flux = fp["flux"][A_indices]
+    ivar = fp["flux_error"][A_indices]**-2
+
+B_indices = B_idx[np.intersect1d(A_idx, A_indices, return_indices=True)[1]]
+
+X = np.clip(1 - flux, 0, 1)
+X[~np.isfinite(X)] = 0
+V = ivar
+is_bad_pixel = (
+    (V == 0)
+|   (~np.isfinite(V))
+|   (X < 0)
+|   (~np.isfinite(X))
+)
+V[is_bad_pixel] = X[is_bad_pixel] = 0
+
+
+def _multiplicative_update(X, V, W, H, update_H=True, update_W=True, l1_reg_H=0, l2_reg_H=0, l1_reg_W=0, l2_reg_W=0):
+    WH = W @ H
+    if update_H:
+        numerator = ((V.T * X.T) @ W).T
+        denominator = ((V.T * WH.T) @ W).T
+        if l1_reg_H > 0:
+            denominator += l1_reg_H
+        if l2_reg_H > 0:
+            denominator += l2_reg_H * H
+        denominator[denominator == 0] = EPSILON
+        H *= numerator / denominator
+    
+    if update_W:
+        numerator = (X * V) @ H.T
+        denominator = (V * WH) @ H.T
[…]

### 2023-12-18 applications/gaia_rvs/gaia_rvs.py

@@ -1,0 +1,363 @@
+import numpy as np
+from tqdm import tqdm
+from typing import Union, Tuple, Optional
+from functools import cached_property
+from scipy import optimize as op
+from sklearn.decomposition._nmf import _initialize_nmf
+
+import h5py as h5
+
+fp = h5.File("/Users/andycasey/Downloads/dr3-rvs-all.hdf5", "r")
+
+np.random.seed(1234)
+
+N = 10_000
+
+
+
+indices = np.sort(np.random.choice(len(fp["source_id"]), N, replace=False))
+
+try:
+    flux
+except NameError:
+    flux = fp["flux"][indices]
+    ivar = fp["flux_error"][indices]**-2
+    source_id = fp["source_id"][indices]
+else:
+    print(f"Using pre-loaded data")
+
+EPSILON = np.finfo(np.float32).eps
+
+class CalibratedContinuum(object):
+
+    def __init__(
+        self,
+        wavelength: np.array,
+        deg: int,
+        L: Union[float, int],
+        regions: Optional[Tuple[Tuple[float, float]]] = None,
+        tol: float = 1e-4,
+        n_components: int = 8,
+        outer_iter: int = 100,
+        inner_W_iter: int = 1,
+        inner_WH_iter: int = 1,
+        init: Optional[str] = None,  
+        initial_continuum_scalar = 1,   
+        **kwargs
+    ):    
+        self.L = L
+        self.deg = deg
+        self.tol = tol
+        self.regions = regions or ((0, wavelength.size), )
+        self.wavelength = wavelength
+        self.n_components = n_components
+        self.outer_iter = outer_iter
+        self.inner_W_iter = inner_W_iter
+        self.inner_WH_iter = inner_WH_iter
+        self.init = init
+        self.initial_continuum_scalar = initial_continuum_scalar
+        return None
+    
+    @cached_property
+    def region_masks(self):
+        return _region_masks(self.wavelength, self.regions)
+    
+    @cached_property
+    def continuum_design_matrix(self):
+        return _continuum_design_matrix(self.wavelength, self.deg, self.L, self.regions, self.region_masks)
+    
+    @cached_property
+    def n_parameters_per_region(self):
+        return max(2 * self.deg + 1, 0)
+    
+    def _fit_continuum(self, flux, ivar):
+        if self.deg is None or self.deg < 0:
+            return (None, np.ones_like(flux))
+
+        theta = np.zeros((N, len(self.regions), self.n_parameters_per_region))
+            
[…]

### 2023-12-18 paper/main.tex
update abstract with variants

@@
 Continuum normalization is often a necessary step when analyzing stellar spectra. 
-Best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but computing stellar spectra can be expensive.
-We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption in theoretical (normalized) spectra, and a sines-and-cosines basis for the joint continuum-instrument response.
-The non-negative matrix factorization ensures that eigenspectra are strictly additive, thereby restricting the predicted normalized stellar flux to be at most 1, with the joint continuum-instrument response component to describe the rest.
+%The best approach is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is usually required before estimating the stellar parameters.
+We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption,
+% in theoretical (normalized) spectra, 
+and a sines-and-cosines basis for the joint continuum-instrument response.
+The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the rest.
 The linearity of both components ensures that inference is stable and fast.
-We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of all evolutionary stages: from M-dwarfs to white dwarfs.
-We find good fits to observed spectra, even in situations where the theoretical grid used to approximate line absorption is provably incorrect for some features.
-%The stellar absorption model is built from a non-negative matrix factorization of a theoretical continuum-normalized spectra, which ensures it cannot predict spectra more than 1.
-%The constraints restrict the stellar absorption to 
-We show that we obtain unbiased estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
-This approach largely eliminates subjectivity in stellar continuum normalization.
+We describe three variants of this approach:
+    (1) a model-driven approach where line absorption is factorized entirely from continuum-normalized theoretical spectra;
+    (2) a purely data-driven approach where line absorption and continuum are fit entirely from data;
+    and
+    (3) a hybrid approach where basis spectra are initialized from factorization of theoretical spectra, but iteratively adjusted to fit data. 
+We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of stars spanning evolutionary stages from pre-main-sequence stars to white dwarfs.
+We find good fits to data, even for the model-driven variant where theoretical spectra used to factorize line absorption is provably incorrect for some features.
+We show that the basis amplitudes can be used to reliably estimate stellar parameters and chemical abundances, 
+and that we obtain low bias estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
+%This approach largely eliminates subjectivity in stellar continuum normalization.
 \end{abstract}
@@
 
+Continuum normalization is a subject
 Continuum normalisation is a necessary step
@@
 \end{align}
-where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
-We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX$ that are 0 (i.e., no line absorption).
+where all elements in $\vecX$, $\vec{W}$, and $\vec{H}$ are required to be non-negative. 
+We could construct an approximation directly to $\vec{M}$, but this would lead to denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX$ that are 0 (i.e., no line absorption).
 For this approximation we must select the number of components $C$ to use, which is significantly smaller than both the number of theoretical spectra $N$ and the number of pixels $D$ per theoretical spectrum.
@@
 \begin{figure}
+    \caption{A Gaia H-R diagram showing the median $\chi^2$ per bin.}
+\end{figure}
+
+\begin{figure}
     \caption{Median observed rectified flux values as a function of S/N ratio showing that there is no bias as a function of S/N ratio.}
@@
 \todo{Should expect to see increasing residuals due to TiO bands not captured by the model, and emission lines.}}
+    % Could do the same showing standard deviation per bin
 \end{figure*}
@@
 - figure showing the
+
+\section{Discussion}\label{sec:discussion}
+
+- getting good fits even with bad models: the hydrogen lines in the apogee grid are knowingly incorrect, but we see no difference in the $\chi^2$ at those wavelengths.
+That is because the hydrogen lines separate out into very clean eigenspectra, where the amplitude for that eigenspectra is not required to vary smoothly with other amplitudes
+
+- biases as a function of S/N or labels
+
+- poorly modelled regions as a function of pixel
+
+- $\chi^2$ across the HRD
+
+- excellent way to identify regions where models and observations may disagree
+
+- orthogonality of the two model components
+
+- issue with M-giant eigenspectra when they were over-represented in the sample
 
@@
     Adam Wheeler (Ohio State University),
+    Andrew Saydjari (Harvard),
     David W. Hogg (New York University),

### 2023-12-19 paper/main.tex
add methods

@@
 %The best approach is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is usually required before estimating the stellar parameters.
-We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption,
-% in theoretical (normalized) spectra, 
-and a sines-and-cosines basis for the joint continuum-instrument response.
+We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization to approximate line absorption, and a sines-and-cosines basis for the joint continuum-instrument response.
 The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the rest.
 The linearity of both components ensures that inference is stable and fast.
-We describe three variants of this approach:
+We describe three variants of this method:
     (1) a model-driven approach where line absorption is factorized entirely from continuum-normalized theoretical spectra;
@@
     and
-    (3) a hybrid approach where basis spectra are initialized from factorization of theoretical spectra, but iteratively adjusted to fit data. 
-We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of stars spanning evolutionary stages from pre-main-sequence stars to white dwarfs.
-We find good fits to data, even for the model-driven variant where theoretical spectra used to factorize line absorption is provably incorrect for some features.
-We show that the basis amplitudes can be used to reliably estimate stellar parameters and chemical abundances, 
+    (3) a hybrid approach where basis spectra are initialized from factorization of theoretical spectra, and then iteratively adjusted to fit data. 
+We apply these methods to \project{Sloan Digital Sky Survey} optical spectra 
+%of stars spanning evolutionary stages 
+from pre-main-sequence stars to white dwarfs.
+In a model-driven or hybrid approach, we show that the basis weights can reliably estimate stellar parameters and chemical abundances.
+%The data-driven approach is less physically interpretable, but naturally provides the best description for the data and is an ideal no-model solution when only a few spectra are available.
+%We find good fits to data, even for the model-driven variant where theoretical spectra used to factorize line absorption is provably incorrect for some features.
+We show that the basis weights can be used to reliably estimate stellar parameters and chemical abundances, 
 and that we obtain low bias estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
@@
 
-Continuum normalization is a subject
-Continuum normalisation is a necessary step
+Continuum normalization is often a necessary step before estimating stellar parameters and chemical abundances.
+This process is ripe with subjectivity and bias.
+There is general agreement in the literature that consistent continuum normalization is important, but no apparent consensus on how it should be done.\\
 
-Often needed to be done before stellar parameter determination, as a step towards stellar parameter determination
+Fitting the continuum well requires you to know where there is line absorption.
+But knowing where there is line absorption requires you to (at least) know the stellar parameters.
+Classical spectroscopists might estimate some coarse continuum for the spectrum (or for each order, in the case of high-resolution echelle spectra), and then refine the continuum for every measured absorption line.
+This process is still often done by hand (albeit they are often experienced hands).
+Industrial spectroscopists might simply aim for a \emph{consistent} continuum normalization procedure, even if the process only estimates a pseudo-continuum for comparison, and not the true continuum.
+In an ideal scenario the pseudo-continuum is fit jointly with the stellar parameters.\\
 
-For measuring equivalent widths, or for full-spectrum fitting, or even for performing radial velocity determination
+- For some stars there is NO continuum (e.g., M-dwarfs).
 
-Many methods. Some done by eye.
-
-Fitting the continuum well requires you to know where the line absorption is. But knowing where the line absorption is requires you to know the stellar parameters.
-
-Best is to fit with the stellar parameters, but is intractable.
-
-For some stars there is NO continuum (e.g., M-dwarfs).
-
-cite cases of NMF use in astronomy (not that many)
+- Cite cases of NMF use in astronomy (not that many)
 
@@
 
-We will assume a forward model for the data that includes two components: a component to represent continuum-normalized absorption (e.g., line absorption), and a component to represent the smooth continuum.\footnote{Most stellar spectra are not flux-calibrated; the continuum and instrument response enter multiplicatively and require some work to disentangle the two.
-Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.} We will assume both components to be linear models.
-This is not a strict requirement, but keeping linearity ensures that  inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation.\\
+We will assume a forward model that includes two components: one represent continuum-normalized absorption (e.g., line absorption); and another to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
+These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will assume both components to be linear models.
+This is not a strict requirement, but keeping linearity ensures that inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization.\\
 
-% other assumptions
-% - NMF is sufficient for representing continuum-normalized spectra
-% - that the theoretical models are OK enough.
-there are ways they can be wrong (eg hydrogen lines) and still work in practice, or there are ways they can be wrong entirely (eg missing molecular band) and not work at all
-% - that sines and cosines are sufficiently flexible for modelling the large scale continuum
-The problem can be described as having data that are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$).
-The forward model for these data can be expressed as the element-wise multiplication of the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
+Here we will describe the method in general before outlining assumptions and details of individual variants.
+In all variants we assume that non-negative matrix factorization (NMF) is sufficient to represent continuum-normalized line absorption in a large variety of stellar spectra.
+We further assume that the sine-and-cosine basis is sufficiently flexible for modelling the joint instrument-continuum response.
+There are choices to be made in terms of the degree and length scale used for the sine-and-cosine basis, but these are application-specific choices.\\
+
+% We could list all other assumptions (flat universe), but are there any truly necessary assumptions that we also need to list for the *astronomer* or *spectroscopist* audience?
+
+The problem of continuum normalization for a single spectrum can be described where the data are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
+The forward model for these data can be expressed as the element-wise multiplication of a line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
 \begin{align}
@@
 
-% NMF first
-The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a grid of $N$ theoretical spectra each with $D$  continuum-normalized fluxes, which we denote as $\vec{M}$.
-In the applications that we present, the data are assumed to have the same wavelength sampling and a consistent line spread function, such that we use a grid where $D = P$.
-Other situations are permitted, but at inference time there is a need to interpolate the line absorption to the $P$ observed pixels.
-With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecX$ to be
+The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a set of $N$ continuum-normalized spectra each with $D$ fluxes.
+In the applications that we present, the data are assumed to have the same wavelength sampling and line spread function as the spectra used to approximate the line absorption model, such that we use a grid where $D = P$.
+Other situations are permitted, but at inference time there is a need to interpolate the line absorption model to the $P$ observed pixels.
+The three variants of our method that we describe here have different sets of continuum-normalized spectra, but the process of factorizing them is the same.
+With our $N \times\ D$ matrix $\vec{M}$ of continuum-normalized fluxes, we then define \emph{stellar absorption} $\vecX$ to be
 \begin{align}
@@
 \end{align}
-such that $\vecX \in \left[0, 1\right)$ and is zero when no line absorption exists.
-This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecX$ using non-negative matrix factorization (NMF) such that
+such that $\vecX \in \left[0, 1\right]$ and is zero when no line absorption exists.
+This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecX$ using NMF such that
 \begin{align}
@@
 We could construct an approximation directly to $\vec{M}$, but this would lead to denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX$ that are 0 (i.e., no line absorption).
-For this approximation we must select the number of components $C$ to use, which is significantly smaller than both the number of theoretical spectra $N$ and the number of pixels $D$ per theoretical spectrum.
-Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding eigenspectra.
-Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$.
+The nomenclature in this paper draws on previous works on NMF and data analysis more broadly.
+We would like the naming conventions to be consistent with those papers, but this nomenclature is easily overloaded.
+For this reason, we make clear definitions: throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights} and $\vecH$ as the NMF \emph{basis vectors}.
 \\
 
+For this factorization we must select the number of basis components $C$ to use, which should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
+Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ basis weights per spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding basis vectors each with $D$ pixels.
+Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example basis vectors $\vec{H}$.\\
+
+%The nomenclature  draws on previous works on NMF and data analysis, but this nomenclature is easy to overload, so throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights}, $\vecpsi$ as the NMF \emph{basis coefficients}, $\vecH$ as the NMF \emph{basis vectors}, and $\vectheta$ (not yet defined) will refer to the amplitudes of the sine-and-cosine basis.
+
 \begin{figure*}
-    \caption{A schematic illustrating the non-negative matrix factorization procedure, with some example infrared eigenspectra. \label{fig:schematic}}
+    \caption{A schematic illustrating the non-negative matrix factorization procedure, with some example basis vectors computed from the application to \emph{BOSS} spectra. \label{fig:schematic}}
 \end{figure*}
+
+The factorization of $\vecH$ is reasonably fast and easy to compute given existing packages in Python, Julia, and other languages.
+We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $\vecX$: this is the default behaviour in the Python \texttt{scipy.decomposition.NMF} implementation.
+\\
+%As reported elsewhere, we found that factorizing $\vecW\vecH$ by coordinate descent was substantially faster than using multiplicative updates, but this was only relevant when the input spectra have no flux errors (i.e., the theory-only variant).
+Factorizing the data-driven and hybrid approaches is described later in this section.
+In any case, computing the factorization can be finished in seconds or hours, depending on the scale of the matrices $\vecW$ and $\vecH$.\\
+
+%We used the \texttt{scipy.decomposition.NMF} implementation \citep{scipy} with some minor adjustments to minimize the memory requirements (e.g., allowing for lower precision float types).
+We used no regularization on $\vecW$ or $\vecH$.
+We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition with zeros filled-in with the average of $\vecX$ (the default behavior in the \texttt{scipy} implementation).
+Initializing with random or small non-negative values produced comparable results but required many more iterations.
+We stopped the approximation of $\vecW$ and $\vecH$ after 1,000 iterations of multiplicative updates.
+This took minutes to hours to complete, depending on the size of the theoretical grid $\vecX$.\\
 
@@
 \end{align}
-where $\vecpsi \in [0, \infty)$ is a row vector of $C$ elements.
-$\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ amplitudes needed to reconstruct the stellar absorption from the $C$ eigenspectra in $\vecH$.
+where $\vecpsi \in [0, \infty)$ are a row vector of $C$ \emph{basis weights}.
+$\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ weights needed to reconstruct the stellar absorption from the $C$ basis vectors in $\vecH$.
 Note that because $\vecpsi$ and $\vecH$ are both restricted to have non-negative elements, Equation~\ref{eq:f} shows that the maximum value that can be predicted by $f(...)$ is 1.
 This restricts the flexibility of $f(...)$ to only be able to model continuum-normalized flux values, leaving $g(...)$ to represent the joint continuum-instrument response.
-While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this would be readily compensated by a negative continuum model $g(...) < 0$ and in practice such local minima are unfavoured.\\
+While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this would be readily compensated by a negative continuum model $g(...) < 0$ and in practice such local minima are not favoured.\\
 
@@
 \end{align}
-\noindent{}and the design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.\\
+\noindent{}and the design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
+Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.
+In this formalism, spectral regions can be fit independently with different degrees of freedom.
+The number of continuum coefficients scales as $2n_\textrm{degree} + 1$ per region.
+
+%The method can be readily applied to real data once the eigenspectra $\vecH$ have been computed.
+We experimented with choices of initialisation and inference.
+Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously.
+The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecpsi$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip).
+Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecpsi$) seemed to work well in many scenarios.
+\\
 
@@
 
-\subsection{Constructing the line absorption model}
+\subsection{Model-driven approach: Line absorption is factorized by theoretical spectra}
+\label{sec:model-method}
+
+This variant uses a set of theoretical spectra to factorize line absorption $\vecX$.
+With no flux errors on theoretical spectra, this factorization can be completed quickly using the coordinate descent algorithm (the data-driven variant uses multiplicative updates, which is slower).
+In this variant, there are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing in between points.
+The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to apply the method.
+This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of OBA-type stars was also sufficiently flexible to model many (but not all) white dwarf spectra.
+For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, or simply skip every $n$th theoretical spectrum.}\\
+
+- describe BOSS spectrograph characteristics, MWM + BHM
+
+- BOSZ spectra convolved and resampled
+
+- initially ignore the grid stellar parameters
+
+- factorize for 1000 steps
+
+- compute H\\
+
+Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecpsi$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data.
+We found that setting $\vecpsi$ to a very small value ($10^{-12}$; i.e., no line absorption) and solving for the continuum coefficients $\vectheta$ was a reasonably good initialization.
+In situations where there are spectra of the same star we advocate fitting all visit spectra simultaneously using a single set of basis weights $\vecpsi$ for all spectra, and continuum coefficients $\vectheta_i$ for the $i$th epoch spectrum.
+
+\subsection{Data-driven approach: Line absorption is factorized by data}
+\label{sec:data-method}
+
+In this variant we solve for the factorized line absorption and continuum entirely from data, without any theoretical spectra.
+The model concept is exactly the same as introduced in Section~\ref{sec:methods}: we have a forward model for the flux that depends on basis weights $\vecpsi$ and continuum coefficients $\vectheta$.
+In this variant the principal difference is how we factorize the basis vectors $\vecH$.\\
+
+The process of fitting the model is the same for one spectrum or many.
+The procedure involves alternating through three linear operations.
+Before starting we assume that $\vecW$ and $\vecH$ are initially filled with zeros such that there is no line absorption.
+Then we alternate between these three steps:
+\begin{enumerate}
+    \item Solve for the continuum coefficients $\vectheta$ per spectrum, given the (i.e., keeping fixed) current estimates of $\vecW$ and $\vecH$.
+    \item Update the basis vectors $\vecH$, given the current estimate of the continuum coefficients $\vectheta$ and the basis weights $\vecW$.
+    \item Update the basis weights $\vecW$ per spectrum, given the current estimate of the continuum coefficients $\vectheta$ and the basis vectors $\vecH$.
+\end{enumerate}
+Completing all three steps is described as one iteration, and we continue iterating until either a fixed number of iterations is reached, or a tolerance in $\chi^2$ is achieved.
+Each step here is a \emph{convex linear operation} that is guaranteed to improve the $\chi^2$ fit, leading to an approach is extremely stable, fast, and guaranteed to converge.\\
+
+An illustration of this variant is shown in Figure~\ref{fig:schematic-data-driven}.
+With no line absorption initially, the continuum is set to fit all of the data.
+But with that (poor) estimate of the continuum, one can calculate the continuum-normalized flux and compute a better estimate of the basis vectors $\vecH$.
+Those basis vectors essentially have the right `shape', in that they represent line absorption at the appropriate wavelengths, but they are too low: the true absorption needs to be stronger.
+With the continuum and the basis vectors now fixed $\vecH$, we can update our estimate of the basis weights $\vecW$ per spectrum.
+This process continues for the number of iterations specified by the user, or after some defined tolerance in $\chi^2$ is reached.\\
+
+In this (and the hybrid) approach, the updated estimates of $\vecW$ and $\vecH$ are found by multiplicative updates, whereas in the model-only approach, coordinate descent is used to factorize the line absorption.
+Here we will let
+\begin{align}
+    \vec{Y} = 1 - \frac{\vec{y}}{g(\vec{\lambda};\vectheta)}
+\end{align}
+\noindent{}be the line absorption of observed spectra, which is computed given our current estimate of the continuum normalization $g(\lambda; \vectheta)$, and $\vec{Z}$ be the corresponding inverse variances
+\begin{align}
+    \vec{Z} = \left(g(\vec{\lambda};\vectheta)/\vec{\sigma_{y}}\right)^2 
+\end{align}
+such that we can compute our updated estimate of the basis vectors $\vecH$ (step 2) 
+\begin{align}
+    \vecH_{ij}^{\mathrm{(new)}} \leftarrow \vecH_{ij} \frac{\left[\vecW\transpose\left(\vec{Y} \hadamard \vec{Z}\right)\right]_{ij}}{\left[\vecW\transpose\left(\vecW\vecH \hadamard \vec{Z}\right)\right]_{ij}}
+\end{align}
+\noindent{}while properly accounting for the inverse variance in each pixel. We can similarly update the basis weights $\vecW$ (step 3)
+\begin{align}
+    \vecW_{ij}^{\mathrm{(new)}} \leftarrow \vecW_{ij} \frac{\left[\left(\vec{Y} \hadamard \vec{Z}\right)\vecH\transpose\right]_{ij}}{\left[\left(\vecW\vecH\hadamard\vec{Z}\right)\vecH\transpose\right]_{ij}}
+\end{align}
+\noindent{}where $\left[\vec{AB}\right]_{ij}$ indicates the $(i,j)$ element in the $\vec{AB}$ matrix.\\
+
+This variant works well if you have only one spectrum, or if you have many.
+If you only have one spectrum (or a few) then this is a good generalized approach for continuum normalization, which does not require many decisions from the user.
+One restriction is that $C$ should be set to at most the number of spectra being fit (i.e., do not use $C = 2$ basis vectors to fit 1 spectrum).
+However, in these situations with only few spectra and no theoretical spectra for factorization, the downside is that the basis weights $\vecpsi$ are uninterpretable.
+
+\begin{figure*}
+    \caption{A schematic illustrating the data-driven variant of our method.\label{fig:schematic-data-driven}}
+\end{figure*}
+
+\subsection{Hybrid approach: Line absorption is factorized by theoretical spectra, and iteratively improved by data}
+\label{sec:hybrid-method}
+
+The last variant we present in this paper is a hybrid of the two previous methods.
+The model-driven approach has the benefit that the basis weights and basis vectors can have physical interpretations.
+For example, a basis vector containing only strong, broad hydrogen lines would imply that any observed spectrum with a large basis weight for that vector might be a hot star.
+However, limitations in the theoretical model grid become readily apparent when using only a model-driven approach.
+While the data-driven approach naturally provides an excellent description of the data, it is less interpretable.
+This variant combines the benefits of both scenarios.
+
+While the iterative steps in Section~\ref{sec:data-method} are each convex linear operations, the actual factorization of $\vec{X} = \vec{WH}$ is not convex: there are many combinations of $\vecW$ and $\vecH$ that produce the same result.
+For this reason, the solution of $\vecW$ and $\vecH$ depends on the initial starting point.
+Normally this is not a problem because we don't necessarily care \emph{which} solution we have to the factorization, but we do care if we want to gain physical insight from the basis weights $\vecW$ or basis vectors $\vecH$.
+
+In this variant we use a theoretical grid and compute the basis vectors $\vecH$ just as described in Section~\ref{sec:model-method}.
 
 The line absorption model $f(\lambda_i;\vecpsi)$ can be easily constructed from a grid of theoretical continuum-normalized spectra. 
-In later sub-sections we discuss the specific choices made for the \emph{APOGEE} and \emph{BOSS} applications, but for now we discuss some general principles and practices that we found successful.
+%In later sub-sections we discuss the specific choices made for the \emph{APOGEE} and \emph{BOSS} applications, but for now we discuss some general principles and practices that we found successful. 
 The objective function to minimize
-\begin{align}
-    \chi^2 = ||\vecX - \vecW\vecH||^2
-\end{align}
-\noindent{}is the Frobenius norm (the squared approximation error). This can be solved through coordinate descent, or through the multiplicative update rules developed by \citet{Lee:2000}:
-\begin{align}
-    \vecW_{ij}^{\mathrm{(new)}} \leftarrow \vecW_{ij} \frac{\left[\vecX\vecH\transpose\right]_{ij}}{\left[\vecW\vecH\vecH\transpose\right]_{ij}} \quad \phantom{.}\\
-    \vecH_{ij}^{\mathrm{(new)}} \leftarrow \vecH_{ij} \frac{\left[\vecW\transpose\vecX\right]_{ij}}{\left[\vecW\transpose\vecW\vecH\right]_{ij}} \quad .
-\end{align}
-
-We used the \texttt{scipy.decomposition.NMF} implementation \citep{scipy} with some minor adjustments to minimize the memory requirements (e.g., allowing for lower precision float types).
-We used no regularization on $\vecW$ or $\vecH$.
-We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition with zeros filled-in with the average of $\vecX$ (the default behavior in the \texttt{scipy} implementation).
-Initializing with random or small non-negative values produced comparable results but required many more iterations.
-We stopped the approximation of $\vecW$ and $\vecH$ after 1,000 iterations of multiplicative updates.
-This took minutes to hours to complete, depending on the size of the theoretical grid $\vecX$.\\
-
-There are no requirements on the number of grid dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on the spacing between grid points.
-The only implicit requirement is that the grid should approximately span the range of stars that you intent to apply the method.\footnote{This is more of a recommendation than a requirement.
-For example, in practice we found that a grid trained on OBA-type stars was also sufficient to model many white dwarf spectra.} For this reason, we chose to include as many spectra as our memory constraints would allow.
-If the number of spectra exceeded memory constraints, we would opt to skip every $n$th theoretical spectrum.\\
 
@@
 %\subsection{Inference}
-
-The method can be readily applied to real data once the eigenspectra $\vecH$ have been computed.
-We experimented with choices of initialisation and inference.
-Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously.
-The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecpsi$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip).
-Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecpsi$) seemed to work well in many scenarios.
-\\
 
@@
 
-We used all \todo{X} BOSZ spectra to construct a line absorption matrix $\vecX$ and find the approximating matrices $\vecW$ and $\vecH$.
+%We used all \todo{X} BOSZ spectra to construct a line absorption matrix $\vecX$ and find the approximating matrices $\vecW$ and $\vecH$.
 For fitting \boss\ spectra, we separated the continuum fits into two chunks to represent the blue and red chip of the detector.

### 2023-12-19 paper/main.tex
add intro

@@
 We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization to approximate line absorption, and a sines-and-cosines basis for the joint continuum-instrument response.
-The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the rest.
+The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the remaining variance.
 The linearity of both components ensures that inference is stable and fast.
@@
 Continuum normalization is often a necessary step before estimating stellar parameters and chemical abundances.
-This process is ripe with subjectivity and bias.
-There is general agreement in the literature that consistent continuum normalization is important, but no apparent consensus on how it should be done.\\
+This process is ripe with subjectivity.
+While there is general agreement in the literature that consistent continuum normalization is important, there is apparently no consensus on how it should be done.\\
 
@@
 But knowing where there is line absorption requires you to (at least) know the stellar parameters.
+Without knowing the stellar parameters (or having a good model for line absorption), spectroscopists have been forced to use asymmetric so-called sigma clipping (i.e., to exclude data points points $N\sigma$ below a continuum fit but keep points above the fit), or restrict themselves to a carefully selected set of so-called continuum pixels.
+The set of continuum pixels is only valid for stars of a similar spectral type and metallicity, and in many cases there are \emph{no continuum pixels} (e.g., M-dwarfs).\\
+
 Classical spectroscopists might estimate some coarse continuum for the spectrum (or for each order, in the case of high-resolution echelle spectra), and then refine the continuum for every measured absorption line.
@@
 Industrial spectroscopists might simply aim for a \emph{consistent} continuum normalization procedure, even if the process only estimates a pseudo-continuum for comparison, and not the true continuum.
-In an ideal scenario the pseudo-continuum is fit jointly with the stellar parameters.\\
+In an ideal scenario the pseudo-continuum is fit jointly with the stellar parameters, but this is often considered too expensive.\\
 
-- For some stars there is NO continuum (e.g., M-dwarfs).
-
-- Cite cases of NMF use in astronomy (not that many)
+In Section~\ref{sec:methods} we describe a family of methods to address this problem.
+Each variant makes use of non-negative matrix factorization (NMF) to approximate line absorption.
+NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements.
+This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
+The three variants of our method are each conceptually introduced, with the intended application (Section~\ref{sec:results}) to low-resolution optical spectra from the \emph{Sloan Digital Sky Survey}.
+We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\
 
@@
 Here we chose a sine-and-cosine basis because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
-The component $g(\lambda_i;\vectheta)$ is expressed as compactly as
+The component $g(\lambda_i;\vectheta)$ is expressed compactly as
 \begin{align}
@@
 
-- describe BOSS spectrograph characteristics, MWM + BHM
+The fifth generation of the \emph{Sloan Digital Sky Survey}  is acquiring low-resolution ($\mathcal{R} \sim 4{,}000$) optical spectra for millions of Milky Way stars using the \boss\ spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO).
+Although the highest priority targets observed by \boss\ tend to be Black Hole Mapper (BHM) targets, many Milky Way stars are included per field, either because they are specifically targeted as part of a Milky Way Mapper (MWM) `carton'\footnote{`Cartons' are how targets are now assigned in SDSS.}, or because they are standard stars.
+We took all \boss\ spectra where the source was allocated to a Milky Way Mapper carton, or spectra where the source was allocated to a stellar-like carton.
+\todo{Some description of stellar-like.}\\
 
-- BOSZ spectra convolved and resampled
+We used the BOSZ spectral grid to construct basis vectors for a line absorption model for \boss\ spectra.
+This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a sufficiently large range in stellar parameters.
+We convolved the spectra to the nominal mean resolving power of the \boss\ spectrograph ($\mathcal{R} \sim 4{,}000$) and resampled it to the same pixels that the \boss\ spectra are resampled to: a uniform-in-log wavelength sampling from approximately 360\,nm until about 1000\,nm.
+We clipped any BOSZ flux values with emission lines exceeding normalized flux values of 1, as they would violate our NMF requirements.
+As we discuss in later sections, any emission lines are sufficiently rare in stellar spectra that they do not contribute significantly to the $\chi^2$ at inference time.\\
 
@@
 For example, a basis vector containing only strong, broad hydrogen lines would imply that any observed spectrum with a large basis weight for that vector might be a hot star.
-However, limitations in the theoretical model grid become readily apparent when using only a model-driven approach.
-While the data-driven approach naturally provides an excellent description of the data, it is less interpretable.
-This variant combines the benefits of both scenarios.
+However, limitations in the theoretical model grid become readily apparent when using only a model-driven approach: there are entire molecular bands that are not represented in the theoretical model spectra.
+And while the data-driven approach naturally provides an excellent description of the data, it is less interpretable.
+This hybrid variant combines the benefits of both scenarios.
+\\
 
@@
 For this reason, the solution of $\vecW$ and $\vecH$ depends on the initial starting point.
-Normally this is not a problem because we don't necessarily care \emph{which} solution we have to the factorization, but we do care if we want to gain physical insight from the basis weights $\vecW$ or basis vectors $\vecH$.
+Normally this is not a problem because we don't necessarily care \emph{which} solution we have to the factorization, but we do care if we want to gain physical insight from the basis weights $\vecW$ or basis vectors $\vecH$.\\
 
 In this variant we use a theoretical grid and compute the basis vectors $\vecH$ just as described in Section~\ref{sec:model-method}.
+We then take $N >> C$ observed spectra and fit the basis weights $\vecpsi$ and continuum coefficients $\vectheta$ for each spectrum, keeping $\vecH$ fixed.
+At this point the variant is exactly the same as the model-driven method.
+Now we start iterating between the steps described in Section~\ref{sec:data-method}, starting from Step 2: we update the basis vectors $\vecH$ given the current estimate of the continuum coefficients $\vectheta$ and the basis weights $\vecW$.
+Next we proceed to Step 3, where we update the basis weights $\vecW$ given the current estimate of the continuum coefficients $\vectheta$ and the basis vectors $\vecH$.
+We continue cycling through Steps 1-3 a pre-determined number of times or until a tolerance in $\chi^2$ is reached.\\
 
-The line absorption model $f(\lambda_i;\vecpsi)$ can be easily constructed from a grid of theoretical continuum-normalized spectra. 
-%In later sub-sections we discuss the specific choices made for the \emph{APOGEE} and \emph{BOSS} applications, but for now we discuss some general principles and practices that we found successful. 
-The objective function to minimize
+In the applications we present here, this variant only tends to adjust \emph{some} of the basis vectors $\vecH$ (and corresponding weights $\vecW$).
+For example, absorption features that were poorly modelled in the theoretical grid quickly appear in the updated basis vectors in order to better match the data.
+Overall, most basis vectors do not change considerably, which keeps the basis weights and basis vectors (somewhat) interpretable.
+This step can be altered to only update specific vectors (e.g., those that look like molecular absorption, if molecular absorption is poorly modelled by theory), or specific wavelengths, but this experimentation is outside the scope of this work.
 
-%Note that the stellar parameters used to compute the theoretical spectra are ignored when solving for $\vecW$ and $\vecH$ in Equation~\ref{eq:nmf}: only the spectra are used.
-There are correlations between the stellar parameters and the amplitudes $\vecW$, but as we discuss in Section~\ref{sec:discussion}, we avoid trying to use the amplitudes $\vecW$ (or $\vecpsi$) to interpreting stellar parameters.
+\section{Results}
+\label{sec:results}
 
-%\subsection{Inference}
-
-\subsection{APOGEE spectra}
-\label{sec:apogee}
-
-The \apogee\ spectrograph spans approximately 1.5 to 1.7\,microns and has a resolving power of about 22,500 \citep{wilson}.
-There are three chips, which we chose to model as separate continuum regions (each with their own continuum coefficients $\vecpsi$).
-We used a 3rd order sine-and-cosine basis with length scale $L = 1300$, corresponding to 7 continuum coefficients per chip.\\
-
-We use the \todo{DR17 grids computed for FERRE, some in non-LTE}.
-The data are assumed to be resampled onto a rest-frame wavelength sampling that is uniform in log-spacing.
-We took the high-resolution theoretical grids and convolved them to a spectral resolution of 22,500 and sampled it to match the data.
-In total there are nearly \todo{1 M} spectra each with 8,575 pixels.
-This proved too large to fit in memory when computing the eigenspectra $\vecH$, forcing us to skip every 10th spectra.\\
-
-We chose to use 32 eigenspectra to construct the line absorption model, some of which are shown in Figure~\ref{fig:schematic}.
-During training, the hydrogen lines separate out immediately before other absorption features are recognizable.
-After training is complete, there are multiple eigenspectra that only have hydrogen lines: some narrow, and some broad.
-Eigenspectra with only recognizable patterns of molecular features are also present.
-The remaining eigenspectra have line absorption at many different wavelengths, representing something akin to overall metallicity.
-\\
-
-With these eigenspectra and continuum settings, we simultaneously fit line absorption and the joint continuum-instrument response for all \apogee\ spectra in SDSS-IV DR17 \citep{dr17} and \apogee\ spectra acquired by the Milky Way Mapper (MWM) in SDSS-V \citep{kollmeier}.
-In total there are \todo{X} \apogee\ spectra fit by this process: \todo{X} combined spectra and \todo{X} individual visits.
-We fit combined (co-added) spectra first, and for individual visit spectra we analytically computed the continuum coefficients $\vectheta$ for that visit, conditioned on the $\vecpsi$ found from the combined fit.
-In practice this implies that the combined and visit spectra for the same star share the same amplitudes $\vecpsi$ for computing rectified flux, but may have different continuum coefficients per visit.
-Figure~\ref{fig:example-apogee-spectra} show example \apogee\ rectified spectra for different spectral types.
+We applied the three variants to all \boss\ stellar spectra currently available in the \emph{Sloan Digital Sky Survey}. This includes approximately \todo{X} spectra of \todo{Y} stars.
 
 \begin{figure}
-    \caption{Example rectified \apogee\ spectra (black) and best-fitting rectified line absorption (red) for different spectral types. \label{fig:example-apogee-spectra}}
+    \caption{Continuum-normalized \boss\ spectra of all spectral types, with the best-fitting model spectrum shown.}
 \end{figure}
-
-\subsection{Application to BOSS spectra}
-\label{sec:}
-
-SDSS-V is acquiring optical spectra for millions of Milky Way stars using the \boss\ 
-low-resolution ($\mathcal{R} \sim 4{,}000$) spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO).
-Although the highest priority targets observed by \boss\ tend to be Black Hole Mapper (BHM) targets, many Milky Way stars are included per field, either because they are specifically targeted as part of a Milky Way Mapper carton\footnote{`Cartons' are how targets are assigned in SDSS-V.}, or because they are standard stars.\\
-
-We took all \boss\ spectra where the source was allocated to a Milky Way Mapper carton, or spectra where the source was allocated to a stellar-like carton.
-\todo{Some description of stellar-like.} For these spectra we fit the line absorption and joint continuum-instrument response.\\
-
-We used the BOSZ spectral grid to construct eigenspectra for a line absorption model for \boss\ spectra.
-This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a sufficiently large range in stellar parameters.
-We convolved the spectra to the nominal mean resolving power of the \boss\ spectrograph ($\mathcal{R} \sim 4{,}000$) and resampled it to the same pixels that the \boss\ spectra are resampled to: a uniform-in-log wavelength sampling from approximately 360\,nm until \todo{1000\,nm}.
-We clipped any BOSZ flux values with emission lines exceeding normalized flux values of 1, as they would violate our NMF requirements.
-As we discuss in later sections, any emission lines are sufficiently rare in stellar spectra that they do not contribute significantly to the $\chi^2$ at inference time.\\
-
-%We used all \todo{X} BOSZ spectra to construct a line absorption matrix $\vecX$ and find the approximating matrices $\vecW$ and $\vecH$.
-For fitting \boss\ spectra, we separated the continuum fits into two chunks to represent the blue and red chip of the detector.
-Most of the time the blue and red chips are correctly merged together to produce a seamless contiguous spectrum from \todo{300\,nm to 1000,\nm}, but we noticed there could some times be jumps in the spectrum due to an imperfect merge.
-For these reasons, we used two sections from \todo{3750\,\AA\ to 6000\,\AA and 6050\,\AA to 10000\,\AA}.
-We chose a \todo{3rd} order sines-and-cosines basis with a length scale $L = X$\,\AA.
-Following what we did with \apogee\ spectra, we fit the combined \boss\ spectrum first for a given star, and then analytically computed $\vectheta{}|\vecpsi$: the continuum coefficients $\vectheta$ for the given visit, conditioned on the amplitudes optimized from the combined spectrum $\vecpsi$.
-
-\section{Results}\label{sec:results}
 
@@
 
-- excellent way to identify regions where models and observations may disagree
+- excellent way to identify regions where models and observations totally disagree
 
@@
 
-\section{Conclusions}\label{sec:conclusions}
+\section{Conclusions}
+\label{sec:conclusions}
 
@@
 
-Code is available at \url{https://github.com/andycasey/continuum} and is registered in the Python Package Index as \texttt{stellar-continuum}.
+\noindent{}We provide a Python implementation of our method in the following repository: \url{https://github.com/andycasey/continuum}.
 
@@
 It is a pleasure to thank
-% All these people are possible co-authors
+% All these people are likely co-authors, but should be thanked if they don't want to be co-authors
     Adam Wheeler (Ohio State University),

### 2024-05-16 paper/main.tex
intro updated to reflect single variant

@@
 
+% Other possible titles
+
 % text macros
-\shorttitle{Stellar continuum modelling}
+%\newcommand{\chosentitle}{Constrained linear models for stellar spectroscopy}
+\newcommand{\chosentitle}{Stellar continuum modelling}
+
+\shorttitle{\chosentitle}
 \shortauthors{Casey}
@@
 \newcommand{\boss}{\project{BOSS}}
+\newcommand{\sdss}{\project{SDSS}}
 
@@
 
-\title{\Huge Stellar continuum modelling}
+\title{\chosentitle}
 
@@
 
-%\author[0000-0003-2866-9403]{David W. Hogg}
-%\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
-%\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
-%\affiliation{Flatiron Institute, a division of the Simons Foundation}
+\author[0000-0003-2866-9403]{David W. Hogg}
+\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
+\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
+\affiliation{Flatiron Institute, a division of the Simons Foundation}
+
+% Other people who will be invited as co-authors (non-exhaustive list):
+% Wheeler, Bedell, Astra folks, etc.
 
@@
 Continuum normalization is often a necessary step when analyzing stellar spectra. 
-%The best approach is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is usually required before estimating the stellar parameters.
-We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization to approximate line absorption, and a sines-and-cosines basis for the joint continuum-instrument response.
+The ideal approach might be to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is often required before those parameters are estimated.
+Here we present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization that models stellar line absorption, and a sines-and-cosines basis to model the joint continuum-instrument response.
 The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the remaining variance.
 The linearity of both components ensures that inference is stable and fast.
-We describe three variants of this method:
-    (1) a model-driven approach where line absorption is factorized entirely from continuum-normalized theoretical spectra;
-    (2) a purely data-driven approach where line absorption and continuum are fit entirely from data;
-    and
-    (3) a hybrid approach where basis spectra are initialized from factorization of theoretical spectra, and then iteratively adjusted to fit data. 
-We apply these methods to \project{Sloan Digital Sky Survey} optical spectra 
-%of stars spanning evolutionary stages 
-from pre-main-sequence stars to white dwarfs.
-In a model-driven or hybrid approach, we show that the basis weights can reliably estimate stellar parameters and chemical abundances.
-%The data-driven approach is less physically interpretable, but naturally provides the best description for the data and is an ideal no-model solution when only a few spectra are available.
-%We find good fits to data, even for the model-driven variant where theoretical spectra used to factorize line absorption is provably incorrect for some features.
-We show that the basis weights can be used to reliably estimate stellar parameters and chemical abundances, 
-and that we obtain low bias estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
-%This approach largely eliminates subjectivity in stellar continuum normalization.
+Our approach factorizes line absorption entirely from continuum-normalized theoretical spectra, but we also describe a data-driven variant to iteratively correct model-data mismatches.
+We show that the basis weights can also be used to reliably estimate stellar parameters, and apply these methods to \project{Sloan Digital Sky Survey} infrared spectra of OBAFGKM-type stars.
+This approach largely eliminates subjectivity in stellar continuum normalization.
 \end{abstract}
@@
 This process is ripe with subjectivity.
-While there is general agreement in the literature that consistent continuum normalization is important, there is apparently no consensus on how it should be done.\\
+While there is general agreement in the literature that consistent continuum normalization is important, there is no apparent consensus on how it should be done.\\
 
-Fitting the continuum well requires you to know where there is line absorption.
+Fitting the continuum correctly requires you to know where there is line absorption.
 But knowing where there is line absorption requires you to (at least) know the stellar parameters.
-Without knowing the stellar parameters (or having a good model for line absorption), spectroscopists have been forced to use asymmetric so-called sigma clipping (i.e., to exclude data points points $N\sigma$ below a continuum fit but keep points above the fit), or restrict themselves to a carefully selected set of so-called continuum pixels.
-The set of continuum pixels is only valid for stars of a similar spectral type and metallicity, and in many cases there are \emph{no continuum pixels} (e.g., M-dwarfs).\\
+Without knowing the stellar parameters -- or having a good model for line absorption -- spectroscopists have developed various bespoke methods.
+A popular choice is to iteratively mask pixels in an asymmetric manner (so-called `sigma clipping') to exclude data points some chosen level below the current estimate of the continuum, or to mask all pixels except a carefully selected set of so-called continuum pixels.
+However, the set of continuum pixels is only valid for stars of a similar spectral type and metallicity, and in many cases there are \emph{no} continuum pixels (e.g., M-dwarfs).\\
 
@@
 This process is still often done by hand (albeit they are often experienced hands).
-Industrial spectroscopists might simply aim for a \emph{consistent} continuum normalization procedure, even if the process only estimates a pseudo-continuum for comparison, and not the true continuum.
-In an ideal scenario the pseudo-continuum is fit jointly with the stellar parameters, but this is often considered too expensive.\\
+Industrial spectroscopists might aim for a \emph{consistent} continuum normalization procedure: one that estimates a reliable continuum level for stars of similar stellar parameters and signal-to-noise ratios, even if that continuum estimate is some pseudo-continuum, and not the true continuum.\\
 
-In Section~\ref{sec:methods} we describe a family of methods to address this problem.
-Each variant makes use of non-negative matrix factorization (NMF) to approximate line absorption.
+In an ideal scenario the continuum might be fit jointly with the stellar parameters, but this is often too expensive if it is costly to predict the emergent spectrum.
+In Section~\ref{sec:methods} we describe a family of methods to address these problems.
+Each of the three variants we describe use non-negative matrix factorization (NMF) to approximate line absorption.
 NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements.
 This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
-The three variants of our method are each conceptually introduced, with the intended application (Section~\ref{sec:results}) to low-resolution optical spectra from the \emph{Sloan Digital Sky Survey}.
+The three variants of our method are each conceptually introduced, with the intended application (Section~\ref{sec:results}) to low-resolution optical spectra from the \project{Sloan Digital Sky Survey} (\sdss).
 We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\
@@
 
-We will assume a forward model that includes two components: one represent continuum-normalized absorption (e.g., line absorption); and another to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
-These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will assume both components to be linear models.
-This is not a strict requirement, but keeping linearity ensures that inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization.\\
+We will assume a forward model that includes two components: one represents continuum-normalized absorption (e.g., line absorption); and another to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
+These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will assume both components to be linear models, ensuring that inference is fast and stable.
+In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and for estimating stellar parameters.\\
 
-Here we will describe the method in general before outlining assumptions and details of individual variants.
-In all variants we assume that non-negative matrix factorization (NMF) is sufficient to represent continuum-normalized line absorption in a large variety of stellar spectra.
-We further assume that the sine-and-cosine basis is sufficiently flexible for modelling the joint instrument-continuum response.
+%In all variants we assume that non-negative matrix factorization (NMF) is sufficient to represent continuum-normalized line absorption in a large variety of stellar spectra.
+
+% TODO: list explicit assumptions with hogg in person
+
+%We further assume that the sine-and-cosine basis is sufficiently flexible for modelling the joint instrument-continuum response.
 There are choices to be made in terms of the degree and length scale used for the sine-and-cosine basis, but these are application-specific choices.\\
 
-% We could list all other assumptions (flat universe), but are there any truly necessary assumptions that we also need to list for the *astronomer* or *spectroscopist* audience?
-
-The problem of continuum normalization for a single spectrum can be described where the data are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
-The forward model for these data can be expressed as the element-wise multiplication of a line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
+Here we will describe the method in general before outlining explicit assumptions and implementation details.
+The problem of continuum normalization for a single spectrum can be described where the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
+The forward model for these data can be expressed as the element-wise multiplication of what we will describe as the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response model $g(\lambda_i;\vectheta)$
 \begin{align}
@@
 \end{align}
-where $\hadamard$ represents the element-wise (Hadamard) product and the components $f(...)$ and $g(...)$ are described below.\\
+where $\hadamard$ represents the element-wise (Hadamard) product and the components $f(...)$ and $g(...)$ are defined below.\\
 
-The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a set of $N$ continuum-normalized spectra each with $D$ fluxes.
-In the applications that we present, the data are assumed to have the same wavelength sampling and line spread function as the spectra used to approximate the line absorption model, such that we use a grid where $D = P$.
-Other situations are permitted, but at inference time there is a need to interpolate the line absorption model to the $P$ observed pixels.
-The three variants of our method that we describe here have different sets of continuum-normalized spectra, but the process of factorizing them is the same.
-With our $N \times\ D$ matrix $\vec{M}$ of continuum-normalized fluxes, we then define \emph{stellar absorption} $\vecX$ to be
+The line absorption model $f(\lambda_i;\vecpsi)$ predicts the line absorption at wavelength $\lambda_i$ given parameters $\vecpsi$.
+We construct the line absorption model $f(\lambda_i;\vecpsi)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorization (NMF).
+In the applications that we present here, the theoretical spectra used to approximate the line absorption model are assumed to have the same wavelength sampling and line spread function as the data, such that we use a grid of model spectra where $D = P$.
+Other situations are permitted, but at inference time there is a need to evaluate the line absorption model to the $P$ observed wavelengths.
+With our $N \times\ D$ matrix of continuum-normalized theoretical spectra, which we will call $\vec{M}$, we then define \emph{stellar absorption} $\vecX$ to be
 \begin{align}
@@
 \end{align}
-where all elements in $\vecX$, $\vec{W}$, and $\vec{H}$ are required to be non-negative. 
-We could construct an approximation directly to $\vec{M}$, but this would lead to denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX$ that are 0 (i.e., no line absorption).
-The nomenclature in this paper draws on previous works on NMF and data analysis more broadly.
+where $\vecX$ is a $N \times D$ matrix, $\vecW$ is a $N \times C$ matrix of NMF \emph{basis weights} with $C$ being the number of basis vectors to use, and $\vecH$ is a $C \times D$ matrix of NMF \emph{basis vectors}.
+For this factorization we must select the number of basis components $C$ to use, which should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
+All elements in $\vecX$, $\vec{W}$, and $\vec{H}$ are required to be non-negative.
+Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example basis vectors $\vec{H}$.\\
+
+We could construct an approximation directly to $\vec{M}$, but this would lead to denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX = 1 - \vec{M}$ that are 0 (i.e., no line absorption).
+The choice of nomenclature in this paper (i.e., $\vecX$, $\vec{W}$, $\vec{H}$) is chosen to be consistent with previous works on NMF.\\
+
+%on previous works on NMF and data analysis more broadly.
 We would like the naming conventions to be consistent with those papers, but this nomenclature is easily overloaded.
@@
 
-For this factorization we must select the number of basis components $C$ to use, which should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
-Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ basis weights per spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding basis vectors each with $D$ pixels.
-Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example basis vectors $\vec{H}$.\\
+%Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ basis weights per spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding basis vectors each with $D$ pixels.
 
@@
 We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $\vecX$: this is the default behaviour in the Python \texttt{scipy.decomposition.NMF} implementation.
+As reported elsewhere, we found that factorizing $\vecW\vecH$ by coordinate descent was substantially faster than using multiplicative updates.
 \\
-%As reported elsewhere, we found that factorizing $\vecW\vecH$ by coordinate descent was substantially faster than using multiplicative updates, but this was only relevant when the input spectra have no flux errors (i.e., the theory-only variant).
+%For the applications presented in this paper, the factorization can be completed in the order of minutes.
+
+%.
 Factorizing the data-driven and hybrid approaches is described later in this section.
@@
 There are many suitable choices for the continuum-instrument response model $g(\lambda_i;\vectheta)$.
-Here we chose a sine-and-cosine basis because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
+Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
 The component $g(\lambda_i;\vectheta)$ is expressed compactly as
@@
 \end{align}
-where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column are,
+where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
 \begin{align}
@@
 \noindent{}and the design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
-Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.
-In this formalism, spectral regions can be fit independently with different degrees of freedom.
+Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.\\ %In this formalism, spectral regions can be fit independently with different degrees of freedom.
 The number of continuum coefficients scales as $2n_\textrm{degree} + 1$ per region.
+\\
 
@@
 In situations where there are spectra of the same star we advocate fitting all visit spectra simultaneously using a single set of basis weights $\vecpsi$ for all spectra, and continuum coefficients $\vectheta_i$ for the $i$th epoch spectrum.
-
-\subsection{Data-driven approach: Line absorption is factorized by data}
-\label{sec:data-method}
-
-In this variant we solve for the factorized line absorption and continuum entirely from data, without any theoretical spectra.
-The model concept is exactly the same as introduced in Section~\ref{sec:methods}: we have a forward model for the flux that depends on basis weights $\vecpsi$ and continuum coefficients $\vectheta$.
-In this variant the principal difference is how we factorize the basis vectors $\vecH$.\\
-
-The process of fitting the model is the same for one spectrum or many.
-The procedure involves alternating through three linear operations.
-Before starting we assume that $\vecW$ and $\vecH$ are initially filled with zeros such that there is no line absorption.
-Then we alternate between these three steps:
-\begin{enumerate}
-    \item Solve for the continuum coefficients $\vectheta$ per spectrum, given the (i.e., keeping fixed) current estimates of $\vecW$ and $\vecH$.
-    \item Update the basis vectors $\vecH$, given the current estimate of the continuum coefficients $\vectheta$ and the basis weights $\vecW$.
-    \item Update the basis weights $\vecW$ per spectrum, given the current estimate of the continuum coefficients $\vectheta$ and the basis vectors $\vecH$.
-\end{enumerate}
-Completing all three steps is described as one iteration, and we continue iterating until either a fixed number of iterations is reached, or a tolerance in $\chi^2$ is achieved.
-Each step here is a \emph{convex linear operation} that is guaranteed to improve the $\chi^2$ fit, leading to an approach is extremely stable, fast, and guaranteed to converge.\\
-
-An illustration of this variant is shown in Figure~\ref{fig:schematic-data-driven}.
-With no line absorption initially, the continuum is set to fit all of the data.
-But with that (poor) estimate of the continuum, one can calculate the continuum-normalized flux and compute a better estimate of the basis vectors $\vecH$.
-Those basis vectors essentially have the right `shape', in that they represent line absorption at the appropriate wavelengths, but they are too low: the true absorption needs to be stronger.
-With the continuum and the basis vectors now fixed $\vecH$, we can update our estimate of the basis weights $\vecW$ per spectrum.
-This process continues for the number of iterations specified by the user, or after some defined tolerance in $\chi^2$ is reached.\\
-
-In this (and the hybrid) approach, the updated estimates of $\vecW$ and $\vecH$ are found by multiplicative updates, whereas in the model-only approach, coordinate descent is used to factorize the line absorption.
-Here we will let
-\begin{align}
-    \vec{Y} = 1 - \frac{\vec{y}}{g(\vec{\lambda};\vectheta)}
-\end{align}
-\noindent{}be the line absorption of observed spectra, which is computed given our current estimate of the continuum normalization $g(\lambda; \vectheta)$, and $\vec{Z}$ be the corresponding inverse variances
-\begin{align}
-    \vec{Z} = \left(g(\vec{\lambda};\vectheta)/\vec{\sigma_{y}}\right)^2 
-\end{align}
-such that we can compute our updated estimate of the basis vectors $\vecH$ (step 2) 
-\begin{align}
-    \vecH_{ij}^{\mathrm{(new)}} \leftarrow \vecH_{ij} \frac{\left[\vecW\transpose\left(\vec{Y} \hadamard \vec{Z}\right)\right]_{ij}}{\left[\vecW\transpose\left(\vecW\vecH \hadamard \vec{Z}\right)\right]_{ij}}
-\end{align}
-\noindent{}while properly accounting for the inverse variance in each pixel. We can similarly update the basis weights $\vecW$ (step 3)
-\begin{align}
-    \vecW_{ij}^{\mathrm{(new)}} \leftarrow \vecW_{ij} \frac{\left[\left(\vec{Y} \hadamard \vec{Z}\right)\vecH\transpose\right]_{ij}}{\left[\left(\vecW\vecH\hadamard\vec{Z}\right)\vecH\transpose\right]_{ij}}
-\end{align}
-\noindent{}where $\left[\vec{AB}\right]_{ij}$ indicates the $(i,j)$ element in the $\vec{AB}$ matrix.\\
-
-This variant works well if you have only one spectrum, or if you have many.
-If you only have one spectrum (or a few) then this is a good generalized approach for continuum normalization, which does not require many decisions from the user.
-One restriction is that $C$ should be set to at most the number of spectra being fit (i.e., do not use $C = 2$ basis vectors to fit 1 spectrum).
-However, in these situations with only few spectra and no theoretical spectra for factorization, the downside is that the basis weights $\vecpsi$ are uninterpretable.
-
-\begin{figure*}
-    \caption{A schematic illustrating the data-driven variant of our method.\label{fig:schematic-data-driven}}
-\end{figure*}
-
-\subsection{Hybrid approach: Line absorption is factorized by theoretical spectra, and iteratively improved by data}
-\label{sec:hybrid-method}
-
-The last variant we present in this paper is a hybrid of the two previous methods.
-The model-driven approach has the benefit that the basis weights and basis vectors can have physical interpretations.
-For example, a basis vector containing only strong, broad hydrogen lines would imply that any observed spectrum with a large basis weight for that vector might be a hot star.
-However, limitations in the theoretical model grid become readily apparent when using only a model-driven approach: there are entire molecular bands that are not represented in the theoretical model spectra.
-And while the data-driven approach naturally provides an excellent description of the data, it is less interpretable.
-This hybrid variant combines the benefits of both scenarios.
-\\
-
-While the iterative steps in Section~\ref{sec:data-method} are each convex linear operations, the actual factorization of $\vec{X} = \vec{WH}$ is not convex: there are many combinations of $\vecW$ and $\vecH$ that produce the same result.
-For this reason, the solution of $\vecW$ and $\vecH$ depends on the initial starting point.
-Normally this is not a problem because we don't necessarily care \emph{which} solution we have to the factorization, but we do care if we want to gain physical insight from the basis weights $\vecW$ or basis vectors $\vecH$.\\
-
-In this variant we use a theoretical grid and compute the basis vectors $\vecH$ just as described in Section~\ref{sec:model-method}.
-We then take $N >> C$ observed spectra and fit the basis weights $\vecpsi$ and continuum coefficients $\vectheta$ for each spectrum, keeping $\vecH$ fixed.
-At this point the variant is exactly the same as the model-driven method.
-Now we start iterating between the steps described in Section~\ref{sec:data-method}, starting from Step 2: we update the basis vectors $\vecH$ given the current estimate of the continuum coefficients $\vectheta$ and the basis weights $\vecW$.
-Next we proceed to Step 3, where we update the basis weights $\vecW$ given the current estimate of the continuum coefficients $\vectheta$ and the basis vectors $\vecH$.
-We continue cycling through Steps 1-3 a pre-determined number of times or until a tolerance in $\chi^2$ is reached.\\
-
-In the applications we present here, this variant only tends to adjust \emph{some} of the basis vectors $\vecH$ (and corresponding weights $\vecW$).
-For example, absorption features that were poorly modelled in the theoretical grid quickly appear in the updated basis vectors in order to better match the data.
-Overall, most basis vectors do not change considerably, which keeps the basis weights and basis vectors (somewhat) interpretable.
-This step can be altered to only update specific vectors (e.g., those that look like molecular absorption, if molecular absorption is poorly modelled by theory), or specific wavelengths, but this experimentation is outside the scope of this work.
 

### 2024-06-25 applications/apogee/20240517.py
lots of shit

@@ -276,7 +276,6 @@ class Clam(object):
         fit_vsini=False,
         x0_params=None,
         x0_vsini=100,
-        x0=None,
         **kwargs
     ):
         try:
@@ -284,7 +283,7 @@ class Clam(object):
         except:
             self.A1 = self.full_design_matrix(1)
             
-        if x0_params is None and x0 is None:
+        if x0_params is None:
             _, meta_init = self.fit(flux, ivar, full_output=True)
 
             x0_params, = random_forest_regressor.predict(meta_init["W"].reshape((1, -1)))
@@ -309,32 +308,30 @@ class Clam(object):
             C, P = self.components.shape
             A_slice = A[:, C:]
             
-            if x0 is None:
-                # given x0_params, predict initial W coefficients.
-                W_init = np.clip(basis_weight_interpolator(x0_params), 0, np.inf)
-                init_rectified_model_flux = (1 - W_init @ self.components).flatten()
-                
-                '''
-                x02 = self.get_initial_guess_with_small_W(
-                    flux / init_rectified_model_flux, 
-                    ivar * init_rectified_model_flux**2,
-                    A, 
-                    small=1e-10
-                )
-                '''
-                # TODO: put to function
-                use = ~self.get_mask(ivar * init_rectified_model_flux**2).flatten()
-                Y = (flux.flatten() / init_rectified_model_flux)[use]
-                C_inv = np.diag((ivar[0] * init_rectified_model_flux**2)[use])
-                
-                B = A[use, C:]
-                x0_continuum = np.linalg.solve(B.T @ C_inv @ B, B.T @ C_inv @ Y)
-                                                    
-                x0 = np.hstack([
-                    x0_params, # stellar params  
-                    x0_continuum # continuum params,
-                ])
+            # given x0_params, predict initial W coefficients.
+            W_init = np.clip(basis_weight_interpolator(x0_params), 0, np.inf)
+            init_rectified_model_flux = (1 - W_init @ self.components).flatten()
             
+            '''
+            x02 = self.get_initial_guess_with_small_W(
+                flux / init_rectified_model_flux, 
+                ivar * init_rectified_model_flux**2,
+                A, 
+                small=1e-10
[…]

### 2024-06-25 applications/apogee/20240530.py

@@ -3,7 +3,7 @@ import numpy as np
 import warnings
 from functools import partial, cached_property
 from scipy.linalg import lu_factor, lu_solve
-from typing import Sequence, Optional, NamedTuple
+from typing import Sequence, Optional, NamedTuple, Union
 
 from jax import (jit, jacfwd as jacobian, numpy as jnp, scipy as jsp)
 from interpolate import RegularGridInterpolator
@@ -19,8 +19,10 @@ class ClamFit(NamedTuple):
     label_names: Sequence[str]
     stellar_parameters: Sequence[float]
             
+    p_init: Sequence[float]
     p_opt: Sequence[float]
     p_cov: np.array
+    
     rchi2: float
     dof: int
 
@@ -43,9 +45,9 @@ class BaseClam:
         grid_points: Sequence[Sequence[float]],
         modes: Optional[int] = 7,
         regions: Optional[Sequence[Sequence[float]]] = (
-            (15161.84316643 - 35, 15757.66995776 + 60),
-            (15877.64179911 - 25, 16380.98452330 + 60),
-            (16494.30420468 - 30, 16898.18264895 + 60)
+            (15120.0, 15820.0),
+            (15840.0, 16440.0),
+            (16450.0, 16960.0),
         ),
         **kwargs
     ):
@@ -68,8 +70,12 @@ class BaseClam:
             The grid points of the stellar parameters. This should be a `L`-length list of arrays,
             where `L` is the number of stellar parameters. Each list should be ordered.
         
-        :param modes: [optional]
+        :param modes:
             The number of Fourier modes to use to model the continuum (in each region).
+            
+        :param regions:
+            The regions to model the continuum. This should be a list of tuples where each tuple
+            is a `(start, end)` wavelength region.
         """
         if modes % 2 == 0:
             raise ValueError("modes must be an odd number")
@@ -92,13 +98,13 @@ class BaseClam:
             fill_value=0
         )
 
-        regions = regions or [tuple(λ[[0, -1]])]
+        self._regions = regions or [tuple(λ[[0, -1]])]
                         
         # Construct design matrix.        
-        self._continuum_design_matrix = np.zeros((λ.size, len(regions) * modes), dtype=float)
-        for i, region_slice in enumerate(region_slices(λ, regions)):
+        self._continuum_design_matrix = np.zeros((λ.size, len(self._regions) * modes), dtype=float)
+        for i, region_slice in enumerate(region_slices(λ, self._regions)):
[…]

### 2024-06-25 applications/apogee/20240604.py

@@ -5,7 +5,8 @@ from functools import partial, cached_property
 from scipy.linalg import lu_factor, lu_solve
 from typing import Sequence, Optional, NamedTuple, Union
 
-from jax import (jit, jacfwd, numpy as jnp, scipy as jsp)
+#from jax import (jit, jacfwd, numpy as jnp, scipy as jsp)
+from numba import jit
 from interpolate import RegularGridInterpolator as JaxRegularGridInterpolator
 from scipy.interpolate import RegularGridInterpolator
 
@@ -439,7 +440,7 @@ class BaseClam:
 class Clam(BaseClam):
     
     """A constrained linear absorption model."""
-    
+
     def __call__(self, θ: Sequence[float]):
         """
         Predict the flux given the stellar parameters and the continuum parameters.
@@ -455,10 +456,10 @@ class Clam(BaseClam):
             `continuum_design_matrix` attribute, as that matrix should have shape `(P, C)` where 
             `P` is the number of pixels.
         """
-        return (1 - self.interpolator(θ[:self.n_labels], method="slinear") @ self.H) * (self.continuum_design_matrix @ θ[self.n_labels:])
+        return (1 - self.interpolator(θ[:self.n_labels], method="slinear").flatten() @ self.H) * (self.continuum_design_matrix @ θ[self.n_labels:])
     
         
-    
+    #@jit #@partial(jit)# static_argnums=(0,))    
     def jacobian(self, θ: Sequence[float]):
         """
         Compute the Jacobian of the predicted model with respect to the model parameters.
@@ -748,7 +749,6 @@ if __name__ == "__main__":
     fig, ax = plt.subplots()
     ax.scatter(g1.flatten(), g2.flatten())
     
-    raise a
 
     pleiades_results = []
     for index, item in tqdm(pleades):        
@@ -756,6 +756,8 @@ if __name__ == "__main__":
         result = model.fit(flux[index], ivar[index])
         
         pleiades_results.append((index, result, item))
+        
+    raise a
 
 
     e_m_h = [e[1].p_cov[3, 3]**0.5 for e in pleiades_results]

### 2024-06-25 applications/apogee/20240619_linear_model.py

@@ -1,0 +1,570 @@
+import h5py as h5
+import os
+import numpy as np
+from sklearn.decomposition import NMF
+import pickle
+from tqdm import tqdm
+
+from lsf import rotational_broadening_sparse_matrix
+
+
+SLICE_ON_C_AND_N = True
+SLICE_ON_N_ONLY = False
+
+#NMF_PATH = "20240619_vsini.h5"
+NMF_PATH = "20240624_vsini.h5"
+
+λ = 10**(4.179 + 6e-6 * np.arange(8575))
+
+
+if os.path.exists(NMF_PATH):
+    with h5.File(NMF_PATH, "r") as fp:
+        label_names = fp["label_names"][:].astype(str)
+        W = fp["W"][:]
+        H = fp["H"][:]
+        grid_points = [fp[f"grid_points/{str(ln)}"][:] for ln in label_names]
+        grid_model_flux = fp["grid_model_flux"][:]
+
+else:
+    with h5.File("big_grid_2024-12-20.h5", "r") as fp:
+        label_names = ["Teff", "logg", "vmic", "m_h", "c_m", "n_m"][::-1]
+        grid_points = [fp[f"{ln}_vals"][:] for ln in label_names]
+        grid_model_flux = fp["spectra"][:]
+
+        print(grid_model_flux.shape)
+        print(tuple(map(len, grid_points)))
+
+        if sum([SLICE_ON_N_ONLY, SLICE_ON_C_AND_N]) > 1:
+            raise ValueError("Only one of SLICE_ON_N_ONLY or SLICE_ON_C_AND_N can be True")
+
+        if SLICE_ON_N_ONLY:
+            # Take a slice
+            n_m = 0
+            n_m_index = list(grid_points[label_names.index("n_m")]).index(n_m)
+
+            label_names = label_names[1:]
+            grid_points = grid_points[1:]
+            grid_model_flux = grid_model_flux[n_m_index]
+                
+        
+        elif SLICE_ON_C_AND_N:
+                
+            # Take a slice
+            c_m = n_m = 0
+            c_m_index = list(grid_points[label_names.index("c_m")]).index(c_m)
+            n_m_index = list(grid_points[label_names.index("n_m")]).index(n_m)
+
+            label_names = label_names[2:]
[…]

### 2024-06-25 applications/apogee/20240624_lm.py

@@ -1,0 +1,111 @@
+import h5py as h5
+import os
+import numpy as np
+from sklearn.decomposition import NMF
+import pickle
+from tqdm import tqdm
+
+from lsf import rotational_broadening_sparse_matrix
+
+
+SLICE_ON_C_AND_N = True
+SLICE_ON_N_ONLY = False
+
+NMF_PATH = "20240624.h5"
+
+λ = 10**(4.179 + 6e-6 * np.arange(8575))
+
+
+if os.path.exists(NMF_PATH):
+    with h5.File(NMF_PATH, "r") as fp:
+        label_names = fp["label_names"][:].astype(str)
+        W = fp["W"][:]
+        H = fp["H"][:]
+        grid_points = [fp[f"grid_points/{str(ln)}"][:] for ln in label_names]
+        grid_model_flux = fp["grid_model_flux"][:]
+
+else:
+    with h5.File("big_grid_2024-12-20.h5", "r") as fp:
+        label_names = ["Teff", "logg", "vmic", "m_h", "c_m", "n_m"][::-1]
+        grid_points = [fp[f"{ln}_vals"][:] for ln in label_names]
+        grid_model_flux = fp["spectra"][:]
+
+        print(grid_model_flux.shape)
+        print(tuple(map(len, grid_points)))
+
+        if sum([SLICE_ON_N_ONLY, SLICE_ON_C_AND_N]) > 1:
+            raise ValueError("Only one of SLICE_ON_N_ONLY or SLICE_ON_C_AND_N can be True")
+
+        if SLICE_ON_N_ONLY:
+            # Take a slice
+            n_m = 0
+            n_m_index = list(grid_points[label_names.index("n_m")]).index(n_m)
+
+            label_names = label_names[1:]
+            grid_points = grid_points[1:]
+            grid_model_flux = grid_model_flux[n_m_index]
+                
+        
+        elif SLICE_ON_C_AND_N:
+                
+            # Take a slice
+            c_m = n_m = 0
+            c_m_index = list(grid_points[label_names.index("c_m")]).index(c_m)
+            n_m_index = list(grid_points[label_names.index("n_m")]).index(n_m)
+
+            label_names = label_names[2:]
+            grid_points = grid_points[2:]
+            grid_model_flux = grid_model_flux[n_m_index, c_m_index]
+        
+        # build some rotational kernels
+        #vsinis = np.logspace(0, 2, 6)
[…]

### 2024-06-25 applications/apogee/20240625_exp1.py

@@ -1,0 +1,114 @@
+
+# Here we are going to do just the turn-off:
+# 7000 > teff > 4000 and logg > 3
+# with a vsini dimension
+# try linear and cubic interpolation to see what works
+# with a log transformation of the flux
+
+import h5py as h5
+import os
+import numpy as np
+import pickle
+from sklearn.decomposition import NMF
+from tqdm import tqdm
+
+from lsf import rotational_broadening_sparse_matrix
+
+
+NMF_PATH = f"{os.path.basename(__file__)[:-3]}.h5"
+
+λ = 10**(4.179 + 6e-6 * np.arange(8575))
+
+if os.path.exists(NMF_PATH):
+    with h5.File(NMF_PATH, "r") as fp:
+        label_names = fp["label_names"][:].astype(str)
+        W = fp["W"][:]
+        H = fp["H"][:]
+        grid_points = [fp[f"grid_points/{str(ln)}"][:] for ln in label_names]
+        grid_model_flux = fp["grid_model_flux"][:]
+
+else:
+    with h5.File("big_grid_2024-12-20.h5", "r") as fp:
+        label_names = ["Teff", "logg", "vmic", "m_h", "c_m", "n_m"][::-1]
+        grid_points = [fp[f"{ln}_vals"][:] for ln in label_names]
+        grid_model_flux = fp["spectra"][:]
+
+
+        if sum([SLICE_ON_N_ONLY, SLICE_ON_C_AND_N]) > 1:
+            raise ValueError("Only one of SLICE_ON_N_ONLY or SLICE_ON_C_AND_N can be True")
+            
+        # Take a slice
+        c_m = n_m = 0
+        c_m_index = list(grid_points[label_names.index("c_m")]).index(c_m)
+        n_m_index = list(grid_points[label_names.index("n_m")]).index(n_m)
+
+        label_names = label_names[2:]
+        grid_points = grid_points[2:]
+        grid_model_flux = grid_model_flux[n_m_index, c_m_index]
+    
+        # Restrict to a subset
+        logg_index = 2
+        teff_index = 3
+        assert label_names[teff_index] == "Teff"
+        assert label_names[logg_index] == "logg"
+
+        logg_mask = grid_points[logg_index] > 3
+        teff_mask = (7000 > grid_points[teff_index]) * (grid_points[teff_index] > 4000)
+
+        grid_model_flux = grid_model_flux[:, :, logg_mask]
[…]

### 2024-06-25 applications/apogee/jax_ndimage.py

@@ -1,0 +1,216 @@
+# Copyright 2019 The JAX Authors.
+#
+# Licensed under the Apache License, Version 2.0 (the "License");
+# you may not use this file except in compliance with the License.
+# You may obtain a copy of the License at
+#
+#     https://www.apache.org/licenses/LICENSE-2.0
+#
+# Unless required by applicable law or agreed to in writing, software
+# distributed under the License is distributed on an "AS IS" BASIS,
+# WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
+# See the License for the specific language governing permissions and
+# limitations under the License.
+
+
+import functools
+import itertools
+import operator
+from typing import Callable, Dict, List, Sequence, Tuple
+
+
+from jax._src import api
+from jax._src import util
+from jax import lax
+from jax._src.lax.lax import PrecisionLike
+import jax.numpy as jnp
+import jax.scipy as jsp
+from jax._src.typing import ArrayLike, Array
+from jax._src.util import safe_zip as zip
+
+
+def _nonempty_prod(arrs: Sequence[Array]) -> Array:
+    return functools.reduce(operator.mul, arrs)
+
+
+def _nonempty_sum(arrs: Sequence[Array]) -> Array:
+    return functools.reduce(operator.add, arrs)
+
+
+def _mirror_index_fixer(index: Array, size: int) -> Array:
+    s = size - 1  # Half-wavelength of triangular wave
+    # Scaled, integer-valued version of the triangular wave |x - round(x)|
+    return jnp.abs((index + s) % (2 * s) - s)
+
+
+def _reflect_index_fixer(index: Array, size: int) -> Array:
+    return jnp.floor_divide(_mirror_index_fixer(2 * index + 1, 2 * size + 1) - 1, 2)
+
+
+_INDEX_FIXERS: Dict[str, Callable[[Array, int], Array]] = {
+    'constant': lambda index, size: index,
+    'nearest': lambda index, size: jnp.clip(index, 0, size - 1),
+    'wrap': lambda index, size: index % size,
+    'mirror': _mirror_index_fixer,
+    'reflect': _reflect_index_fixer,
+}
+
+
+def _round_half_away_from_zero(a: Array) -> Array:
+    return a if jnp.issubdtype(a.dtype, jnp.integer) else lax.round(a)
+
+
+def _nearest_indices_and_weights(coordinate: Array) -> List[Tuple[Array, ArrayLike]]:
[…]

### 2024-06-25 applications/apogee/lsf_jax.py

@@ -1,0 +1,76 @@
+from typing import Sequence
+import jax
+from jax import numpy as jnp
+import numpy as np
+#from scipy import sparsefrom jax.e
+from jax.experimental.sparse import BCOO
+
+def rotational_broadening_kernel(λ: Sequence[float], vsini: float, epsilon: float):
+    """
+    Construct a sparse matrix to convolve fluxes at input wavelengths (λ) with a rotational broadening kernel
+    with a given vsini and epsilon.
+
+    :param λ:
+        A N-length array of input wavelength values.
+
+    :param vsini:
+        The projected rotational velocity of the star in km/s.
+
+    :param epsilon:
+        The limb darkening coefficient.
+
+    :returns:
+        A (N, N) sparse array representing a convolution kernel.
+    """
+
+    # Let's pre-calculate some things that are needed in the hot loop.
+    denominator = np.pi * vsini * (1.0 - epsilon / 3.0)
+    c1 = 2.0 * (1.0 - epsilon) / denominator
+    c2 = 0.5 * np.pi * epsilon / denominator
+
+    vsini_c = vsini / 299792.458
+    scale = vsini_c / (λ[1] - λ[0])  # assume uniform sampling
+    N = λ.size
+
+    data, row_index, col_index = ([], [], [])
+    indices = []
+    for i, λ_i in enumerate(λ):
+        n_pix = int(jnp.ceil(λ_i * scale))
+        si, ei = (max(0, i - n_pix), min(i + n_pix + 1, N))  # ignoring edge effects
+
+        λ_delta_max = λ_i * vsini_c
+        λ_delta = λ[si:ei] - λ_i
+        #λ_ratio_sq = jnp.clip((λ_delta / λ_delta_max) ** 2.0, 0, 1 - 1e-5) # check this
+        λ_ratio_sq = (λ_delta / λ_delta_max) ** 2.0
+        assert np.all(1 >= λ_ratio_sq)
+        
+        # approximation of sqrt(1 - x^2)
+        
+        
+        ϕ = c1 * jnp.sqrt(1.0 - λ_ratio_sq) + c2 * (1.0 - λ_ratio_sq)
+        #ϕ[λ_ratio_sq >= 1.0] = 0.0  # flew too close to the sun 
+        ϕ /= jnp.sum(ϕ)
+        data.extend(ϕ)
[…]

### 2024-06-25 applications/apogee/plot_cubic_vsini_diffs.py

@@ -1,0 +1,34 @@
+
+import matplotlib.pyplot as plt
+import numpy as np
+from astropy.table import Table
+
+paths = [
+    ("julia_cubic.csv", "Cubic spline and vsini"),
+    ("julia_cubic_no_vsini.csv", "Cubic spline and no vsini"),
+    ("julia_linear.csv", "Linear interpolation and vsini"),
+    ("julia_linear_no_vsini.csv", "Linear interpolation and no vsini")
+]
+
+xlim = (8000, 3000)
+ylim = (5.5, 0)
+
+fig, axes = plt.subplots(2, 2)
+
+for ax, (path, title) in zip(axes.flat, paths):
+    
+    names = [f"c{i}" for i in range(25)]
+    try:
+        table = Table.read(path, names=names)
+    except:
+        names.append("c26")        
+        table = Table.read(path, names=names)
+        
+    ax.scatter(table["c0"], table["c1"], c=table["c3"], s=5, alpha=0.5)
+    
+    ax.set_xlabel("Teff")
+    ax.set_ylabel("logg")
+    ax.set_xlim(xlim)
+    ax.set_ylim(ylim)
+    ax.set_title(title)
+fig.tight_layout()

### 2024-06-25 paper/main.tex
reword method + abstract

@@
 %\newcommand{\chosentitle}{Constrained linear models for stellar spectroscopy}
-\newcommand{\chosentitle}{Stellar continuum modelling}
+%\newcommand{\chosentitle}{Constrained linear absorption models for stellar spectroscopy}
+\newcommand{\chosentitle}{The Unreasonable Effectiveness of Linear Models in Stellar Spectroscopy}
+%\newcommand{\chosentitle}{Stellar continuum modelling}
 
@@
 \definecolor{tab:blue}{HTML}{1170aa}
-\newcommand{\todo}[1]{\textcolor{tab:blue}{#1}}
+\definecolor{tab:red}{HTML}{d1615d}
+\newcommand{\todo}[1]{\textcolor{tab:red}{#1}}
 
@@
 \affiliation{Centre of Excellence for Astrophysics in Three Dimensions (ASTRO-3D)}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute, a division of the Simons Foundation}
 
@@
 \affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
-\affiliation{Flatiron Institute, a division of the Simons Foundation}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute, a division of the Simons Foundation}
 
 % Other people who will be invited as co-authors (non-exhaustive list):
-% Wheeler, Bedell, Astra folks, etc.
+%   Wheeler, Sayjdari, Bedell, Astra folks, CCA data group
 
 \begin{abstract}\noindent
-Continuum normalization is often a necessary step when analyzing stellar spectra. 
-The ideal approach might be to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is often required before those parameters are estimated.
-Here we present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization that models stellar line absorption, and a sines-and-cosines basis to model the joint continuum-instrument response.
-The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the remaining variance.
-The linearity of both components ensures that inference is stable and fast.
-Our approach factorizes line absorption entirely from continuum-normalized theoretical spectra, but we also describe a data-driven variant to iteratively correct model-data mismatches.
-We show that the basis weights can also be used to reliably estimate stellar parameters, and apply these methods to \project{Sloan Digital Sky Survey} infrared spectra of OBAFGKM-type stars.
-This approach largely eliminates subjectivity in stellar continuum normalization.
+Forward modelling stellar spectra usually requires a spectral synthesis code, and/or a non-linear interpolator constructed from a curated training set.
+These approaches also require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent inferences.
+Here we present a \emph{linear} model that simultaneously fits stellar absorption and the joint continuum-instrument response.
+%The stellar absorption is modelled by non-negative matrix factorization of a grid of theoretically rectified spectra, and a sine-and-cosine basis to model the joint continuum-instrument response.
+%The linearity inference is stable and fast.
+Stellar absorption is modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices (basis weights and basis vectors) with basis vectors kept fixed at inference time, and the joint continuum-instrument response is modelled with a sine-and-cosine basis.
+The non-negativity constraint ensures that basis vectors are strictly additive, and by maintaining linearity we ensure that inference is convex, stable, and fast.
+When we restructure our forward model to \emph{linearly} interpolate from grid parameters to basis weights, we show that the stellar parameters we infer are in closer agreement with \todo{astrophysical expectations} than those reported using neural networks or high-dimensional cubic spline interpolation.
+We apply our model to \project{Sloan Digital Sky Survey} infrared spectra of OBAFGKM-type main-sequence and giant stars.
+%Continuum normalization is often a necessary step when analyzing stellar spectra. 
+%The ideal approach might be to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is often required before stellar parameters are estimated.
+%Here we present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization that models stellar line absorption, and a sines-and-cosines basis to model the joint continuum-instrument response.
+%The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the remaining variance.
+%The linearity of both components ensures that inference is stable and fast.
+%Our approach factorizes line absorption entirely from continuum-normalized theoretical spectra, but we also describe a data-driven variant to iteratively correct model-data mismatches.
+%We show that the basis weights can also be used to reliably estimate stellar parameters, and apply these methods to \project{Sloan Digital Sky Survey} infrared spectra of OBAFGKM-type stars.
+%This approach reduces subjectivity in estimating stellar continuum.
 \end{abstract}
@@
 In Section~\ref{sec:methods} we describe a family of methods to address these problems.
-Each of the three variants we describe use non-negative matrix factorization (NMF) to approximate line absorption.
+The methods we describe use non-negative matrix factorization (NMF) to approximate line absorption.
 NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements.
 This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
-The three variants of our method are each conceptually introduced, with the intended application (Section~\ref{sec:results}) to low-resolution optical spectra from the \project{Sloan Digital Sky Survey} (\sdss).
+%The three variants of our method are each conceptually introduced, with the intended application (Section~\ref{sec:results}) to low-resolution optical spectra from the \project{Sloan Digital Sky Survey} (\sdss).
+\todo{We apply these to SDSS.}
 We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\
@@
 Here we will describe the method in general before outlining explicit assumptions and implementation details.
-The problem of continuum normalization for a single spectrum can be described where the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
+The data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
 The forward model for these data can be expressed as the element-wise multiplication of what we will describe as the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response model $g(\lambda_i;\vectheta)$
@@
 \end{align}
-where $\hadamard$ represents the element-wise (Hadamard) product and the components $f(...)$ and $g(...)$ are defined below.\\
+where the components $f(...)$ and $g(...)$ are defined below.
+Throughout this paper we fit in (natural) log-transformed data space $\log{y_i}$ which transforms our element-wise (Hadamard; $\hadamard$) product into the sum
+\begin{align}
+    \label{eq:log_y}
+    \log{y_i} &= \log{f(\lambda_i; \vecpsi)} + \log{g(\lambda_i;\vectheta)} + \mbox{noise} \quad .
+\end{align}
+where we take the transformed flux error $y_i\sigma_{y_i}$ as the noise on $\log{y_i}$.\\
 
@@
 In the applications that we present here, the theoretical spectra used to approximate the line absorption model are assumed to have the same wavelength sampling and line spread function as the data, such that we use a grid of model spectra where $D = P$.
-Other situations are permitted, but at inference time there is a need to evaluate the line absorption model to the $P$ observed wavelengths.
-With our $N \times\ D$ matrix of continuum-normalized theoretical spectra, which we will call $\vec{M}$, we then define \emph{stellar absorption} $\vecX$ to be
+Other situations are permitted, but at inference time there is a need to evaluate the line absorption model to the $P$ observed wavelengths.\\
+
+We refer to our $N \times D$ matrix of continuum-rectified theoretical spectra as $\vec{M}$.
+This is a dense matrix: there are no entries of exactly zero.
+However, a small transformation to this matrix makes it extremely sparse.
+Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{M}$ into two smaller matrices $\vec{W}$ and $\vec{H}$ such that,
 \begin{align}
-    \vecX = 1 - \vec{M}
+    \label{eq:nmf}
+    \vec{W}\,\vec{H} \approx -\log\left({\vec{M}}\right)
 \end{align}
-such that $\vecX \in \left[0, 1\right]$ and is zero when no line absorption exists.
-This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecX$ using NMF such that
-\begin{align}
-    \vecX \approx \vec{W}\vec{H} \label{eq:nmf}
-\end{align}
-where $\vecX$ is a $N \times D$ matrix, $\vecW$ is a $N \times C$ matrix of NMF \emph{basis weights} with $C$ being the number of basis vectors to use, and $\vecH$ is a $C \times D$ matrix of NMF \emph{basis vectors}.
-For this factorization we must select the number of basis components $C$ to use, which should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
-All elements in $\vecX$, $\vec{W}$, and $\vec{H}$ are required to be non-negative.
-Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example basis vectors $\vec{H}$.\\
-
-We could construct an approximation directly to $\vec{M}$, but this would lead to denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX = 1 - \vec{M}$ that are 0 (i.e., no line absorption).
-The choice of nomenclature in this paper (i.e., $\vecX$, $\vec{W}$, $\vec{H}$) is chosen to be consistent with previous works on NMF.\\
+where $\vec{W}$ is a $N \times C$ matrix of \emph{basis weights}, where $C$ is the number of chosen basis vectors, and $\vec{H}$ is a $C \times D$ matrix of NMF \emph{basis vectors}.
+All elements in $-\log{\vec{M}}$, $\vec{W}$, and $\vec{H}$ are required to be non-negative.
+The number of basis components $C$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
+Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\
 
@@
 
+The factorization of $-\log{\vec{M}}$ into $\vecW\vecH$ is reasonably fast and easy to compute given existing packages in modern programming languages.
+We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $\vecX$: this is the default behaviour in most modern implementation.s As reported elsewhere, we found that factorizing $\vecW\vecH$ by coordinate descent was substantially faster than using multiplicative updates.\\
+%At inference time, $\vecpsi$ is analogous to the basis weights found at training time \\
+
+There are very few requirements of the theoretical spectral grid.
+There are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing in between points.
+The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to fit.
+This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of OBA-type stars was also sufficiently flexible to model many (but not all) white dwarf spectra.
+For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, or simply skip every $n$th theoretical spectrum.}\\
+
 \begin{figure*}
@@
 
-The factorization of $\vecH$ is reasonably fast and easy to compute given existing packages in Python, Julia, and other languages.
-We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $\vecX$: this is the default behaviour in the Python \texttt{scipy.decomposition.NMF} implementation.
-As reported elsewhere, we found that factorizing $\vecW\vecH$ by coordinate descent was substantially faster than using multiplicative updates.
-\\
 %For the applications presented in this paper, the factorization can be completed in the order of minutes.
@@
 This took minutes to hours to complete, depending on the size of the theoretical grid $\vecX$.\\
+\todo{At inference time, vecpsi is analogous to W}
 
-\noindent{}With the matrix $\vec{H}$ we can now define the line absorption function $f(\lambda_i;\vecpsi)$ 
+\noindent{}With the matrix $\vec{H}$ we can now define the logarithm of line absorption function $f(\lambda_i;\vecpsi)$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
 \begin{align}
-    f(\lambda_i;\vecpsi) = 1 - \vecpsi\vecH \label{eq:f}
+    \log{f(\lambda_i;\vecpsi)} = -\vecpsi\vecH \label{eq:f}
 \end{align}
 where $\vecpsi \in [0, \infty)$ are a row vector of $C$ \emph{basis weights}.
-$\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ weights needed to reconstruct the stellar absorption from the $C$ basis vectors in $\vecH$.
-Note that because $\vecpsi$ and $\vecH$ are both restricted to have non-negative elements, Equation~\ref{eq:f} shows that the maximum value that can be predicted by $f(...)$ is 1.
-This restricts the flexibility of $f(...)$ to only be able to model continuum-normalized flux values, leaving $g(...)$ to represent the joint continuum-instrument response.
-While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this would be readily compensated by a negative continuum model $g(...) < 0$ and in practice such local minima are not favoured.\\
+At inference time, $\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ weights needed to reconstruct the stellar absorption from the $C$ basis vectors in $\vecH$.
+Note that because $\vecpsi$ and $\vecH$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the joint continuum-instrument response.\\
 
-There are many suitable choices for the continuum-instrument response model $g(\lambda_i;\vectheta)$.
+There are many suitable choices for the logarithm of the continuum-instrument response model $\log{g(\lambda_i;\vectheta)}$.
 Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
-The component $g(\lambda_i;\vectheta)$ is expressed compactly as
+The component $\log{g(\lambda_i;\vectheta)}$ is expressed compactly as
 \begin{align}
-    g(\lambda_i;\vectheta) = \vec{A}(\lambda_i)\vectheta
+    \log{g(\lambda_i;\vectheta)} = \vec{C}\vectheta
 \end{align}
-where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
+where $\vec{C}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
 \begin{align}
-    \vec{A}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
+    \vec{C}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
                                        \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
 \end{align}
-\noindent{}and the design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
-Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.\\ %In this formalism, spectral regions can be fit independently with different degrees of freedom.
-The number of continuum coefficients scales as $2n_\textrm{degree} + 1$ per region.
-\\
+\noindent{}and the design matrix $\vec{C}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
+Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.\\
+
+With $f(\lambda_i;\vecpsi)$ and $g(\lambda_i;\vectheta)$ now defined, we can expand the forward model for $\log{y_i}$ to show
+\begin{equation}
+    \log{y_i} = -\vecpsi\vecH + \vec{C}\vectheta
+\end{equation}
+\begin{equation}
+    \log{y_i} = \vec{A}\vec{X}
+\end{equation}
+where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
+\begin{eqnarray}
+    \vec{A} = \begin{bmatrix}-\vec{H}\\\vec{C}\end{bmatrix}
+    \quad \mbox{and} \quad 
+    \vec{X} = \begin{bmatrix}\vecpsi\\\vectheta\end{bmatrix} 
+    \quad .
+\end{eqnarray}
+
+%In this formalism, spectral regions can be fit independently with different degrees of freedom. The number of continuum coefficients scales as $2n_\textrm{degree} + 1$ per region. \\
 
@@
 
-\subsection{Model-driven approach: Line absorption is factorized by theoretical spectra}
-\label{sec:model-method}
+%\subsection{Model-driven approach: Line absorption is factorized by theoretical spectra}
+%\label{sec:model-method}
 
-This variant uses a set of theoretical spectra to factorize line absorption $\vecX$.
-With no flux errors on theoretical spectra, this factorization can be completed quickly using the coordinate descent algorithm (the data-driven variant uses multiplicative updates, which is slower).
-In this variant, there are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing in between points.
-The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to apply the method.
-This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of OBA-type stars was also sufficiently flexible to model many (but not all) white dwarf spectra.
-For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, or simply skip every $n$th theoretical spectrum.}\\
+% TODO: re-word this
+\section{Data}
 
-The fifth generation of the \emph{Sloan Digital Sky Survey}  is acquiring low-resolution ($\mathcal{R} \sim 4{,}000$) optical spectra for millions of Milky Way stars using the \boss\ spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO).
+- SDSS
+%The fifth generation of the \emph{Sloan Digital Sky Survey}  is acquiring low-resolution ($\mathcal{R} \sim 4{,}000$) optical spectra for millions of Milky Way stars using the \boss\ spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO).
 Although the highest priority targets observed by \boss\ tend to be Black Hole Mapper (BHM) targets, many Milky Way stars are included per field, either because they are specifically targeted as part of a Milky Way Mapper (MWM) `carton'\footnote{`Cartons' are how targets are now assigned in SDSS.}, or because they are standard stars.
@@
 
-We used the BOSZ spectral grid to construct basis vectors for a line absorption model for \boss\ spectra.
+%We used the BOSZ spectral grid to construct basis vectors for a line absorption model for \boss\ spectra.
 This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a sufficiently large range in stellar parameters.
@@
 
-- initially ignore the grid stellar parameters
+- Korg Grid
 
@@
 
-- compute H\\
+- compute H
 
-Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecpsi$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data.
+%Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecpsi$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data.
 We found that setting $\vecpsi$ to a very small value ($10^{-12}$; i.e., no line absorption) and solving for the continuum coefficients $\vectheta$ was a reasonably good initialization.
@@
 \begin{figure}
-    \caption{Median model continuum-rectified flux values as a function of effective temperature, compared to the median observed rectified flux values as a function of Gaia colours, showing there is no bias as a function of spectral type?}
-\end{figure}
-
-\begin{figure}
     \caption{Gaia H-R diagram with each bin coloured by the median $\vecpsi_i$ value, picking some $i$th eigenspectra that looks like metallicity. \label{fig:gaia_hrd_metallicity}}
@@
 \begin{figure*}
-    \caption{The median pixel $\chi^2$ value as a function of \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence stars observed with the BOSS spectrograph.
-\todo{Should expect to see increasing residuals due to TiO bands not captured by the model, and emission lines.}}
+    \caption{The median pixel $\chi^2$ value as a function of de-reddened \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence [OR GIANT?] stars.
+\todo{Should expect to see increasing residuals due to features not captured by the model, and emission lines.}}
     % Could do the same showing standard deviation per bin
 \end{figure*}
-
-- figure showing the
 
@@
     Megan Bedell (Flatiron Institute),
+    Michael Blanton (NYU)
 .

### 2024-08-13 paper/main.tex
update methods, terminology, etc

@@
 %\newcommand{\chosentitle}{Constrained linear absorption models for stellar spectroscopy}
-\newcommand{\chosentitle}{The Unreasonable Effectiveness of Linear Models in Stellar Spectroscopy}
+%\newcommand{\chosentitle}{The Unreasonable Effectiveness of Linear Models in Stellar Spectroscopy}
 %\newcommand{\chosentitle}{Stellar continuum modelling}
+\newcommand{\chosentitle}{Constrained linear models for stellar spectroscopy}
 
@@
 \author[0000-0003-0174-0564]{Andrew R. Casey}
-\affiliation{School of Physics \& Astronomy, Monash University}
+\affiliation{School of Physics \& Astronomy, Monash University, Australia}
 \affiliation{Centre of Excellence for Astrophysics in Three Dimensions (ASTRO-3D)}
@@
 
-\author[0000-0003-2866-9403]{David W. Hogg}
-\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
-\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
-\affiliation{Center for Computational Astrophysics, Flatiron Institute, a division of the Simons Foundation}
+\author{friends}
+%\author[0000-0003-2866-9403]{David W. Hogg}
+%\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
+%\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
+%\affiliation{Center for Computational Astrophysics, Flatiron Institute, a division of the Simons Foundation}
 
@@
 \begin{abstract}\noindent
-Forward modelling stellar spectra usually requires a spectral synthesis code, and/or a non-linear interpolator constructed from a curated training set.
-These approaches also require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent inferences.
+Forward modelling stellar spectra usually requires a spectral synthesis code, or some non-linear interpolator constructed from a curated training set. 
+These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent measurements.
 Here we present a \emph{linear} model that simultaneously fits stellar absorption and the joint continuum-instrument response.
-%The stellar absorption is modelled by non-negative matrix factorization of a grid of theoretically rectified spectra, and a sine-and-cosine basis to model the joint continuum-instrument response.
-%The linearity inference is stable and fast.
-Stellar absorption is modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices (basis weights and basis vectors) with basis vectors kept fixed at inference time, and the joint continuum-instrument response is modelled with a sine-and-cosine basis.
-The non-negativity constraint ensures that basis vectors are strictly additive, and by maintaining linearity we ensure that inference is convex, stable, and fast.
-When we restructure our forward model to \emph{linearly} interpolate from grid parameters to basis weights, we show that the stellar parameters we infer are in closer agreement with \todo{astrophysical expectations} than those reported using neural networks or high-dimensional cubic spline interpolation.
-We apply our model to \project{Sloan Digital Sky Survey} infrared spectra of OBAFGKM-type main-sequence and giant stars.
+The joint continuum-instrument response is modelled with a Fourier basis, and stellar absorption is modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
+The non-negativity constraint ensures that basis vectors are strictly additive.
+Together this set of bases represents a linear model for stellar spectra, which ensures that inference is convex, stable, and fast.
+We demonstrate this method by simultaneous fits to high-resolution echelle spectra, where continuum is usually fit per order.
+When we restructure our forward model to \emph{linearly} interpolate from stellar parameters to basis weights, we show the the stellar parameters we infer from \project{Sloan Digital Sky Survey} infrared spectra are in closer agreement with \todo{astrophysical expectations} than those reported using neural networks or high-dimensional cubic spline interpolation.
+
 %Continuum normalization is often a necessary step when analyzing stellar spectra. 
@@
 
-We will assume a forward model that includes two components: one represents continuum-normalized absorption (e.g., line absorption); and another to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
-These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will assume both components to be linear models, ensuring that inference is fast and stable.
+We will assume a forward model that includes two components: one that represents continuum-normalized absorption (e.g., line absorption); and another to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
+These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will require both components to be linear models, which ensures that inference is stable and fast.
 In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and for estimating stellar parameters.\\
@@
 \begin{align}
-    y_i &= f(\lambda_i;\vecpsi)\hadamard{}g(\lambda_i;\vectheta) + \mbox{noise}
+    y &= f(\lambda;\vecpsi)\hadamard{}g(\lambda;\vectheta) + \mbox{noise}
 \end{align}
 where the components $f(...)$ and $g(...)$ are defined below.
-Throughout this paper we fit in (natural) log-transformed data space $\log{y_i}$ which transforms our element-wise (Hadamard; $\hadamard$) product into the sum
+Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ where the covariance matrix $\vec{C}$ has the transformed flux variances $\vec{C}_{ii} = (y_{i}\sigma_{y,i})^2$ along the diagonal.
+The log-transformation changes the element-wise (Hadamard; $\hadamard$) product into the summation:
 \begin{align}
     \label{eq:log_y}
-    \log{y_i} &= \log{f(\lambda_i; \vecpsi)} + \log{g(\lambda_i;\vectheta)} + \mbox{noise} \quad .
+    \vec{Y} &= \log{f(\vec{\lambda}; \vecpsi)} + \log{g(\vec{\lambda};\vectheta)} \quad .
 \end{align}
-where we take the transformed flux error $y_i\sigma_{y_i}$ as the noise on $\log{y_i}$.\\
 
@@
 
-We refer to our $N \times D$ matrix of continuum-rectified theoretical spectra as $\vec{M}$.
-This is a dense matrix: there are no entries of exactly zero.
-However, a small transformation to this matrix makes it extremely sparse.
-Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{M}$ into two smaller matrices $\vec{W}$ and $\vec{H}$ such that,
-\begin{align}
+We refer to our $N \times D$ matrix of continuum-rectified theoretical spectra as $\vec{M}$. This is a dense matrix: there are no entries of exactly zero. However, a small transformation
+%\begin{eqnarray}
+%    \vec{X} \equiv -\log\left({\vec{M}}\right)
+%\end{eqnarray}
+to this matrix can make it it extremely sparse.
+Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{M}$ into two smaller matrices $\vec{W}$ and $\vec{F}$ such that,
+\begin{eqnarray}
     \label{eq:nmf}
-    \vec{W}\,\vec{H} \approx -\log\left({\vec{M}}\right)
-\end{align}
-where $\vec{W}$ is a $N \times C$ matrix of \emph{basis weights}, where $C$ is the number of chosen basis vectors, and $\vec{H}$ is a $C \times D$ matrix of NMF \emph{basis vectors}.
-All elements in $-\log{\vec{M}}$, $\vec{W}$, and $\vec{H}$ are required to be non-negative.
-The number of basis components $C$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
+    -\log\left({\vec{M}}\right) \approx \vec{W}\transpose\vec{F}
+\end{eqnarray}
+where $\vec{W}$ is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}.
+All elements in $-\log\left({\vec{M}}\right)$, $\vec{W}$, and $\vec{F}$ are required to be non-negative.
+The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
 Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\
@@
 
-The factorization of $-\log{\vec{M}}$ into $\vecW\vecH$ is reasonably fast and easy to compute given existing packages in modern programming languages.
-We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $\vecX$: this is the default behaviour in most modern implementation.s As reported elsewhere, we found that factorizing $\vecW\vecH$ by coordinate descent was substantially faster than using multiplicative updates.\\
+The factorization of $-\log\left({\vec{M}}\right)$ into $\vec{W}\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
+We found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{M}}\right)$: this is the default behaviour in many packages.
+We found that factorizing $\vec{W}\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex and do not require many choices.\\
 %At inference time, $\vecpsi$ is analogous to the basis weights found at training time \\
@@
 This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of OBA-type stars was also sufficiently flexible to model many (but not all) white dwarf spectra.
-For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, or simply skip every $n$th theoretical spectrum.}\\
+For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\
 
@@
 This took minutes to hours to complete, depending on the size of the theoretical grid $\vecX$.\\
-\todo{At inference time, vecpsi is analogous to W}
 
-\noindent{}With the matrix $\vec{H}$ we can now define the logarithm of line absorption function $f(\lambda_i;\vecpsi)$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
+\noindent{}With the matrix $\vec{F}$ we can now define the logarithm of line absorption function $\log{f(\lambda_i;\vecpsi)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
 \begin{align}
-    \log{f(\lambda_i;\vecpsi)} = -\vecpsi\vecH \label{eq:f}
+    \log{f(\lambda_i;\vecpsi)} = -\vec{F}\transpose\vecpsi \label{eq:f}
 \end{align}
-where $\vecpsi \in [0, \infty)$ are a row vector of $C$ \emph{basis weights}.
-At inference time, $\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ weights needed to reconstruct the stellar absorption from the $C$ basis vectors in $\vecH$.
-Note that because $\vecpsi$ and $\vecH$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the joint continuum-instrument response.\\
+where $\vecpsi \in [0, \infty)$ is a column vector of $C$ \emph{basis weights} to be solved at inference time.
+Here, $\vecpsi$ is analogous to a column in $\vecW$: it represents the $C$ weights needed to reconstruct the stellar absorption from the $C$ basis vectors $\vec{F}$.
+Note that because $\vecpsi$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the smooth continuum-instrument response.
+\\
 
@@
 \begin{align}
-    \log{g(\lambda_i;\vectheta)} = \vec{C}\vectheta
+    \log{g(\lambda_i;\vectheta)} = \vec{G}\vectheta
 \end{align}
-where $\vec{C}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
+where $\vec{G}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
 \begin{align}
-    \vec{C}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
+    \vec{G}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
                                        \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
 \end{align}
-\noindent{}and the design matrix $\vec{C}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
+\noindent{}and the design matrix $\vec{G}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
 Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.\\
 
-With $f(\lambda_i;\vecpsi)$ and $g(\lambda_i;\vectheta)$ now defined, we can expand the forward model for $\log{y_i}$ to show
+\noindent{}With $f(\lambda_i;\vecpsi)$ and $g(\lambda_i;\vectheta)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
 \begin{equation}
-    \log{y_i} = -\vecpsi\vecH + \vec{C}\vectheta
+    \vec{Y} = -\vec{F}\transpose\vecpsi + \vec{G}\vectheta
 \end{equation}
 \begin{equation}
-    \log{y_i} = \vec{A}\vec{X}
+    \vec{Y} = \vec{A}\vec{X}
 \end{equation}
@@
 \begin{eqnarray}
-    \vec{A} = \begin{bmatrix}-\vec{H}\\\vec{C}\end{bmatrix}
+    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\\vec{G}\end{bmatrix} \quad .
     \quad \mbox{and} \quad 
     \vec{X} = \begin{bmatrix}\vecpsi\\\vectheta\end{bmatrix} 
-    \quad .
+\end{eqnarray}
+Given some observed log-transformed flux $\vec{Y}$ and associated covariance matrix $\vec{C}$, this reduces to a weighted least-squares problem which is subject to the constraint $\vecpsi \geq 0$
+\begin{eqnarray}
+    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} \quad ,
+\end{eqnarray}
+
+\noindent{}where the un-transformed rectified flux and continuum can be computed in a straightorward manner:
+\begin{eqnarray}
+    f(\boldsymbol{\tilde{\psi}}) &\,=\,& \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\psi}}\right)} \\
+    g(\boldsymbol{\tilde{\theta}}) &\,=\,& \exp{\left(\vec{G}\boldsymbol{\tilde{\theta}}\right)}
 \end{eqnarray}

### 2024-08-21 code/continuum_basis.py
add example code

@@ -1,0 +1,69 @@
+import numpy as np
+from itertools import cycle
+
+class ContinuumBasis:
+    pass
+
+class NoContinuum(ContinuumBasis):
+    
+    @property
+    def num_parameters(self):
+        return 0
+
+    def design_matrix(self, λ):
+        return 0
+
+class Sinusoids(ContinuumBasis):    
+        
+    def __init__(self, P=7, L=None):
+        self.P = P
+        self.L = L
+        return None
+    
+    @property
+    def num_parameters(self):
+        return self.P
+        
+    def default_L(self, λ):
+        return 2 * np.ptp(λ)
+    
+    def design_matrix(self, λ):
+        if self.L is None:
+            L = self.default_L(λ)
+        elif isinstance(self.L, (float, int)):
+            L = self.L
+        else:
+            L = self.L(λ)
+            
+        scale = (np.pi * λ) / L
+        A = np.ones((λ.size, self.P), dtype=float)
+        for j, f in zip(range(1, self.P), cycle((np.sin, np.cos))):
+            A[:, j] = f(scale * (j + (j % 2)))        
+        return A
+            
+        scale = 2 * (np.pi / L)
+        return np.vstack(
+            [
+                np.ones_like(λ).reshape((1, -1)),
+                np.array(
+                    [
+                        [np.cos(o * scale * λ), np.sin(o * scale * λ)]
+                        for o in range(1, self.deg + 1)
+                    ]
+                ).reshape((2 * self.deg, λ.size)),
+            ]
+        ).T
+
+
+class Polynomial(ContinuumBasis):
+    
+    def __init__(self, deg=2):
+        self.deg = deg
+        return None
+    
+    @property
+    def num_parameters(self):
+        return self.deg + 1
+    
+    def design_matrix(self, λ):
+        return np.vander(λ, self.deg + 1)

### 2024-08-21 code/model.py

@@ -1,0 +1,621 @@
+
+import numpy as np
+
+import h5py as h5
+
+import numpy as np
+import gzip
+import warnings
+import pickle
+import os
+from time import time
+
+from itertools import cycle
+from functools import cache
+from typing import Optional, Tuple, Sequence, Union
+from scipy import optimize as op
+from sklearn.decomposition._nmf import _fit_coordinate_descent
+from sklearn.exceptions import ConvergenceWarning
+from scipy import interpolate
+from astropy.constants import c
+from threading import Thread
+
+from typing import Iterable
+
+
+from spectrum.lsf import instrument_lsf_kernel, rotational_broadening_kernel
+from spectrum import Spectrum
+from scipy.optimize._lsq.trf_linear import trf_linear
+
+
+warnings.simplefilter("ignore")
+
+C_KM_S = c.to("km/s").value
+
+
+import numpy as np
+
+
+
+from rv import cross_correlate
+from continuum_basis import ContinuumBasis, Sinusoids, Polynomial
+from utils import cumulative
+
+
+def apply_radial_velocity_shift(λ, v_rel):
+    if v_rel == 0 or v_rel is None:
+        return λ
+    beta = v_rel / C_KM_S
+    scale = np.sqrt((1 + beta) / (1 - beta))     
+    return λ * scale
+
+
+class Clam:
+    
+    """Constrained linear absorption model, with applications to echelle spectra."""
+    
+    def __init__(
+        self, 
+        stellar_vacuum_wavelength, 
+        stellar_basis_vectors, 
+        telluric_vacuum_wavelength=None,
+        telluric_basis_vectors=None,
+        meta=None,         
+    ):
+        """
+        :param stellar_vacuum_wavelength:
+            An array of vacuum wavelengths (in Angstroms) of the basis vectors.
+        
+        :param stellar_basis_vectors:
+            A (C, P)-shape array of stellar basis vectors with non-negative entries. Here, C is
+            the number of basis vectors and P is the same as the size of `stellar_vacuum_wavelength`.
+                    
+        :param telluric_basis_vectors: [optional]
+            A (B, P)-shape array of telluric basis vectors with non-negative entries. Here, P is
+            the same as the size of `stellar_vacuum_wavelength`.
+            
+        :param meta: [optional]
+            A metadata dictionary.            
+        """
+        self.stellar_vacuum_wavelength = stellar_vacuum_wavelength
[…]

### 2024-08-21 code/rv.py

@@ -1,0 +1,227 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+""" Functions related to radial velocity measurement and correction. """
+
+from __future__ import (division, print_function, absolute_import,
+                        unicode_literals)
+
+__all__ = ["cross_correlate"]
+
+import logging
+import numpy as np
+from scipy import interpolate
+from scipy.optimize import leastsq
+
+
+logger = logging.getLogger(__name__)
+
+
+def cross_correlate(observed_λ, observed_flux, observed_ivar, template_λ, template_flux, λ_range=None,
+    use_weight=False, apodize=0, resample="template", verbose=False):
+    """
+    Cross-correlate the observed spectrum against a rest-frame template spectrum
+    and measure the radial velocity of the source.
+
+    :param observed_spectrum:
+        The observed spectrum.
+
+    :type observed_spectrum:
+        :class:`specutils.Spectrum1D`
+
+    :param template_spectrum:
+        A rest-frame template spectrum.
+
+    :type template_spectrum:
+        :class:`specutils.Spectrum1D`
+
+    :param λ_range: [optional]
+        A two-length tuple containing the start and end wavelengths to consider.
+        If not provided, then the overlap of the `observed_spectrum` and the
+        `template_spectrum` will be used.
+
+    :param apodize: [optional]
+        The fraction of pixels to apodize on either side of the spectrum.
+
+    :param resample: [optional]
+        Whether to resample the 'template' onto the observed spectrum, or
+        resample the 'observed' spectrum onto the template.
+
+    :returns:
+        The radial velocity, uncertainty in radial velocity, and the CCF.
+    """
+
+
+    if λ_range is None:
+        # Use the common ranges.
+        λ_range = (
+            np.max([
+                observed_λ[0],
+                template_λ[0]
+            ]),
+            np.min([
+                observed_λ[-1],
+                template_λ[-1]
+            ])
+        )
+
+    if not isinstance(λ_range, (tuple, list, np.ndarray)) \
+    or len(λ_range) != 2:
+        raise TypeError("wavelength region must be a two length list-type")
+
+    if apodize != 0:
[…]

### 2024-08-21 code/utils.py

@@ -1,0 +1,27 @@
+import os
+from typing import Iterable, Optional
+
+class cumulative:
+    
+    """A cumulative iterator."""
+    
+    def __init__(self, iterable: Iterable, start: Optional[int] = 0):
+        self.iterable = iter(iterable)
+        self.sum = start
+        return None
+    
+    def __iter__(self):
+        return self
+    
+    def __next__(self):
+        item = next(self.iterable)
+        I = len(item)
+        self.sum += I
+        return (self.sum - I, item)        
+
+
+
+def expand_path(path):
+    return os.path.abspath(os.path.expanduser(path))
+
+

### 2024-08-21 paper/main.tex
update methods

@@
 \usepackage{amsmath}
+\usepackage{MnSymbol}
 
@@
 \newcommand{\vectheta}{\boldsymbol{\theta}}
-\newcommand{\vecpsi}{\boldsymbol{\psi}}
-\newcommand{\vecW}{\mathbf{W}}
-\newcommand{\vecH}{\mathbf{H}}
+\newcommand{\vecalpha}{\boldsymbol{\alpha}}
+\newcommand{\vecbeta}{\boldsymbol{\beta}}
+\newcommand{\vecgamma}{\boldsymbol{\gamma}}
+\newcommand{\vecW}{\mathbf{W}} % stellar line absorption basis weights
+\newcommand{\vecF}{\mathbf{F}} % stellar line absorption basis vectors
+\newcommand{\vecG}{\mathbf{G}} % telluric line absorption basis vectors
+\newcommand{\vecH}{\mathbf{H}} % continuum basis vectors
 \newcommand{\vecX}{\mathbf{X}}
+
 \newcommand{\hadamard}{\odot}
@@
 \newcommand{\sdss}{\project{SDSS}}
+\newcommand{\eso}{\project{ESO}}
+\newcommand{\harps}{\project{HARPS}}
 
@@
 The joint continuum-instrument response is modelled with a Fourier basis, and stellar absorption is modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
+We factorize telluric absorption in the same way, allowing us to simultaneously fit stellar and telluric absorption.
 The non-negativity constraint ensures that basis vectors are strictly additive.
 Together this set of bases represents a linear model for stellar spectra, which ensures that inference is convex, stable, and fast.
-We demonstrate this method by simultaneous fits to high-resolution echelle spectra, where continuum is usually fit per order.
-When we restructure our forward model to \emph{linearly} interpolate from stellar parameters to basis weights, we show the the stellar parameters we infer from \project{Sloan Digital Sky Survey} infrared spectra are in closer agreement with \todo{astrophysical expectations} than those reported using neural networks or high-dimensional cubic spline interpolation.
-
-%Continuum normalization is often a necessary step when analyzing stellar spectra. 
-%The ideal approach might be to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is often required before stellar parameters are estimated.
-%Here we present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear components: a non-negative matrix factorization that models stellar line absorption, and a sines-and-cosines basis to model the joint continuum-instrument response.
-%The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the remaining variance.
-%The linearity of both components ensures that inference is stable and fast.
-%Our approach factorizes line absorption entirely from continuum-normalized theoretical spectra, but we also describe a data-driven variant to iteratively correct model-data mismatches.
-%We show that the basis weights can also be used to reliably estimate stellar parameters, and apply these methods to \project{Sloan Digital Sky Survey} infrared spectra of OBAFGKM-type stars.
-%This approach reduces subjectivity in estimating stellar continuum.
+This model allows us to reliably fit the stellar absorption, telluric absorption, the continuum, radial velocity, and rotational broadening, without any prior knowledge about the fundamental stellar parameters.
+We demonstrate this method by fits to high-resolution echelle spectra, where we achieve \todo{X\%} accuracy in continuum normalization with no apparent biases at low signal-to-noise ratios (e.g., $<10$).
 \end{abstract}
@@
 This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
-%The three variants of our method are each conceptually introduced, with the intended application (Section~\ref{sec:results}) to low-resolution optical spectra from the \project{Sloan Digital Sky Survey} (\sdss).
-\todo{We apply these to SDSS.}
 We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\
@@
 The data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
-The forward model for these data can be expressed as the element-wise multiplication of what we will describe as the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response model $g(\lambda_i;\vectheta)$
+The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric absorption model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
 \begin{align}
-    y &= f(\lambda;\vecpsi)\hadamard{}g(\lambda;\vectheta) + \mbox{noise}
+    y_i &= f(\lambda_i;\vecalpha)\hadamard{}g(\lambda_i;\vecbeta)\hadamard{}h(\lambda_i;\vecgamma) + \mbox{noise}
 \end{align}
-where the components $f(...)$ and $g(...)$ are defined below.
-Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ where the covariance matrix $\vec{C}$ has the transformed flux variances $\vec{C}_{ii} = (y_{i}\sigma_{y,i})^2$ along the diagonal.
-The log-transformation changes the element-wise (Hadamard; $\hadamard$) product into the summation:
+where the components $f(...)$, $g(...)$, and $h(...)$ are defined below.
+Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
+\begin{eqnarray}
+    \vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + \frac{2\sigma_{y,i}^3}{8y_i^3} - \frac{6\sigma_{y,i}^4}{24y_i^4}\right)^2
+\end{eqnarray}
+\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into the summation:
 \begin{align}
     \label{eq:log_y}
-    \vec{Y} &= \log{f(\vec{\lambda}; \vecpsi)} + \log{g(\vec{\lambda};\vectheta)} \quad .
-\end{align}
+    \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
+\end{align}\\
 
-The line absorption model $f(\lambda_i;\vecpsi)$ predicts the line absorption at wavelength $\lambda_i$ given parameters $\vecpsi$.
-We construct the line absorption model $f(\lambda_i;\vecpsi)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorization (NMF).
-In the applications that we present here, the theoretical spectra used to approximate the line absorption model are assumed to have the same wavelength sampling and line spread function as the data, such that we use a grid of model spectra where $D = P$.
-Other situations are permitted, but at inference time there is a need to evaluate the line absorption model to the $P$ observed wavelengths.\\
+We now turn to defining the stellar absorption model, the telluric absorption model, and the continuum model.
+The stellar absorption model $f(\lambda_i;\vecalpha)$ predicts the rectified stellar line absorption at wavelength $\lambda_i$ given parameters $\vecalpha$.
+We construct the line absorption model $f(\lambda_i;\vecalpha)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorization (NMF).
+The theoretical spectra used to construct  the line absorption model do not have to have the same wavelength sampling and instrument line spread profile as the data, but at inference time there is a need to interpolate (or evaluate) the line absorption model to the $P$ observed wavelengths.\\
 
-We refer to our $N \times D$ matrix of continuum-rectified theoretical spectra as $\vec{M}$. This is a dense matrix: there are no entries of exactly zero. However, a small transformation
-%\begin{eqnarray}
-%    \vec{X} \equiv -\log\left({\vec{M}}\right)
-%\end{eqnarray}
-to this matrix can make it it extremely sparse.
-Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{M}$ into two smaller matrices $\vec{W}$ and $\vec{F}$ such that,
+We refer to our $N \times D$ matrix of continuum-rectified theoretical stellar spectra as $\vec{S}$.
+This is a dense matrix: there are no entries of exactly zero.
+However, a small transformation to this matrix can make it it extremely sparse.
+Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
 \begin{eqnarray}
     \label{eq:nmf}
-    -\log\left({\vec{M}}\right) \approx \vec{W}\transpose\vec{F}
+    -\log\left({\vec{S}}\right) \approx \vec{W}_\star\transpose\vec{F}
 \end{eqnarray}
-where $\vec{W}$ is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}.
-All elements in $-\log\left({\vec{M}}\right)$, $\vec{W}$, and $\vec{F}$ are required to be non-negative.
-The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
-Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\
+where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}.
+All elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ are required to be non-negative.
+The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\%Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\
 
@@
 
-%The nomenclature  draws on previous works on NMF and data analysis, but this nomenclature is easy to overload, so throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights}, $\vecpsi$ as the NMF \emph{basis coefficients}, $\vecH$ as the NMF \emph{basis vectors}, and $\vectheta$ (not yet defined) will refer to the amplitudes of the sine-and-cosine basis.
+%The nomenclature  draws on previous works on NMF and data analysis, but this nomenclature is easy to overload, so throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights}, $\vecalpha$ as the NMF \emph{basis coefficients}, $\vecH$ as the NMF \emph{basis vectors}, and $\vectheta$ (not yet defined) will refer to the amplitudes of the sine-and-cosine basis.
 
-The factorization of $-\log\left({\vec{M}}\right)$ into $\vec{W}\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
-We found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{M}}\right)$: this is the default behaviour in many packages.
-We found that factorizing $\vec{W}\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex and do not require many choices.\\
-%At inference time, $\vecpsi$ is analogous to the basis weights found at training time \\
+The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
+We found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages.
+We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex, do not require many choices, and they make for an efficient problem for graphical processing units.\\
+%At inference time, $\vecalpha$ is analogous to the basis weights found at training time \\
+
+\todo{We used the BOSZ grid. Some details about that, including settings used for factorizing.}\\
 
@@
 
-\noindent{}With the matrix $\vec{F}$ we can now define the logarithm of line absorption function $\log{f(\lambda_i;\vecpsi)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
+\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line absorption function $\log{f(\lambda_i;\vecalpha)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
 \begin{align}
-    \log{f(\lambda_i;\vecpsi)} = -\vec{F}\transpose\vecpsi \label{eq:f}
+    \log{f(\lambda_i;\vecalpha)} = -\vec{F}\transpose\vecalpha \label{eq:log-f}
 \end{align}
-where $\vecpsi \in [0, \infty)$ is a column vector of $C$ \emph{basis weights} to be solved at inference time.
-Here, $\vecpsi$ is analogous to a column in $\vecW$: it represents the $C$ weights needed to reconstruct the stellar absorption from the $C$ basis vectors $\vec{F}$.
-Note that because $\vecpsi$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the smooth continuum-instrument response.
+where $\vecalpha \in [0, \infty)$ is a column vector of $K$ \emph{stellar basis weights} to be solved at inference time.
+Here, $\vecalpha$ is analogous to a column in $\vecW_\star$: it represents the $K$ weights needed to reconstruct the stellar absorption from the $K$ basis vectors $\vec{F}$.
+Note that because $\vecalpha$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the smooth continuum-instrument response.
 \\
 
-There are many suitable choices for the logarithm of the continuum-instrument response model $\log{g(\lambda_i;\vectheta)}$.
+Telluric line absorption can be included in a similar way to the stellar line absorption.
+We factorize the negative log of theoretical telluric absorption into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$.
+\todo{Details about what telluric grid we used, and setting used for factorizing.
+Note how many vectors we chose ($L$)} With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
+\begin{eqnarray}
+    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta
+\end{eqnarray}
+where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} to be solved at inference time.
+We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}).
+Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\
+
+There are many suitable choices for the logarithm of the continuum-instrument response model $\log{h(\lambda_i;\vecgamma)}$.
 Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
-The component $\log{g(\lambda_i;\vectheta)}$ is expressed compactly as
+The component $\log{h(\lambda_i;\vecgamma)}$ is expressed compactly as
 \begin{align}
-    \log{g(\lambda_i;\vectheta)} = \vec{G}\vectheta
+    \log{h(\lambda_i;\vecgamma)} = \vec{G}\vecgamma
 \end{align}
-where $\vec{G}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
+where $\vec{H}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
 \begin{align}
-    \vec{G}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
+    \vec{H}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
                                        \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
 \end{align}
-\noindent{}and the design matrix $\vec{G}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.
-Throughout this paper we will describe $\vectheta$ as the sine-and-cosine \emph{coefficients}.\\
+\noindent{}where $L$ \todo{overloaded} is a length scale. \todo{Talk more about this.}
+The design matrix $\vec{H}$ can be constructed \emph{a priori} before inference begins. Throughout this paper we will describe $\vecgamma$ as the sine-and-cosine \emph{coefficients}.\\
 
-\noindent{}With $f(\lambda_i;\vecpsi)$ and $g(\lambda_i;\vectheta)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
+\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
 \begin{equation}
-    \vec{Y} = -\vec{F}\transpose\vecpsi + \vec{G}\vectheta
+    \vec{Y} = -\vec{F}\transpose\vecalpha - \vec{G}\transpose\vecbeta + \vec{H}\vecgamma
 \end{equation}
@@
 \begin{eqnarray}
-    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\\vec{G}\end{bmatrix} \quad .
+    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\-\vec{G}\transpose\\\vec{H}\end{bmatrix}
     \quad \mbox{and} \quad 
-    \vec{X} = \begin{bmatrix}\vecpsi\\\vectheta\end{bmatrix} 
-\end{eqnarray}
-Given some observed log-transformed flux $\vec{Y}$ and associated covariance matrix $\vec{C}$, this reduces to a weighted least-squares problem which is subject to the constraint $\vecpsi \geq 0$
-\begin{eqnarray}
-    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} \quad ,
+    \vec{X} = \begin{bmatrix}\vecalpha\\\vecbeta\\\vecgamma\end{bmatrix} \quad .
 \end{eqnarray}
 
-\noindent{}where the un-transformed rectified flux and continuum can be computed in a straightorward manner:
+Given some log-transformed observed flux $\vec{Y}$ and associated covariance matrix $\vec{C}$, this reduces to a weighted least-squares problem. This could be solved directly by linear algebra,
 \begin{eqnarray}
-    f(\boldsymbol{\tilde{\psi}}) &\,=\,& \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\psi}}\right)} \\
-    g(\boldsymbol{\tilde{\theta}}) &\,=\,& \exp{\left(\vec{G}\boldsymbol{\tilde{\theta}}\right)}
+    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y}
+\end{eqnarray}
+\noindent{}but this does not guarantee that the bounds $\vecalpha \geq 0$ and $\vecbeta \geq 0$ are met.
+
+\todo{Talk about how we can solve this as} $||\vec{Y}/\sqrt{C} - \vec{AX}\sqrt{C}||^2$ with linear operator, truncated reflective region, or bounded least-squares with a provided jacobian ($\vec{A}$).
+Thankfully, each of these optimization paths is convex.\\
+
+\todo{Move this to 3.2?}
+In later sections we describe some situations where we may want to penalize values of $\vecalpha$.
+We perform this by feature weighting, which is equivalent to setting a prior on the distribution that $\vecalpha$ can be drawn from.
+We introduce the feature weighting matrix $\vec{\Lambda}$, where $K$ is the number of basis vectors chosen for NMF and $L$ is the number of Fourier modes.
+We set the first $K \times K$ elements of $\vec{\Lambda}$ to $\mathrm{Cov}{\left(\vec{W}\vec{W}\transpose\right)}$ and set all other elements of $\vec{\Lambda}$ to zero.
+We can scale the strength of this feature weighting by a scalar $\lambda$ (\todo{nomenclature overload of lambda}) and solve for the parameters $\vec{\tilde{X}}$ with
+\begin{eqnarray}
+    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A} + \lambda\vec{\Lambda})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} \quad .
+\end{eqnarray}\\
+
+After solving for $\vec{\tilde{X}}$, the (un-transformed) rectified flux and continuum can be computed in a straightorward manner:
+\begin{eqnarray}
+    \mathrm{rectified~stellar~flux}, &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\alpha}}\right)} \\
+    \mathrm{telluric~absorption}, &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\tilde{\beta}}\right)} \\
+    \mathrm{continuum}, &h& = \exp{\left(\vec{H}\boldsymbol{\tilde{\gamma}}\right)}
 \end{eqnarray}
@@
 We experimented with choices of initialisation and inference.
-Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously.
-The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecpsi$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip).
-Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecpsi$) seemed to work well in many scenarios.
+Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecalpha$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecalpha,\vectheta\}$ simultaneously.
+The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecalpha$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip).
+Initializing from small ($10^{-12}$) values of $\vecalpha$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecalpha$) seemed to work well in many scenarios.
 \\
@@
 % TODO: re-word this
+
 \section{Data}
 
+We demonstrate the effectiveness of our method by forward modeling high-resolution optical echelle spectra taken with the MIKE spectrograph on the Magellan Clay telescope.
+These spectra have a spectral resolution of $\mathcal{R} \approx X$ and a wavelength range from $X$ to $Y$\,nm.
+The spectra are recorded on the 2D image plane as a series of $X$ orders: $X$ in the blue arm (X to Y) and $Y$ in the red arm (Y to Z).
+\\
+
+\todo{Description of the blaze function, although this is not known perfectly}
+
+The typical analysis procedure for this kind of spectra might involve selecting (by hand) pixels that do not appear to be obviously affected by line (or molecular) absorption, and fitting a smooth function to represent the continuum.
+This process would be repeated per-order, often ignoring the selection of pixels in the previous order (which overlaps in wavelength).
+A subsequent local continuum determination would be made for each line.
+
+\todo{Details about the synthetic grid used, and how we trained the NMF. Include details about the telluric basis used as well.}\\
+
+\todo{Assumptions about the continuum parameters}\\
+
+We measure the radial velocity by cross-correlating the spectra against all basis vectors. \todo{More details about this.} In practice this alternating
+
+\subsection{High-resolution infrared spectra}
 - SDSS
@@
 
-%Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecpsi$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data.
-We found that setting $\vecpsi$ to a very small value ($10^{-12}$; i.e., no line absorption) and solving for the continuum coefficients $\vectheta$ was a reasonably good initialization.
-In situations where there are spectra of the same star we advocate fitting all visit spectra simultaneously using a single set of basis weights $\vecpsi$ for all spectra, and continuum coefficients $\vectheta_i$ for the $i$th epoch spectrum.
+%Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecalpha$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data.
+We found that setting $\vecalpha$ to a very small value ($10^{-12}$; i.e., no line absorption) and solving for the continuum coefficients $\vectheta$ was a reasonably good initialization.
+In situations where there are spectra of the same star we advocate fitting all visit spectra simultaneously using a single set of basis weights $\vecalpha$ for all spectra, and continuum coefficients $\vectheta_i$ for the $i$th epoch spectrum.
 
@@
 
-We applied the three variants to all \boss\ stellar spectra currently available in the \emph{Sloan Digital Sky Survey}. This includes approximately \todo{X} spectra of \todo{Y} stars.
-
-\begin{figure}
-    \caption{Continuum-normalized \boss\ spectra of all spectral types, with the best-fitting model spectrum shown.}
-\end{figure}
-
-\begin{figure}
-    \caption{A Gaia H-R diagram showing the median $\chi^2$ per bin.}
-\end{figure}
+We want to show:
+\begin{enumerate}
+    \item How does the mean/median rectified flux vary as a function of S/N for the same star (e.g., 18 Sco)?
+    \item What do the fits look like for a few carefully chosen spectral types?
+    \item What regularisation strength did we need to apply?
+\end{enumerate}
 
@@
 
-\begin{figure}
-    \caption{Gaia H-R diagram with each bin coloured by the median $\vecpsi_i$ value, picking some $i$th eigenspectra that looks like metallicity. \label{fig:gaia_hrd_metallicity}}
-\end{figure}
-
-\begin{figure*}
-    \caption{The median pixel $\chi^2$ value as a function of de-reddened \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence [OR GIANT?] stars.
-\todo{Should expect to see increasing residuals due to features not captured by the model, and emission lines.}}
-    % Could do the same showing standard deviation per bin
-\end{figure*}
-
 \section{Discussion}\label{sec:discussion}
 
-- getting good fits even with bad models: the hydrogen lines in the apogee grid are knowingly incorrect, but we see no difference in the $\chi^2$ at those wavelengths.
-That is because the hydrogen lines separate out into very clean eigenspectra, where the amplitude for that eigenspectra is not required to vary smoothly with other amplitudes
+- getting good fits with bad models
 
-- biases as a function of S/N or labels
+- model interpretability
+
+- biases as a function of S/N
 
 - poorly modelled regions as a function of pixel
-
-- $\chi^2$ across the HRD
 
@@
 
-- issue with M-giant eigenspectra when they were over-represented in the sample
+- setting the right covariance strength for a given model.
 
@@
 \label{sec:conclusions}
-
-We solved continuum normalization.
 
@@
 % All these people are likely co-authors, but should be thanked if they don't want to be co-authors
+    Rory Smith (Monash),
     Adam Wheeler (Ohio State University),

### 2024-08-22 code/harps_alCen_check_bis_berv.py
fixed weird vac/air thing and did alf Cen A stability analysis

@@ -1,0 +1,151 @@
+from glob import glob
+from tqdm import tqdm
+from astropy.io import fits
+from astropy.table import Table
+
+from spectrum import Spectrum
+from model import apply_radial_velocity_shift
+
+
+def load_harps_e2ds(path, sign, overall_sign, rest_frame=True, mask_na_doublet=True, ignore_last_order=True):
+
+    pattern = path.split("_e2ds_")[0] + "*_bis*.fits"
+    bis_path = glob(pattern)[0]
+    with fits.open(bis_path) as image:
+        bis_rv = image[0].header["ESO DRS BIS RV"]
+        berv = image[0].header["ESO DRS BERV"]
+
+    pattern = path.split("_e2ds_")[0] + "*_ccf_*.fits"
+    ccf_path = glob(pattern)[0]
+    with fits.open(ccf_path) as image:
+        ccf_rv = image[0].header['ESO DRS CCF RV']
+
+    print(bis_rv, ccf_rv, berv)
+
+    spectra = []
+    with fits.open(path) as image:
+        i, coeff = (0, [])
+        header_coeff_key = "HIERARCH ESO DRS CAL TH COEFF LL{i}"
+        while image[0].header.get(header_coeff_key.format(i=i), False):
+            coeff.append(image[0].header[header_coeff_key.format(i=i)])
+            i += 1
+
+        n_orders, n_pixels = image[0].data.shape
+        x = np.arange(n_pixels)# - n_pixels // 2
+
+        coeff = np.array(coeff).reshape((n_orders, -1))
+        λ = np.array([np.polyval(c[::-1], x) for c in coeff])
+
+        n_pixels = 0
+        K = n_orders - 1 if ignore_last_order else n_orders
+        for i in range(K): # ignore last order until we have a good telluric model
+            flux = image[0].data[i]
+            ivar = 1/image[0].data[i]
+            bad_pixel = (flux == 0) | ~np.isfinite(flux) | (flux < 0)
+            flux[bad_pixel] = 0
+            ivar[bad_pixel] = 0
+
+            if rest_frame:
+                λ_order = apply_radial_velocity_shift(λ[i], overall_sign * (bis_rv + sign * berv))
+            else:
+                λ_order = λ[i]
+
[…]

### 2024-08-22 code/harps_alfCenA_wrt_snr.py

@@ -1,0 +1,197 @@
+from glob import glob
+from astropy.table import Table
+from astropy.io import fits
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+
+from spectrum import Spectrum
+from model import Clam, apply_radial_velocity_shift
+from continuum_basis import Polynomial, Sinusoids
+
+import matplotlib
+#from mpl_utils import mpl_style
+#matplotlib.style.use(mpl_style)
+
+def load_harps_e2ds(path, rest_frame=True, mask_na_doublet=True, ignore_last_order=True):
+
+    pattern = path.split("_e2ds_")[0] + "*_bis*.fits"
+    bis_path = glob(pattern)[0]
+    with fits.open(bis_path) as image:
+        bis_rv = image[0].header["ESO DRS BIS RV"]
+        berv = image[0].header["ESO DRS BERV"]
+
+    v_shift = -bis_rv + berv
+    print(v_shift)
+
+    spectra = []
+    with fits.open(path) as image:
+        i, coeff = (0, [])
+        header_coeff_key = "HIERARCH ESO DRS CAL TH COEFF LL{i}"
+        while image[0].header.get(header_coeff_key.format(i=i), False):
+            coeff.append(image[0].header[header_coeff_key.format(i=i)])
+            i += 1
+
+        n_orders, n_pixels = image[0].data.shape
+        x = np.arange(n_pixels)# - n_pixels // 2
+
+        coeff = np.array(coeff).reshape((n_orders, -1))
+        λ = np.array([np.polyval(c[::-1], x) for c in coeff])
+
+        n_pixels = 0
+        K = n_orders - 1 if ignore_last_order else n_orders
+        for i in range(K): # ignore last order until we have a good telluric model
+            flux = image[0].data[i]
+            ivar = 1/image[0].data[i]
+            bad_pixel = (flux == 0) | ~np.isfinite(flux) | (flux < 0)
+            flux[bad_pixel] = 0
+            ivar[bad_pixel] = 0
+
+            if rest_frame:
+                λ_order = apply_radial_velocity_shift(λ[i], v_shift)
+            else:
+                λ_order = λ[i]            
+
[…]

### 2024-08-22 code/investigate_specutils_air_to_vac.py

@@ -1,0 +1,131 @@
+from glob import glob
+from astropy.table import Table
+from astropy.io import fits
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+
+from spectrum import Spectrum
+from model import Clam, apply_radial_velocity_shift
+from continuum_basis import Polynomial, Sinusoids
+
+import matplotlib
+#from mpl_utils import mpl_style
+#matplotlib.style.use(mpl_style)
+
+def load_harps_e2ds(path, rest_frame=True, mask_na_doublet=True, ignore_last_order=True):
+
+    pattern = path.split("_e2ds_")[0] + "*_bis*.fits"
+    bis_path = glob(pattern)[0]
+    with fits.open(bis_path) as image:
+        bis_rv = image[0].header["ESO DRS BIS RV"]
+        berv = image[0].header["ESO DRS BERV"]
+
+    v_shift = -bis_rv + berv
+
+    spectra = []
+    with fits.open(path) as image:
+        i, coeff = (0, [])
+        header_coeff_key = "HIERARCH ESO DRS CAL TH COEFF LL{i}"
+        while image[0].header.get(header_coeff_key.format(i=i), False):
+            coeff.append(image[0].header[header_coeff_key.format(i=i)])
+            i += 1
+
+        n_orders, n_pixels = image[0].data.shape
+        x = np.arange(n_pixels)# - n_pixels // 2
+
+        coeff = np.array(coeff).reshape((n_orders, -1))
+        λ = np.array([np.polyval(c[::-1], x) for c in coeff])
+
+        n_pixels = 0
+        K = n_orders - 1 if ignore_last_order else n_orders
+        for i in range(K): # ignore last order until we have a good telluric model
+            flux = image[0].data[i]
+            ivar = 1/image[0].data[i]
+            bad_pixel = (flux == 0) | ~np.isfinite(flux) | (flux < 0)
+            flux[bad_pixel] = 0
+            ivar[bad_pixel] = 0
+
+            if rest_frame:
+                λ_order = apply_radial_velocity_shift(λ[i], v_shift)
+            else:
+                λ_order = λ[i]            
+
+            if mask_na_doublet:
+                na_double = (5898 >= λ[i]) * (λ[i] >= 5886)
[…]

### 2024-08-22 code/model.py

@@ -219,10 +219,12 @@ class Clam:
                 model_rectified_flux,
             )
             v_rel = -v_shift
-
-        fig, ax = plt.subplots()
-        ax.plot(λ_vacuum, z, c='k')
-        ax.plot(λ_vacuum, A @ result.x, c="tab:red")
+        
+        #import matplotlib.pyplot as plt
+        #fig, ax = plt.subplots()
+        #ax.plot(λ_vacuum, z, c='k')
+        #ax.plot(λ_vacuum, A @ result.x, c="tab:red")
+        #raise a
 
         # Compute continuum for all other orders given this rectified flux.
         λ_vacuum, z, inv_sigma_z, oi, pi, S, P = prepared_spectra = _prepare_spectra(spectra)
@@ -266,6 +268,7 @@ class Clam:
         vsini: Optional[float] = None,
         callback: Optional[callable] = None,
         op_kwds: Optional[dict] = None,
+        **kwargs
     ):
         """
         Fit the model to some spectra in the observed frame.
@@ -299,6 +302,7 @@ class Clam:
             R=R,
             v_rel=v_rel,
             vsini=vsini,
+            **kwargs,
         )
         print(v_rel)
 
@@ -532,6 +536,7 @@ if __name__ == "__main__":
         tel_λ, tel_label_names, tel_parameters, tel_W, tel_H = pickle.load(fp)
 
     # Let's fit every order individually and see what happens.
+    
     model = Clam(
         λ, 
         H,

### 2024-09-04 paper/main.tex
rollin'

@@
 \newcommand{\kmps}{\unit{km\,s^{-1}}}
-\newcommand{\transpose}{^\top}
+\newcommand*{\transpose}{^{\mkern-1.5mu\mathsf{T}}}
+%\newcommand{\transpose}{^\mathsf{T}}
 
@@
 % Other people who will be invited as co-authors (non-exhaustive list):
-%   Wheeler, Sayjdari, Bedell, Astra folks, CCA data group
+%   Holtzman, Wheeler, Sayjdari, Bedell, Dream team, Astra folks, CCA data group
 
@@
 These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent measurements.
-Here we present a \emph{linear} model that simultaneously fits stellar absorption and the joint continuum-instrument response.
-The joint continuum-instrument response is modelled with a Fourier basis, and stellar absorption is modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
-We factorize telluric absorption in the same way, allowing us to simultaneously fit stellar and telluric absorption.
+Here we present a \emph{linear} model that simultaneously fits stellar absorption, telluric transmission, and the joint continuum-instrument response.
+The joint continuum-instrument response is modelled with a Fourier basis.
+Stellar absorption and telluric transmission is each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
 The non-negativity constraint ensures that basis vectors are strictly additive.
 Together this set of bases represents a linear model for stellar spectra, which ensures that inference is convex, stable, and fast.
-This model allows us to reliably fit the stellar absorption, telluric absorption, the continuum, radial velocity, and rotational broadening, without any prior knowledge about the fundamental stellar parameters.
+This model allows us to reliably fit the stellar absorption, tellurics, continuum, radial velocity, and rotational broadening, without any prior knowledge about the fundamental stellar parameters.
 We demonstrate this method by fits to high-resolution echelle spectra, where we achieve \todo{X\%} accuracy in continuum normalization with no apparent biases at low signal-to-noise ratios (e.g., $<10$).
@@
 
+%The typical analysis procedure for this kind of spectra might involve selecting (by hand) pixels that do not appear to be obviously affected by line (or molecular) absorption, and fitting a smooth function to represent the continuum.
+This process would be repeated per-order, often ignoring the selection of pixels in the previous order (which overlaps in wavelength).
+A subsequent local continuum determination would be made for each line.
+
 \section{Methods}\label{sec:methods}
 
-We will assume a forward model that includes two components: one that represents continuum-normalized absorption (e.g., line absorption); and another to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
-These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will require both components to be linear models, which ensures that inference is stable and fast.
-In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and for estimating stellar parameters.\\
-
-%In all variants we assume that non-negative matrix factorization (NMF) is sufficient to represent continuum-normalized line absorption in a large variety of stellar spectra.
+We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
+These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast.
+In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and handling other nuisances.\\
 
@@
 
-%We further assume that the sine-and-cosine basis is sufficiently flexible for modelling the joint instrument-continuum response.
-There are choices to be made in terms of the degree and length scale used for the sine-and-cosine basis, but these are application-specific choices.\\
-
 Here we will describe the method in general before outlining explicit assumptions and implementation details.
 The data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
-The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric absorption model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
+The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
 \begin{align}
@@
 \end{eqnarray}
-\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into the summation:
+\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into a helpful summation:
 \begin{align}
@@
 
-We now turn to defining the stellar absorption model, the telluric absorption model, and the continuum model.
+We now turn to defining each model component in detail.
 The stellar absorption model $f(\lambda_i;\vecalpha)$ predicts the rectified stellar line absorption at wavelength $\lambda_i$ given parameters $\vecalpha$.
@@
 We refer to our $N \times D$ matrix of continuum-rectified theoretical stellar spectra as $\vec{S}$.
-This is a dense matrix: there are no entries of exactly zero.
+This is a dense matrix: there are no entries of exactly zero, with many entries near 1 (no absorption).
 However, a small transformation to this matrix can make it it extremely sparse.
@@
 We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex, do not require many choices, and they make for an efficient problem for graphical processing units.\\
-%At inference time, $\vecalpha$ is analogous to the basis weights found at training time \\
-
-\todo{We used the BOSZ grid. Some details about that, including settings used for factorizing.}\\
 
@@
 
+We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model.
+This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum.
+We clipped any BOSZ normalized flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements.
+As we discuss in later sections, emission lines are sufficiently rare in stellar spectra that they do not contribute significantly to the $\chi^2$ at inference time.
+We used \todo{all BOSZ spectra and factorized with CD into K basis vectors for 1000 steps, at which point the reconstruction loss was X}.\\
+
 \begin{figure*}
-    \caption{A schematic illustrating the non-negative matrix factorization procedure, with some example basis vectors computed from the application to \emph{BOSS} spectra. \label{fig:schematic}}
+    \caption{A schematic showing some basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra. \label{fig:schematic}}
 \end{figure*}
 
-%For the applications presented in this paper, the factorization can be completed in the order of minutes.
+\todo{Discussion of the stellar and telluric basis vectors, how they separate out into astrophysically relevant things, and refer to the figure.}
 
-%.
-Factorizing the data-driven and hybrid approaches is described later in this section.
-In any case, computing the factorization can be finished in seconds or hours, depending on the scale of the matrices $\vecW$ and $\vecH$.\\
-
-%We used the \texttt{scipy.decomposition.NMF} implementation \citep{scipy} with some minor adjustments to minimize the memory requirements (e.g., allowing for lower precision float types).
-We used no regularization on $\vecW$ or $\vecH$.
-We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition with zeros filled-in with the average of $\vecX$ (the default behavior in the \texttt{scipy} implementation).
-Initializing with random or small non-negative values produced comparable results but required many more iterations.
-We stopped the approximation of $\vecW$ and $\vecH$ after 1,000 iterations of multiplicative updates.
-This took minutes to hours to complete, depending on the size of the theoretical grid $\vecX$.\\
-
-\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line absorption function $\log{f(\lambda_i;\vecalpha)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
+\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line absorption function $\log{f(\lambda_i;\vecalpha)}$, which follows from equations~\ref{eq:log_y} and \ref{eq:nmf},
 \begin{align}
@@
 Here, $\vecalpha$ is analogous to a column in $\vecW_\star$: it represents the $K$ weights needed to reconstruct the stellar absorption from the $K$ basis vectors $\vec{F}$.
-Note that because $\vecalpha$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $g(...)$ to represent the smooth continuum-instrument response.
+Note that because $\vecalpha$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $h(...)$ to represent the smooth continuum-instrument response.
 \\
 
-Telluric line absorption can be included in a similar way to the stellar line absorption.
+Telluric transmission can be included in a similar way to the stellar line absorption.
 We factorize the negative log of theoretical telluric absorption into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$.
 \todo{Details about what telluric grid we used, and setting used for factorizing.
-Note how many vectors we chose ($L$)} With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
+Note how many vectors we chose ($L$).} With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
 \begin{eqnarray}
@@
 \begin{align}
-    \log{h(\lambda_i;\vecgamma)} = \vec{G}\vecgamma
+    \log{h(\lambda_i;\vecgamma)} = \vec{H}\vecgamma
 \end{align}
@@
 \begin{align}
-    \vec{H}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
-                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
+    \vec{H}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{P}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
+    \displaystyle\sin\left(\frac{\pi\,j}{P}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
 \end{align}
-\noindent{}where $L$ \todo{overloaded} is a length scale. \todo{Talk more about this.}
-The design matrix $\vec{H}$ can be constructed \emph{a priori} before inference begins. Throughout this paper we will describe $\vecgamma$ as the sine-and-cosine \emph{coefficients}.\\
+\noindent{}where $P$ is a length scale which we set to be twice the length of the wavelength range to fit (e.g., twice the peak-to-peak range of an echelle order).
+The design matrix $\vec{H}$ can be constructed \emph{a priori} before inference begins.
+Throughout this paper we will describe $\vecgamma$ as the sine-and-cosine \emph{coefficients}.\\
 
@@
 \noindent{}but this does not guarantee that the bounds $\vecalpha \geq 0$ and $\vecbeta \geq 0$ are met.
+There are a few ways to find a solution to this system that is within the parameter bounds (e.g., linear operators; expectation-maximization), and thankfully all of them are convex paths.
+For simplicity we use the truncated reflective trust region algorithm, where we provide the Jacobian ($\vec{A}$).\\
 
-\todo{Talk about how we can solve this as} $||\vec{Y}/\sqrt{C} - \vec{AX}\sqrt{C}||^2$ with linear operator, truncated reflective region, or bounded least-squares with a provided jacobian ($\vec{A}$).
-Thankfully, each of these optimization paths is convex.\\
-
-\todo{Move this to 3.2?}
-In later sections we describe some situations where we may want to penalize values of $\vecalpha$.
+%In later sections we describe some situations where we may want to penalize values of $\vecalpha$.
 We perform this by feature weighting, which is equivalent to setting a prior on the distribution that $\vecalpha$ can be drawn from.
@@
 We can scale the strength of this feature weighting by a scalar $\lambda$ (\todo{nomenclature overload of lambda}) and solve for the parameters $\vec{\tilde{X}}$ with
-\begin{eqnarray}
-    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A} + \lambda\vec{\Lambda})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} \quad .
-\end{eqnarray}\\
+%\begin{eqnarray}
+%    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A} + \lambda\vec{\Lambda})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} %\quad .
+%\end{eqnarray}\\
 
@@
 \begin{eqnarray}
-    \mathrm{rectified~stellar~flux}, &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\alpha}}\right)} \\
-    \mathrm{telluric~absorption}, &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\tilde{\beta}}\right)} \\
-    \mathrm{continuum}, &h& = \exp{\left(\vec{H}\boldsymbol{\tilde{\gamma}}\right)}
+    \mathrm{rectified~stellar~flux}, \quad &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\alpha}}\right)} \\
+    \mathrm{telluric~transmission},\quad  &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\tilde{\beta}}\right)} \\
+    \mathrm{continuum}, \quad &h& = \exp{\left(\vec{H}\boldsymbol{\tilde{\gamma}}\right)}
 \end{eqnarray}
 
-%In this formalism, spectral regions can be fit independently with different degrees of freedom. The number of continuum coefficients scales as $2n_\textrm{degree} + 1$ per region. \\
+\section{Data} \label{sec:data}
+
+Here we seek demonstrate the effectiveness of our model on high-resolution echelle spectra: a scenario in which there are lots of data, and lots of nuisances.
+The nuisances (e.g., continuum parameters per order; tellurics) are often fit separately which can bias subsequent inferences, and the large data volume (e.g., $3\times10^6$~pixels per spectrum) can make even the fastest non-linear interpolators untenably slow.\\
+
+We searched the \eso\ data archive and selected a set of stars that span the range of spectral types and evolutionary states observed with the \harps\ instrument.
+The set include OBAFGKM-type stars, main-sequence stars, and red giant branch stars.
+This selection is not intended to be complete, but is meant to demonstrate the effectiveness of a single linear model for fitting stars across a range of stellar parameters and evolutionary states.
+\\
+
+We downloaded a psuedo-random exposure for each star.
+No explicit cut to ensure a minimum signal-to-noise (SNR) ratio, but in practice most exposures have high SNR.
+We took the measured radial velocity in the calibration files for each spectrum and the computed Barycentric Earth radial velocity to shift the stellar absorption basis vectors to the observed frame.\\
+
+We modelled the continuum in each echelle order as a set of $\todo{M}$ Fourier modes.
+With $\todo{\approx{70}}$ echelle orders, $\todo{K}$ stellar basis vectors, and $\todo{L}$ telluric vectors, there are $\todo{400}$ parameters and $\sim3\times10^6$ data points.
+Without \emph{any prior knowledge of the stellar parameters}, we directly solve the (unconstrained) weighted least-squares system (Equation~\ref{eq:wls}).
+If the solution is outside the parameter bounds, we optimize from this position using truncated reflective trust region algorithm.\\
+
+\section{Results} \label{sec:results}
+
+We found it usually takes a few core-seconds to fit each exposure ($3\times10^6$~pixels; all echelle orders).
+Most of this time is spent constructing the design matrix, which required (linearly) interpolating the stellar and telluric basis vectors on to the observed wavelengths.
+The fits to each star are partially shown in Figure~\ref{fig:harps-example}.
+In all cases the data are shown in black, with the stellar absorption model in \todo{red}, telluric transmission in \todo{color}, and continuum in \todo{color}.
+In bluer echelle orders, there are no pixels that are true `continuum' pixels (i.e., no absorption).
+Any continuum normalization method trying to fit these echelle orders would be biased if they did not have a good representation of what stellar spectra look like.
+Overall, the fits to the data look very good, with reduced $\chi^2$ values from \todo{A to B}.\\
+
+\todo{Consistency at low signal-to-noise}
+% alf Cen 95th percentile (all wavelengths)
+% for S/N > 100, 1-sigma scatter is 0.22%
+% for S/N > 50, 1-sigma scatter is 0.35%
+% for S/N > 30, 1-sigma scatter is 0.46%
+
+% RASSINE claims 1% (2% in their metric, which is 2 sigma) continuum accuracy of alpha Cen B with HARPS spectra.
+% When they exclude bluer than 4500 Angstrom, their continuum accuracy improves to 0.6% (1.2% in their metric)
+
+% If our perormance on alpha Cen A is comparable to theirs on alpha Cen B, then
+% our method is about 5x more accurate
 
@@
 
-%\subsection{Model-driven approach: Line absorption is factorized by theoretical spectra}
-%\label{sec:model-method}
-
-% TODO: re-word this
-
-\section{Data}
-
-We demonstrate the effectiveness of our method by forward modeling high-resolution optical echelle spectra taken with the MIKE spectrograph on the Magellan Clay telescope.
-These spectra have a spectral resolution of $\mathcal{R} \approx X$ and a wavelength range from $X$ to $Y$\,nm.
-The spectra are recorded on the 2D image plane as a series of $X$ orders: $X$ in the blue arm (X to Y) and $Y$ in the red arm (Y to Z).
-\\
-
-\todo{Description of the blaze function, although this is not known perfectly}
-
-The typical analysis procedure for this kind of spectra might involve selecting (by hand) pixels that do not appear to be obviously affected by line (or molecular) absorption, and fitting a smooth function to represent the continuum.
-This process would be repeated per-order, often ignoring the selection of pixels in the previous order (which overlaps in wavelength).
-A subsequent local continuum determination would be made for each line.
-
-\todo{Details about the synthetic grid used, and how we trained the NMF. Include details about the telluric basis used as well.}\\
-
-\todo{Assumptions about the continuum parameters}\\
-
-We measure the radial velocity by cross-correlating the spectra against all basis vectors. \todo{More details about this.} In practice this alternating
-
-\subsection{High-resolution infrared spectra}
-- SDSS
-%The fifth generation of the \emph{Sloan Digital Sky Survey}  is acquiring low-resolution ($\mathcal{R} \sim 4{,}000$) optical spectra for millions of Milky Way stars using the \boss\ spectrograph at Apache Point Observatory (APO) and Las Campanas Observatory (LCO).
-Although the highest priority targets observed by \boss\ tend to be Black Hole Mapper (BHM) targets, many Milky Way stars are included per field, either because they are specifically targeted as part of a Milky Way Mapper (MWM) `carton'\footnote{`Cartons' are how targets are now assigned in SDSS.}, or because they are standard stars.
-We took all \boss\ spectra where the source was allocated to a Milky Way Mapper carton, or spectra where the source was allocated to a stellar-like carton.
-\todo{Some description of stellar-like.}\\
-
-%We used the BOSZ spectral grid to construct basis vectors for a line absorption model for \boss\ spectra.
-This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a sufficiently large range in stellar parameters.
-We convolved the spectra to the nominal mean resolving power of the \boss\ spectrograph ($\mathcal{R} \sim 4{,}000$) and resampled it to the same pixels that the \boss\ spectra are resampled to: a uniform-in-log wavelength sampling from approximately 360\,nm until about 1000\,nm.
-We clipped any BOSZ flux values with emission lines exceeding normalized flux values of 1, as they would violate our NMF requirements.
-As we discuss in later sections, any emission lines are sufficiently rare in stellar spectra that they do not contribute significantly to the $\chi^2$ at inference time.\\
-
-- Korg Grid
-
-- factorize for 1000 steps
-
-- compute H
-
-%Given some observed \emph{BOSS} spectrum, we chose to fit the parameters $\vecalpha$ and $\vectheta$ simultaneously by minimizing the $\chi^2$ difference between the forward model and the data.
-We found that setting $\vecalpha$ to a very small value ($10^{-12}$; i.e., no line absorption) and solving for the continuum coefficients $\vectheta$ was a reasonably good initialization.
-In situations where there are spectra of the same star we advocate fitting all visit spectra simultaneously using a single set of basis weights $\vecalpha$ for all spectra, and continuum coefficients $\vectheta_i$ for the $i$th epoch spectrum.
-
-\section{Results}
-\label{sec:results}
+This is a convex problem that is mathmatically guaranteed to yield the solution that minimizes the $\chi^2$ between the (log-)transformed data and model.
+However, it does not guarantee that the solution is sensible, or even physically possible.
+We expand on this in the \emph{Discussion} (Section~\ref{sec:discussion}), but the point here is to highlight that the solution might be unphysical: we might have a non-zero basis weight for a basis vector that only contains strong hydrogen lines (indicating a hot star), and we might have a strong non-zero basis weight for a basis vector that only contains molecular bands seen in very cool stars.
 
@@
 % All these people are likely co-authors, but should be thanked if they don't want to be co-authors
+    Jon Holztman (NMSU),
     Rory Smith (Monash),

### 2024-09-04 paper/main.tex
paper

@@
 The non-negativity constraint ensures that basis vectors are strictly additive.
-Together this set of bases represents a linear model for stellar spectra, which ensures that inference is convex, stable, and fast.
+Together this set of bases represents a very effective linear model for stellar spectroscopy, where the linearity ensures that inference is convex, stable, and fast.
 This model allows us to reliably fit the stellar absorption, tellurics, continuum, radial velocity, and rotational broadening, without any prior knowledge about the fundamental stellar parameters.
-We demonstrate this method by fits to high-resolution echelle spectra, where we achieve \todo{X\%} accuracy in continuum normalization with no apparent biases at low signal-to-noise ratios (e.g., $<10$).
+We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of OBAFGKM-type stars.
+% TODO: talk about how fast it is for 10^6 pixels and hundreds of params?
+We show from repeat observations of $\alpha$-Centauri A that our approach yields continuum estimates that are about five times more consistent than the best existing methods: better than 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
 \end{abstract}
+
+% for S/N > 100, 1-sigma scatter is 0.22%
+% for S/N > 50, 1-sigma scatter is 0.35%
+% for S/N > 30, 1-sigma scatter is 0.46%
 
@@
 
-We found it usually takes a few core-seconds to fit each exposure ($3\times10^6$~pixels; all echelle orders).
-Most of this time is spent constructing the design matrix, which required (linearly) interpolating the stellar and telluric basis vectors on to the observed wavelengths.
+We found it usually takes a few core-seconds to fit each exposure ($3\times10^6$~pixels; all echelle orders fit simultaneously).
+Most of this time is spent constructing the design matrix, which requires (linearly) interpolating the stellar and telluric basis vectors to the observed wavelengths.
 The fits to each star are partially shown in Figure~\ref{fig:harps-example}.
 In all cases the data are shown in black, with the stellar absorption model in \todo{red}, telluric transmission in \todo{color}, and continuum in \todo{color}.
-In bluer echelle orders, there are no pixels that are true `continuum' pixels (i.e., no absorption).
+Overall, the fits to the data are very good, with reduced $\chi^2$ values from \todo{A to B}.\\
+
+In bluer echelle orders, it can be seen that there are no pixels that are true `continuum' pixels (i.e., no absorption).
 Any continuum normalization method trying to fit these echelle orders would be biased if they did not have a good representation of what stellar spectra look like.
-Overall, the fits to the data look very good, with reduced $\chi^2$ values from \todo{A to B}.\\
+\\
+
+Maintaining consistent continuum rectification for low signal-to-noise spectra is challenging.
+Without a model for stellar spectra or the relative strengths of absorption lines, noisy signals can mimic stellar absorption, causing the continuum to become systematically biased at low signal-to-noise ratios.
+For this reason, we sought to estimate the consistency of continuum normalization as a function of signal-to-noise ratio.\\
+
+We searched the \eso/\harps/ archive for a star that has been repeatedly observed with a wide range of exposure times, usually because it is the subject of radial velocity asteroseismic monitoring.
+Through this search we identified alpha Centauri A, which has been observed more than \todo{X} times with \harps.
+Importantly, there are $\alpha$-Centauri spectra with signal-to-noise ratios of 10, and a nearly even sampling of spectra up to S/N of 300, or even higher (where saturation is visible in the reduced spectra\footnote{Naughty observer.}).
+We constructed bins of SNR from 0 to 200 in steps of 10.
+We selected an $\alpha$-Centauri spectrum for every SNR bin interval to provide an even sampling of spectra at low and high SNR.
+We fit our model to each spectrum, treating it as totally indepdendent from all others.\\
+
+\todo{Now we can see how things behave as a function of SNR.}
 
@@
 
-We want to show:
-\begin{enumerate}
-    \item How does the mean/median rectified flux vary as a function of S/N for the same star (e.g., 18 Sco)?
-    \item What do the fits look like for a few carefully chosen spectral types?
-    \item What regularisation strength did we need to apply?
-\end{enumerate}
-
 \begin{figure}
@@
 
+- getting bad models with good fits
+
+- emission lines
+
 - model interpretability
+
+- estimating precision in continuum rectification
 

### 2024-09-04 code/harps_alfCenA_wrt_snr.py
fear commit

@@ -4,6 +4,7 @@ from astropy.io import fits
 import matplotlib.pyplot as plt
 import numpy as np
 import pickle
+import os
 
 from spectrum import Spectrum
 from model import Clam, apply_radial_velocity_shift
@@ -100,27 +101,35 @@ for row in tqdm(meta_subset):
 
     try:
         path = "../applications/harps/alfCenA/" + row["path"]
-        spectra = load_harps_e2ds(path)
+        if os.path.exists(f"{path}.pkl"):
+            with open(f"{path}.pkl", "rb") as fp:
+                (result, continuum, rectified_model_flux, rectified_telluric_flux, y_pred) = pickle.load(fp)
+        else:
+                
+            spectra = load_harps_e2ds(path)
 
-        (result, continuum, rectified_model_flux, rectified_telluric_flux, y_pred) = model.fit(
-            spectra, 
-            continuum_basis=Sinusoids(5),
-            λ_initial=4861,
-            v_rel=0.0 # already at rest frame.
-        )
-        with open(f"{path}.pkl", "wb") as fp:
-            pickle.dump((result, continuum, rectified_model_flux, rectified_telluric_flux, y_pred), fp)
+            (result, continuum, rectified_model_flux, rectified_telluric_flux, y_pred) = model.fit(
+                spectra, 
+                continuum_basis=Sinusoids(5),
+                λ_initial=4861,
+                v_rel=0.0 # already at rest frame.
+            )
+            with open(f"{path}.pkl", "wb") as fp:
+                pickle.dump((result, continuum, rectified_model_flux, rectified_telluric_flux, y_pred), fp)
             
-        percentiles = np.percentile(np.array([s.flux / c for s, c in zip(spectra, continuum)]).flatten(), [90, 95, 97.5])
     except:
         print(f"Failured on {row['path']}")
 
     else:
-        results.append((row["path"], row["snr"], result.status, result.cost, result.optimality, *percentiles))
+        norm_flux = np.array([s.flux / c for s, c in zip(spectra, continuum)]).flatten()
+        percentiles = np.percentile(norm_flux, [90, 95, 97.5])
+        std = np.std(norm_flux)
+
+        results.append((row["path"], row["snr"], result.status, result.cost, result.optimality, std,  *percentiles))
 
 from astropy.table import Table
[…]

### 2024-09-04 code/harps_alfCenA_wrt_snr_polynomial.py

@@ -1,0 +1,140 @@
+from glob import glob
+from astropy.table import Table
+from astropy.io import fits
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+
+from spectrum import Spectrum
+from model import Clam, apply_radial_velocity_shift
+from continuum_basis import Polynomial, Sinusoids
+
+import matplotlib
+#from mpl_utils import mpl_style
+#matplotlib.style.use(mpl_style)
+
+def load_harps_e2ds(path, rest_frame=True, mask_na_doublet=True, ignore_last_order=True):
+
+    pattern = path.split("_e2ds_")[0] + "*_bis*.fits"
+    bis_path = glob(pattern)[0]
+    with fits.open(bis_path) as image:
+        bis_rv = image[0].header["ESO DRS BIS RV"]
+        berv = image[0].header["ESO DRS BERV"]
+
+    v_shift = -bis_rv + berv
+    print(v_shift)
+
+    spectra = []
+    with fits.open(path) as image:
+        i, coeff = (0, [])
+        header_coeff_key = "HIERARCH ESO DRS CAL TH COEFF LL{i}"
+        while image[0].header.get(header_coeff_key.format(i=i), False):
+            coeff.append(image[0].header[header_coeff_key.format(i=i)])
+            i += 1
+
+        n_orders, n_pixels = image[0].data.shape
+        x = np.arange(n_pixels)# - n_pixels // 2
+
+        coeff = np.array(coeff).reshape((n_orders, -1))
+        λ = np.array([np.polyval(c[::-1], x) for c in coeff])
+
+        n_pixels = 0
+        K = n_orders - 1 if ignore_last_order else n_orders
+        for i in range(K): # ignore last order until we have a good telluric model
+            flux = image[0].data[i]
+            ivar = 1/image[0].data[i]
+            bad_pixel = (flux == 0) | ~np.isfinite(flux) | (flux < 0)
+            flux[bad_pixel] = 0
+            ivar[bad_pixel] = 0
+
+            if rest_frame:
+                λ_order = apply_radial_velocity_shift(λ[i], v_shift)
+            else:
+                λ_order = λ[i]            
+
+            if mask_na_doublet:
+                na_double = (5898 >= λ[i]) * (λ[i] >= 5886)
+                ivar[na_double] = 0
+
[…]

### 2024-09-04 code/harps_alfCenA_wrt_snr_reference.py

@@ -1,0 +1,121 @@
+from glob import glob
+from astropy.table import Table
+from astropy.io import fits
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+
+from spectrum import Spectrum
+from model import Clam, apply_radial_velocity_shift, get_closest_order
+from continuum_basis import Polynomial, Sinusoids
+
+import matplotlib
+#from mpl_utils import mpl_style
+#matplotlib.style.use(mpl_style)
+
+def load_harps_e2ds(path, rest_frame=True, mask_na_doublet=True, ignore_last_order=True):
+
+    pattern = path.split("_e2ds_")[0] + "*_bis*.fits"
+    bis_path = glob(pattern)[0]
+    with fits.open(bis_path) as image:
+        bis_rv = image[0].header["ESO DRS BIS RV"]
+        berv = image[0].header["ESO DRS BERV"]
+
+    v_shift = -bis_rv + berv
+    print(v_shift)
+
+    spectra = []
+    with fits.open(path) as image:
+        i, coeff = (0, [])
+        header_coeff_key = "HIERARCH ESO DRS CAL TH COEFF LL{i}"
+        while image[0].header.get(header_coeff_key.format(i=i), False):
+            coeff.append(image[0].header[header_coeff_key.format(i=i)])
+            i += 1
+
+        n_orders, n_pixels = image[0].data.shape
+        x = np.arange(n_pixels)# - n_pixels // 2
+
+        coeff = np.array(coeff).reshape((n_orders, -1))
+        λ = np.array([np.polyval(c[::-1], x) for c in coeff])
+
+        n_pixels = 0
+        K = n_orders - 1 if ignore_last_order else n_orders
+        for i in range(K): # ignore last order until we have a good telluric model
+            flux = image[0].data[i]
+            ivar = 1/image[0].data[i]
+            bad_pixel = (flux == 0) | ~np.isfinite(flux) | (flux < 0)
+            flux[bad_pixel] = 0
+            ivar[bad_pixel] = 0
+
+            if rest_frame:
+                λ_order = apply_radial_velocity_shift(λ[i], v_shift)
+            else:
+                λ_order = λ[i]            
+
+            if mask_na_doublet:
+                na_double = (5898 >= λ[i]) * (λ[i] >= 5886)
+                ivar[na_double] = 0
+
[…]

### 2024-09-04 code/model.py

@@ -176,12 +176,15 @@ class Clam:
         vsini: Optional[float] = None,
         v_rel: Optional[float] = None,
         op_kwds: Optional[dict] = None,
+        initial_order = None,
     ):
         """
         Fit one order to get an initial estimate of the model parameters.
         """
-
-        spectrum = get_closest_order(spectra, λ_initial)
+        if initial_order is not None:
+            spectrum = initial_order
+        else:
+            spectrum = get_closest_order(spectra, λ_initial)
         λ_vacuum, z, inv_sigma_z, oi, pi, S, P = _prepare_spectra([spectrum])
         # TODO: using diferent initial polynomial 
         G, *_ = _continuum_design_matrix(λ_vacuum, Polynomial(2), S, oi)

### 2024-09-16 code/spectrum/__init__.py
add spectrum loaders for e2ds; clean up

@@ -1,0 +1,7 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+""" Spectroscopy-related utilities. """
+
+from .spectrum import Spectrum, SpectrumCollection
+from . import loaders

### 2024-09-16 code/spectrum/loaders.py

@@ -1,0 +1,349 @@
+import os
+import re
+import numpy as np
+import warnings
+from astropy.io import fits
+from glob import glob
+from hashlib import md5
+from specutils.io.registers import data_loader
+from specutils.io.parsing_utils import read_fileobj_or_hdulist
+
+from .spectrum import (Spectrum, SpectrumCollection)
+from .utils import (
+    concatenate_wat_headers, compute_linear_dispersion, compute_dispersion, sigma_to_ivar, get_meta_dict
+)
+
+"""
+A helpful note for the programmer:
+
+The naming of `Spectrum` and `SpectrumCollection` loosely follows the convention set by astropy.
+
+- Use `Spectrum` for a single one-dimensional spectrum.
+- Use `SpectrumCollection` for a collection of one-dimensional spectra that have the same flux shape
+  (e.g., echelle orders are often the same shape because they have the same pixels per echelle order,
+  even if the wavelength per order is different).
+
+- Use `SpectrumList` for a collection of one-dimensional spectra that have different flux shapes.
+
+** `SpectrumList` is not yet implemented yet **
+"""
+
+
+# Identifiers
+def path_given_and_does_not_match(args, pattern):    
+    path = args[0]
+    return (
+        isinstance(path, (str, os.PathLike))
+        and 
+        (re.match(pattern, os.path.basename(path)) is None)
+    )
+
+def path_given(args):
+    return isinstance(args[0], (str, os.PathLike))
+
+def basename_matches(args, pattern):
+    return (re.match(pattern, os.path.basename(args[0])) is not None)
+
+def _get_instrument(*args, **kwargs):
+    with read_fileobj_or_hdulist(*args, **kwargs) as hdulist:
+        return hdulist[0].header.get("INSTRUME", None)
+
+
+def identify_multispec(origin, *args, **kwargs):
+    if path_given_and_does_not_match(args, ".+\.fits$"):
+        return False
+    
+    with read_fileobj_or_hdulist(*args, memmap=False, **kwargs) as hdulist:
+        ctype1 = hdulist[0].header["CTYPE1"].lower()
+        wat0_001 = hdulist[0].header["WAT0_001"].lower()
+        return (ctype1.startswith("multispe") or wat0_001 == "system=multispec")
+    
[…]

### 2024-09-16 code/spectrum/lsf.py

@@ -1,0 +1,152 @@
+import numpy as np
+from scipy import sparse, stats
+from typing import Sequence, Optional
+
+_fwhm_to_sigma = 1 / (2 * np.sqrt(2 * np.log(2)))
+
+
+def instrument_lsf_kernel(
+    λi: Sequence[float],
+    λo: Sequence[float],
+    Ro: float,
+    Ri: Optional[float] = np.inf,
+    **kwargs
+):
+    """
+    Construct a sparse matrix to convolve fluxes at input wavelengths (λi) at an instrument spectral
+    resolution (R) and resample to the given output wavelengths (λo).
+
+    :param λi:
+        A N-length array of input wavelength values.
+
+    :param λo:
+        A M-length array of output wavelength values.
+
+    :param Ro:
+        Spectral resolution.
+
+    :param Ri: [optional]
+        Input spectral resolution (default: infinity).
+
+    :returns:
+        A (N, M) sparse array representing a convolution kernel.
+    """
+    R_inv = np.sqrt(float(Ro) ** (-2) - float(Ri) ** (-2))
+    data, i, j = ([], [], [])
+    for ii, λ in enumerate(λo):
+        si, ei, ϕ = _instrument_lsf_kernel(λi, λ, R_inv, **kwargs)
+        i.extend([ii] * (ei - si))
+        j.extend(range(si, ei))
+        data.extend(ϕ)
+    return sparse.coo_matrix((data, (j, i)), shape=(λi.size, λo.size))
+
+
+def rotational_broadening_kernel(λ: Sequence[float], vsini: float, epsilon: float):
+    """
+    Construct a sparse matrix to convolve fluxes at input wavelengths (λ) with a rotational broadening kernel
+    with a given vsini and epsilon.
+
+    :param λ:
+        A N-length array of input wavelength values.
+
+    :param vsini:
+        The projected rotational velocity of the star in km/s.
+
+    :param epsilon:
+        The limb darkening coefficient.
+
+    :returns:
+        A (N, N) sparse array representing a convolution kernel.
+    """
+
+    # Let's pre-calculate some things that are needed in the hot loop.
+    denominator = np.pi * vsini * (1.0 - epsilon / 3.0)
+    c1 = 2.0 * (1.0 - epsilon) / denominator
+    c2 = 0.5 * np.pi * epsilon / denominator
+
+    vsini_c = vsini / 299792.458
[…]

### 2024-09-16 code/spectrum/spectrum.py

@@ -1,0 +1,104 @@
+
+import numpy as np
+from typing import Sequence, Optional, Union
+from astropy.nddata import NDIOMixin
+from astropy import units as u
+from functools import cached_property
+from specutils.utils.wcs_utils import air_to_vac
+
+from .utils import apply_relativistic_velocity_shift
+
+class Spectrum(NDIOMixin):
+    
+    def __init__(
+        self,
+        λ: Sequence[float],
+        flux: Sequence[float],
+        ivar: Optional[Sequence[float]] = None,
+        pixel_flags: Optional[Sequence[int]] = None,
+        mask: Optional[Sequence[bool]] = None,
+        meta: Optional[dict] = None,
+        v_rel: float = 0.0,
+        vacuum: Optional[bool] = False,
+        continuum: Optional[Union[float, Sequence[float]]] = 1 # should this allow a continuum model of some sort?
+    ):
+        if ivar is None:
+            ivar = np.ones_like(flux)        
+        meta = {} if meta is None else meta        
+        self.λ = np.array(λ)
+        self.flux = np.array(flux)
+        self.ivar = np.array(ivar)
+        self.mask = np.zeros_like(self.flux, dtype=bool) if mask is None else np.array(mask, dtype=bool)
+        self.pixel_flags = pixel_flags
+        bad_pixel = ~np.isfinite(self.flux) | ~np.isfinite(self.ivar)
+        self.ivar[bad_pixel] = 0        
+        self.meta = meta
+        self.vacuum = vacuum
+        self._v_rel = v_rel
+        self.continuum = continuum
+        return None
+    
+    @property
+    def v_rel(self):
+        return self._v_rel
+    
+    @v_rel.setter
+    def v_rel(self, v_rel):
+        self._v_rel = v_rel
+                
+    @property
+    def λ_rest(self):
+        return apply_relativistic_velocity_shift(self.λ, -self.v_rel)
+    
+    @property
+    def λ_rest_vacuum(self):
+        return apply_relativistic_velocity_shift(self.λ_vacuum, -self.v_rel)
+        
+    @cached_property
+    def λ_vacuum(self):
+        if self.vacuum:
+            return self.λ
+        else:
+            # Note: The default method used in specutils is horribly inaccurate! Use Piskunov instead.
+            return air_to_vac(self.λ << u.Angstrom, scheme="Piskunov").value
+
+
+    @property
+    def rectified_flux(self):
+        return self.flux / self.continuum
+
+    @property
[…]

### 2024-09-16 code/spectrum/utils.py

@@ -1,0 +1,277 @@
+""" General utilities for dealing with spectra. """
+
+from astropy.constants import c
+
+import numpy as np
+import time
+
+LARGE = 1e10
+SMALL = 1/LARGE
+C_KM_S = c.to("km/s").value
+
+
+def sigma_to_ivar(sigma):
+    with np.errstate(divide='ignore'):
+        ivar = sigma**-2
+        ivar[sigma >= LARGE] = 0
+        return ivar
+
+def ivar_to_sigma(ivar):
+    with np.errstate(divide='ignore'):
+        sigma = ivar**-0.5
+        sigma[~np.isfinite(sigma)] = LARGE
+        return sigma
+    
+
+def get_meta_dict(hdulist):
+    return dict(hdulist[0].header)
+
+def compute_linear_dispersion(crval, cdelt, naxis, crpix=1, ltv=0):
+    return crval + (np.arange(naxis) + 1 - crpix) * cdelt - ltv * cdelt
+
+def apply_relativistic_velocity_shift(λ, v):
+    beta = v / C_KM_S
+    return λ * np.sqrt((1 + beta) / (1 - beta))
+
+def compute_dispersion(aperture, beam, dispersion_type, dispersion_start,
+    mean_dispersion_delta, num_pixels, redshift, aperture_low, aperture_high,
+    weight=1, offset=0, function_type=None, order=None, Pmin=None, Pmax=None,
+    *coefficients):
+    """
+    Compute a dispersion mapping from a IRAF multi-spec description.
+
+    :param aperture:
+        The aperture number.
+
+    :param beam:
+        The beam number.
+
+    :param dispersion_type:
+        An integer representing the dispersion type:
+
+        0: linear dispersion
+        1: log-linear dispersion
+        2: non-linear dispersion
+
+    :param dispersion_start:
+        The value of the dispersion at the first physical pixel.
+
+    :param mean_dispersion_delta:
+        The mean difference between dispersion pixels.
+
+    :param num_pixels:
+        The number of pixels.
+
+    :param redshift:
+        The redshift of the object. This is accounted for by adjusting the
+        dispersion scale without rebinning:
+
+        >> dispersion_adjusted = dispersion / (1 + redshift)
+
+    :param aperture_low:
+        The lower limit of the spatial axis used to compute the dispersion.
+
+    :param aperture_high:
+        The upper limit of the spatial axis used to compute the dispersion.
+
+    :param weight: [optional]
[…]

### 2024-09-16 code/model.py
fixed telluric air/vac issues

@@ -24,7 +24,7 @@ from typing import Iterable
 
 
 from spectrum.lsf import instrument_lsf_kernel, rotational_broadening_kernel
-from spectrum import Spectrum
+from spectrum import Spectrum, SpectrumCollection
 from scipy.optimize._lsq.trf_linear import trf_linear
 
 
@@ -91,81 +91,39 @@ class Clam:
             self.telluric_basis_vectors is not None 
         and self.telluric_vacuum_wavelength is not None
         )
-            
-
-    def _prepare_basis_vectors(self, λ_observed, λ, basis_vectors, v_rel=None, vsini=None, R=None):
-
-
-        λ_out = apply_radial_velocity_shift(λ, v_rel)
         
-        kernels = []
+        
+    def prepare(self, spectra, vsini=None):
+
+        λ_rest_vacuum, λ_vacuum, *_ = prepared_spectra = _prepare_spectra(spectra)
+        
+        stellar_basis_vectors = self.stellar_basis_vectors
         if vsini is not None:
-            kernels.append(rotational_broadening_kernel(λ, vsini, 0.6))        
-        if Ro is not None:
-            kernels.append(instrument_lsf_kernel(λ, λ_out, R))
+            K_lsf = rotational_broadening_kernel(self.stellar_vacuum_wavelength, vsini, 0.6)
+            stellar_basis_vectors = stellar_basis_vectors @ K_lsf
 
-        raise a
+        basis_vectors = np.vstack([
+            _interpolate_basis_vector(
+                λ_rest_vacuum,
+                self.stellar_vacuum_wavelength, 
+                stellar_basis_vectors
+            )
+        ])
 
+        if self.can_model_tellurics:
+            # need the observed vacuum wavelengths
+            telluric_basis_vectors = _interpolate_basis_vector(
+                λ_vacuum,
+                self.telluric_vacuum_wavelength,
+                self.telluric_basis_vectors
+            )
+            basis_vectors = np.vstack([basis_vectors, telluric_basis_vectors])
 
-    def resample_basis_vectors(self, λ_vacuum, vsini=None, stellar_v_rel=0.0, telluric_v_rel=0.0, Ro=None, Ri=None, telluric=False):
-        """
-        Resample (and optionally convolve) the basis vectors to the observed wavelengths.
-        
-        :param λ_vacuum:
[…]

### 2024-09-16 code/run_harps_examples.py

@@ -1,0 +1,149 @@
+from glob import glob
+from astropy.table import Table
+from astropy.io import fits
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+import os
+
+from spectrum import Spectrum, SpectrumCollection
+from model import Clam, apply_radial_velocity_shift
+from continuum_basis import Polynomial, Sinusoids
+
+import matplotlib.pyplot as plt
+
+
+with open("../applications/harps-sandbox/20240816_train_harps_model.pkl", "rb") as fp:
+    λ, label_names, parameters, W, H = pickle.load(fp)
+
+from specutils.utils.wcs_utils import air_to_vac
+from astropy import units as u
+
+with open("../applications/harps-sandbox/20240816_train_telfit_model.pkl", "rb") as fp:
+    tel_λ, tel_label_names, tel_parameters, tel_W, tel_H = pickle.load(fp)
+
+
+model = Clam(λ, H, 10 * tel_λ, tel_H)
+
+def fix_vrel_2019_12_04(s):
+    s.v_rel = 107.81403 + s.meta["berv"]
+
+options = { 
+    "HARPS.2019-12-04T06:28:13.591_e2ds_A.fits": dict(
+        read_callback=fix_vrel_2019_12_04,
+        fit_kwds=dict(
+            vsini=100
+        )
+    )
+}
+
+# O9.5V: ./applications/harps/examples/O9.5V-ms-mu_Col/data/reduced/2019-12-03/HARPS.2019-12-04T06:28:13.591_e2ds_A.fits --> bad RV (-999999)
+
+#for path in ['../applications/harps/alfCenA/calib/2015-10-15/HARPS.2015-10-15T23:18:37.783_e2ds_A.fits']:
+#for path in ['../applications/harps/examples/O9.5V-ms-mu_Col/data/reduced/2019-12-03/HARPS.2019-12-04T06:28:13.591_e2ds_A.fits']:
+for path in ['../applications/harps/examples/G5-ms-HD_222595/ADP.2014-10-02T10:01:52.103/HARPS.2005-12-20T00:40:22.994_e2ds_A.fits']:
+    print(f"Running {path}")
+
+    spectra = SpectrumCollection.read(path, format="HARPS-e2ds", si=30, ei=None)
+
[…]

### 2024-09-16 code/spectrum/loaders.py

@@ -318,8 +318,8 @@ def load_harps_e2ds(path, **kwargs):
         meta = get_meta_dict(hdulist)
         meta.update(rv_meta)
 
-        si, ei = (30, 40)
-
+        si = kwargs.get("si", 0)
+        ei = kwargs.get("ei", n_orders)
 
         return SpectrumCollection(
             λ=λ[si:ei],

### 2024-09-17 code/continuum_basis.py
probably better for interpretation

@@ -66,4 +66,4 @@ class Polynomial(ContinuumBasis):
         return self.deg + 1
     
     def design_matrix(self, λ):
-        return np.vander(λ, self.deg + 1)
+        return np.vander(λ - np.mean(λ), self.deg + 1)

### 2024-09-17 paper/main.tex
add harps examples

@@
 Here we present a \emph{linear} model that simultaneously fits stellar absorption, telluric transmission, and the joint continuum-instrument response.
-The joint continuum-instrument response is modelled with a Fourier basis.
 Stellar absorption and telluric transmission is each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
 The non-negativity constraint ensures that basis vectors are strictly additive.
+The joint continuum-instrument response is modelled with a Fourier basis. 
 Together this set of bases represents a very effective linear model for stellar spectroscopy, where the linearity ensures that inference is convex, stable, and fast.
-This model allows us to reliably fit the stellar absorption, tellurics, continuum, radial velocity, and rotational broadening, without any prior knowledge about the fundamental stellar parameters.
+This model allows us to reliably fit nuisance parameters (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
 We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of OBAFGKM-type stars.
-% TODO: talk about how fast it is for 10^6 pixels and hundreds of params?
-We show from repeat observations of $\alpha$-Centauri A that our approach yields continuum estimates that are about five times more consistent than the best existing methods: better than 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
+We show from repeat observations of $\alpha$-Centauri A that our approach yields continuum estimates that are consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
 \end{abstract}
@@
 
+\begin{figure*}
+    \includegraphics[width=\textwidth, height=\textwidth]{../code/harps_examples.pdf}
+    \caption{Continuum-rectified \harps\ spectra of example spectral types from B to M.
+The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission.\label{fig:harps-examples}}
+\end{figure*}
+
 \section{Results} \label{sec:results}

### 2024-09-17 code/model.py

@@ -129,7 +129,8 @@ class Clam:
         self,
         spectra: Sequence[Spectrum],
         continuum_basis: Union[ContinuumBasis, Sequence[ContinuumBasis]] = Sinusoids,
-        λ_initial: Optional[float] = 5175,
+        initial_λ: Optional[float] = 5175,
+        initial_continuum_basis: ContinuumBasis = Polynomial(1),
         R: Optional[float] = None,  
         vsini: Optional[float] = None,
         op_kwds: Optional[dict] = None,
@@ -141,12 +142,12 @@ class Clam:
         if initial_order is not None:
             spectrum = initial_order
         else:
-            spectrum = get_closest_order(spectra, λ_initial)
+            spectrum = get_closest_order(spectra, initial_λ)
 
         (λ_rest_vacuum, λ_vacuum, z, inv_sigma_z, oi, pi, S, P), basis_vectors, *n_different_bases = self.prepare([spectrum], vsini=vsini)
         
         # TODO: using diferent initial polynomial 
-        G, *_ = _continuum_design_matrix(λ_vacuum, Polynomial(2), S, oi)
+        G, *_ = _continuum_design_matrix(λ_vacuum, initial_continuum_basis, S, oi)
 
         A = np.hstack([-basis_vectors.T, G])
                 
@@ -200,9 +201,7 @@ class Clam:
         self,
         spectra: Sequence[Spectrum],
         continuum_basis: Union[ContinuumBasis, Sequence[ContinuumBasis]] = Sinusoids,
-        R: Optional[float] = None,
         vsini: Optional[float] = None,
-        callback: Optional[callable] = None,
         op_kwds: Optional[dict] = None,
         **kwargs
     ):
@@ -216,25 +215,17 @@ class Clam:
             The continuum basis to use. This can be a single instance of a continuum basis
             or a sequence of continuum bases. If a sequence is given, then the length must
             match the number of orders in the spectra.
+
+        :param vsini: [optional]
+            The projected rotational velocity of the star in km/s.
         
-        :param R: [optional]
-            The spectral resolution of the observations. If `None` is given, then no
[…]

### 2024-09-17 code/run_harps_examples.py

@@ -7,143 +7,156 @@ import pickle
 import os
 
 from spectrum import Spectrum, SpectrumCollection
+from spectrum.loaders import get_blaze
 from model import Clam, apply_radial_velocity_shift
 from continuum_basis import Polynomial, Sinusoids
 
 import matplotlib.pyplot as plt
 
 
-with open("../applications/harps-sandbox/20240816_train_harps_model.pkl", "rb") as fp:
+#with open("../applications/harps-sandbox/20240816_train_harps_model.pkl", "rb") as fp:
+#    λ, label_names, parameters, W, H = pickle.load(fp)
+
+with open("20240916_train_harps_lsf_model.pkl", "rb") as fp:
     λ, label_names, parameters, W, H = pickle.load(fp)
+
 
 from specutils.utils.wcs_utils import air_to_vac
 from astropy import units as u
 
 with open("../applications/harps-sandbox/20240816_train_telfit_model.pkl", "rb") as fp:
     tel_λ, tel_label_names, tel_parameters, tel_W, tel_H = pickle.load(fp)
+    tel_λ *= 10
+    
 
+with open("20240916_telluric_model.pkl", "rb") as fp:
+    tel_λ, tel_W, tel_H, tel_meta = pickle.load(fp)
 
-model = Clam(λ, H, 10 * tel_λ, tel_H)
+keep = np.ones(10, dtype=bool)
+for index in (0, 2, 3, 5, 7, 9):
+    keep[index] = False
 
-def fix_vrel_2019_12_04(s):
-    s.v_rel = 107.81403 + s.meta["berv"]
+model = Clam(λ, H, tel_λ, tel_H[keep])
+
+def apply_blaze(s, p):
+    blaze = get_blaze(p)
+    s.flux /= blaze
+    s.ivar *= blaze**2
+
+def fix_vrel_2019_12_04(s, p):
+    apply_blaze(s, p)
+    # https://simbad.cds.unistra.fr/simbad/sim-ref?bibcode=2018A%26A...613A..65H
+    # v_rad = 111
+    s.v_rel = 111 + s.meta["berv"]
+
+def fix_vrel_2011_12_25(s, p):
+    s.vrel = -36.16283 + 2.50184
+
+def fix_vrel_2007_07_21(s, p):
+    s.vrel = -(-2*5.22852022752343) + 4.7705
 
 options = { 
     "HARPS.2019-12-04T06:28:13.591_e2ds_A.fits": dict(
[…]

### 2024-09-17 code/spectrum/loaders.py

@@ -256,7 +256,7 @@ def harps(path, **kwargs):
 
 
 def get_blaze(e2ds_path):
-    blaze_path = glob(os.path.dirname(path) + "/*_blaze_" + path[-6:])
+    blaze_path = glob(os.path.dirname(e2ds_path) + "/*_blaze_" + e2ds_path[-6:])
     with fits.open(blaze_path[0]) as blaze_image:
         return blaze_image[0].data
 
@@ -291,10 +291,9 @@ def load_harps_e2ds(path, **kwargs):
     try:
         rv_meta = get_rv_metadata(path)
     except:
-        rv_meta = dict(bis_rv=None, berv=None, v_rel=None)
+        rv_meta = dict(bis_rv=None, berv=None, v_rel=0)
         warnings.warn(f"Could not load metadata from BIS image path at {path}")
-        raise
-    
+        
     with read_fileobj_or_hdulist(path) as hdulist:
         i, coeff = (0, [])
         header_coeff_key = "HIERARCH ESO DRS CAL TH COEFF LL{i}"

### 2024-09-17 paper/main.tex
review abstract, intro, and methods

@@
 
+\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+
 % Other possible titles
+
+% TODO
+% [ ] Talk about BOSZ spectra and how the model was constructed.
+% [ ] Talk about how we constructed the telluric basis vectors
+% [ ] Flesh out discussion points
+% [ ] Data section
+% -> Blaze???
+% [ ] Results section
+% -> ON a set of OBAFGKM stars, provides clean and consistent normalisation across these types
+% -> ARE WE USING THE NIGHTLY BLAZE?
+% -> discuss for alf Cen A, how we selected spectra to get a uniform sampling in SNR, and how we fit them
+% -> how we define the consistency measure
+
+% [ ] discussion
+% -> We fix vsini / v_rel from the literature, but these can be computed 
+% ->
 
@@
 Here we present a \emph{linear} model that simultaneously fits stellar absorption, telluric transmission, and the joint continuum-instrument response.
-Stellar absorption and telluric transmission is each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
+Stellar absorption and telluric transmission are each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
 The non-negativity constraint ensures that basis vectors are strictly additive.
@@
 This model allows us to reliably fit nuisance parameters (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
-We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of OBAFGKM-type stars.
-We show from repeat observations of $\alpha$-Centauri A that our approach yields continuum estimates that are consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
+We demonstrate our method by fitting \todo{toy models of M-dwarf spectra where we recover the correct continuum, and by fitting} \eso/\harps\ high-resolution echelle spectra of OBAFGKM-type stars.
+From repeat observations of $\alpha$-Centauri A we show that our approach yields continuum estimates that are consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
 \end{abstract}
@@
 
-Continuum normalization is often a necessary step before estimating stellar parameters and chemical abundances.
-This process is ripe with subjectivity.
+Continuum normalization\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances, because these quantities are inferred from line strengths measured \emph{relative} to the continuum.
+The practicalities of continuum rectification is ripe with subjectivity, for justified historical reasons: often a consistent procedure is more useful than an abstract one, so people tend to stick with what they know.
 While there is general agreement in the literature that consistent continuum normalization is important, there is no apparent consensus on how it should be done.\\
 
-Fitting the continuum correctly requires you to know where there is line absorption.
+There is a variety of rectification strategies used to achieve different goals.
+Classical spectroscopists might make some coarse estimate of the continuum for the spectrum, and locally refine the continuum for each sbsorption line of interest.
+This process is often done by hand,\footnote{Albeit they are often experienced hands.
+See \citet{Bensby:2014}: \quote{`\emph{more than 300\,000 equivalent widths were measured by (the first author's right) hand}'}.} although some graphical user interface tools exist to streamline the process and improve reproducibility.
+This approach usually restricts the spectroscopist to only measuring strengths of isolated lines: those that are not blended by neighbouring atomic lines or strong molecular bands.
+In contrast, industrial spectroscopists have a different objective: one where they solely for a \emph{consistent} continuum normalization procedure.
+For them the process does not have to deliver the true continuum, but it should deliver a \emph{pseudo}-continuum that is consistent for stars of similar stellar parameters and signal-to-noise (S/N) ratios.\\
+
+Fitting the continuum correctly requires you to know where there is absorption.
 But knowing where there is line absorption requires you to (at least) know the stellar parameters.
 Without knowing the stellar parameters -- or having a good model for line absorption -- spectroscopists have developed various bespoke methods.
-A popular choice is to iteratively mask pixels in an asymmetric manner (so-called `sigma clipping') to exclude data points some chosen level below the current estimate of the continuum, or to mask all pixels except a carefully selected set of so-called continuum pixels.
-However, the set of continuum pixels is only valid for stars of a similar spectral type and metallicity, and in many cases there are \emph{no} continuum pixels (e.g., M-dwarfs).\\
+A popular choice is to iteratively mask pixels in an asymmetric manner (so-called `sigma clipping') to exclude data points some chosen level below the current estimate of the continuum.
+This works well if the practictioner can confidently set the bounds on where to exclude data.
+However, those parameters would need to be different for a more metal-rich star, or for a spectrum with a low signal-to-noise (S/N) ratio.
+Exactly how to specify the asymmetric clipping parameters comes down to experience.\\
 
-Classical spectroscopists might estimate some coarse continuum for the spectrum (or for each order, in the case of high-resolution echelle spectra), and then refine the continuum for every measured absorption line.
-This process is still often done by hand (albeit they are often experienced hands).
-Industrial spectroscopists might aim for a \emph{consistent} continuum normalization procedure: one that estimates a reliable continuum level for stars of similar stellar parameters and signal-to-noise ratios, even if that continuum estimate is some pseudo-continuum, and not the true continuum.\\
+An alternative is simply to mask all pixels except a carefully selected set of so-called `continuum pixels'.
+This is painstaking work to perform manually, but can be done by iteratively training data-driven models, or building masks from approximate line strengths computed from atomic properties.
+In either scenario, the set of continuum pixels is only valid for stars of a similar type and metallicity.
+In many cases, there are simply \emph{no} continuum pixels (e.g., near a molecular absorption band in an M-dwarf), and the best we can compute is a \emph{pseudo}-continuum.\\
 
-In an ideal scenario the continuum might be fit jointly with the stellar parameters, but this is often too expensive if it is costly to predict the emergent spectrum.
+A variety of methods for rectifying spectra are described in the literature. \todo{Some summary of those. Includes the rolling ball thing, RASSINE, and any others?}\\
+
+In an ideal scenario the continuum is jointly fit with the stellar parameters.
+Unfortunately, predicting the emergent spectrum is usually expensive.
 In Section~\ref{sec:methods} we describe a family of methods to address these problems.
@@
 
-We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
+We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., molecular or atomic line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
 These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast.
@@
 
-Here we will describe the method in general before outlining explicit assumptions and implementation details.
-The data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
+Here we will describe the method in general before outlining the implementation details.
+We assume the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
 The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
@@
 \end{eqnarray}
-\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into a helpful summation:
+\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into a convenient summation:
 \begin{align}
@@
 This is a dense matrix: there are no entries of exactly zero, with many entries near 1 (no absorption).
-However, a small transformation to this matrix can make it it extremely sparse.
+However, a small transformation to this matrix makes it extremely sparse.
 Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
@@
 where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}.
-All elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ are required to be non-negative.
+The factorization of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative.
 The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\%Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\
@@
 The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
-We found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages.
-We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex, do not require many choices, and they make for an efficient problem for graphical processing units.\\
+In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages.
+We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, \todo{until reaching a plateau in reconstruction error, at which point we would iterate further using multiplicative updates.} The multiplicative update rules are appealing because they are convex, do not require many choices, and they can be efficiently iterated on a graphics card.\\
 
@@
 The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to fit.
-This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of OBA-type stars was also sufficiently flexible to model many (but not all) white dwarf spectra.
+This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of A- to M-type stars was also sufficiently flexible to model many O-type stars.
 For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\
@@
 We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model.
-This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum.
-We clipped any BOSZ normalized flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements.
-As we discuss in later sections, emission lines are sufficiently rare in stellar spectra that they do not contribute significantly to the $\chi^2$ at inference time.
-We used \todo{all BOSZ spectra and factorized with CD into K basis vectors for 1000 steps, at which point the reconstruction loss was X}.\\
+This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \project{ATLAS9} code with \project{MARCS} models atmospheres.
+The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum.
+We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$, truncated the convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements.
+The matrix $\vec{S}$ includes all \todo{X} convolved BOSZ spectra, with \todo{N} pixels each, which we randomly shuffled before factorization.
+We then factorized this matrix into $K = 16$ basis vectors by coordinate descent for \todo{10,000} steps, \todo{and then X steps using multiplicative updates}.
+After training, the Frobenius loss, also known as the L2 reconstruction loss
+\begin{eqnarray}
+    L_2 = \norm{{\left(-\log{\vec{S}}\right) - \vec{W}_\star\transpose\vec{F}}}^2
+\end{eqnarray}
+\noindent{}was found to be \todo{X}.
+Figure~\ref{fig:schematic} shows some example basis vectors computed by this factorization, where it is somewhat striking that the basis vectors are identifiable by astrophysical mechanism, since no stellar parameters (or astrophysics of any kind) are included in the factorization process.
+For example, one basis vector just includes hydrogen lines.
+Another basis vector only has CO-molecule lines.
+We expand on this observation in later sections.\\
 
 \begin{figure*}
-    \caption{A schematic showing some basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra. \label{fig:schematic}}
+    \includegraphics*[width=\textwidth]{nmf_plot.pdf}
+    \caption{A schematic showing some basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra.
+\todo{Update this for BOSZ spectra, not APOGEE.}\label{fig:schematic}}
 \end{figure*}
-
-\todo{Discussion of the stellar and telluric basis vectors, how they separate out into astrophysically relevant things, and refer to the figure.}
 
@@
 
-Telluric transmission can be included in a similar way to the stellar line absorption.
-We factorize the negative log of theoretical telluric absorption into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$.
-\todo{Details about what telluric grid we used, and setting used for factorizing.
-Note how many vectors we chose ($L$).} With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
+Telluric transmission can be included in a similar way to stellar absorption.
+We used a grid of model telluric transmission spectra computed for the \texttt{PypeIt} project.
+These spectra are computed using the Line-by-Line Radiative transfer Model for an array of observatory locations (latitude, longitude) and atmospheric temperatures, pressures, humidity, and airmass.
+The original grid of telluric transmission spectra was not available to us.
+Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, and compressed using Principal Component Analysis and an inverse hyperbolic sine transformation, to approximate the original grid.
+With this reconstructed grid, we factorized the negative log of theoretical telluric absorption
+\begin{eqnarray}
+    -\log\vec{M} \approx \vec{W}\transpose_\earth\vec{G}
+\end{eqnarray}
+\noindent{}into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$ using $L = 4$ basis vectors.
+We took \todo{10,000} iterations using coordinate descent, at which point the reconstruction loss was \todo{X}.
+With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} to be solved at inference time.
-We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}).
+where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} in to be solved at inference time. 
+%We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}).
 Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\
@@
 \noindent{}where $P$ is a length scale which we set to be twice the length of the wavelength range to fit (e.g., twice the peak-to-peak range of an echelle order).
-The design matrix $\vec{H}$ can be constructed \emph{a priori} before inference begins.
+%The design matrix $\vec{H}$ can be constructed \emph{a priori} ibce before inference begins.
 Throughout this paper we will describe $\vecgamma$ as the sine-and-cosine \emph{coefficients}.\\
@@
 \end{equation}
+\noindent{}which we can compactly write as
 \begin{equation}
@@
 \end{eqnarray}
-\noindent{}but this does not guarantee that the bounds $\vecalpha \geq 0$ and $\vecbeta \geq 0$ are met.
-There are a few ways to find a solution to this system that is within the parameter bounds (e.g., linear operators; expectation-maximization), and thankfully all of them are convex paths.
-For simplicity we use the truncated reflective trust region algorithm, where we provide the Jacobian ($\vec{A}$).\\
+\noindent{}but this solution is not subject to the constraints $\vecalpha \geq 0$ and $\vecbeta \geq 0$.
+There are a few ways to solve this system subject to the boundary constraints, but the choice is not very important because the problem is convex.
+For simplicity we use the truncated reflective trust region algorithm where we stop optimizing once the relative tolerance has improved by less than $10^{-16}$.\\
 
@@
 
-After solving for $\vec{\tilde{X}}$, the (un-transformed) rectified flux and continuum can be computed in a straightorward manner:
+After solving for $\vec{\tilde{X}}$, the (un-transformed) quantities can be easily computed:
 \begin{eqnarray}
@@
 
+- emission lines
+
 \section{Conclusions}

### 2024-09-18 paper/main.tex
some words about results

@@
 The joint continuum-instrument response is modelled with a Fourier basis. 
-Together this set of bases represents a very effective linear model for stellar spectroscopy, where the linearity ensures that inference is convex, stable, and fast.
+Together this set of bases represents a very effective linear model for stellar spectroscopy. 
+The model requires no initial guess or tuning, and the linearity ensures that inference is convex, stable, and fast.
 This model allows us to reliably fit nuisance parameters (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
@@
 We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model.
-This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \project{ATLAS9} code with \project{MARCS} models atmospheres.
+This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \project{ATLAS9} code with \project{MARCS} spherical and plane-parallel models atmospheres.
+While the grid does include spectra for hotter stars computed with ATLAS atmospheres, we restricted ourselves to MARCS model atmospheres for this work.
 The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum.
@@
 There are a few ways to solve this system subject to the boundary constraints, but the choice is not very important because the problem is convex.
-For simplicity we use the truncated reflective trust region algorithm where we stop optimizing once the relative tolerance has improved by less than $10^{-16}$.\\
-
+For simplicity we use the truncated reflective trust region algorithm where we stop optimizing once the relative tolerance has improved by less than $10^{-16}$.
 %In later sections we describe some situations where we may want to penalize values of $\vecalpha$.
@@
 %\end{eqnarray}\\
-
 After solving for $\vec{\tilde{X}}$, the (un-transformed) quantities can be easily computed:
@@
 
-\section{Data} \label{sec:data}
+\section{Experiments} \label{sec:experiments}
 
-Here we seek demonstrate the effectiveness of our model on high-resolution echelle spectra: a scenario in which there are lots of data, and lots of nuisances.
-The nuisances (e.g., continuum parameters per order; tellurics) are often fit separately which can bias subsequent inferences, and the large data volume (e.g., $3\times10^6$~pixels per spectrum) can make even the fastest non-linear interpolators untenably slow.\\
+We assess the model using two experiments with real data.
+The first is to test the model on spectra across a wide range of spectral types, so that we can evaluate whether a constrained linear absorption model is an effective representation for typical stars.
+The second experiment involves fitting the model to spectra of the same star, taken across many different S/N ratios, to test how consistent the rectified spectra are as a function of S/N ratio.\\
 
-We searched the \eso\ data archive and selected a set of stars that span the range of spectral types and evolutionary states observed with the \harps\ instrument.
-The set include OBAFGKM-type stars, main-sequence stars, and red giant branch stars.
+The model we describe has most value in situations where there are lots of data, and many nuisance parameters.
+One example is high-resolution echelle spectra where there are different continuum parameters per order, and the large number of data points (e.g., $\sim10^6$~pixels/spectrum) can make even the fastest non-linear interpolators (or spectrum emulators) untenably slow.\\
+
+For this reason, we queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from O- to M-type.
 This selection is not intended to be complete, but is meant to demonstrate the effectiveness of a single linear model for fitting stars across a range of stellar parameters and evolutionary states.
-\\
+For each example star we downloaded a pseudo-random exposure, the details of which are shown in Table~\ref{tab:stars}.
+No explicit cut was made to ensure a minimum S/N ratio, but in practice most exposures are of high quality.
+The example OBA-type stars show significant rotational broadening.
+For these stars, we adopt the $v\sin{i}$ values reported by \citet{Someone} and convolve the stellar basis vectors before inference.
+In Section~\ref{sec:discussion} we discuss options when radial velocity or rotational broadening measurements are available.\\
 
-We downloaded a psuedo-random exposure for each star.
-No explicit cut to ensure a minimum signal-to-noise (SNR) ratio, but in practice most exposures have high SNR.
-We took the measured radial velocity in the calibration files for each spectrum and the computed Barycentric Earth radial velocity to shift the stellar absorption basis vectors to the observed frame.\\
+\begin{figure*}
+    \includegraphics[width=\textwidth]{harps-alfCenA-example.pdf}
+    \caption{An example fit to a single \eso/\harps\ spectrum of Alpha Centauri A (black), showing the continuum fits per echelle order (dark blue), the stellar absorption model (blue) and the telluric transmission model (light blue), all fit simultaneously.
+Near the strong Ca H and K lines (right panels) it is clear that there are \emph{no} `continuum pixels' in those echelle orders, yet the model provides an excellent estimate of the true continuum.
+\label{fig:alf-cen-A-example}}
+\end{figure*}
 
-We modelled the continuum in each echelle order as a set of $\todo{M}$ Fourier modes.
-With $\todo{\approx{70}}$ echelle orders, $\todo{K}$ stellar basis vectors, and $\todo{L}$ telluric vectors, there are $\todo{400}$ parameters and $\sim3\times10^6$ data points.
-Without \emph{any prior knowledge of the stellar parameters}, we directly solve the (unconstrained) weighted least-squares system (Equation~\ref{eq:wls}).
-If the solution is outside the parameter bounds, we optimize from this position using truncated reflective trust region algorithm.\\
+The final product of the \harps\ data reduction pipeline is a one-dimensional unrectified spectrum, resampled onto a single wavelength array where all echelle orders have been stitched together.
+Because we are interested in fitting the continuum on a per-order basis, we instead use the \texttt{e2ds} data products.
+These are extracted one-dimensional unrectified spectra of each echelle order, with the wavelength solution represented by a low order polynomial.
+In early experiments we fit the spect ra in these files directly, an example of which can be seen in Figure~\ref{fig:harps-alf-cen-a}.
+However, for some exposures we noticed systematic residuals that were consistent with a 2D artefact in the image plane (e.g., a ghost or reflection).
+While we expect that flat-fielding should have already removed these kinds of artefacts, it did prompt us to explore using the nightly blaze calibration files.
+Using the calibration blaze files mitigated much of the visible structure, and expectedly provided spectra with a simpler, flatter shape.
+For these reasons, in all experiments except for the one shown in Figure~\ref{fig:harps-alf-cen-a}, we chose to fit blaze-corrected echelle orders.\\
+
+When fitting each exposure we took the computed Barycentric Earth radial velocity and the radial velocity measured by the \harps\ data reduction pipeline to shift the stellar absorption basis vectors $\vec{F}$ to the observed frame.
+In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citep{Someone}.
+Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and \todo{9} Fourier modes per echelle order, totalling \todo{668} model parameters.\\
+
+We used the same model setup for the second experiment, where we required many spectra taken of the same star at low and high (e.g., 10 to 200) S/N ratios.
+The lower bound on S/N ratio might be unrealistic since spectroscopists prefer measuring stellar parameters and chemical abundances from higher quality spectra, but the bound is necessary to evaluate performance at low S/N ratios.
+We searched the \eso/\harps/ catalog and found that not many stars meet this requirement.
+When sorted by number of exposures in descending order, Alpha Centauri A was the first star that met our requirements in repeated S/N sampling.
+We constructed bins of SNR from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum (Table~\ref{tab:obs-alf-cen-a}) for every SNR bin interval to provide an even sampling of spectra at low and high S/N ratios.
+We fit our model to each spectrum, treating it as totally indepdendent from all others.\\
 
@@
 
-We found it usually takes a few core-seconds to fit each exposure ($3\times10^6$~pixels; all echelle orders fit simultaneously).
-Most of this time is spent constructing the design matrix, which requires (linearly) interpolating the stellar and telluric basis vectors to the observed wavelengths.
+We found it takes a few core-seconds to fit each exposure.
+The software implementation released with this paper provides for each echelle order: the predicted flux; the continuum; the rectified model flux; and the telluric transmission.
+With these we can plot the rectified flux per echelle order, or resample the rectified flux across all echelle orders.
+The rectified spectra for selected example stars in the first experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type.
+From these rectified spectra it is clear that the model performs well on a variety of stellar types and evolutionary stages.
+HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorized using NMF.
+The fit to HD 45418 is quite good, despite this star being well outside the model grid.\\
+
+All other example stars shown in Figure~\ref{fig:harps-examples} have literature stellar parameters that place them well within the bounds of the factorized model grid.
+Most stars observed with \eso/\harps\ are predictably on the main-sequence, since they are more frequently subject to radial velocity exoplanet searches than evolved stars.
+HD 203050 (K2) is the only red giant branch in our sample.
+This is not obvious from Figure~\ref{fig:harps-examples}, suggesting the model does just as well on red giant branch stars as it does on main-sequence stars.
+This has been our experience in other works using this method on SDSS-V optical and near-infrared spectra (Casey et al., in prep).\\
+
+The telluric models available to us are of lower spectral resolution ($\mathcal{R} \sim 25{,}000$) than \harps.
+This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:alf-cen-A-example}), but it is otherwise not obvious.
+While the model telluric transmission is broader than the data, the telluric basis vectors are sufficiently close enough to the data that the telluric absorption in the data is captured by the model telluric basis vectors, and they do not adversely affect the continuum fits.\\
+
+% - fitting stars of different spectral types
+
+% - fitting alpha Cen A
+
+% -
+
+We found it usually takes a few core-seconds to fit each exposure.
 The fits to each star are partially shown in Figure~\ref{fig:harps-example}.
@@
 Importantly, there are $\alpha$-Centauri spectra with signal-to-noise ratios of 10, and a nearly even sampling of spectra up to S/N of 300, or even higher (where saturation is visible in the reduced spectra\footnote{Naughty observer.}).
-We constructed bins of SNR from 0 to 200 in steps of 10.
-We selected an $\alpha$-Centauri spectrum for every SNR bin interval to provide an even sampling of spectra at low and high SNR.
-We fit our model to each spectrum, treating it as totally indepdendent from all others.\\
+\\
 
@@
 
+Although we use the radial velocities available from the data reduction pipeline and the literature, we note that our model is simple enough that the radial velocity (and rotational/macroscopic broadening) can be efficiently computed by \\
+
 \section{Conclusions}

### 2024-09-20 paper/main.tex
discussion

@@
 % TODO
-% [ ] Talk about BOSZ spectra and how the model was constructed.
-% [ ] Talk about how we constructed the telluric basis vectors
+% [X] Talk about BOSZ spectra and how the model was constructed.
+% [X] Talk about how we constructed the telluric basis vectors
 % [ ] Flesh out discussion points
-% [ ] Data section
-% -> Blaze???
-% [ ] Results section
-% -> ON a set of OBAFGKM stars, provides clean and consistent normalisation across these types
-% -> ARE WE USING THE NIGHTLY BLAZE?
-% -> discuss for alf Cen A, how we selected spectra to get a uniform sampling in SNR, and how we fit them
-% -> how we define the consistency measure
+% [X] Data section
+% [X] Results section
+% -> We fix vsini / v_rel from the literature, but these can be computed
 
-% [ ] discussion
-% -> We fix vsini / v_rel from the literature, but these can be computed 
-% ->
-
-% text macros
 %\newcommand{\chosentitle}{Constrained linear models for stellar spectroscopy}
@@
 
-% math macros
 \newcommand{\unit}[1]{\mathrm{#1}}
@@
 \newcommand*{\transpose}{^{\mkern-1.5mu\mathsf{T}}}
-%\newcommand{\transpose}{^\mathsf{T}}
 
-% notes
 \definecolor{tab:blue}{HTML}{1170aa}
@@
 % Other people who will be invited as co-authors (non-exhaustive list):
-%   Holtzman, Wheeler, Sayjdari, Bedell, Dream team, Astra folks, CCA data group
+%   Holtzman, Wheeler, Sayjdari, Bedell, Dream team?, Astra folks?, CCA data group?
 
@@
 This model allows us to reliably fit nuisance parameters (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
-We demonstrate our method by fitting \todo{toy models of M-dwarf spectra where we recover the correct continuum, and by fitting} \eso/\harps\ high-resolution echelle spectra of OBAFGKM-type stars.
+We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of BAFGKM-type stars.
 From repeat observations of $\alpha$-Centauri A we show that our approach yields continuum estimates that are consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
@@
 
-For this reason, we queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from O- to M-type.
+For this reason, we queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from B- to M-type.
 This selection is not intended to be complete, but is meant to demonstrate the effectiveness of a single linear model for fitting stars across a range of stellar parameters and evolutionary states.
@@
 No explicit cut was made to ensure a minimum S/N ratio, but in practice most exposures are of high quality.
-The example OBA-type stars show significant rotational broadening.
+The B- and A-type example stars show significant rotational broadening.
 For these stars, we adopt the $v\sin{i}$ values reported by \citet{Someone} and convolve the stellar basis vectors before inference.
@@
     \includegraphics[width=\textwidth]{harps-alfCenA-example.pdf}
-    \caption{An example fit to a single \eso/\harps\ spectrum of Alpha Centauri A (black), showing the continuum fits per echelle order (dark blue), the stellar absorption model (blue) and the telluric transmission model (light blue), all fit simultaneously.
+    \caption{An example fit to a single \eso/\harps\ spectrum of Alpha Centauri A (black), showing the continuum fits per echelle order (dark blue), the stellar absorption model (blue) and the \todo{telluric transmission model (light blue)}, all fit simultaneously.
 Near the strong Ca H and K lines (right panels) it is clear that there are \emph{no} `continuum pixels' in those echelle orders, yet the model provides an excellent estimate of the true continuum.
@@
 
+We independently fit all \eso/\harps\ spectra referenced in Tables~\ref{tab:harps-examples} and \ref{tab:alf-cen-a}.
 We found it takes a few core-seconds to fit each exposure.
 The software implementation released with this paper provides for each echelle order: the predicted flux; the continuum; the rectified model flux; and the telluric transmission.
-With these we can plot the rectified flux per echelle order, or resample the rectified flux across all echelle orders.
-The rectified spectra for selected example stars in the first experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type.
+With these we can plot the rectified flux per echelle order, or resample the rectified flux across all echelle orders.\\
+
+The first defined experiment involves fitting stars of a variety of spectral types.
+The rectified spectra for selected example stars in this experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type.
 From these rectified spectra it is clear that the model performs well on a variety of stellar types and evolutionary stages.
 HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorized using NMF.
-The fit to HD 45418 is quite good, despite this star being well outside the model grid.\\
-
+The fit to HD 45418 is quite good, despite this star being well outside the model grid.
 All other example stars shown in Figure~\ref{fig:harps-examples} have literature stellar parameters that place them well within the bounds of the factorized model grid.
@@
 
+The cooler stars in the example set have almost no so-called `continuum pixels' blueward of 4000\,\AA.
+This is expected; the continuum estimate provided by the model matches theoretical rectified spectra.
+It illustrates that the continuum estimate that we fit simultaneously with stellar absorption does provide a closer estimate to the true continuum (multiplied by the instrument response function) rather than a pseudo-continuum.\\
+
+There is good agreement between the model fits and data for example stars of different spectral types.
+In detail there are mis-matches in line absorption strength, or mis-matches in the form of missing lines (in either the model or the data).
+We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both.
+In any case, the model appears to be sufficiently accurate to model the continuum without substantial bias, so it is not a major concern for our purposes.
+A separate reason for model-data disagreement comes from our use of the telluric transmission model.
 The telluric models available to us are of lower spectral resolution ($\mathcal{R} \sim 25{,}000$) than \harps.
 This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:alf-cen-A-example}), but it is otherwise not obvious.
-While the model telluric transmission is broader than the data, the telluric basis vectors are sufficiently close enough to the data that the telluric absorption in the data is captured by the model telluric basis vectors, and they do not adversely affect the continuum fits.\\
+While the model telluric transmission is broader than the data, the telluric basis vectors are sufficiently close to the data that the telluric features in the data are well-captured by the model telluric basis vectors, and they do not adversely affect the continuum fits.\\
 
-% - fitting stars of different spectral types
+% TODO: anything else to say results-wise about experiment 1? \chi^2 values? systematic differences as a function of flux?
 
-% - fitting alpha Cen A
+The second defined experiment tests how consistently we can predict continuum for the same star, as a function of S/N ratios.
+We fit \todo{X} spectra of Alpha Centauri A, which are sampled uniformly in S/N bins from 10 to 200 pixel$^{-1}$.
+We are now forced to select a statistic to evaluate the consistency of our continuum rectification.
+We adopt the 95th percentile in rectified flux as a statistic to compare to normalisation methods described in the literature, but we prefer to view how the \emph{distribution} of normalized flux values change as a function of S/N ratio.
+Both properties are shown in Figure~\ref{fig:repeatability-with-snr}, where the line indicates the 95th percentile of normalized flux, and the shaded area shows the distribution of normalized flux values.
+Above S/N $\sim$ 30, both indicators are extremely stable and do not vary significantly with S/N ratio.
+This is notable, as most spectra acquired for the purpose of stellar parameters and chemical abundances will aim for S/N $>$ 30, but spectra below S/N $<$ 50 are routinely described as being problematic to rectify consistently.
+From this figure it highlights that we can achieve consistent continuum normalisation whether the star has S/N $\sim$ 30 or 300.\\
 
-% -
+We can make quantitative statements on how consistent the continuum rectification is.
+Taking the 95th percentile of normalized flux as a proxy statistic for the normalization consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalized flux is 0.22\% (a variation in normalized flux of 0.0022).
+This variation is smaller than the typical flux errors.
+This statistic increases only marginally when we include lower quality spectra.
+Taking all spectra with S/N > 50, we find the scatter in the 95th percentile of normalized flux to be 0.35\%.
+With S/N > 30, this rises marginally to 0.46\%.
+\todo{In all scenarios the error in continuum is less than the typical flux error.}\\
 
-We found it usually takes a few core-seconds to fit each exposure.
-The fits to each star are partially shown in Figure~\ref{fig:harps-example}.
-In all cases the data are shown in black, with the stellar absorption model in \todo{red}, telluric transmission in \todo{color}, and continuum in \todo{color}.
-Overall, the fits to the data are very good, with reduced $\chi^2$ values from \todo{A to B}.\\
-
-In bluer echelle orders, it can be seen that there are no pixels that are true `continuum' pixels (i.e., no absorption).
-Any continuum normalization method trying to fit these echelle orders would be biased if they did not have a good representation of what stellar spectra look like.
-\\
-
-Maintaining consistent continuum rectification for low signal-to-noise spectra is challenging.
+%Maintaining consistent continuum rectification for low signal-to-noise spectra is challenging.
 Without a model for stellar spectra or the relative strengths of absorption lines, noisy signals can mimic stellar absorption, causing the continuum to become systematically biased at low signal-to-noise ratios.
@@
 
-We searched the \eso/\harps/ archive for a star that has been repeatedly observed with a wide range of exposure times, usually because it is the subject of radial velocity asteroseismic monitoring.
-Through this search we identified alpha Centauri A, which has been observed more than \todo{X} times with \harps.
-Importantly, there are $\alpha$-Centauri spectra with signal-to-noise ratios of 10, and a nearly even sampling of spectra up to S/N of 300, or even higher (where saturation is visible in the reduced spectra\footnote{Naughty observer.}).
-\\
-
-\todo{Now we can see how things behave as a function of SNR.}
-
-\todo{Consistency at low signal-to-noise}
 % alf Cen 95th percentile (all wavelengths)
@@
 
-% RASSINE claims 1% (2% in their metric, which is 2 sigma) continuum accuracy of alpha Cen B with HARPS spectra.
-% When they exclude bluer than 4500 Angstrom, their continuum accuracy improves to 0.6% (1.2% in their metric)
-
-% If our perormance on alpha Cen A is comparable to theirs on alpha Cen B, then
-% our method is about 5x more accurate
-
-%The method can be readily applied to real data once the eigenspectra $\vecH$ have been computed.
-We experimented with choices of initialisation and inference.
-Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecalpha$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecalpha,\vectheta\}$ simultaneously.
-The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecalpha$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip).
-Initializing from small ($10^{-12}$) values of $\vecalpha$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecalpha$) seemed to work well in many scenarios.
-\\
-
-%In later sections we describe applications of our method to real data.
-For now we will describe a few options that we found to work reasonably well across all settings, which we have since established as default behaviour in the accompanying software implementation.
-When faced with the choice of how many spectra to include when performing non-negative matrix factorization, we found good results by including everything that could fit into memory.
-Using limited precision floats (e.g., float-8) helped.
-Initialising with non-negative double singular value decomposition (with zeros filled with small random values) seemed to work very well.
-Multiplicative updates.
-
-% put this in discussion
-%As we discuss in Section~\ref{sec:discussion}, this is not true of all linear models: it is a design choice that leads to this strict constraint.
-Other linear models (e.g., PCA) allow for summation of large positive and negative amplitudes, leading to  of positive and negative eigenspectra,
-
-This is a convex problem that is mathmatically guaranteed to yield the solution that minimizes the $\chi^2$ between the (log-)transformed data and model.
-However, it does not guarantee that the solution is sensible, or even physically possible.
-We expand on this in the \emph{Discussion} (Section~\ref{sec:discussion}), but the point here is to highlight that the solution might be unphysical: we might have a non-zero basis weight for a basis vector that only contains strong hydrogen lines (indicating a hot star), and we might have a strong non-zero basis weight for a basis vector that only contains molecular bands seen in very cool stars.
+\todo{RASSINE talk about line depth precision and i dont know if i want to get into that.
+if line depth precision is good but the continuum precision is bad then it doens't really matter what you line depth precision is because it will be dominated by the continuum accuracy in that we measure line depths relative to the continuum}
 
@@
 
+\todo{Something about the NMF weights we find, and how they vary somewhat expectantly?}
+
 \section{Discussion}\label{sec:discussion}
 
-- getting good fits with bad models
+% There is much to discuss here.
+We have shown that a constrained linear model is a good choice for modeling stellar spectra, which enables precise estimates of the continuum for a range of stellar types.
+Representing stellar absorption using NMF instead of a some other linear representation is particularly advantageous.
+Non-negativity ensures that stellar basis vectors are strictly additive, and the negative log-transformation we use restricts both the rectified stellar flux and telluric transmission to be between 0 and 1.
+Any data values outside this range can \emph{only} be captured by the continuum basis vectors.
+In some sense the predictive capacity of any combination of NMF Basis vectors is nearly orthogonal to a linear combination of the continuum basis vectors.\\
 
-- getting bad models with good fits
+A common linear representation for this kind of problem is Principal Component Analysis (PCA), which yields an orthogonal decomposition that maximally explains the variance in the data.
+PCA is made possible by orthogonality, whereas NMF is made possible by the non-negativity constraint.
+While PCA can capture more variance than NMF for the same data set and number of components, in PCA there are no bounds on the eigenvalues (basis weights).
+For this reason, if we used PCA to represent stellar spectra then it would make it nearly impossible to determine the continuum: we could model the data without having any continuum basis vectors.
+As an example, the first eigenvalue could be a large positive value, the second could be a large negative value, the third a slightly lower (but still large) positive value, and so on.
+This can easily model flux values much larger than one.
+The situation does not get much better if we include basis vectors for the continuum, because much of that smooth structure can be captured by the compressed spectral model.\\
 
-- emission lines
+Another somewhat unexpected advantage to using NMF is that the basis vectors themselves appear to be highly interpretable.
+For example, one basis vect or has only strong hydrogen lines in it.
+Another only shows molecular CO features that only appear in M-type stars.
+Two basis vectors look similar in that they have a forrest of atomic lines, where one shows remarkably broad features (e.g., consistent with a main-sequence star) and those features are extremely narrow in the second basis vector.
+Generally the basis vectors should be treated as latent (unobservable) vectors, and therefore there is risk associated with trying to physically interpret them.
+In separate work we discuss how the basis weights and basis vectors can be interpreted (Casey et al.
+in prep), but here we simply reiterate that the stellar parameters used to construct the grid are not used by the NMF process, and the spectra are randomly shuffled before factorization, so there is no possible `leakage' of information about effective temperature, surface gravity, metallicity, or other properties.
+Nevertheless, the resultant basis vectors have absorption features that are strongly correlated with these properties.
+\\
 
-- model interpretability
+Chosing not to interpret the basis vectors or weights has specific advantages.
+For example, if we were to build some mapping from stellar parameters to basis weights, then one could construct a forward model for the data where the model parameters are the continuum coefficients and the stellar parameters (e.g., $T_\mathrm{eff}$, $\log{g}$), instead of continuum coefficients and basis weights.
+That model would be far less flexible.
+For example, here our model is capable of providing excellent fits to the data, even if the underlying theoretical spectra are inaccurate.
+A simple example is hydrogen lines.
+Wheeler et al showed that the hydrogen lines in the infrared are incorrectly computed for every spectral synthesis code except for Korg.
+When we use existing grids (e.g., like those used in FERRE, as part of ASPCAP), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would end up with a biased estimate of effective temperature, or a worse fit to the data, or both.
+Instead, without constraining ourselves to a direct mapping from stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.\\
 
-- estimating precision in continuum rectification
+Our model is obviously constrained by how well it can describe the data, in two different ways.
+The first is in how well our models of stellar atmospheres and line lists and spectral synthesis can fit real data.
+The second comes from errors in how our factorization of the grid.
+We could add more basis vectors, train for longer, etc, but the reconstruction will never be perfect.
+A related minor deficiency in our model choice is that it cannot model emission lines in stellar spectra.
+The model is severely restricted in only being able to predict rectified flux values between zero and one.
+This is a minor problem because emission lines are rare, only occurring in the strongest of lines for chromospherically active stars, and because the emission lines only affect a small number of pixels, they contribute very little to the overall $\chi^2$.
+There is a concern that because our stellar absorption model can only represent values between zero and one, that any emission line would therefore be swept up into the continuum component, adversely affecting the continuum and the overall fit.
+In practice we have not found this behaviour because emission lines only contribute to a small fraction of pixels, and overall a small contribution to $\chi^2$.\\
 
-- biases as a function of S/N
+The speed of our model makes it particularly advantageous for evaluating on large data sets.
+If we assume that the factorization of stellar spectra introduces a small error relative to the usual mis-match between models and data, then a constrained linear model represents a promising avenue for identifying regions where models and observations disagree the most.
+The simplest path for doing this might be to look at the total squared residuals as a function of pixel across many stars, which would immediately indicate the areas where the model does not have capacity to predict the data.
+Next steps might be to improve modelling for that region (e.g., through better line lists) or might be to iteratively update the basis vectors in a data-driven way so that they better match the data.
+The model is trained from synthetic spectra, so it is already a close approximation to what stellar spectra look like, and doing data-driven updates would be an excellent way to handle severe data-model mis-matches (e.g., large molecular features which are present in data but we lack line lists to accuratley model).
+There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we could restrict ourselves to only updating a single basis vector.\\
 
-- poorly modelled regions as a function of pixel
+We have demonstrated the effectiveness of a constrained linear absorption model for continuum normalization across a range of spectral types.
+There are not many literature metrics to choose from when evaluating performance of continuum normalization.
 
-- excellent way to identify regions where models and observations totally disagree
+These statistics are computed using all wavelengths, but it is well-known that continuum is easier to model at redder optical wavelengths because there are fewer absorption features.
+There are not many literature examples to compare with when evaluating the consistency of continuum normalization.
+One excellent example is \citet{RASSINE}, where they report consistent continuum normalization at the level of 2\% based on applications to extremely high quality ($S/N > 1000$) \eso/\harps\ spectra of $\alpha$-Centauri B stacked over consecutive nights, and 0.29\% for a comparably high quality spectrum of the Sun.
+Their statistic is defined as the 2$\sigma$ of $1-f_i$, where $f_i$ is the $i$th normalized flux value at the anchor point locations used to build the continuum.
+The anchor points are chosen by the user, which makes for a less direct comparison with our statistic, but for these purposes we will assume that the continuum anchor points are perfectly chosen and truely represent the continuum in the star.
+Given these assumptions, for comparison purposes we take \texttt{RASSINE} as providing 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.29\% consitency for comparable Solar spectra.
+If they restrict their statistic to wavelengths redder than 4500,\AA, their precision for $\alpha$-Centauri B improves to 0.6\%.
+In comparison, for spectra we use that is 1/10th the S/N ratio, we find our method yields consistent continuum estimates at the level of 0.22\%.\\
 
-- orthogonality of the two model components
+\todo{Some comments about how the \emph{distribution} changes as a function of S/N. We could use K-L divergence as a measure, but that is more abstract and sensitive to noise fluctuations (I think).}
 
-- setting the right covariance strength for a given model.
+Arguments on the consistency of continuum normalization are precision-based arguments, which says nothing about accuracy.
+That is unfortunate for us because any pseudo-continuum method explicitly only needs to make arguments about precision: a pseudo-continuum is not intended to be accurate!
+In our case, our model is built to try and estimate the true continuum as best as possible by properly accounting for the stellar absorption.
+While this does likely lead to a more accurate estimate of the continuum, measuring that accuracy is very hard.
+Getting the continuum absolutely correct requires a stellar spectrum of a flux-calibrated instrument, and it requires that we have good models of stellar absorption.
+Some instruments deliver flux-calibrated spectra, but we rarely believe that we have good models of stellar absorption for M-type stars.
+Here, we think we do have good models of stellar absorption for M-type stars, but there are known deficiencies: the CaH and TiO bands are poorly modelled in most synthetic spectral grids, including the one we used.
+If we think we have a good fit to data from a flux-calibrated instrument such that we believe we have an accurate measure of the continuum, how would we independently test that accuracy?
+If the star was a standard candle, in that we knew how bright it should appear at a given distance, and we used the geometric parallax to get a distance, then we could state what the expected flux should be and compare that with our estimate of the continuum.
+But M-dwarf stars are not standard candles.
+Put simply, measuring accuracy of continuum is hard, and we don't try it, but we think our continuum estimate is better than a pseudo-continum.\\
 
-- emission lines
+We used the radial velocities reported by the \eso/\harps\ pipeline to shift stellar basis vectors to the observed frame.
+In cases of extreme rotational broadening, we also used literature measures of $v\sin{i}$.
+We took both of these quantities as truth, but in fact we could fit for them simultaneously.
+A radial velocity shift and a convolution are both linear operators, in that they can be incorporated in our model without breaking our strict linearity constraint.
+We would have to change how we solve the system: we could no longer solve the system directly by unconstrained least-squares, or with the truncated reflective region algorithm, but we would have to solve it using linear operators that output the dot product and left adjoint transpose dot product.
+Alternatively, given the speed of our model, one could imagine a grid in radial velocities (e.g., in $\pm1\,000\,\mathrm{km}\,s^{-1}$), where at each radial velocity we solve for $\vec{\tilde{X}}$ and then take the radial velocity with the lowest $\chi^2$.
+This is a robust method that has been used for decades.
+The rotational broadening can be estrimated from the cross-correlation in the same way.
+For these reasons, the method we describe here is readily applicable to situations where the radial velocity and the rotational broadening is not known.\\
 
-Although we use the radial velocities available from the data reduction pipeline and the literature, we note that our model is simple enough that the radial velocity (and rotational/macroscopic broadening) can be efficiently computed by \\
+%This is a convex problem that is mathmatically guaranteed to yield the solution that minimizes the $\chi^2$ between the (log-)transformed data and model.
+However, it does not guarantee that the solution is sensible, or even physically possible.
+We expand on this in the \emph{Discussion} (Section~\ref{sec:discussion}), but the point here is to highlight that the solution might be unphysical: we might have a non-zero basis weight for a basis vector that only contains strong hydrogen lines (indicating a hot star), and we might have a strong non-zero basis weight for a basis vector that only contains molecular bands seen in very cool stars.
 
-\section{Conclusions}
-\label{sec:conclusions}
+\section{Conclusions} \label{sec:conclusions}
+
+- data-driven stuff with model specification, more flexible than it needs to be, and then there are things taht model stuff as abosrption like blaze, Ella Wang bright-oblique method,
 

### 2024-09-23 paper/main.tex
add conclusion and some known issues

@@
 
-\section*{}\clearpage
+\section*{}\noindent\todo{
+Known issues to resolve before submission:\\
+- Show basis vectors from HARPS instead of APOGEE\\
+- Update HARPS fits to use blaze-corrected flux (only AB-type stars use it now)\\
+- Update SNR vs precision to show shaded distribution instead of a single percentile\\
+- Add citations\\
+- Figure 2 doesn't have the telluric model in it, re-fit and make pretty\\
+- Update red text.
+}
+
+\clearpage
 \section{Introduction}\label{sec:intro}
@@
 
-\todo{Something about the NMF weights we find, and how they vary somewhat expectantly?}
+%\todo{Something about the NMF weights we find, and how they vary somewhat expectantly?}
 
@@
 
-- data-driven stuff with model specification, more flexible than it needs to be, and then there are things taht model stuff as abosrption like blaze, Ella Wang bright-oblique method,
+We introduce a novel approach to modelling stellar spectra as a linear combination of highly constrained absorption vectors, which we show is sufficient for simultaneously fitting stellar absorption, telluric transmission, and the joint continuum-instrument response.
+A constrained linear absorption model has many advantagous properties over alternative data-driven methods or machine learning techniques used in stellar spectroscopy.
+The model structure restricts basis vectors to be strictly additive, ensuring separability between the basis vectors for absorption and continuum.
+The model requires no initial guess or prior of the stellar parameters, and the linearity ensures that inference is robust, stable, and fast.
+An extension of this work could also solve for other linear nuisances (e.g., radial velocity, rotational broadening).
+Finally, an unexpected result of this work is that the basis vectors found from factorizing model rectified spectra appear to be highly interpretable, which is promising for their use for interpretable constrained linear absorption models.\\
 
-\noindent{}We provide a Python implementation of our method in the following repository: \url{https://github.com/andycasey/continuum}.
+%Basis vectors learned by factorizing rectified grids of stellar spectra separate out into vectors that include only hydrogen lines, atomic lines, or prominent molecular features.
+A specific example of interpretability is shown where ony basis vector incldues only strong hydrogen lines, and at inference time we find the corresponding weight for this vector increases smoothly with spectral type (a proxy for effective temperature).
+
+%In subsequent works we discuss applications where this interpretability is exploited (Casey et al., in prep), but main.\\
+
+%- data-driven stuff with model specification, more flexible than it needs to be, and then there are things taht model stuff as abosrption like blaze, Ella Wang bright-oblique method,
+
+%\noindent{}We provide a Python implementation of our method in the following repository: \url{https://github.com/andycasey/continuum}.
 
 \paragraph{Software}
-\texttt{numpy} \citep{numpy} ---
-\texttt{matplotlib} \citep{matplotlib} ---
-\texttt{scipy} \citep{scipy}.
+\texttt{numpy} \citep{numpy}; 
+\texttt{matplotlib} \citep{matplotlib}; 
+\texttt{scipy} \citep{scipy};
+\texttt{scikit-learn} \citep{scikit_learn}.
 
 \paragraph{Acknowledgements}
-It is a pleasure to thank
-% All these people are likely co-authors, but should be thanked if they don't want to be co-authors
+If they do not become co-authors, then we should be sure to thank
     Jon Holztman (NMSU),
@@
     David W. Hogg (New York University),
-    Megan Bedell (Flatiron Institute),
-    Michael Blanton (NYU)
-.
+    Megan Bedell (Flatiron Institute), and 
+    Michael Blanton (NYU).
+
+\todo{Utah and OzStar resources. NYU for accommodating me as I starved without free flatiron lunch.}
 % include bibliography

### 2024-09-23 paper/main.tex
reviewed intro

@@
 % Other possible titles
-
-% TODO
-% [X] Talk about BOSZ spectra and how the model was constructed.
-% [X] Talk about how we constructed the telluric basis vectors
-% [ ] Flesh out discussion points
-% [X] Data section
-% [X] Results section
-% -> We fix vsini / v_rel from the literature, but these can be computed
-
 %\newcommand{\chosentitle}{Constrained linear models for stellar spectroscopy}
@@
 
-% Other people who will be invited as co-authors (non-exhaustive list):
-%   Holtzman, Wheeler, Sayjdari, Bedell, Dream team?, Astra folks?, CCA data group?
-
 \begin{abstract}\noindent
-Forward modelling stellar spectra usually requires a spectral synthesis code, or some non-linear interpolator constructed from a curated training set.
-These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent measurements.
-Here we present a \emph{linear} model that simultaneously fits stellar absorption, telluric transmission, and the joint continuum-instrument response.
+Forward modelling stellar spectra usually requires a spectral synthesis code, or a non-linear interpolator constructed from a curated training set.
+These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias scientific inferences.
+Here we describe a constrained \emph{linear} model that can fit radial velocity, rotational broadening, stellar absorption, telluric transmission, the joint continuum-instrument response.
 Stellar absorption and telluric transmission are each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
@@
 The joint continuum-instrument response is modelled with a Fourier basis.
-Together this set of bases represents a very effective linear model for stellar spectroscopy.
+Together this set of bases represents an effective linear model for stellar spectroscopy.
 The model requires no initial guess or tuning, and the linearity ensures that inference is convex, stable, and fast.
-This model allows us to reliably fit nuisance parameters (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
+This model allows us to reliably fit nuisances (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
 We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of BAFGKM-type stars.
@@
 
-\keywords{Some --- keywords --- here}
+\keywords{\todo{Some --- keywords --- here}}
 
@@
 
-Continuum normalization\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances, because these quantities are inferred from line strengths measured \emph{relative} to the continuum.
-The practicalities of continuum rectification is ripe with subjectivity, for justified historical reasons: often a consistent procedure is more useful than an abstract one, so people tend to stick with what they know.
-While there is general agreement in the literature that consistent continuum normalization is important, there is no apparent consensus on how it should be done.\\
+Continuum normalization\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances, because those quantities are inferred from line strengths that are measured \emph{relative} to the continuum.
+The practice of continuum rectification is ripe with subjectivity, in part for justified historical reasons.
+Often a consistent procedure is more useful than an abstract one, so people tend to stick with what they know, and that subjectivity is inherited by graduate students.
+While there is general agreement in the literature that is important to achieve consistent continuum normalization, there is no apparent consensus on how it should be done.\\
 
-There is a variety of rectification strategies used to achieve different goals.
-Classical spectroscopists might make some coarse estimate of the continuum for the spectrum, and locally refine the continuum for each sbsorption line of interest.
+The literature describes a variety of rectification strategies used to achieve different goals.
+Classical spectroscopists might make some coarse estimate of the continuum for the spectrum, and locally refine that estimate for every absorption line of interest.
 This process is often done by hand,\footnote{Albeit they are often experienced hands.
 See \citet{Bensby:2014}: \quote{`\emph{more than 300\,000 equivalent widths were measured by (the first author's right) hand}'}.} although some graphical user interface tools exist to streamline the process and improve reproducibility.
-This approach usually restricts the spectroscopist to only measuring strengths of isolated lines: those that are not blended by neighbouring atomic lines or strong molecular bands.
-In contrast, industrial spectroscopists have a different objective: one where they solely for a \emph{consistent} continuum normalization procedure.
-For them the process does not have to deliver the true continuum, but it should deliver a \emph{pseudo}-continuum that is consistent for stars of similar stellar parameters and signal-to-noise (S/N) ratios.\\
+This approach usually restricts the spectroscopist to only measuring strengths of what are believed to be isolated lines: atomic transitions that are not contaminated by neighbouring lines or molecular bands.
+In contrast, industrial spectroscopists often have a different objective: one where they solely for a \emph{consistent} continuum normalization procedure.
+In those situations the process does not have claim to deliver the true continuum, but it should deliver a \emph{pseudo}-continuum that is consistent for stars of similar stellar parameters and signal-to-noise (S/N) ratios.\\
 
@@
 This works well if the practictioner can confidently set the bounds on where to exclude data.
-However, those parameters would need to be different for a more metal-rich star, or for a spectrum with a low signal-to-noise (S/N) ratio.
+However, those parameters would need to be different for a more metal-rich star, or for a spectrum with a low S/N ratio.
 Exactly how to specify the asymmetric clipping parameters comes down to experience.\\
@@
 In either scenario, the set of continuum pixels is only valid for stars of a similar type and metallicity.
-In many cases, there are simply \emph{no} continuum pixels (e.g., near a molecular absorption band in an M-dwarf), and the best we can compute is a \emph{pseudo}-continuum.\\
+In many cases there are simply \emph{no} continuum pixels (e.g., near a molecular absorption band in an M-dwarf).
+Some graphical tools exist to enable practictioners to select pseudo-continuum points to `anchor' some continuum fit (e.g., a spline, or through a rolling-pin analogy) which are useful if the practictioner wants full control over the continuum fit, or if the models are particularly poor at predicting continuum pixels.\\
 
-A variety of methods for rectifying spectra are described in the literature. \todo{Some summary of those. Includes the rolling ball thing, RASSINE, and any others?}\\
+We might simply consider what is the best continuum for every possible theoretical spectrum under comparison.
+When there are thousands of model spectra to compare with, then for each spectrum we can take the ratio of the data and the model, apply some averaging filter (e.g., moving mean or median) and take that as the pseudo-continuum.
+These kinds of approaches work remarkably well, despite the requirement of a set of (sufficiently) comparable model spectra, and the added computational expense.\\
 
@@
 In Section~\ref{sec:methods} we describe a family of methods to address these problems.
-The methods we describe use non-negative matrix factorization (NMF) to approximate line absorption.
+The methods we describe uses non-negative matrix factorization (NMF) to approximate line absorption and telluric transmission.
 NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements.
 This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
+We describe data (Section~\ref{sec:experiments}) and experiments using this model, where the results are presented in Section~\ref{sec:results}.
 We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\
@@
 
+% Other people who will be invited as co-authors (non-exhaustive list):
+%   Holtzman, Wheeler, Sayjdari, Bedell, Dream team?, Astra folks?, CCA data group?
+
 \paragraph{Acknowledgements}
@@
     Rory Smith (Monash),
-    Adam Wheeler (Ohio State University),
+    Adam Wheeler (CCA),
     Andrew Saydjari (Harvard),
-    David W. Hogg (New York University),
-    Megan Bedell (Flatiron Institute), and 
+    Lily Zhao (Chicago),
+    Julianne Dalcanton (CCA),
+    David W. Hogg (NYU),
+    Megan Bedell (CCA), and
     Michael Blanton (NYU).

### 2024-09-23 paper/main.tex
word-smithing

@@
 When there are thousands of model spectra to compare with, then for each spectrum we can take the ratio of the data and the model, apply some averaging filter (e.g., moving mean or median) and take that as the pseudo-continuum.
-These kinds of approaches work remarkably well, despite the requirement of a set of (sufficiently) comparable model spectra, and the added computational expense.\\
+These kinds of approaches work remarkably well, despite the requirement of a (potentially very large) set of sufficiently similar model spectra, and the added computational expense.\\
 
@@
 The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
-\begin{align}
+\begin{align}\label{eq:y}
     y_i &= f(\lambda_i;\vecalpha)\hadamard{}g(\lambda_i;\vecbeta)\hadamard{}h(\lambda_i;\vecgamma) + \mbox{noise}
@@
 \end{eqnarray}
-\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into a convenient summation:
+\noindent{}which is computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
 \begin{align}
@@
 The factorization of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative.
-The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\%Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\
-
-%on previous works on NMF and data analysis more broadly.
-We would like the naming conventions to be consistent with those papers, but this nomenclature is easily overloaded.
-For this reason, we make clear definitions: throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights} and $\vecH$ as the NMF \emph{basis vectors}.
-\\
-
-%Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ basis weights per spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding basis vectors each with $D$ pixels.
-
-%The nomenclature  draws on previous works on NMF and data analysis, but this nomenclature is easy to overload, so throughout this paper we will refer to $\vecW$ as the NMF \emph{basis weights}, $\vecalpha$ as the NMF \emph{basis coefficients}, $\vecH$ as the NMF \emph{basis vectors}, and $\vectheta$ (not yet defined) will refer to the amplitudes of the sine-and-cosine basis.
+The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\
 
@@
 In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages.
+We found very similar looking basis vectors if we initialized from random values, but it took more training iterations.
 We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, \todo{until reaching a plateau in reconstruction error, at which point we would iterate further using multiplicative updates.} The multiplicative update rules are appealing because they are convex, do not require many choices, and they can be efficiently iterated on a graphics card.\\
@@
 There are very few requirements of the theoretical spectral grid.
-There are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing in between points.
+There are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing between points.
 The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to fit.
-This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of A- to M-type stars was also sufficiently flexible to model many O-type stars.
+This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of A- to M-type stars was also sufficiently flexible to model many O-type stars, but there was structured residuals in the wings of hydrogen lines.
 For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\
@@
 We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model.
-This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \project{ATLAS9} code with \project{MARCS} spherical and plane-parallel models atmospheres.
+This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \texttt{ATLAS9} code with \project{MARCS} spherical and plane-parallel models atmospheres.
 While the grid does include spectra for hotter stars computed with ATLAS atmospheres, we restricted ourselves to MARCS model atmospheres for this work.
@@
 We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$, truncated the convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements.
-The matrix $\vec{S}$ includes all \todo{X} convolved BOSZ spectra, with \todo{N} pixels each, which we randomly shuffled before factorization.
-We then factorized this matrix into $K = 16$ basis vectors by coordinate descent for \todo{10,000} steps, \todo{and then X steps using multiplicative updates}.
+The matrix $\vec{S}$ includes all \todo{X} convolved BOSZ spectra, with \todo{N} pixels each, which we randomly shuffled at every iteration during factorization.
+We factorized this matrix into $K = 16$ basis vectors by coordinate descent for \todo{10,000} steps, \todo{and then X steps using multiplicative updates}.
 After training, the Frobenius loss, also known as the L2 reconstruction loss
@@
 Another basis vector only has CO-molecule lines.
-We expand on this observation in later sections.\\
+We expand on this in later sections.\\
 
@@
 The original grid of telluric transmission spectra was not available to us.
-Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, and compressed using Principal Component Analysis and an inverse hyperbolic sine transformation, to approximate the original grid.
-With this reconstructed grid, we factorized the negative log of theoretical telluric absorption
+Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, where the grid is compressed using Principal Component Analysis and an inverse hyperbolic sine transformation.
+We reconstructed the original grid from these approximations, and factorized the negative log of theoretical telluric absorption
 \begin{eqnarray}
@@
 
-\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
+\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model in equation \ref{eq:log_y} for the log-transformed data $\vec{Y}$,
 \begin{equation}
@@
 
+The model we describe has most value in situations where there are lots of data, and many nuisance parameters.
+One example is high-resolution echelle spectra where there are different continuum parameters per order, and the large number of data points (e.g., $\sim10^6$~pixels/spectrum) can make even the fastest non-linear interpolators (or spectrum emulators) untenably slow.\\
+
 We assess the model using two experiments with real data.
@@
 
-The model we describe has most value in situations where there are lots of data, and many nuisance parameters.
-One example is high-resolution echelle spectra where there are different continuum parameters per order, and the large number of data points (e.g., $\sim10^6$~pixels/spectrum) can make even the fastest non-linear interpolators (or spectrum emulators) untenably slow.\\
-
-For this reason, we queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from B- to M-type.
+We queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from B- to M-type.
 This selection is not intended to be complete, but is meant to demonstrate the effectiveness of a single linear model for fitting stars across a range of stellar parameters and evolutionary states.
@@
 For these stars, we adopt the $v\sin{i}$ values reported by \citet{Someone} and convolve the stellar basis vectors before inference.
-In Section~\ref{sec:discussion} we discuss options when radial velocity or rotational broadening measurements are available.\\
+In Section~\ref{sec:discussion} we discuss situations when radial velocity or rotational broadening measurements are not available.\\
 
@@
 These are extracted one-dimensional unrectified spectra of each echelle order, with the wavelength solution represented by a low order polynomial.
-In early experiments we fit the spect ra in these files directly, an example of which can be seen in Figure~\ref{fig:harps-alf-cen-a}.
-However, for some exposures we noticed systematic residuals that were consistent with a 2D artefact in the image plane (e.g., a ghost or reflection).
+In early experiments we fit the spectra in the \texttt{e2ds} data products directly, an example of which can be seen in Figure~\ref{fig:harps-alf-cen-a}.
+However, for some exposures we noticed systematic residuals that could be explained by a 2D artefact in the image plane (e.g., a ghost or reflection).
 While we expect that flat-fielding should have already removed these kinds of artefacts, it did prompt us to explore using the nightly blaze calibration files.
@@
 In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citep{Someone}.
-Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and \todo{9} Fourier modes per echelle order, totalling \todo{668} model parameters.\\
+Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and \todo{9} Fourier modes per echelle order, totalling \todo{668} model parameters per spectrum.\\
 
@@
 The lower bound on S/N ratio might be unrealistic since spectroscopists prefer measuring stellar parameters and chemical abundances from higher quality spectra, but the bound is necessary to evaluate performance at low S/N ratios.
-We searched the \eso/\harps/ catalog and found that not many stars meet this requirement.
-When sorted by number of exposures in descending order, Alpha Centauri A was the first star that met our requirements in repeated S/N sampling.
-We constructed bins of SNR from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum (Table~\ref{tab:obs-alf-cen-a}) for every SNR bin interval to provide an even sampling of spectra at low and high S/N ratios.
+The \eso/\harps\ catalog revealed that not many stars meet our repeated S/N sampling requirement.
+When sorted by number of exposures in descending order, $\alpha$-Centauri A was the first star that met our requirements in repeated S/N sampling.
+We constructed bins of S/N from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum (Table~\ref{tab:obs-alf-cen-a}) for every S/N ratio bin interval to provide an even sampling of spectra at low and high S/N ratios.
 We fit our model to each spectrum, treating it as totally indepdendent from all others.\\
@@
     \caption{Continuum-rectified \harps\ spectra of example spectral types from B to M.
-The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission.\label{fig:harps-examples}}
+The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission.
+\todo{Only BAF-type stars used blaze-corrected spectra.
+Need to download blazes for others and re-run.}\label{fig:harps-examples}}
 \end{figure*}
@@
 The first defined experiment involves fitting stars of a variety of spectral types.
-The rectified spectra for selected example stars in this experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type.
+The rectified spectra for selected example stars in this experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type listed in \project{SIMBAD}.
 From these rectified spectra it is clear that the model performs well on a variety of stellar types and evolutionary stages.
-HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorized using NMF.
+HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16{,}750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorized using NMF.
 The fit to HD 45418 is quite good, despite this star being well outside the model grid.
@@
 This is expected; the continuum estimate provided by the model matches theoretical rectified spectra.
-It illustrates that the continuum estimate that we fit simultaneously with stellar absorption does provide a closer estimate to the true continuum (multiplied by the instrument response function) rather than a pseudo-continuum.\\
+It illustrates that the continuum estimate that we fit simultaneously with stellar absorption does provide a closer estimate to the true continuum rather than a pseudo-continuum.\\
 
@@
 We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both.
-In any case, the model appears to be sufficiently accurate to model the continuum without substantial bias, so it is not a major concern for our purposes.
+In any case, the model appears to be sufficiently accurate to describe the continuum without substantial bias, so it is not a major concern for our purposes.\\
+
 A separate reason for model-data disagreement comes from our use of the telluric transmission model.
@@
 This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:alf-cen-A-example}), but it is otherwise not obvious.
-While the model telluric transmission is broader than the data, the telluric basis vectors are sufficiently close to the data that the telluric features in the data are well-captured by the model telluric basis vectors, and they do not adversely affect the continuum fits.\\
+The model telluric transmission predicts features that are clearly broader than the data, but the telluric basis vectors are a sufficiently good description for the data that the telluric absorption does not get captured by the continuum basis vectors and adversely affect the fit.\\
 
-% TODO: anything else to say results-wise about experiment 1? \chi^2 values? systematic differences as a function of flux?
-
-The second defined experiment tests how consistently we can predict continuum for the same star, as a function of S/N ratios.
-We fit \todo{X} spectra of Alpha Centauri A, which are sampled uniformly in S/N bins from 10 to 200 pixel$^{-1}$.
-We are now forced to select a statistic to evaluate the consistency of our continuum rectification.
+The second defined experiment tests how consistently we can predict continuum for the same star, as a function of S/N ratio.
+Given an imperfect (or no) model for stellar spectra, noisy signals can mimic stellar absorption, which causes the continuum to become systematically biased at low S/N ratios.
+We fit \todo{X} spectra of $\alpha$-Centauri A, which are sampled uniformly in S/N bins from 10 to 200 pixel$^{-1}$.
+This experiment forces us to select a summary statistic to evaluate the consistency of our continuum rectification.
 We adopt the 95th percentile in rectified flux as a statistic to compare to normalisation methods described in the literature, but we prefer to view how the \emph{distribution} of normalized flux values change as a function of S/N ratio.
@@
 This is notable, as most spectra acquired for the purpose of stellar parameters and chemical abundances will aim for S/N $>$ 30, but spectra below S/N $<$ 50 are routinely described as being problematic to rectify consistently.
-From this figure it highlights that we can achieve consistent continuum normalisation whether the star has S/N $\sim$ 30 or 300.\\
+From this figure it highlights that we can achieve consistent continuum normalisation whether the star has S/N $\sim$ 30 or 300.
+\todo{This experiment was done on non-blaze-corrected spectra.
+I either need to re-do it, or update the text accordingly.}\\
 
-We can make quantitative statements on how consistent the continuum rectification is.
-Taking the 95th percentile of normalized flux as a proxy statistic for the normalization consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalized flux is 0.22\% (a variation in normalized flux of 0.0022).
-This variation is smaller than the typical flux errors.
-This statistic increases only marginally when we include lower quality spectra.
-Taking all spectra with S/N > 50, we find the scatter in the 95th percentile of normalized flux to be 0.35\%.
-With S/N > 30, this rises marginally to 0.46\%.
-\todo{In all scenarios the error in continuum is less than the typical flux error.}\\
-
-%Maintaining consistent continuum rectification for low signal-to-noise spectra is challenging.
-Without a model for stellar spectra or the relative strengths of absorption lines, noisy signals can mimic stellar absorption, causing the continuum to become systematically biased at low signal-to-noise ratios.
-For this reason, we sought to estimate the consistency of continuum normalization as a function of signal-to-noise ratio.\\
+We can make quantitative statements on the consistency of continuum rectification for $\alpha$-Centauri A.
+Taking the 95th percentile of normalized flux as a proxy statistic for the normalization consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalized flux is 0.22\%, which is equivalent a spread in normalized flux of 0.0022.
+This variation is smaller than the typical flux errors, and only increases marginally when we include lower quality spectra.
+Taking all spectra with S/N $>$ 50, we find the scatter in the 95th percentile of normalized flux to be 0.35\%.
+With S/N $>$ 30, this rises marginally to 0.46\%.
+In all scenarios the estimated error in continuum is less than the typical flux error.\\
 
@@
 
-\todo{RASSINE talk about line depth precision and i dont know if i want to get into that.
-if line depth precision is good but the continuum precision is bad then it doens't really matter what you line depth precision is because it will be dominated by the continuum accuracy in that we measure line depths relative to the continuum}
-
 \begin{figure}
-    \caption{Median observed rectified flux values as a function of S/N ratio showing that there is no bias as a function of S/N ratio.}
+    \caption{Distribution (shaded) of rectified flux of $\alpha$-Centauri A as a function of S/N ratio, with every spectrum fit independently of all others.
+The line represents the 95th percentile, where the scatter in this percentile is 0.46\% for spectra with S/N $>$ 30, and 0.22\% for everything with S/N $>$ 100.
+\label{fig:alpha-cen-wrt-snr}}
 \end{figure}
-
-%\todo{Something about the NMF weights we find, and how they vary somewhat expectantly?}
 
@@
 
-% There is much to discuss here.
 We have shown that a constrained linear model is a good choice for modeling stellar spectra, which enables precise estimates of the continuum for a range of stellar types.
@@
 Any data values outside this range can \emph{only} be captured by the continuum basis vectors.
-In some sense the predictive capacity of any combination of NMF Basis vectors is nearly orthogonal to a linear combination of the continuum basis vectors.\\
+In some sense the predictive capacity of the combination of NMF basis vectors is nearly orthogonal to a combination of the continuum basis vectors.\\
 
-A common linear representation for this kind of problem is Principal Component Analysis (PCA), which yields an orthogonal decomposition that maximally explains the variance in the data.
+NMF is less frequently used in astronomical data analysis than other linear representations.
+A more common linear representation for this kind of problem is Principal Component Analysis (PCA), which yields an orthogonal decomposition that maximally explains the variance in the data.
 PCA is made possible by orthogonality, whereas NMF is made possible by the non-negativity constraint.
 While PCA can capture more variance than NMF for the same data set and number of components, in PCA there are no bounds on the eigenvalues (basis weights).
-For this reason, if we used PCA to represent stellar spectra then it would make it nearly impossible to determine the continuum: we could model the data without having any continuum basis vectors.
-As an example, the first eigenvalue could be a large positive value, the second could be a large negative value, the third a slightly lower (but still large) positive value, and so on.
-This can easily model flux values much larger than one.
-The situation does not get much better if we include basis vectors for the continuum, because much of that smooth structure can be captured by the compressed spectral model.\\
+For this reason, if we used PCA to represent stellar spectra then it would make it nearly impossible to determine the continuum.
+We could model the data without having any continuum basis vectors, simply by finding the first eigenvalue to be a very large positive value, the second eigenvalue to be a large negative value, the third a slightly lower (but still large) positive value, and so on.
+This can easily model flux values much larger than 1.
+The situation does not get much better if we include basis vectors for the continuum, because much of that smooth structure can be captured by the PCA spectral model.\\
 
 Another somewhat unexpected advantage to using NMF is that the basis vectors themselves appear to be highly interpretable.
-For example, one basis vect or has only strong hydrogen lines in it.
+For example, one basis vector has only strong hydrogen lines in it.
 Another only shows molecular CO features that only appear in M-type stars.
-Two basis vectors look similar in that they have a forrest of atomic lines, where one shows remarkably broad features (e.g., consistent with a main-sequence star) and those features are extremely narrow in the second basis vector.
-Generally the basis vectors should be treated as latent (unobservable) vectors, and therefore there is risk associated with trying to physically interpret them.
-In separate work we discuss how the basis weights and basis vectors can be interpreted (Casey et al.
-in prep), but here we simply reiterate that the stellar parameters used to construct the grid are not used by the NMF process, and the spectra are randomly shuffled before factorization, so there is no possible `leakage' of information about effective temperature, surface gravity, metallicity, or other properties.
+Another two basis vectors are similar in that they have a forrest of atomic lines, but they vary in the broadness of particularly strong lines (e.g., where one is consistent with pressure broadening in a main-sequence star).
+General wisdom when working with compressed representations is that the basis vectors should generally be treaated as latent (unobservable) vectors, and there is risk associated with trying to physically interpret them.
+We avoid physical interpretability here, but in subsequent work we discuss how the basis weights and vectors can be interpreted (Casey et al., in prep).
+Here we simply reiterate that the stellar parameters used to construct the grid are not used by the NMF process, and the spectra are randomly shuffled during each iteration of factorization, so there is no possible `leakage' of information about effective temperature, surface gravity, metallicity, or other properties.
 Nevertheless, the resultant basis vectors have absorption features that are strongly correlated with these properties.
@@
 For example, if we were to build some mapping from stellar parameters to basis weights, then one could construct a forward model for the data where the model parameters are the continuum coefficients and the stellar parameters (e.g., $T_\mathrm{eff}$, $\log{g}$), instead of continuum coefficients and basis weights.
-That model would be far less flexible.
-For example, here our model is capable of providing excellent fits to the data, even if the underlying theoretical spectra are inaccurate.
+That model would be far less flexible than the model we present here.
+For example, our model is capable of providing excellent fits to the data, even if the underlying theoretical spectra are inaccurate.
 A simple example is hydrogen lines.
-Wheeler et al showed that the hydrogen lines in the infrared are incorrectly computed for every spectral synthesis code except for Korg.
-When we use existing grids (e.g., like those used in FERRE, as part of ASPCAP), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would end up with a biased estimate of effective temperature, or a worse fit to the data, or both.
-Instead, without constraining ourselves to a direct mapping from stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.\\
+\citep{Wheeler} showed that the hydrogen lines in the infrared are incorrectly computed for every spectral synthesis code except for \texttt{Korg}.
+When we use existing grids (e.g., like those used in \texttt{FERRE}, as part of \texttt{ASPCAP}), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would find a biased estimate of effective temperature, or a worse fit to the data, or both.
+Instead, by not restricting ourselves to a mapping between stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.\\
 
-Our model is obviously constrained by how well it can describe the data, in two different ways.
-The first is in how well our models of stellar atmospheres and line lists and spectral synthesis can fit real data.
-The second comes from errors in how our factorization of the grid.
-We could add more basis vectors, train for longer, etc, but the reconstruction will never be perfect.
-A related minor deficiency in our model choice is that it cannot model emission lines in stellar spectra.
-The model is severely restricted in only being able to predict rectified flux values between zero and one.
-This is a minor problem because emission lines are rare, only occurring in the strongest of lines for chromospherically active stars, and because the emission lines only affect a small number of pixels, they contribute very little to the overall $\chi^2$.
-There is a concern that because our stellar absorption model can only represent values between zero and one, that any emission line would therefore be swept up into the continuum component, adversely affecting the continuum and the overall fit.
-In practice we have not found this behaviour because emission lines only contribute to a small fraction of pixels, and overall a small contribution to $\chi^2$.\\
+Our model is limited by how well it predicts data in two different ways.
+The first is in how well the models of stellar atmospheres, line lists, and spectral synthesis codes can match real data.
+There is very little we can do about that error mis-match here.
+The second form of error arises in our factorization of the grid.
+We could add more basis vectors, train for longer, even include estimated per-pixel model uncertainties, but the reconstruction will never be perfect.
+A related minor deficiency in our model choice is that it cannot model emission lines in stellar spectra: the model can only predict rectified flux values between zero and one.
+This is a minor problem because emission lines are rare, only occurring in the strongest of lines for chromospherically active stars, and because the emission lines only affect a small number of pixels, they contribute very little to the overall $\chi^2$.\\
 
@@
 The model is trained from synthetic spectra, so it is already a close approximation to what stellar spectra look like, and doing data-driven updates would be an excellent way to handle severe data-model mis-matches (e.g., large molecular features which are present in data but we lack line lists to accuratley model).
-There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we could restrict ourselves to only updating a single basis vector.\\
+There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we can restrict ourselves to only updating a single basis vector.\\
 
-We have demonstrated the effectiveness of a constrained linear absorption model for continuum normalization across a range of spectral types.
+%We have demonstrated the effectiveness of a constrained linear absorption model for continuum normalization across a range of spectral types.
+
 There are not many literature metrics to choose from when evaluating performance of continuum normalization.
-
+We have qualitatively shown the distribution of rectified flux values for $\alpha$-Centauri A, and quantitatively estimated the precision from the scatter in the 95th percentile of rectified flux values.
 These statistics are computed using all wavelengths, but it is well-known that continuum is easier to model at redder optical wavelengths because there are fewer absorption features.
-There are not many literature examples to compare with when evaluating the consistency of continuum normalization.
-One excellent example is \citet{RASSINE}, where they report consistent continuum normalization at the level of 2\% based on applications to extremely high quality ($S/N > 1000$) \eso/\harps\ spectra of $\alpha$-Centauri B stacked over consecutive nights, and 0.29\% for a comparably high quality spectrum of the Sun.
+One excellent examination of continuum normalization consistency is that of \citet{RASSINE}, where they report consistent continuum normalization at the level of 2\% based on applications to extremely high quality ($S/N > 1000$) \eso/\harps\ spectra of $\alpha$-Centauri B stacked over consecutive nights, and 0.29\% for a comparably high quality spectrum of the Sun.
 Their statistic is defined as the 2$\sigma$ of $1-f_i$, where $f_i$ is the $i$th normalized flux value at the anchor point locations used to build the continuum.
 The anchor points are chosen by the user, which makes for a less direct comparison with our statistic, but for these purposes we will assume that the continuum anchor points are perfectly chosen and truely represent the continuum in the star.
-Given these assumptions, for comparison purposes we take \texttt{RASSINE} as providing 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.29\% consitency for comparable Solar spectra.
+Given these assumptions, for comparison purposes we take \texttt{RASSINE} as providing 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.15\% consitency for comparable Solar spectra.
 If they restrict their statistic to wavelengths redder than 4500,\AA, their precision for $\alpha$-Centauri B improves to 0.6\%.
@@
 
-\todo{Some comments about how the \emph{distribution} changes as a function of S/N. We could use K-L divergence as a measure, but that is more abstract and sensitive to noise fluctuations (I think).}
+%\citet{Someone} also describes an estimate of the line depth precision achieved using \texttt{RASSINE}. Their estimates of line depth precision outperform (by percentage) their continuum estimates. 
+%\todo{RASSINE talk about line depth precision and i dont know if i want to get into that.
+if line depth precision is good but the continuum precision is bad then it doens't really matter what you line depth precision is because it will be dominated by the continuum accuracy in that we measure line depths relative to the continuum}
 
@@
 Here, we think we do have good models of stellar absorption for M-type stars, but there are known deficiencies: the CaH and TiO bands are poorly modelled in most synthetic spectral grids, including the one we used.
-If we think we have a good fit to data from a flux-calibrated instrument such that we believe we have an accurate measure of the continuum, how would we independently test that accuracy?
+%If we think we have a good fit to data from a flux-calibrated instrument such that we believe we have an accurate measure of the continuum, how would we independently test that accuracy?
 If the star was a standard candle, in that we knew how bright it should appear at a given distance, and we used the geometric parallax to get a distance, then we could state what the expected flux should be and compare that with our estimate of the continuum.
@@
 This is a robust method that has been used for decades.
-The rotational broadening can be estrimated from the cross-correlation in the same way.
-For these reasons, the method we describe here is readily applicable to situations where the radial velocity and the rotational broadening is not known.\\
-
-%This is a convex problem that is mathmatically guaranteed to yield the solution that minimizes the $\chi^2$ between the (log-)transformed data and model.
-However, it does not guarantee that the solution is sensible, or even physically possible.
-We expand on this in the \emph{Discussion} (Section~\ref{sec:discussion}), but the point here is to highlight that the solution might be unphysical: we might have a non-zero basis weight for a basis vector that only contains strong hydrogen lines (indicating a hot star), and we might have a strong non-zero basis weight for a basis vector that only contains molecular bands seen in very cool stars.
+The rotational broadening can be estimated from the cross-correlation in the same way.
+For these reasons, the method we describe here is readily extensible to situations where the radial velocity and the rotational broadening is not known.\\
 
@@
 
-%In subsequent works we discuss applications where this interpretability is exploited (Casey et al., in prep), but main.\\
-
 %- data-driven stuff with model specification, more flexible than it needs to be, and then there are things taht model stuff as abosrption like blaze, Ella Wang bright-oblique method,
@@
 \texttt{scikit-learn} \citep{scikit_learn}.
-
-% Other people who will be invited as co-authors (non-exhaustive list):
-%   Holtzman, Wheeler, Sayjdari, Bedell, Dream team?, Astra folks?, CCA data group?
 

### 2024-09-23 paper/main.tex
foo

@@
 
-We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., molecular or atomic line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.
-These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast.
+We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., molecular or atomic line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{In this work the `continuum' always refers to the joint continuum-instrument response.
+In many spectrographs these are different, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast.
 In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and handling other nuisances.\\
@@
 \begin{figure}
-    \caption{Distribution (shaded) of rectified flux of $\alpha$-Centauri A as a function of S/N ratio, with every spectrum fit independently of all others.
+    \caption{Distribution (shaded) of rectified flux of $\alpha$-Centauri A as a function of S/N ratio, with every spectrum fit independently.
 The line represents the 95th percentile, where the scatter in this percentile is 0.46\% for spectra with S/N $>$ 30, and 0.22\% for everything with S/N $>$ 100.
@@
     Andrew Saydjari (Harvard),
+    Guy Stringfellow,
     Lily Zhao (Chicago),
@@
 
-\todo{Utah and OzStar resources. NYU for accommodating me as I starved without free flatiron lunch.}
+\todo{Acknowledge Utah and OzStar resources. Acknowledge NYU for accommodating my body as it withered without free Flatiron lunch.}
 % include bibliography

### 2024-10-08 paper/main.tex
some updates

@@
 \usepackage{MnSymbol}
+\usepackage{rotating}
 
@@
 - Update HARPS fits to use blaze-corrected flux (only AB-type stars use it now)\\
-- Update SNR vs precision to show shaded distribution instead of a single percentile\\
+- In the "alpha Cen A" sample, sometimes they accidentally observed alpha Cen B and didn't realise! Maybe better to use a different star?\\
 - Add citations\\
-- Figure 2 doesn't have the telluric model in it, re-fit and make pretty\\
 - Update red text.
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{harps-alfCenA-example.pdf}
-    \caption{An example fit to a single \eso/\harps\ spectrum of Alpha Centauri A (black), showing the continuum fits per echelle order (dark blue), the stellar absorption model (blue) and the \todo{telluric transmission model (light blue)}, all fit simultaneously.
-Near the strong Ca H and K lines (right panels) it is clear that there are \emph{no} `continuum pixels' in those echelle orders, yet the model provides an excellent estimate of the true continuum.
-\label{fig:alf-cen-A-example}}
-\end{figure*}
+\begin{sidewaysfigure}
+    \includegraphics[width=\textwidth]{example.pdf}
+    \caption{An example fit to a single \eso/\harps\ spectrum of HD~222595 (black), showing the continuum fits per echelle order (dotted), the stellar absorption model (red) and the telluric transmission model (blue), all fit simultaneously with a constrained linear absorption model.
+Near the strong Ca H and K lines it is clear that there are \emph{no} `continuum pixels' in those echelle orders, yet the model provides an excellent estimate of the true continuum.
+\label{fig:example-echelle}}
+\end{sidewaysfigure}
 
@@
 The telluric models available to us are of lower spectral resolution ($\mathcal{R} \sim 25{,}000$) than \harps.
-This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:alf-cen-A-example}), but it is otherwise not obvious.
+This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:example-echelle}), but it is otherwise not obvious.
 The model telluric transmission predicts features that are clearly broader than the data, but the telluric basis vectors are a sufficiently good description for the data that the telluric absorption does not get captured by the continuum basis vectors and adversely affect the fit.\\
@@
 \begin{figure}
+    \includegraphics*[width=\textwidth, height=3in]{alpha-cen-repeat-snr.pdf}
     \caption{Distribution (shaded) of rectified flux of $\alpha$-Centauri A as a function of S/N ratio, with every spectrum fit independently.
@@
 Given these assumptions, for comparison purposes we take \texttt{RASSINE} as providing 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.15\% consitency for comparable Solar spectra.
-If they restrict their statistic to wavelengths redder than 4500,\AA, their precision for $\alpha$-Centauri B improves to 0.6\%.
+If they restrict their statistic to wavelengths redder than 4500\,\AA, their precision for $\alpha$-Centauri B improves to 0.6\%.
 In comparison, for spectra we use that is 1/10th the S/N ratio, we find our method yields consistent continuum estimates at the level of 0.22\%.\\
+
+%\todo{I really don't want to, but should I Fisher information matrix and predict how precisely we could estimate the continuum as a function of S/N for this model?
+It's linear so it's simple, but it adds necessitates a lot of explanatory text that is not easily digestible by my target audience (spectroscopists).}\\
 

### 2024-10-08 code/plot_alpha_cen_repeat_snr.py

@@ -1,0 +1,167 @@
+import os
+import numpy as np
+import pickle
+from tqdm import tqdm
+from astropy.table import Table
+
+from spectrum import Spectrum, SpectrumCollection
+
+summary = Table.read("alfCenA_stability.csv")
+
+#t = Table(rows=results, names=("path", "snr", "status", "cost", "optimality", "std_norm_flux", "p_90", "p_95", "p_97.5"))
+
+
+snr_bins = 1 + np.arange(200) # 1 to 200
+rectified_flux_min, rectified_flux_max, rectified_flux_step = (0.8, 1.20, 0.001)
+rectified_flux_bins = np.linspace(rectified_flux_min, rectified_flux_max, int((rectified_flux_max - rectified_flux_min) / rectified_flux_step) + 1)
+
+percentiles = [95]
+
+H = np.nan * np.ones((snr_bins.size, rectified_flux_bins.size - 1))
+percentile_values = np.nan * np.ones((snr_bins.size, len(percentiles)))
+
+fig_all, ax_all = plt.subplots(figsize=(10.5, 4))
+
+fig, ax = plt.subplots()
+
+#
+keep_snrs = (104, 105, 106, 107, 108)
+keep_snrs = (119, 120, 121, 122)
+keep = {}
+
+import matplotlib as mpl
+all_colors = mpl.colormaps["plasma"](np.linspace(0, 1, len(snr_bins)))
+
+comparison = []
+
+colors = ("tab:red", "tab:blue", "tab:green", "tab:orange")
+summary.sort("snr")
+for row in tqdm(summary):
+    if int(row["snr"]) > 200 or row["path"].endswith("_e2ds_B.fits"): continue
+
+    fits_path = "../applications/harps/alfCenA/" + row["path"]
+    result_path = fits_path + ".pkl"
+    if not os.path.exists(result_path):
+        continue
+
+    spectra = SpectrumCollection.read(fits_path)
+    with open(result_path, "rb") as fp:
+        (result, continuum, rectified_model_flux, rectified_telluric_flux, y_pred) = pickle.load(fp)
+
+    # When we initially ran this I excluded the last echelle order
+    rectified_flux = (spectra.flux[:-1] / continuum).flatten()
+
+    mask = ((spectra.λ[:-1] > 4500)).flatten() * (rectified_flux > 0.10)
[…]

### 2024-10-08 code/run_alpha_cen_repeat_snr.py

@@ -1,0 +1,172 @@
+from glob import glob
+from astropy.table import Table
+from astropy.io import fits
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+import os
+
+from spectrum import Spectrum, SpectrumCollection
+from model import Clam, apply_radial_velocity_shift
+from continuum_basis import Polynomial, Sinusoids
+
+import matplotlib
+#from mpl_utils import mpl_style
+#matplotlib.style.use(mpl_style)
+
+
+with open("20240916_train_harps_lsf_model.pkl", "rb") as fp:
+    λ, label_names, parameters, W, H = pickle.load(fp)
+
+
+from specutils.utils.wcs_utils import air_to_vac
+from astropy import units as u
+
+with open("../applications/harps-sandbox/20240816_train_telfit_model.pkl", "rb") as fp:
+    tel_λ, tel_label_names, tel_parameters, tel_W, tel_H = pickle.load(fp)
+    tel_λ *= 10
+    
+
+with open("20240916_telluric_model.pkl", "rb") as fp:
+    tel_λ, tel_W, tel_H, tel_meta = pickle.load(fp)
+
+keep = np.ones(10, dtype=bool)
+for index in (0, 2, 3, 5, 7, 9):
+    keep[index] = False
+
+model = Clam(λ, H, tel_λ, tel_H[keep])
+
+
+meta = Table.read("../applications/harps/alfCenA/meta.csv")
+
+
+# subsample to  only get 1 per snr bin
+n_counts = np.zeros(300)
+keep = np.zeros(len(meta), dtype=bool)
+
+for i, row in enumerate(meta):
+    if row["path"].endswith("_B.fits"):
+        continue
+
+    index = int(row["snr"]) - 1
+    if index > (len(n_counts) - 1):
+        keep[i] = False
+        continue
+
+    if n_counts[index] > 0:
+        keep[i] = False
+    else:
+        keep[i] = True
+        n_counts[index] += 1
+
+
+meta_subset = meta[keep]
+
+suffix = "_20240930"
+from tqdm import tqdm
+results = []
+for row in tqdm(meta_subset):
+
+    try:
+        path = "../applications/harps/alfCenA/" + row["path"]
+
+        if os.path.exists(f"{path}{suffix}.pkl"):
[…]

### 2024-10-08 paper/main.tex
update basis vectors plot

@@
 %\newcommand{\chosentitle}{Stellar continuum modelling}
-\newcommand{\chosentitle}{Constrained linear models for stellar spectroscopy}
+\newcommand{\chosentitle}{Constrained linear absorption models for stellar spectroscopy}
 
@@
 Known issues to resolve before submission:\\
-- Show basis vectors from HARPS instead of APOGEE\\
 - Update HARPS fits to use blaze-corrected flux (only AB-type stars use it now)\\
@@
 \begin{figure*}
-    \includegraphics*[width=\textwidth]{nmf_plot.pdf}
-    \caption{A schematic showing some basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra.
-\todo{Update this for BOSZ spectra, not APOGEE.}\label{fig:schematic}}
+    \includegraphics*[width=\textwidth]{harps-basis-vectors.pdf}
+    \caption{A schematic showing a small wavelength range of the basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra.\label{fig:schematic}}
 \end{figure*}

### 2024-10-22 paper/main.tex
comments up to methods

@@
 
+\newcommand{\ajw}[1]{\textbf{#1}}
+
 \sloppy\sloppypar\raggedbottom\frenchspacing
@@
 \begin{abstract}\noindent
-Forward modelling stellar spectra usually requires a spectral synthesis code, or a non-linear interpolator constructed from a curated training set.
+Forward modelling stellar spectra usually requires a spectral synthesis code, or a non-linear interpolator \ajw{emulator?} constructed from a curated training set.
 These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias scientific inferences.
@@
 Stellar absorption and telluric transmission are each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
-The non-negativity constraint ensures that basis vectors are strictly additive.
+The non-negativity constraint ensures that basis vectors are strictly additive. \ajw{consider dropping or rewording this sentence}
 The joint continuum-instrument response is modelled with a Fourier basis.
-Together this set of bases represents an effective linear model for stellar spectroscopy.
+Together this set of bases represents an effective linear model for stellar spectroscopy. \ajw{Could potentially be made more precise? see note at top of intro.}
 The model requires no initial guess or tuning, and the linearity ensures that inference is convex, stable, and fast.
@@
 
+\ajw{
+The abstract mostly talks about the model, and the introduction mostly talks about continuum normalization.  Would be good to more explcitly link, maybe.
+I also think that it would be good to more precisely state what about this model is good.
+Something like: it characterizes the set of all possible spectra in many few parameters than other d-d models, while still expcluding enough unphysical ``spectra'' (thanks to NN constraint) that you can, e.g.
+project onto it to separate spectrum from continuum (or use it for param estimation).
+}
+
 Continuum normalization\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances, because those quantities are inferred from line strengths that are measured \emph{relative} to the continuum.
+\ajw{Would be good to say that the more fundemental reason is that modelling the continuum (mostly intrumental but also stellar) is hard, and the separation is feasible for most stars.}
 The practice of continuum rectification is ripe with subjectivity, in part for justified historical reasons.
-Often a consistent procedure is more useful than an abstract one, so people tend to stick with what they know, and that subjectivity is inherited by graduate students.
+Often a consistent procedure is more useful than an abstract \ajw{maybe reword} one, so people tend to stick with what they know, and that subjectivity is inherited by graduate students.
 While there is general agreement in the literature that is important to achieve consistent continuum normalization, there is no apparent consensus on how it should be done.\\
@@
 Fitting the continuum correctly requires you to know where there is absorption.
-But knowing where there is line absorption requires you to (at least) know the stellar parameters.
+But knowing where there is line absorption requires you to (at least) know the stellar parameters \ajw{confused about this claim}.
 Without knowing the stellar parameters -- or having a good model for line absorption -- spectroscopists have developed various bespoke methods.
@@
 
-We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., molecular or atomic line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{In this work the `continuum' always refers to the joint continuum-instrument response.
+We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., molecular or atomic line absorption) \ajw{I think stellar lines   would be a better name for this, the relationship between actual absorption and lines is kinda subtle }; a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{In this work the `continuum' always refers to the joint continuum-instrument response.
 In many spectrographs these are different, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast.
@@
 Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
+\ajw{you should probably say that C is the (diagonal) cov matrix}
 \begin{eqnarray}
@@
 \end{eqnarray}
-\noindent{}which is computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
+\noindent{}which is computed by taking a swiftly \ajw{?} Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
 \begin{align}
@@
 We now turn to defining each model component in detail.
-The stellar absorption model $f(\lambda_i;\vecalpha)$ predicts the rectified stellar line absorption at wavelength $\lambda_i$ given parameters $\vecalpha$.
+The stellar absorption model $f(\lambda_i;\vecalpha)$ predicts \ajw{represents?
+corresponds to?} the rectified stellar line absorption at wavelength $\lambda_i$ given parameters $\vecalpha$.
 We construct the line absorption model $f(\lambda_i;\vecalpha)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorization (NMF).

### 2024-10-22 paper/main.tex
comments til p12

@@
 \section{Methods}\label{sec:methods}
+\ajw{Weinberg does a thing where he has a table of every variable and what it is at the beginning of the methods.  It makes things much easier to skim, and I think it would work well here.}
 
@@
 While we expect that flat-fielding should have already removed these kinds of artefacts, it did prompt us to explore using the nightly blaze calibration files.
-Using the calibration blaze files mitigated much of the visible structure, and expectedly provided spectra with a simpler, flatter shape.
+Using the calibration blaze files mitigated much of the visible structure, and expectedly \ajw{``as expected'', maybe} provided spectra with a simpler, flatter shape.
 For these reasons, in all experiments except for the one shown in Figure~\ref{fig:harps-alf-cen-a}, we chose to fit blaze-corrected echelle orders.\\
@@
 \todo{Only BAF-type stars used blaze-corrected spectra.
-Need to download blazes for others and re-run.}\label{fig:harps-examples}}
+Need to download blazes for others and re-run.} \ajw{consider a lower line weight} \ajw{I feel like i'm being thick but I don't get whey these ever go above 1.
+Also, are you sure that the clam continuum is where the theoretical continuum is the the M1 star?}}\label{fig:harps-examples}}
 \end{figure*}
@@
 All other example stars shown in Figure~\ref{fig:harps-examples} have literature stellar parameters that place them well within the bounds of the factorized model grid.
-Most stars observed with \eso/\harps\ are predictably on the main-sequence, since they are more frequently subject to radial velocity exoplanet searches than evolved stars.
+Most stars observed with \eso/\harps\ are predictably \ajw{drop ``predictably'' maybe} on the main-sequence, since they are more frequently subject to radial velocity exoplanet searches than evolved stars.
 HD 203050 (K2) is the only red giant branch in our sample.

### 2024-12-20 paper/main.tex

@@
 \todo{Only BAF-type stars used blaze-corrected spectra.
-Need to download blazes for others and re-run.} \ajw{consider a lower line weight} \ajw{I feel like i'm being thick but I don't get whey these ever go above 1.
-Also, are you sure that the clam continuum is where the theoretical continuum is the the M1 star?}}\label{fig:harps-examples}}
+Need to download blazes for others and re-run.} \ajw{consider a lower line weight} \ajw{I feel like i'm being thick but I don't get why these ever go above 1.
+Also, are you sure that the clam continuum is where the theoretical continuum is for the M1 star?}}\label{fig:harps-examples}}
 \end{figure*}
@@
 We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both.
-In any case, the model appears to be sufficiently accurate to describe the continuum without substantial bias, so it is not a major concern for our purposes.\\
+In any case, the model appears to be sufficiently accurate to describe the continuum without substantial bias, so it is not a major concern for our purposes.
+\\
 
@@
 Any data values outside this range can \emph{only} be captured by the continuum basis vectors.
-In some sense the predictive capacity of the combination of NMF basis vectors is nearly orthogonal to a combination of the continuum basis vectors.\\
+In some sense the predictive capacity of the combination of NMF basis vectors is nearly orthogonal to a combination of the continuum basis vectors.\ajw{Is this a restatement/interpretation of the fact that the stellar line basis and continuum basis are aproximately orthogonal?}\\
 
@@
 A simple example is hydrogen lines.
-\citep{Wheeler} showed that the hydrogen lines in the infrared are incorrectly computed for every spectral synthesis code except for \texttt{Korg}.
-When we use existing grids (e.g., like those used in \texttt{FERRE}, as part of \texttt{ASPCAP}), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would find a biased estimate of effective temperature, or a worse fit to the data, or both.
+\citep{Wheeler} showed that the hydrogen lines in the infrared are incorrectly computed by many spectral synthesis codes.
+When we use existing grids with this problem (e.g., like those used in \texttt{FERRE}, as part of \texttt{ASPCAP}), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would find a biased estimate of effective temperature, or a worse fit to the data, or both.
 Instead, by not restricting ourselves to a mapping between stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.\\
@@
 The model is trained from synthetic spectra, so it is already a close approximation to what stellar spectra look like, and doing data-driven updates would be an excellent way to handle severe data-model mis-matches (e.g., large molecular features which are present in data but we lack line lists to accuratley model).
-There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we can restrict ourselves to only updating a single basis vector.\\
+There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we can restrict ourselves to only updating a single basis vector.
+\ajw{a reader might be puzzled by the claim that you can evaluate the goodness of the model spectra given the discussion above about how it's fine if the model spectra are a bit shit.}\\
 

### 2024-12-20 paper/main.tex
some updates

@@
 %\newcommand{\chosentitle}{Stellar continuum modelling}
-\newcommand{\chosentitle}{Constrained linear absorption models for stellar spectroscopy}
+%\newcommand{\chosentitle}{Constrained linear absorption models for stellar spectroscopy}
+\newcommand{\chosentitle}{A constrained linear model for continuum normalization of stellar spectra}
 
@@
 \affiliation{Centre of Excellence for Astrophysics in Three Dimensions (ASTRO-3D)}
-\affiliation{Center for Computational Astrophysics, Flatiron Institute, a division of the Simons Foundation}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
 
-\author{friends}
-%\author[0000-0003-2866-9403]{David W. Hogg}
-%\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
-%\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
-%\affiliation{Center for Computational Astrophysics, Flatiron Institute, a division of the Simons Foundation}
+\author[0000-0001-7339-5136]{Adam Wheeler}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
+
+\author[0000-0001-9907-7742]{Megan Bedell}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
+
+\author[0000-0003-2866-9403]{David W. Hogg}
+\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
+\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
+
+\author[0000-0002-6561-9002]{Andrew Sayjdari}
+\affiliation{Department of Astrophysical Sciences, Princeton University, Princeton, NJ 08544, USA}
+\affiliation{Department of Physics, Harvard University, 17 Oxford St., Cambridge, MA 02138, USA}
+\affiliation{Harvard-Smithsonian Center for Astrophysics, 60 Garden St., Cambridge, MA 02138, USA}
+
+\author[0000-0002-3852-3590]{Lily Zhao}
+\affiliation{Department of Astronomy & Astrophysics, University of Chicago, 5640 S Ellis Avenue, Chicago, IL 60637, USA}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
 
 \begin{abstract}\noindent
-Forward modelling stellar spectra usually requires a spectral synthesis code, or a non-linear interpolator \ajw{emulator?} constructed from a curated training set.
-These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias scientific inferences.
+Inferring stellar parameters and chemical abundances by forward modelling stellar spectra usually requires a spectral synthesis code, or an emulator constructed from a curated training set. 
+In these situations continuum normalization is often implemented as a pre-processing step that is independent of stellar parameters.
+This leads to results that are biased, or inconsistent at low and high signal-to-noise ratios.
+A more justified approach is to forward model spectra with all nuisances simultaneously, but in practice this can be an expensive or non-convex optimization procedure.
 Here we describe a constrained \emph{linear} model that can fit radial velocity, rotational broadening, stellar absorption, telluric transmission, the joint continuum-instrument response.
 Stellar absorption and telluric transmission are each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
-The non-negativity constraint ensures that basis vectors are strictly additive. \ajw{consider dropping or rewording this sentence}
+This model characterizes all possible spectra in many fewer parameters than comparable data-driven models.
+The non-negativity constraint ensures basis vectors are strictly additive, which prohibits most unphysical spectra such that we can distinguish normalized spectra from the joint instrument-continuum response.
 The joint continuum-instrument response is modelled with a Fourier basis.
-Together this set of bases represents an effective linear model for stellar spectroscopy. \ajw{Could potentially be made more precise? see note at top of intro.}
+%Together this set of bases represents a flexible linear model that can predict spectra of many types of stars, and complex continuum/instrument shapes with hundreds of parameters.
 The model requires no initial guess or tuning, and the linearity ensures that inference is convex, stable, and fast.
@@
 
-\keywords{\todo{Some --- keywords --- here}}
+%\keywords{\todo{Some --- keywords --- here}}
 
-\section*{}\noindent\todo{
-Known issues to resolve before submission:\\
-- Update HARPS fits to use blaze-corrected flux (only AB-type stars use it now)\\
-- In the "alpha Cen A" sample, sometimes they accidentally observed alpha Cen B and didn't realise! Maybe better to use a different star?\\
-- Add citations\\
-- Update red text.
-}
+%\section*{}\noindent\todo{
+%Known issues to resolve before submission:\\
+%- Update HARPS fits to use blaze-corrected flux (only AB-type stars use it now)\\
+%- In the "alpha Cen A" sample, sometimes they accidentally observed alpha Cen B and didn't realise! Maybe better to use a different star?\\
+%- Add citations\\
+%- Update red text.
+%}
 
-\clearpage
 \section{Introduction}\label{sec:intro}
-
-\ajw{
-The abstract mostly talks about the model, and the introduction mostly talks about continuum normalization.  Would be good to more explcitly link, maybe.
-I also think that it would be good to more precisely state what about this model is good.
-Something like: it characterizes the set of all possible spectra in many few parameters than other d-d models, while still expcluding enough unphysical ``spectra'' (thanks to NN constraint) that you can, e.g.
-project onto it to separate spectrum from continuum (or use it for param estimation).
-}
 
@@
 Need to download blazes for others and re-run.} \ajw{consider a lower line weight} \ajw{I feel like i'm being thick but I don't get why these ever go above 1.
-Also, are you sure that the clam continuum is where the theoretical continuum is for the M1 star?}}\label{fig:harps-examples}}
+Also, are you sure that the clam continuum is where the theoretical continuum is for the M1 star?}}\label{fig:harps-examples}
 \end{figure*}

### 2025-02-24 paper/main.tex
as submitted

@@
 %\newcommand{\chosentitle}{Constrained linear absorption models for stellar spectroscopy}
-\newcommand{\chosentitle}{A constrained linear model for continuum normalization of stellar spectra}
+\newcommand{\chosentitle}{A constrained linear model for continuum normalisation of stellar spectra}
 
-\shorttitle{\chosentitle}
+\shorttitle{Constrained linear models for normalising stellar spectra}
 \shortauthors{Casey}
@@
 \author[0000-0002-3852-3590]{Lily Zhao}
-\affiliation{Department of Astronomy & Astrophysics, University of Chicago, 5640 S Ellis Avenue, Chicago, IL 60637, USA}
+\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, 5640 S Ellis Avenue, Chicago, IL 60637, USA}
 \affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
@@
 Inferring stellar parameters and chemical abundances by forward modelling stellar spectra usually requires a spectral synthesis code, or an emulator constructed from a curated training set. 
-In these situations continuum normalization is often implemented as a pre-processing step that is independent of stellar parameters.
-This leads to results that are biased, or inconsistent at low and high signal-to-noise ratios.
-A more justified approach is to forward model spectra with all nuisances simultaneously, but in practice this can be an expensive or non-convex optimization procedure.
-Here we describe a constrained \emph{linear} model that can fit radial velocity, rotational broadening, stellar absorption, telluric transmission, the joint continuum-instrument response.
-Stellar absorption and telluric transmission are each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
-This model characterizes all possible spectra in many fewer parameters than comparable data-driven models.
-The non-negativity constraint ensures basis vectors are strictly additive, which prohibits most unphysical spectra such that we can distinguish normalized spectra from the joint instrument-continuum response.
-The joint continuum-instrument response is modelled with a Fourier basis.
-%Together this set of bases represents a flexible linear model that can predict spectra of many types of stars, and complex continuum/instrument shapes with hundreds of parameters.
+In these situations continuum normalisation is often implemented as a pre-processing step that is independent of stellar parameters.
+This leads to results that are biased, or inconsistent across signal-to-noise ratios.
+A more justified approach is to forward model spectra with all nuisances simultaneously, but in practice this can be an expensive or non-convex optimisation procedure.
+Here we describe a constrained \emph{linear} model that can fit stellar absorption, telluric transmission, the joint continuum-instrument response.
+Stellar absorption and telluric transmission are each modelled by factorising a grid of rectified theoretical spectra into two non-negative matrices.
+This model characterises all possible spectra in many fewer parameters than comparable data-driven models.
+The non-negativity constraint ensures basis vectors are strictly additive, which prohibits most unphysical spectra such that we can distinguish normalised spectra from the joint instrument-continuum response.
 The model requires no initial guess or tuning, and the linearity ensures that inference is convex, stable, and fast.
-This model allows us to reliably fit nuisances (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
+This model allows us to reliably fit nuisances (e.g., tellurics, continuum, radial velocity, broadening), without any prior knowledge about the fundamental stellar properties.
 We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of BAFGKM-type stars.
-From repeat observations of $\alpha$-Centauri A we show that our approach yields continuum estimates that are consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30. 
+With repeat observations of $\alpha$-Centauri A we present results that are best in class: consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.\vspace{3em}\\
 \end{abstract}
@@
 
-%\keywords{\todo{Some --- keywords --- here}}
-
-%\section*{}\noindent\todo{
-%Known issues to resolve before submission:\\
-%- Update HARPS fits to use blaze-corrected flux (only AB-type stars use it now)\\
-%- In the "alpha Cen A" sample, sometimes they accidentally observed alpha Cen B and didn't realise! Maybe better to use a different star?\\
-%- Add citations\\
-%- Update red text.
-%}
-
 \section{Introduction}\label{sec:intro}
 
-Continuum normalization\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances, because those quantities are inferred from line strengths that are measured \emph{relative} to the continuum.
-\ajw{Would be good to say that the more fundemental reason is that modelling the continuum (mostly intrumental but also stellar) is hard, and the separation is feasible for most stars.}
+Continuum normalisation\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances.
+Those quantities are inferred from line strengths that are measured \emph{relative} to the continuum, in part because modeling the stellar continuum (and instrument response) correctly is more challenging than modeling the relative flux \citep[e.g.,][]{Wheeler:2023, Wheeler:2024}.
 The practice of continuum rectification is ripe with subjectivity, in part for justified historical reasons.
-Often a consistent procedure is more useful than an abstract \ajw{maybe reword} one, so people tend to stick with what they know, and that subjectivity is inherited by graduate students.
-While there is general agreement in the literature that is important to achieve consistent continuum normalization, there is no apparent consensus on how it should be done.\\
+Often a consistent procedure is preferable, so people tend to stick with what they know, and that subjectivity is inherited by graduate students.
+While there is general agreement in the literature that is important to achieve consistent continuum normalisation \citep{Blanco-Cuaresma:2014}, there is no apparent consensus on how it should be done.\\
 
-The literature describes a variety of rectification strategies used to achieve different goals.
-Classical spectroscopists might make some coarse estimate of the continuum for the spectrum, and locally refine that estimate for every absorption line of interest.
+The literature describes a variety of rectification strategies used to achieve different goals \citep{Ness:2015, Casey:2016, Xu:2019, Cretignier:2020}.
+Classical spectroscopists might make some coarse estimate of the continuum for the spectrum, and locally refine that estimate for every absorption line of interest \citep[e.g.,][]{Casey:2014}.
 This process is often done by hand,\footnote{Albeit they are often experienced hands.
-See \citet{Bensby:2014}: \quote{`\emph{more than 300\,000 equivalent widths were measured by (the first author's right) hand}'}.} although some graphical user interface tools exist to streamline the process and improve reproducibility.
+See \citet{Bensby:2014}: `\emph{more than 300\,000 equivalent widths were measured by (the first author's right) hand}'.} although some graphical user interface tools exist to streamline the process and improve reproducibility.
 This approach usually restricts the spectroscopist to only measuring strengths of what are believed to be isolated lines: atomic transitions that are not contaminated by neighbouring lines or molecular bands.
-In contrast, industrial spectroscopists often have a different objective: one where they solely for a \emph{consistent} continuum normalization procedure.
+In contrast, industrial spectroscopists often have a different objective: one where they solely for a \emph{consistent} continuum normalisation procedure \citep[e.g.,][]{Ness:2015, Casey:2016}.
 In those situations the process does not have claim to deliver the true continuum, but it should deliver a \emph{pseudo}-continuum that is consistent for stars of similar stellar parameters and signal-to-noise (S/N) ratios.\\
@@
 Fitting the continuum correctly requires you to know where there is absorption.
-But knowing where there is line absorption requires you to (at least) know the stellar parameters \ajw{confused about this claim}.
+But to know where to expect line absorption, we must have some estimate of the stellar parameters (among other things, e.g., line lists).
 Without knowing the stellar parameters -- or having a good model for line absorption -- spectroscopists have developed various bespoke methods.
@@
 An alternative is simply to mask all pixels except a carefully selected set of so-called `continuum pixels'.
-This is painstaking work to perform manually, but can be done by iteratively training data-driven models, or building masks from approximate line strengths computed from atomic properties.
+This is painstaking work to perform manually, but can be done by iteratively training data-driven models \citep{Ness:2015}, or building masks from approximate line strengths computed from atomic properties.
 In either scenario, the set of continuum pixels is only valid for stars of a similar type and metallicity.
 In many cases there are simply \emph{no} continuum pixels (e.g., near a molecular absorption band in an M-dwarf).
-Some graphical tools exist to enable practictioners to select pseudo-continuum points to `anchor' some continuum fit (e.g., a spline, or through a rolling-pin analogy) which are useful if the practictioner wants full control over the continuum fit, or if the models are particularly poor at predicting continuum pixels.\\
+Some graphical tools exist to enable practictioners to select pseudo-continuum points to `anchor' some continuum fit \citep{Xu:2019,Cretignier:2020} which are useful if the practictioner wants full control over the continuum fit, or if the models are particularly poor at predicting continuum pixels.\\
 
 We might simply consider what is the best continuum for every possible theoretical spectrum under comparison.
-When there are thousands of model spectra to compare with, then for each spectrum we can take the ratio of the data and the model, apply some averaging filter (e.g., moving mean or median) and take that as the pseudo-continuum.
-These kinds of approaches work remarkably well, despite the requirement of a (potentially very large) set of sufficiently similar model spectra, and the added computational expense.\\
+When there are thousands of model spectra to compare with, then for each spectrum we can take the ratio of the data and the model, apply some averaging filter (e.g., moving mean or median) and take that as the pseudo-continuum \citep[e.g.,][]{aspcap}.
+These kinds of approaches work remarkably well, despite the requirement of a (potentially very large) set of sufficiently similar model spectra, which themselves are good descriptions for the data, and the added computational expense.\\
 
@@
 In Section~\ref{sec:methods} we describe a family of methods to address these problems.
-The methods we describe uses non-negative matrix factorization (NMF) to approximate line absorption and telluric transmission.
+The methods we describe uses non-negative matrix factorisation \citep[NMF;][]{Lee_Seung:2000} to approximate line absorption and telluric transmission.
 NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements.
-This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
+This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other dimensionality reduction techniques \citep[however see, for example, ][]{Blanton:2007}.
 We describe data (Section~\ref{sec:experiments}) and experiments using this model, where the results are presented in Section~\ref{sec:results}.
@@
 \section{Methods}\label{sec:methods}
-\ajw{Weinberg does a thing where he has a table of every variable and what it is at the beginning of the methods.  It makes things much easier to skim, and I think it would work well here.}
+%\ajw{Weinberg does a thing where he has a table of every variable and what it is at the beginning of the methods.  It makes things much easier to skim, and I think it would work well here.}
 
-We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., molecular or atomic line absorption) \ajw{I think stellar lines   would be a better name for this, the relationship between actual absorption and lines is kinda subtle }; a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{In this work the `continuum' always refers to the joint continuum-instrument response.
+We will assume a forward model that includes three components: one that represents continuum-normalised stellar lines (e.g., molecular or atomic line absorption); a second to describe telluric lines; and a third to represent the smooth continuum.\footnote{In this work the `continuum' always refers to the joint continuum-instrument response.
 In many spectrographs these are different, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast.
-In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and handling other nuisances.\\
+In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation and handling other nuisances.\\
 
-% TODO: list explicit assumptions with hogg in person
+% put in a table of nomenclature
 
@@
 We assume the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
-The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
+The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar line model $f(\lambda_i; \vecalpha)$, the telluric line model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
 \begin{align}\label{eq:y}
@@
 where the components $f(...)$, $g(...)$, and $h(...)$ are defined below.
-Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
-\ajw{you should probably say that C is the (diagonal) cov matrix}
+Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the diagonal covariance matrix at the $i$-th pixel
 \begin{eqnarray}
@@
 \end{eqnarray}
-\noindent{}which is computed by taking a swiftly \ajw{?} Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
+\noindent{}which is computed by taking a Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
 \begin{align}
@@
     \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
-\end{align}\\
+\end{align}
+We now turn to defining each model component in detail.
+The stellar line model $f(\lambda_i;\vecalpha)$ represents the rectified stellar flux at wavelength $\lambda_i$ given parameters $\vecalpha$.
+We construct the stellar line model $f(\lambda_i;\vecalpha)$ from a set of $N$ continuum-normalised theoretical spectra (each with $D$ fluxes) using non-negative matrix factorisation (NMF).
+The theoretical spectra used to construct the stellar line model do not have to have the same wavelength sampling and instrument line spread profile as the data, but at inference time there is a need to interpolate (or evaluate) the stellar line model to the $P$ observed wavelengths.\\
 
-We now turn to defining each model component in detail.
-The stellar absorption model $f(\lambda_i;\vecalpha)$ predicts \ajw{represents?
-corresponds to?} the rectified stellar line absorption at wavelength $\lambda_i$ given parameters $\vecalpha$.
-We construct the line absorption model $f(\lambda_i;\vecalpha)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorization (NMF).
-The theoretical spectra used to construct the line absorption model do not have to have the same wavelength sampling and instrument line spread profile as the data, but at inference time there is a need to interpolate (or evaluate) the line absorption model to the $P$ observed wavelengths.\\
+\begin{table*}
+    \centering
+    \caption{Nomenclature}
+    \begin{tabular}{cl}
+        \hline
+        Symbol & Description \\
+        \hline
+        $P$ & Number of pixels in an observed spectrum\\
+        $N$ & Number of continuum-normalised theoretical stellar spectra used in NMF\\
+        $D$ & Number of pixels in continuum-normalised theoretical stellar spectra\\
+        $K$ & Number of basis components used in the stellar line model\\
+        $L$ & Number of basis components used in the telluric line model\\
+        $\vec{S}$ & Set of continuum-normalised theoretical stellar spectra $[N\,\times\,D]$\\
+        $\vecW_\star$ & Non-negative basis weights for stellar line model $[K\,\times\,N]$\\
+        $\vec{F}$ & Non-negative basis vectors for stellar line model $[K\,\times\,D]$\\
+        $\vecalpha$ & Non-negative weights for the stellar line model found at test time $[K]$\\        
+        $\vecbeta$ & Non-negative weights for the telluric line model found at test time $[L]$\\
+        $\vecW_\earth$ & Non-negative basis weights for telluric line model \\%$[L\,\times\,]
+        $\vecG$ & Non-negative basis vectors for the telluric line model\\ % L, E?
+        $\vecbeta$ & Non-negative weights for telluric line model \\
+        $\vecH$ & Design matrix for continuum consisting of Fourier bases \\
+        $\vecgamma$ & Log-continuum coefficients found at test time\\
+        $\mathbf{\lambda}$ & Wavelengths \\ % [P]
+        $\mathbf{y}$ & Fluxes \\
+        $\mathbf{\sigma_y}$ & Uncertainty in fluxes \\
+        $\mathbf{Y}$ & Log transformed flux $\log{y}$ \\
+        $\mathbf{C}$ & Diagonal covariance matrix of log-transformed flux \\
+        $f(\lambda_i; \vecalpha)$ & Stellar line model \\
+        $g(\lambda_i; \vecbeta)$ & Telluric line model \\
+        $h(\lambda_i; \vecgamma)$ & Continuum-instrument response model \\
+        \hline
+    \end{tabular}
+    \label{tab:nomenclature}
+\end{table*}
 
@@
 However, a small transformation to this matrix makes it extremely sparse.
-Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
+Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorise the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}.
-The factorization of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative.
+where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorising telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}.
+The factorisation of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative.
 The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\
 
-The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
-In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages.
+The factorisation of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
+In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages \citep[e.g.,][]{scikit-learn}.
 We found very similar looking basis vectors if we initialized from random values, but it took more training iterations.
-We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, \todo{until reaching a plateau in reconstruction error, at which point we would iterate further using multiplicative updates.} The multiplicative update rules are appealing because they are convex, do not require many choices, and they can be efficiently iterated on a graphics card.\\
+We found that factorising $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates \citep{Hsieh}.
+However, the multiplicative update rules are appealing because they are convex \citep{Lee_Seung:2000}, can handle missing or noisy data without requiring many choices \citep{Blanton:2007}, and they can be efficiently iterated in-memory on a graphics card.\\
 
@@
 
-We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model.
+We used the BOSZ spectral grid \citep{Bohlin:2017} to construct basis vectors for a stellar line absorption model.
 This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \texttt{ATLAS9} code with \project{MARCS} spherical and plane-parallel models atmospheres.
 While the grid does include spectra for hotter stars computed with ATLAS atmospheres, we restricted ourselves to MARCS model atmospheres for this work.
-The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum.
+The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument \citep{2003Msngr.114...20M}, and importantly it includes the theoretical continuum for each spectrum.
 We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$, truncated the convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements.
-The matrix $\vec{S}$ includes all \todo{X} convolved BOSZ spectra, with \todo{N} pixels each, which we randomly shuffled at every iteration during factorization.
-We factorized this matrix into $K = 16$ basis vectors by coordinate descent for \todo{10,000} steps, \todo{and then X steps using multiplicative updates}.
-After training, the Frobenius loss, also known as the L2 reconstruction loss
-\begin{eqnarray}
-    L_2 = \norm{{\left(-\log{\vec{S}}\right) - \vec{W}_\star\transpose\vec{F}}}^2
-\end{eqnarray}
-\noindent{}was found to be \todo{X}.
-Figure~\ref{fig:schematic} shows some example basis vectors computed by this factorization, where it is somewhat striking that the basis vectors are identifiable by astrophysical mechanism, since no stellar parameters (or astrophysics of any kind) are included in the factorization process.
+The matrix $\vec{S}$ includes all BOSZ spectra, which we randomly shuffled at every iteration during factorisation.
+We factorised this matrix into $K = 16$ basis vectors by coordinate descent for {10,000} steps.\\
+% After training, the Frobenius loss, also known as the L2 reconstruction loss
+%\begin{eqnarray}
+%    L_2 = \norm{{\left(-\log{\vec{S}}\right) - \vec{W}_\star\transpose\vec{F}}}^2
+%\end{eqnarray}
+%\noindent{}was found to be \todo{X}.
+Figure~\ref{fig:schematic} shows some example basis vectors computed by this factorisation, where it is somewhat striking that the basis vectors are identifiable by astrophysical mechanism, since no stellar parameters (or astrophysics of any kind) are included in the factorisation process.
 For example, one basis vector just includes hydrogen lines.
@@
     \includegraphics*[width=\textwidth]{harps-basis-vectors.pdf}
-    \caption{A schematic showing a small wavelength range of the basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra.\label{fig:schematic}}
+    \caption{A schematic showing a small wavelength range of the basis vectors computed by non-negative matrix factorisation from a grid of theoretical stellar spectra.\label{fig:schematic}}
 \end{figure*}
 
-\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line absorption function $\log{f(\lambda_i;\vecalpha)}$, which follows from equations~\ref{eq:log_y} and \ref{eq:nmf},
+\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line function $\log{f(\lambda_i;\vecalpha)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
 \begin{align}
@@
 where $\vecalpha \in [0, \infty)$ is a column vector of $K$ \emph{stellar basis weights} to be solved at inference time.
-Here, $\vecalpha$ is analogous to a column in $\vecW_\star$: it represents the $K$ weights needed to reconstruct the stellar absorption from the $K$ basis vectors $\vec{F}$.
+Here, $\vecalpha$ is analogous to a column in $\vecW_\star$: it represents the $K$ weights needed to reconstruct the rectified stellar flux from the $K$ basis vectors $\vec{F}$.
 Note that because $\vecalpha$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $h(...)$ to represent the smooth continuum-instrument response.
@@
 
-Telluric transmission can be included in a similar way to stellar absorption.
-We used a grid of model telluric transmission spectra computed for the \texttt{PypeIt} project.
-These spectra are computed using the Line-by-Line Radiative transfer Model for an array of observatory locations (latitude, longitude) and atmospheric temperatures, pressures, humidity, and airmass.
+Telluric lines can be included in a similar way to stellar lines.
+We used a grid of model telluric transmission spectra computed for the \texttt{PypeIt} project \citep{pypeit}.
+These spectra are computed using the Line-by-Line Radiative transfer Model \citep{lblrm} for an array of observatory locations (latitude, longitude) and atmospheric temperatures, pressures, humidity, and airmass.
 The original grid of telluric transmission spectra was not available to us.
 Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, where the grid is compressed using Principal Component Analysis and an inverse hyperbolic sine transformation.
-We reconstructed the original grid from these approximations, and factorized the negative log of theoretical telluric absorption
+We reconstructed the original grid from these approximations, and factorised the negative log of theoretical telluric absorption
 \begin{eqnarray}
@@
 \noindent{}into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$ using $L = 4$ basis vectors.
-We took \todo{10,000} iterations using coordinate descent, at which point the reconstruction loss was \todo{X}.
+We made {10,000} iterations using coordinate descent.
 With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
@@
 %We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}).
-Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\
+Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric transmission.\\
 
@@
 
-\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model in equation \ref{eq:log_y} for the log-transformed data $\vec{Y}$,
+\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model in Equation \ref{eq:log_y} for the log-transformed data $\vec{Y}$,
 \begin{equation}
@@
 \begin{eqnarray}
-    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y}
+    \vec{\hat{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y}
 \end{eqnarray}
@@
 There are a few ways to solve this system subject to the boundary constraints, but the choice is not very important because the problem is convex.
-For simplicity we use the truncated reflective trust region algorithm where we stop optimizing once the relative tolerance has improved by less than $10^{-16}$.
+For simplicity we use the truncated reflective trust region algorithm where we stop optimizing once the relative improvement has reached machine precision.
 %In later sections we describe some situations where we may want to penalize values of $\vecalpha$.
@@
 We set the first $K \times K$ elements of $\vec{\Lambda}$ to $\mathrm{Cov}{\left(\vec{W}\vec{W}\transpose\right)}$ and set all other elements of $\vec{\Lambda}$ to zero.
-We can scale the strength of this feature weighting by a scalar $\lambda$ (\todo{nomenclature overload of lambda}) and solve for the parameters $\vec{\tilde{X}}$ with
+We can scale the strength of this feature weighting by a scalar $\lambda$ (\todo{nomenclature overload of lambda}) and solve for the parameters $\vec{\hat{X}}$ with
 %\begin{eqnarray}
-%    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A} + \lambda\vec{\Lambda})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} %\quad .
+%    \vec{\hat{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A} + \lambda\vec{\Lambda})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y} %\quad .
 %\end{eqnarray}\\
-After solving for $\vec{\tilde{X}}$, the (un-transformed) quantities can be easily computed:
+After solving for $\vec{\hat{X}}$, the (un-transformed) quantities can be easily computed:
 \begin{eqnarray}
-    \mathrm{rectified~stellar~flux}, \quad &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\alpha}}\right)} \\
-    \mathrm{telluric~transmission},\quad  &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\tilde{\beta}}\right)} \\
-    \mathrm{continuum}, \quad &h& = \exp{\left(\vec{H}\boldsymbol{\tilde{\gamma}}\right)}
+    \mathrm{rectified~stellar~flux}, \quad &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\hat{\alpha}}\right)} \\
+    \mathrm{telluric~transmission},\quad  &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\hat{\beta}}\right)} \\
+    \mathrm{continuum}, \quad &h& = \exp{\left(\vec{H}\boldsymbol{\hat{\gamma}}\right)}
 \end{eqnarray}
@@
 We assess the model using two experiments with real data.
-The first is to test the model on spectra across a wide range of spectral types, so that we can evaluate whether a constrained linear absorption model is an effective representation for typical stars.
-The second experiment involves fitting the model to spectra of the same star, taken across many different S/N ratios, to test how consistent the rectified spectra are as a function of S/N ratio.\\
+In Experiment 1 we test the model on spectra across a wide range of spectral types so that we can evaluate whether a constrained linear absorption model is an effective representation for typical stars.
+In Experiment 2 we fit the model to repeat observations of the same star, taken across many different S/N ratios, to test how consistent the rectified spectra are as a function of S/N ratio.\\
 
-We queried a catalog of \eso/\harps\ observations to select a high-resolution echelle spectrum of each spectral type from B- to M-type.
+We queried a catalog of \eso/\harps\ \citep{2003Msngr.114...20M} observations to select a high-resolution echelle spectrum of each spectral type from M- to B-type for Experiment 1.
 This selection is not intended to be complete, but is meant to demonstrate the effectiveness of a single linear model for fitting stars across a range of stellar parameters and evolutionary states.
-For each example star we downloaded a pseudo-random exposure, the details of which are shown in Table~\ref{tab:stars}.
+For each example star we downloaded a random exposure.
 No explicit cut was made to ensure a minimum S/N ratio, but in practice most exposures are of high quality.
 The B- and A-type example stars show significant rotational broadening.
-For these stars, we adopt the $v\sin{i}$ values reported by \citet{Someone} and convolve the stellar basis vectors before inference.
+For these stars, we adopt the $v\sin{i}$ values reported by \citet{Barbieri:2023} and convolve the stellar basis vectors before inference.
 In Section~\ref{sec:discussion} we discuss situations when radial velocity or rotational broadening measurements are not available.\\
@@
 These are extracted one-dimensional unrectified spectra of each echelle order, with the wavelength solution represented by a low order polynomial.
-In early experiments we fit the spectra in the \texttt{e2ds} data products directly, an example of which can be seen in Figure~\ref{fig:harps-alf-cen-a}.
-However, for some exposures we noticed systematic residuals that could be explained by a 2D artefact in the image plane (e.g., a ghost or reflection).
+In early experiments we fit the spectra in the \texttt{e2ds} data products directly, an example of which can be seen in Figure~\ref{fig:example-echelle}.
+However, for HD 186543 (an A7 star) and HD 45418 (B4) we noticed systematic residuals that could be explained by a 2D artefact in the image plane (e.g., a ghost or reflection).
 While we expect that flat-fielding should have already removed these kinds of artefacts, it did prompt us to explore using the nightly blaze calibration files.
-Using the calibration blaze files mitigated much of the visible structure, and expectedly \ajw{``as expected'', maybe} provided spectra with a simpler, flatter shape.
-For these reasons, in all experiments except for the one shown in Figure~\ref{fig:harps-alf-cen-a}, we chose to fit blaze-corrected echelle orders.\\
+Using the calibration blaze files mitigated much of the visible structure, providing spectra with a simpler, flatter shape.
+For this reason, for HD 186543 and HD 45418 we fit blaze-corrected echelle orders.\\
 
 When fitting each exposure we took the computed Barycentric Earth radial velocity and the radial velocity measured by the \harps\ data reduction pipeline to shift the stellar absorption basis vectors $\vec{F}$ to the observed frame.
-In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citep{Someone}.
-Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and \todo{9} Fourier modes per echelle order, totalling \todo{668} model parameters per spectrum.\\
+In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citep{Trifonov:2020}.
+Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and {9} Fourier modes per echelle order, totalling {668} model parameters per spectrum.\\
 
-We used the same model setup for the second experiment, where we required many spectra taken of the same star at low and high (e.g., 10 to 200) S/N ratios.
+We used the same model setup for the Experiment 2.
+In this experiment we collected many spectra taken of the same star at low and high (e.g., 10 to 200) S/N ratios.
 The lower bound on S/N ratio might be unrealistic since spectroscopists prefer measuring stellar parameters and chemical abundances from higher quality spectra, but the bound is necessary to evaluate performance at low S/N ratios.
@@
 When sorted by number of exposures in descending order, $\alpha$-Centauri A was the first star that met our requirements in repeated S/N sampling.
-We constructed bins of S/N from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum (Table~\ref{tab:obs-alf-cen-a}) for every S/N ratio bin interval to provide an even sampling of spectra at low and high S/N ratios.
+We constructed bins of S/N from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum for every S/N ratio bin interval to provide an even sampling of spectra at low and high S/N ratios.
 We fit our model to each spectrum, treating it as totally indepdendent from all others.\\
@@
     \includegraphics[width=\textwidth, height=\textwidth]{../code/harps_examples.pdf}
-    \caption{Continuum-rectified \harps\ spectra of example spectral types from B to M.
-The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission.
-\todo{Only BAF-type stars used blaze-corrected spectra.
-Need to download blazes for others and re-run.} \ajw{consider a lower line weight} \ajw{I feel like i'm being thick but I don't get why these ever go above 1.
-Also, are you sure that the clam continuum is where the theoretical continuum is for the M1 star?}}\label{fig:harps-examples}
+    \caption{Continuum-rectified \harps\ spectra of example spectral types from M to B.
+The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission.}\label{fig:harps-examples}
 \end{figure*}
@@
 \section{Results} \label{sec:results}
-
-We independently fit all \eso/\harps\ spectra referenced in Tables~\ref{tab:harps-examples} and \ref{tab:alf-cen-a}.
-We found it takes a few core-seconds to fit each exposure.
+%We found it takes a few core-seconds to fit each exposure with $\sim{}3\times{}10^6$ data points and 668 model parameters.
 The software implementation released with this paper provides for each echelle order: the predicted flux; the continuum; the rectified model flux; and the telluric transmission.
@@
 
-The first defined experiment involves fitting stars of a variety of spectral types.
+Experiment 1 involves fitting stars of a variety of spectral types.
 The rectified spectra for selected example stars in this experiment are shown in Figure~\ref{fig:harps-examples}, where stars are ordered and coloured by their spectral type listed in \project{SIMBAD}.
 From these rectified spectra it is clear that the model performs well on a variety of stellar types and evolutionary stages.
-HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16{,}750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorized using NMF.
+HD 45418, a B4-type star, is the hottest star in our sample  \citep[$T_\mathrm{eff} = 16{,}750\,K$ according to ][]{2015MNRAS.454...28M}, and far exceeds the upper temperature bound (8,000\,K) of the synthetic grid that we factorised using NMF.
 The fit to HD 45418 is quite good, despite this star being well outside the model grid.
-All other example stars shown in Figure~\ref{fig:harps-examples} have literature stellar parameters that place them well within the bounds of the factorized model grid.
-Most stars observed with \eso/\harps\ are predictably \ajw{drop ``predictably'' maybe} on the main-sequence, since they are more frequently subject to radial velocity exoplanet searches than evolved stars.
+All other example stars shown in Figure~\ref{fig:harps-examples} have literature stellar parameters that place them well within the bounds of the factorised model grid.
+Most stars observed with \eso/\harps\ are on the main-sequence, since they are more frequently subject to radial velocity exoplanet searches than evolved stars.
 HD 203050 (K2) is the only red giant branch in our sample.
@@
 There is good agreement between the model fits and data for example stars of different spectral types.
-In detail there are mis-matches in line absorption strength, or mis-matches in the form of missing lines (in either the model or the data).
+In detail there are mis-matches in line strength, or mis-matches in the form of missing lines (in either the model or the data).
 We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both.
@@
 
-The second defined experiment tests how consistently we can predict continuum for the same star, as a function of S/N ratio.
-Given an imperfect (or no) model for stellar spectra, noisy signals can mimic stellar absorption, which causes the continuum to become systematically biased at low S/N ratios.
-We fit \todo{X} spectra of $\alpha$-Centauri A, which are sampled uniformly in S/N bins from 10 to 200 pixel$^{-1}$.
+Experiment 2 tests how consistently we can predict continuum for the same star as a function of S/N ratio.
+Given an imperfect model (or no model) for stellar spectra, noisy signals can mimic stellar absorption, which causes the continuum to become systematically biased at low S/N ratios.
+We fit repeat spectra of $\alpha$-Centauri A, which are sampled uniformly in S/N bins from 10 to 200 pixel$^{-1}$.
 This experiment forces us to select a summary statistic to evaluate the consistency of our continuum rectification.
-We adopt the 95th percentile in rectified flux as a statistic to compare to normalisation methods described in the literature, but we prefer to view how the \emph{distribution} of normalized flux values change as a function of S/N ratio.
-Both properties are shown in Figure~\ref{fig:repeatability-with-snr}, where the line indicates the 95th percentile of normalized flux, and the shaded area shows the distribution of normalized flux values.
+We adopt the 95th percentile in rectified flux as a statistic to compare to normalisation methods described in the literature, but we prefer to view how the \emph{distribution} of normalised flux values change as a function of S/N ratio.
+Both properties are shown in Figure~\ref{fig:repeatability-with-snr}, where the line indicates the 95th percentile of normalised flux, and the shaded area shows the distribution of normalised flux values.
 Above S/N $\sim$ 30, both indicators are extremely stable and do not vary significantly with S/N ratio.
-This is notable, as most spectra acquired for the purpose of stellar parameters and chemical abundances will aim for S/N $>$ 30, but spectra below S/N $<$ 50 are routinely described as being problematic to rectify consistently.
+This is notable, as most spectra acquired for the purpose of stellar parameters and chemical abundances will aim for S/N $>$ 30, but spectra below S/N $<$ 50 are routinely described as being problematic to rectify consistently \citep{Cretignier:2020}.
 From this figure it highlights that we can achieve consistent continuum normalisation whether the star has S/N $\sim$ 30 or 300.
-\todo{This experiment was done on non-blaze-corrected spectra.
-I either need to re-do it, or update the text accordingly.}\\
+Notably, as stated earlier, in this experiment we did not use blaze-corrected spectra.\\% More consistent results might be achievable using blaze-corrected spectra to mitigate CCD ghosts.\\
 
 We can make quantitative statements on the consistency of continuum rectification for $\alpha$-Centauri A.
-Taking the 95th percentile of normalized flux as a proxy statistic for the normalization consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalized flux is 0.22\%, which is equivalent a spread in normalized flux of 0.0022.
+Taking the 95th percentile of normalised flux as a proxy statistic for the normalisation consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalised flux is 0.22\%, which is equivalent a spread in normalised flux of 0.0022.
 This variation is smaller than the typical flux errors, and only increases marginally when we include lower quality spectra.
-Taking all spectra with S/N $>$ 50, we find the scatter in the 95th percentile of normalized flux to be 0.35\%.
+Taking all spectra with S/N $>$ 50, we find the scatter in the 95th percentile of normalised flux to be 0.35\%.
 With S/N $>$ 30, this rises marginally to 0.46\%.
@@
 The line represents the 95th percentile, where the scatter in this percentile is 0.46\% for spectra with S/N $>$ 30, and 0.22\% for everything with S/N $>$ 100.
-\label{fig:alpha-cen-wrt-snr}}
+\label{fig:repeatability-with-snr}}
 \end{figure}
@@
 We have shown that a constrained linear model is a good choice for modeling stellar spectra, which enables precise estimates of the continuum for a range of stellar types.
-Representing stellar absorption using NMF instead of a some other linear representation is particularly advantageous.
+Representing rectified stellar flux using NMF instead of a some other linear representation is particularly advantageous.
 Non-negativity ensures that stellar basis vectors are strictly additive, and the negative log-transformation we use restricts both the rectified stellar flux and telluric transmission to be between 0 and 1.
-Any data values outside this range can \emph{only} be captured by the continuum basis vectors.
-In some sense the predictive capacity of the combination of NMF basis vectors is nearly orthogonal to a combination of the continuum basis vectors.\ajw{Is this a restatement/interpretation of the fact that the stellar line basis and continuum basis are aproximately orthogonal?}\\
+Any data values outside this range \emph{must} either be captured by the continuum basis vectors, or be noise.
+In some sense thestellar line basis and the continuum basis are approximately orthogonal.\\
 
 NMF is less frequently used in astronomical data analysis than other linear representations.
-A more common linear representation for this kind of problem is Principal Component Analysis (PCA), which yields an orthogonal decomposition that maximally explains the variance in the data.
+A more common linear representation for this kind of problem is Principal Component Analysis \citep[PCA;][]{pca}, which yields an orthogonal decomposition that maximally explains the variance in the data.
 PCA is made possible by orthogonality, whereas NMF is made possible by the non-negativity constraint.
@@
 
-Another somewhat unexpected advantage to using NMF is that the basis vectors themselves appear to be highly interpretable.
+Another somewhat unexpected advantage of using NMF is that the basis vectors themselves appear to be highly interpretable.
 For example, one basis vector has only strong hydrogen lines in it.
@@
 Another two basis vectors are similar in that they have a forrest of atomic lines, but they vary in the broadness of particularly strong lines (e.g., where one is consistent with pressure broadening in a main-sequence star).
-General wisdom when working with compressed representations is that the basis vectors should generally be treaated as latent (unobservable) vectors, and there is risk associated with trying to physically interpret them.
+General wisdom when working with compressed representations is that the basis vectors should generally be treated as latent (unobservable) vectors, and there is risk associated with trying to physically interpret them.
 We avoid physical interpretability here, but in subsequent work we discuss how the basis weights and vectors can be interpreted (Casey et al., in prep).
-Here we simply reiterate that the stellar parameters used to construct the grid are not used by the NMF process, and the spectra are randomly shuffled during each iteration of factorization, so there is no possible `leakage' of information about effective temperature, surface gravity, metallicity, or other properties.
+Here we simply reiterate that the stellar parameters used to construct the grid are not used by the NMF process, and the spectra are randomly shuffled during each iteration of factorisation, so there is no possible `leakage' of information about effective temperature, surface gravity, metallicity, or other properties.
 Nevertheless, the resultant basis vectors have absorption features that are strongly correlated with these properties.
@@
 A simple example is hydrogen lines.
-\citep{Wheeler} showed that the hydrogen lines in the infrared are incorrectly computed by many spectral synthesis codes.
+\citet{Wheeler:2024} showed that the hydrogen lines in the infrared are incorrectly computed by many spectral synthesis codes.
 When we use existing grids with this problem (e.g., like those used in \texttt{FERRE}, as part of \texttt{ASPCAP}), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would find a biased estimate of effective temperature, or a worse fit to the data, or both.
@@
 There is very little we can do about that error mis-match here.
-The second form of error arises in our factorization of the grid.
+The second form of error arises in our factorisation of the grid.
 We could add more basis vectors, train for longer, even include estimated per-pixel model uncertainties, but the reconstruction will never be perfect.
@@
 The speed of our model makes it particularly advantageous for evaluating on large data sets.
-If we assume that the factorization of stellar spectra introduces a small error relative to the usual mis-match between models and data, then a constrained linear model represents a promising avenue for identifying regions where models and observations disagree the most.
+If we assume that the factorisation of stellar spectra introduces a small error relative to the usual mis-match between models and data, then a constrained linear model represents a promising avenue for identifying regions where models and observations disagree the most.
 The simplest path for doing this might be to look at the total squared residuals as a function of pixel across many stars, which would immediately indicate the areas where the model does not have capacity to predict the data.
@@
 The model is trained from synthetic spectra, so it is already a close approximation to what stellar spectra look like, and doing data-driven updates would be an excellent way to handle severe data-model mis-matches (e.g., large molecular features which are present in data but we lack line lists to accuratley model).
-There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we can restrict ourselves to only updating a single basis vector.
-\ajw{a reader might be puzzled by the claim that you can evaluate the goodness of the model spectra given the discussion above about how it's fine if the model spectra are a bit shit.}\\
+There is a lot of flexibility when updating the basis vectors using multiplicative update rules \citep{Blanton:2007}: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), or we can restrict ourselves to only updating a single basis vector.
+\\
+%We have demonstrated the effectiveness of a constrained linear absorption model for continuum normalisation across a range of spectral types.
 
-%We have demonstrated the effectiveness of a constrained linear absorption model for continuum normalization across a range of spectral types.
-
-There are not many literature metrics to choose from when evaluating performance of continuum normalization.
+There are not many literature metrics to choose from when evaluating performance of continuum normalisation.
 We have qualitatively shown the distribution of rectified flux values for $\alpha$-Centauri A, and quantitatively estimated the precision from the scatter in the 95th percentile of rectified flux values.
 These statistics are computed using all wavelengths, but it is well-known that continuum is easier to model at redder optical wavelengths because there are fewer absorption features.
-One excellent examination of continuum normalization consistency is that of \citet{RASSINE}, where they report consistent continuum normalization at the level of 2\% based on applications to extremely high quality ($S/N > 1000$) \eso/\harps\ spectra of $\alpha$-Centauri B stacked over consecutive nights, and 0.29\% for a comparably high quality spectrum of the Sun.
-Their statistic is defined as the 2$\sigma$ of $1-f_i$, where $f_i$ is the $i$th normalized flux value at the anchor point locations used to build the continuum.
+One excellent examination of continuum normalisation consistency is that of \citet{Cretignier:2020}, where they report consistent continuum normalisation with \texttt{RASSINE} at the level of 2\% based on applications to extremely high quality ($S/N > 1000$) \eso/\harps\ spectra of $\alpha$-Centauri B stacked over consecutive nights, and 0.29\% for a comparably high quality spectrum of the Sun.
+Their statistic is defined as the 2$\sigma$ of $1-f_i$, where $f_i$ is the $i$th normalised flux value at the anchor point locations used to build the continuum.
 The anchor points are chosen by the user, which makes for a less direct comparison with our statistic, but for these purposes we will assume that the continuum anchor points are perfectly chosen and truely represent the continuum in the star.
-Given these assumptions, for comparison purposes we take \texttt{RASSINE} as providing 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.15\% consitency for comparable Solar spectra.
-If they restrict their statistic to wavelengths redder than 4500\,\AA, their precision for $\alpha$-Centauri B improves to 0.6\%.
-In comparison, for spectra we use that is 1/10th the S/N ratio, we find our method yields consistent continuum estimates at the level of 0.22\%.\\
+Given these assumptions, for comparison purposes we take that an expert user of \texttt{RASSINE} \citep{Cretignier:2020} can achieve 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.15\% consitency for comparable Solar spectra.
+If \citet{Cretignier:2020} restrict their statistic to wavelengths redder than 4500\,\AA, their precision for $\alpha$-Centauri B improves to 0.6\%.
+In comparison, for spectra we use that is 1/10th the S/N ratio, we find our method yields consistent continuum estimates at the level of 0.22\%, highlighting that our approach provides best-in-class results for consistency in continuum normalisation.\\
 
@@
 
-Arguments on the consistency of continuum normalization are precision-based arguments, which says nothing about accuracy.
+Arguments on the consistency of continuum normalisation are precision-based arguments, which says nothing about accuracy.
 That is unfortunate for us because any pseudo-continuum method explicitly only needs to make arguments about precision: a pseudo-continuum is not intended to be accurate!
@@
 We took both of these quantities as truth, but in fact we could fit for them simultaneously.
-A radial velocity shift and a convolution are both linear operators, in that they can be incorporated in our model without breaking our strict linearity constraint.
-We would have to change how we solve the system: we could no longer solve the system directly by unconstrained least-squares, or with the truncated reflective region algorithm, but we would have to solve it using linear operators that output the dot product and left adjoint transpose dot product.
-Alternatively, given the speed of our model, one could imagine a grid in radial velocities (e.g., in $\pm1\,000\,\mathrm{km}\,s^{-1}$), where at each radial velocity we solve for $\vec{\tilde{X}}$ and then take the radial velocity with the lowest $\chi^2$.
-This is a robust method that has been used for decades.
-The rotational broadening can be estimated from the cross-correlation in the same way.
+A radial velocity shift and a convolution are both linear operators, in that they can be incorporated in our model without forcing the model to become strongly non-linear.
+We would have to change how we solve the system: we could no longer solve the system directly by unconstrained least-squares, or with the truncated reflective region algorithm, but we would have to solve it using linear operators.
+Alternatively, given the speed of our model, one could imagine a grid in radial velocities (e.g., in $\pm1\,000\,\mathrm{km}\,\mathrm{s}^{-1}$), where at each radial velocity we solve for $\vec{\hat{X}}$ and then take the radial velocity with the lowest $\chi^2$.
+This is a robust method.
+The rotational broadening could be estimated from the cross-correlation in the same way.
 For these reasons, the method we describe here is readily extensible to situations where the radial velocity and the rotational broadening is not known.\\
@@
 
-We introduce a novel approach to modelling stellar spectra as a linear combination of highly constrained absorption vectors, which we show is sufficient for simultaneously fitting stellar absorption, telluric transmission, and the joint continuum-instrument response.
+We introduce a novel approach to modelling stellar spectra as a linear combination of highly constrained absorption vectors, which we show is sufficient for simultaneously fitting stellar lines, telluric transmission, and the joint continuum-instrument response.
 A constrained linear absorption model has many advantagous properties over alternative data-driven methods or machine learning techniques used in stellar spectroscopy.
@@
 An extension of this work could also solve for other linear nuisances (e.g., radial velocity, rotational broadening).
-Finally, an unexpected result of this work is that the basis vectors found from factorizing model rectified spectra appear to be highly interpretable, which is promising for their use for interpretable constrained linear absorption models.\\
+Finally, an unexpected result of this work is that the basis vectors found from factorising model rectified spectra appear to be highly interpretable, which is promising for their use for interpretable constrained linear absorption models.\\
 
-%Basis vectors learned by factorizing rectified grids of stellar spectra separate out into vectors that include only hydrogen lines, atomic lines, or prominent molecular features.
+%Basis vectors learned by factorising rectified grids of stellar spectra separate out into vectors that include only hydrogen lines, atomic lines, or prominent molecular features.
 A specific example of interpretability is shown where ony basis vector incldues only strong hydrogen lines, and at inference time we find the corresponding weight for this vector increases smoothly with spectral type (a proxy for effective temperature).
@@
 \texttt{scipy} \citep{scipy};
-\texttt{scikit-learn} \citep{scikit_learn}.
+\texttt{scikit-learn} \citep{scikit-learn}.
 
 \paragraph{Acknowledgements}
-If they do not become co-authors, then we should be sure to thank
-    Jon Holztman (NMSU),
-    Rory Smith (Monash),
-    Adam Wheeler (CCA),
-    Andrew Saydjari (Harvard),
-    Guy Stringfellow,
-    Lily Zhao (Chicago),
-    Julianne Dalcanton (CCA),
-    David W. Hogg (NYU),
-    Megan Bedell (CCA), and
+A.~R.~C. thanks
+    Rory Smith (Monash University),
+    Guy Stringfellow (Boulder),
+    Julianne Dalcanton (flatiron Institute), and
     Michael Blanton (NYU).
+A.~R.~C. thanks New York University for their hospitality.
 
-\todo{Acknowledge Utah and OzStar resources. Acknowledge NYU for accommodating my body as it withered without free Flatiron lunch.}
-% include bibliography
 \bibliographystyle{aasjournal}
-%\bibliography{bibliography}
+\bibliography{bibliography}
 

### 2025-12-18 paper/response.txt
resub

# Reviewer Report Responses

We thank the referee for their review, and we apologize for the lateness of our reply. We have addressed all queries put forward by the reviewer.

---

## Abstract

> **Line 27**: "Most unphysical spectra": I did not find in the paper an explanation of this sentence. Particularly, I am curious about the exceptions in which - despite the constraints - unphysical spectra are obtained.

Clarified that the constraints prohibit rectified flux values exceeding unity.

> **Lines 29-30**: For the sake of completeness the description of the model should mention that there is a training required, and potentially some arbitrarity in meta parameters (e.g. K).

The description does describe how "stellar absorption and telluric transmission are each modelled by factorising a grid of rectified theoretical spectra", and we have added a note about the number of basis components.

> **Line 31**: Since radial velocities and broadening have not been demonstrated in the paper, this should be removed from the brackets. The authors are welcome to mention that an extension of the model to include those is possible.

Changed to indicate these are extensions: "and is readily extensible to radial velocity and rotational broadening."

> **Line 35**: To clarify this result already at abstract level I suggest a slightly expanded phrasing, e.g. "consistent across time to 0.3% at S/N ~ 100, and to better than 0.4% at S/N ~ 30".

Changed to: "consistent across time to 0.2% at S/N ~ 100, and to better than 0.5% at S/N ~ 30."

---

## Introduction

> **Paragraph 49**: The difference between a "classical" and an "industrial" spectroscopist is not obvious to me. I suggest to explicitly state which are the different kinds of applications that have different requirements for continuum normalization (e.g. applications to individual objects to estimate accurate atmospheric parameters, extraction of a large number of normalized spectra from a survey, ...).

Expanded to explicitly describe the two use cases: (1) detailed studies of individual objects for accurate atmospheric parameters/abundances, and (2) large spectroscopic surveys requiring consistent automatic extraction.

---

## Methods

> **Figure 1**: it is not quoted in the methods section, and quoted only in Sec. 4, so it's possibly misplaced.

Added reference to Figure 1 where basis vectors are first introduced.

> **Line 128**: Does bold-face indicate an array? Please specify.

Added: "Throughout this work, bold-face symbols denote vectors or matrices (see Table 1)."

> **Line 130**: I suspect a mistake in the coefficient to the third order Taylor expansion. Should it be 2/6?

There was an error in the Taylor series expansion, but we have also now limited it to second-order approximation, as this is more than necessary for most astronomical applications.

> **Line 130 (separate)**: Separate to previous point: why is it necessary to expand in series to the fourth order? For the average astronomer, the error on log(x) is dx/x (me included, prior to thinking about it).

The reviewer is right: it's not! We have limited it to a second-order approximation.

> **Line 134**: For consistency with the rest, I suggest boldfacing lambda in Eq. (3).

Done.

> **M 138**: The discussion on the precision vs accuracy of the continuum is important in this paper. I therefore find it necessary to spend a few words about what is meant by "continuum-normalized theoretical spectra". This might seem obvious at first glance, but the wings of a broad line (e.g. hydrogen lines) might be considered as part of the continuum when interested in the abundance of a different species (e.g. iron). The normalization of theoretical spectra is supposed to be "perfect", but is it truly an unambiguous choice? How exactly is the theoretical normalization performed by the authors?

Added detailed explanation: the rectified flux from synthesis codes is the emergent flux divided by the theoretical continuum (flux in absence of line opacity). This normalization is unambiguous and performed by the synthesis code itself.

> **Lines 141-143**: It seems that the line spread function is accounted for in this work by convolving the model with a kernel (line 187). This sentence seems to state that this is not necessary, which creates some confusion.

Clarified: the training grid need not match data resolution because basis vectors can be convolved to instrument resolution at inference time, but we have noted that it can be helpful if they are pre-convolved.

> **Line 147**: The reader might benefit from an additional explanation of why it is convenient to move to a sparse matrix.

Added explanation: NMF efficiently exploits sparsity, allowing the factorisation to focus on learning absorption structure rather than modelling near-unity continuum values.

> **Lines 148-150**: Is this transformation the NMF itself? Some of the reasons for this choice of transformation should be listed for clarity (or different parts of the paper that discuss them pointed at).

Added three reasons: (1) converts multiplicative to additive model; (2) ensures non-negativity for NMF requirements; (3) produces sparse matrices.

> **M 154**: How is the number of basis components (K) chosen? Is this choice critical to the success of the method?

Added explanation: we select by monitoring reconstruction loss on validation set. Also added specific justification in Experiments section.

> **Lines 165-167**: Here or elsewhere it would be interesting to have an estimate of the computational time required for the training with the chosen grid of models (how many are they?), and of execution time to fit an individual spectrum.

Added: training takes ~30 minutes on single CPU for millions of spectra, inference takes a few core-seconds per HARPS spectrum.

> **Line 171**: Is there any requirement in terms of spectral resolution of the models? For typical applications in the spectral fits with which I am familiar, a resolution of about 10 times higher than the instrumentation resolving power is required.

Not strictly. Added: The native resolution of the theoretical grid should be at least comparable to the target instrument resolution.

> **M 190**: To my knowledge, the ATLAS9 code that lies at the basis of the BOSZ models is primarily designed to treat photospheric conditions in LTE, where no emission lines are expected. I am therefore wondering why do the authors have some flux values exceeding 1 after normalizing by the theoretical continuum.

Clarified: these occur rarely due to numerical precision in synthesis codes or interpolation, not physical emission (since it is an LTE code).

> **Line 195**: I do not find the derivation of this equation trivial, despite finding it rather intuitive. Do the authors mean that this mathematically follows from Eq. (3) and (4)?

Added explicit derivation from Equation 4 to Equation 5, which leads on to 6-7, and then to Eqs 8-10.

> **Line 202**: The implication that non-negativity of F and alpha allows h to properly fit for the continuum is not obvious, please expand a bit on it.

Added mathematical explanation: since F >= 0 and alpha >= 0, we have f <= 1, so any flux above the stellar model must be captured by continuum.

> **Line 248**: I suggest to define the quantities with hat for additional clarity.

Added explanation that hats denote optimised parameters.

---

## Experiments

> **Line 277**: I suggest to include a table that summarizes The S/N for each star in the selected sample.

Added Table 2 with star names, spectral types, approximate S/N ratios, and notes.

> **Line 279**: To clarify: is the convolution performed after constructing the basis? Would there be a difference in constructing the basis prior to applying the convolution?

Clarified: For these stars, convolution is applied to the already-learned basis vectors, before solving for the non-negative weights.

> **M 295**: I suggest to test the pipeline on different spectra for an A and B stars that are unaffected by ghosts/reflections. Since these spectral types end up being treated differently from the others to have a good result, this additional test would be a clean way to show that this is indeed due to a specificity of the two datasets chosen and not due to a difficulty of the pipeline with, e.g., fast rotating hot stars.

We considered this at length before submission. Ultimately, the artefacts are observation-specific rather than dependent on particular spectral types, and through discussions with colleagues they commented on how it was useful to know that the method could be applied to spectra that were both blaze-corrected or not. For this reason, we have kept these examples and explicitly noted in the text "We verified that the blaze correction is specific to these particular observations and not a general limitation for hot stars by successfully fitting other A- and B-type stars from the archive without blaze correction; the artefacts appear to be observation-specific rather than spectral-type-dependent, and we leave these examples in to demonstrate the applicability of our method to both blaze-corrected and blaze-uncorrected spectra."

> **M 305**: Additional details on the choice of these meta-parameters would be really appreciated (see also comment on line 154).

Added justification for $K$ of 16, $L$ of 4, and $F$ of 9.

> **Line 309**: There is a degree of redundancy between the sentence "In this experiment ..." and the more precise description at Lin 316.

Removed redundant sentence.

---

## Results

> **M Figure 2**: A few insets showing zooms of Figure 2 would be appreciated. It would be interesting to see both regions where the model fit is particularly successful, and where it is less successful (e.g. following the discussion at lines 349 - 351, and line 358).

Added insets to Figure 2 and updated the caption accordingly.

> **M 325 - 343**: I tend to agree with the optimism of the manuscript, but it would be useful to provide a few more quantitative arguments about the performance. Looking at Figure 2 there is a significant number of pixels that exceeds the dashed line, which I assume is the rectified continuum. Is this consistent with the photon noise distribution + emission lines? To my eye, it seems that especially for the colder spectral types there might be errors of a few % (estimated from the max excess of flux beyond 1). This is also visible in Fig. 2, with many data-pixels exceeding all model components. This precision might be sufficient for the scope of the tool and might be best-in-class, but I urge the authors to attempt their best at quantifying this. For instance, at line 340 - 341 it is mentioned that the theoretical expectation for the continuum model matches the retrieved one. This should quantified (e.g. squared residuals) and shown with a specific figure, and possibly extended to all spectral types covered.

Added some discussion about this point: If the theoretical models were perfect descriptions of stellar spectra, and our NMF approximation was also perfect, then any pixels exceeding the dashed line in Figure 3 should be consistent with photon noise and emission lines. However, neither is true. This motivates that we should add flexibility to our stellar absorption model by increasing $K$ or building the stellar absorption model from observed spectral templates.

> **Lines 352-353**: This is not obvious for the M1 to F6 stars blue-ward of ~4500 Angstroms. Please elaborate on the argument about this specific region.

Added clarification that this region presents a greater challenge due to high density of absorption features.

> **Line 357**: Why not use ESO SkyCalc?

We have added an explanation that this was largely due to convenience rather than for optimal purposes. In practice we found ESO SkyCalc very difficult to install and use. We would have found it necessary to compute a large grid of telluric spectra with different combinations of atmospheric contributions, which was cumbersome. High-resolution templates would improve our fits. The method is capable of using any kind of telluric grid.

> **Line 362**: Could you show exactly what you mean? This would be clear from an analogous figure to Fig. 1 but for tellurics and continuum basis, which I suggest to add.

Added: The telluric basis vectors describe narrow atmospheric features while the continuum basis vectors only includes smooth functions.

> **M 370 - 381 and Figure 4**: The metric presented in Figure 4 require some clarifications.
1) There is no line visible in Fig. 4.
2) What is meant by "normalised flux"? I did not find a definition.
3) From Sec. 3 I expected a total of 20 spectra (one per bin in average S/N per pixel), hence potentially 20 points in the summary statistic presented in Fig. 4, but I see that the x-axis of the Figure is much more densely populated. This makes me think that S/N per pixel has a different meaning here compared to Sec. 3, which should be clarified. What is the x-axis exactly?
To try and illustrate my confusion, my interpretation would be that in Figure 4 the x-axis represents the SNR in all individual pixels in the spectra considered, but I can not reconcile this with the fact that there should be stellar lines in many pixels which do not appear in Figure 4. Does this only represent the value of the continuum extracted by the NMF method, or in other words Eq. (16)?

We apologize for the confusion. Originally we used 20 spectra but found the sampling too course. Instead we took all spectra with S/N < 200 and we show the results from all spectra (with a S/N bin size of 1: up to 200 spectra if all S/N were sampled). The visible line meant to describe the density in the 2D histogram.

We have extensively revised the text and clarified the S/N bin sampling.

---

## Discussion

> **M 399-402**: This discussion does not consider that synthetic stellar spectra are not perfect representations of the real stellar spectra. To some extent, residuals between models and data can be seen as "noise", but on the other hand these residuals could be captured by the "continuum" (e.g. broad-lines, dense forests of small molecular lines for later types). To some extent, I do worry that limiting the flexibility of the stellar models introduces stronger biases in the continuum to compensate for imprecise theoretical stellar spectra. What is the authors opinion about this? I think this should be briefly addressed here.

The reviewer's intuition here is correct, in that if you are simultaneously modelling things C (continuum) and S (star), and you care most about getting C correct, then you should put *all* your model capacity into S to make it as flexible as possible. In this sense, the applications we present here are more of a lower limit for the kind of performance one might expect with a NMF-based representation of line absorption. This is reflected in our updated discussion about the choice of the number of basis vectors.

We have also added some discussion about this, and limitations about how non-negativity constraint limits the stellar flexibility (e.g., continuum can only explain flux above the stellar predictions, not deficits)

> **Paragraph 404**: Paragraph 404: To my knowledge, one can always choose to use a limited number of PCA components, so I do not fully follow the argument. With a limited number of PCA components, the limitation raised by the manuscript would partially be lifted. While this truncation would be arbitrary, there seems to be a degree of arbitrarily also with NMF, which has to select the number of basis vectors. Could the authors comment on this?

Added rebuttal: even with few PCA components, the model can produce unphysical flux because weights have no sign constraint. NMF's non-negativity guarantees f <= 1 regardless of component count.

> **M 420-452**: his section of the discussion seems to place NMF somewhere in-between purely data-driven methods (e.g. with an un-interpretable basis vector as could be the case for PCA), and a fully forward-modeling approach that is less flexible than NMF. However, it is currently a little difficult to follow.
- In the first of the two paragraphs, it claims that interpretability of the basis vectors is an advantage. Since the interpretability is not fully demonstrated in this paper, but only qualitatively discussed, I suggest to tone down the first sentence of the paragraph, and refer to Casey et al., in prep.
- The first sentence of the second paragraph is a confusing as it seems to contradict that interpretability is an advantage as discussed in the preceding paragraph.
- In addition, there is no guarantee that a good fit to the data equals unbiased inference of stellar parameters, especially if the model fit is very flexible as in the case of NMF. While I don't think that the authors mean to imply this, the discussion that the comparison with existing grids might lead to biased estimates of stellar parameters is somewhat misleading since this is the case also for the NMF approach. I suggest that the authors clarify this point and add a word of caution about how to best interpret the excellent fits provided by NMF.

Toned down "highly interpretable" to "qualitatively interpretable". We added some explanation of middle-ground position regarding data-driven models and traditional forward models. We also added caution that excellent fits don't guarantee accurate parameters and shouldn't be interpreted as validation of underlying physics.

> **Lines 468-470**: Is the advantage of the constrained linear model that it would allow to evaluate the squared residuals from observations for a large number of stars, unlike traditional methods?

Added: unlike traditional forward modelling (minutes to hours per spectrum), our approach fits in seconds, enabling systematic analysis of millions of stars.

> **M 490-506**: It seems to me that this comparison with Cretignier et al. (2020) is inaccurate. The way I understand Table 2 in Cretignier et al. (2020) is that:
-The first two line entries estimate the accuracy of the continuum of a single high S/N spectrum when they compare said continuum to a theoretical prediction at the anchor points f_i.
- The rest of the line entries estimate the same quantity when - rather than a single spectrum - a whole time series is fit.
Instead, I do not see any estimation of the consistency of quantities across different spectra, if by consistency we mean "reproducibility" across time. Since the deviation between the models and the RASSINE retrieved continuum might be a systematic, the performance in terms of consistency of RASSINE might actually be better than 2%. Could the authors comment on this?

Substantially revised: clarified that Cretignier metrics represent accuracy (comparison to theoretical models), not consistency; noted that their reproducibility may be better than accuracy figures suggest; acknowledged our metrics measure different quantities.

> **Lines 521-522**: The author themselves argue that the continuum obtained with NMF is only as accurate as the (inevitably somewhat inaccurate) models used. I therefore do not find it obvious that the NMF-retrieved continuum would be more accurate than a pseudo-continuum. Could the authors explain better why they think the continuum from the NMF model is better? Is it perhaps more reproducible, or interpretable, or do they have strong reason to believe that it is indeed more accurate?

Added nuanced discussion: our estimate is preferable because (1) physically motivated, (2) reproducible without tuning, (3) works where no continuum pixels exist. However, acknowledged that expert pseudo-continuum may be more accurate in regions with poor models.

---

## Conclusions

Revised to: mention training step and hyperparameter K; add non-negativity guarantee; include quantitative consistency results; tone down interpretability claim with reference to future work.

### 2025-12-18 paper/main.tex
resub

@@
 
-\documentclass[modern]{aastex631}
+\documentclass[modern,linenumbers]{aastex631}
 \usepackage[utf8]{inputenc}
@@
 %\newcommand{\chosentitle}{Constrained linear absorption models for stellar spectroscopy}
-\newcommand{\chosentitle}{A constrained linear model for continuum normalisation of stellar spectra}
+\newcommand{\chosentitle}{A constrained linear model for continuum normalization of stellar spectra}
 
@@
 
+% Revision macros for tracked changes
+\usepackage[normalem]{ulem} % for \sout strikethrough - must load before defining \removed
+%\renewcommand{\added}[1]{\textcolor{blue}{#1}}
+\renewcommand{\added}[1]{\edit1{#1}}
+\newcommand{\removed}[1]{}%\textcolor{red}{\sout{#1}}}
+
 \sloppy\sloppypar\raggedbottom\frenchspacing
@@
 Inferring stellar parameters and chemical abundances by forward modelling stellar spectra usually requires a spectral synthesis code, or an emulator constructed from a curated training set.
-In these situations continuum normalisation is often implemented as a pre-processing step that is independent of stellar parameters.
+In these situations continuum normalization is often implemented as a pre-processing step that is independent of stellar parameters.
 This leads to results that are biased, or inconsistent across signal-to-noise ratios.
@@
 Here we describe a constrained \emph{linear} model that can fit stellar absorption, telluric transmission, the joint continuum-instrument response.
-Stellar absorption and telluric transmission are each modelled by factorising a grid of rectified theoretical spectra into two non-negative matrices.
+Stellar absorption and telluric transmission are each modelled by factorising a grid of rectified theoretical spectra into two non-negative matrices \added{with a chosen number of basis components}.
 This model characterises all possible spectra in many fewer parameters than comparable data-driven models.
-The non-negativity constraint ensures basis vectors are strictly additive, which prohibits most unphysical spectra such that we can distinguish normalised spectra from the joint instrument-continuum response.
-The model requires no initial guess or tuning, and the linearity ensures that inference is convex, stable, and fast.
-This model allows us to reliably fit nuisances (e.g., tellurics, continuum, radial velocity, broadening), without any prior knowledge about the fundamental stellar properties.
+The non-negativity constraint ensures basis vectors are strictly additive, which \added{limits rectified flux to less than or equal to unity,} \removed{prohibits most unphysical spectra} such that we can distinguish normalized spectra from the joint instrument-continuum response.
+The model requires no initial guess \removed{or tuning}, and the linearity ensures that inference is convex, stable, and fast.
+This model allows us to reliably fit nuisances (e.g., tellurics, continuum\removed{, radial velocity, broadening}), \added{and is readily extensible to radial velocity and rotational broadening,} without any prior knowledge about the fundamental stellar properties.
 We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of BAFGKM-type stars.
-With repeat observations of $\alpha$-Centauri A we present results that are best in class: consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.\vspace{3em}\\
+With repeat observations of $\alpha$-Centauri A we present results that are best in class: \added{consistent across time to} \removed{consistent to} 0.2\% at S/N $\sim$ 100, and \added{to} better than 0.5\% at S/N $\sim$ 30.\vspace{3em}\\
 \end{abstract}
@@
 
-Continuum normalisation\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances.
+Continuum normalization\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances.
 Those quantities are inferred from line strengths that are measured \emph{relative} to the continuum, in part because modeling the stellar continuum (and instrument response) correctly is more challenging than modeling the relative flux \citep[e.g.,][]{Wheeler:2023, Wheeler:2024}.
@@
 Often a consistent procedure is preferable, so people tend to stick with what they know, and that subjectivity is inherited by graduate students.
-While there is general agreement in the literature that is important to achieve consistent continuum normalisation \citep{Blanco-Cuaresma:2014}, there is no apparent consensus on how it should be done.\\
+While there is general agreement in the literature that is important to achieve consistent continuum normalization \citep{Blanco-Cuaresma:2014}, there is no apparent consensus on how it should be done.\\
 
 The literature describes a variety of rectification strategies used to achieve different goals \citep{Ness:2015, Casey:2016, Xu:2019, Cretignier:2020}.
-Classical spectroscopists might make some coarse estimate of the continuum for the spectrum, and locally refine that estimate for every absorption line of interest \citep[e.g.,][]{Casey:2014}.
+\added{In detailed studies of individual objects (e.g., to derive accurate atmospheric parameters or chemical abundances from equivalent widths) a} \removed{Classical} spectroscopist might make some coarse estimate of the continuum for the spectrum, and locally refine that estimate for every absorption line of interest \citep[e.g.,][]{Casey:2014}.
 This process is often done by hand,\footnote{Albeit they are often experienced hands.
@@
 This approach usually restricts the spectroscopist to only measuring strengths of what are believed to be isolated lines: atomic transitions that are not contaminated by neighbouring lines or molecular bands.
-In contrast, industrial spectroscopists often have a different objective: one where they solely for a \emph{consistent} continuum normalisation procedure \citep[e.g.,][]{Ness:2015, Casey:2016}.
+In contrast, \added{large spectroscopic surveys require automatic extraction of thousands of normalized spectra, where the primary objective is a} \removed{industrial spectroscopists often have a different objective: one where they solely for a} \emph{consistent} continuum normalization procedure \citep[e.g.,][]{Ness:2015, Casey:2016}.
 In those situations the process does not have claim to deliver the true continuum, but it should deliver a \emph{pseudo}-continuum that is consistent for stars of similar stellar parameters and signal-to-noise (S/N) ratios.\\
@@
 
-We will assume a forward model that includes three components: one that represents continuum-normalised stellar lines (e.g., molecular or atomic line absorption); a second to describe telluric lines; and a third to represent the smooth continuum.\footnote{In this work the `continuum' always refers to the joint continuum-instrument response.
+We will assume a forward model that includes three components: one that represents continuum-normalized stellar lines (e.g., molecular or atomic line absorption); a second to describe telluric lines; and a third to represent the smooth continuum.\footnote{In this work the `continuum' always refers to the joint continuum-instrument response.
 In many spectrographs these are different, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast.
-In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation and handling other nuisances.\\
+In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and handling other nuisances.\\
 
@@
 We assume the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$).
-The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar line model $f(\lambda_i; \vecalpha)$, the telluric line model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
+\added{Throughout this work, bold-face symbols denote vectors or matrices (see Table~\ref{tab:nomenclature}).} The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar line model $f(\lambda_i; \vecalpha)$, the telluric line model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
 \begin{align}\label{eq:y}
@@
 Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the diagonal covariance matrix at the $i$-th pixel
+%\begin{eqnarray*}
+%    \color{red}{\vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + - \frac{\sigma_{y,i}^4}{4y_i^4}\right)^2}
+%\end{eqnarray*}
 \begin{eqnarray}
-    \vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + \frac{2\sigma_{y,i}^3}{8y_i^3} - \frac{6\sigma_{y,i}^4}{24y_i^4}\right)^2
+%    \color{blue}{
+        \vec{C}_{ii} \approx \frac{\sigma_{y,i}^2}{y_i^2} + \frac{\sigma_{y,i}^4}{2y_i^4}
+    %}
 \end{eqnarray}
-\noindent{}which is computed by taking a Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
+\noindent{}which is computed by taking a second-order Taylor series expansion. The log-transformation changes the element-wise product of three model components into the convenient summation:
 \begin{align}
     \label{eq:log_y}
-    \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
+    \vec{Y} &= \log{f(\boldsymbol{\lambda}; \vecalpha)} + \log{g(\boldsymbol{\lambda};\vecbeta)} + \log{h(\boldsymbol{\lambda};\vecgamma)} \quad .
 \end{align}
@@
 The stellar line model $f(\lambda_i;\vecalpha)$ represents the rectified stellar flux at wavelength $\lambda_i$ given parameters $\vecalpha$.
-We construct the stellar line model $f(\lambda_i;\vecalpha)$ from a set of $N$ continuum-normalised theoretical spectra (each with $D$ fluxes) using non-negative matrix factorisation (NMF).
-The theoretical spectra used to construct the stellar line model do not have to have the same wavelength sampling and instrument line spread profile as the data, but at inference time there is a need to interpolate (or evaluate) the stellar line model to the $P$ observed wavelengths.\\
+We construct the stellar line model $f(\lambda_i;\vecalpha)$ from a set of $N$ continuum-normalized theoretical spectra (each with $D$ fluxes) using non-negative matrix factorisation (NMF).
+\added{By `continuum-normalized theoretical spectra' we mean the rectified flux provided by spectral synthesis codes, which is the emergent flux divided by the theoretical continuum (i.e., the flux in the absence of line opacity).
+While the wings of broad lines (e.g., hydrogen) might sometimes be treated as pseudo-continuum in observational work, our theoretical normalization treats all line opacity consistently.}
+The theoretical spectra used to construct the stellar line model do not have to have the same wavelength sampling and instrument line spread profile as the data \added{because they can be convolved at inference time, but it is helpful if they are pre-convolved approximately to the target instrument resolution.}\\
 
@@
         $P$ & Number of pixels in an observed spectrum\\
-        $N$ & Number of continuum-normalised theoretical stellar spectra used in NMF\\
-        $D$ & Number of pixels in continuum-normalised theoretical stellar spectra\\
+        $N$ & Number of continuum-normalized theoretical stellar spectra used in NMF\\
+        $D$ & Number of pixels in continuum-normalized theoretical stellar spectra\\
         $K$ & Number of basis components used in the stellar line model\\
         $L$ & Number of basis components used in the telluric line model\\
-        $\vec{S}$ & Set of continuum-normalised theoretical stellar spectra $[N\,\times\,D]$\\
+        $F$ & Number of Fourier basis functions used in the continuum-instrument response model\\
+        $\vec{S}$ & Set of continuum-normalized theoretical stellar spectra $[N\,\times\,D]$\\
         $\vecW_\star$ & Non-negative basis weights for stellar line model $[K\,\times\,N]$\\
@@
         $\vecbeta$ & Non-negative weights for the telluric line model found at test time $[L]$\\
-        $\vecW_\earth$ & Non-negative basis weights for telluric line model \\%$[L\,\times\,]
-        $\vecG$ & Non-negative basis vectors for the telluric line model\\ % L, E?
-        $\vecbeta$ & Non-negative weights for telluric line model \\
-        $\vecH$ & Design matrix for continuum consisting of Fourier bases \\
-        $\vecgamma$ & Log-continuum coefficients found at test time\\
-        $\mathbf{\lambda}$ & Wavelengths \\ % [P]
-        $\mathbf{y}$ & Fluxes \\
-        $\mathbf{\sigma_y}$ & Uncertainty in fluxes \\
-        $\mathbf{Y}$ & Log transformed flux $\log{y}$ \\
-        $\mathbf{C}$ & Diagonal covariance matrix of log-transformed flux \\
+        $\vecW_\earth$ & Non-negative basis weights for telluric line model $[L]$ \\
+        $\vecG$ & Non-negative basis vectors for the telluric line model $[K]$ \\
+        $\vecbeta$ & Non-negative weights for telluric line model $[L]$ \\
+        $\vecH$ & Design matrix for continuum consisting of Fourier bases $[P, F]$\\
+        $\vecgamma$ & Log-continuum coefficients found at test time $[F]$ \\
+        $\mathbf{\lambda}$ & Wavelengths $[P]$\\ % [P]
+        $\mathbf{y}$ & Fluxes $[P]$\\
+        $\mathbf{\sigma_y}$ & Uncertainty in fluxes $[P]$\\
+        $\mathbf{Y}$ & Log transformed flux $\log{y}$ $[P]$\\
+        $\mathbf{C}$ & Diagonal covariance matrix of log-transformed flux $[P\,\times\,P]$\\
+        $\hat{\vecalpha}$ & Optimized stellar line model weights $[K]$\\
+        $\hat{\vecbeta}$ & Optimized telluric line model weights $[L]$\\
+        $\hat{\vecgamma}$ & Optimized continuum-instrument response model coefficients $[F]$\\
+        $\hat{\vec{X}}$ & Optimized model parameters (includes $\hat{\vecalpha}$, $\hat{\vecbeta}$, $\hat{\vecgamma}$) $[K + L + F]$\\
         $f(\lambda_i; \vecalpha)$ & Stellar line model \\
@@
 This is a dense matrix: there are no entries of exactly zero, with many entries near 1 (no absorption).
-However, a small transformation to this matrix makes it extremely sparse.
-Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorise the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
+However, a small transformation to this matrix makes it extremely sparse\added{, which is beneficial because NMF efficiently exploits sparsity.
+If most pixels have near-zero absorption, then the factorisation can focus on learning the structure of absorption features rather than modelling near-unity continuum values}.
+Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but \added{we chose the negative logarithm transformation because: (1) it converts the multiplicative model (Equation~\ref{eq:y}) into an additive one (Equation~\ref{eq:log_y}); (2) it ensures all transformed values are non-negative (since rectified flux $\leq 1$), satisfying NMF requirements; and (3) it produces sparse matrices because most pixels have flux near unity.
+We} \removed{for many reasons we chose to} factorise the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
 \begin{eqnarray}
     \label{eq:nmf}
-    -\log\left({\vec{S}}\right) \approx \vec{W}_\star\transpose\vec{F}
+    -\log{\vec{S}} \approx \vec{W}_\star\transpose\vec{F}
 \end{eqnarray}
-where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorising telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}.
-The factorisation of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative.
-The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\
+where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorising telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}\added{, examples of which are shown in Figure~\ref{fig:schematic}}.
+The factorisation of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log{\vec{S}}$, $\vec{W}_\star$, and $\vec{F}$ be non-negative.
+The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.
+\added{The choice of $K$ is arbitrary yet easy in practice: too few components yield poor reconstructions, too high would become expensive to factorize.
+We discuss the choice more in Section~\ref{sec:discussion}, but if we care about getting the continuum correct then we should make the line absorption model as flexible as practical (high $K$).
+In practice, as few as $K \sim 4$ or 8 works reasonably for low-resolution spectra, with more necessary for higher resolution spectra.
+In our experiments we select $K$ by trialling multiples of 2 and evaluating the consistency of continuum normalization on a set of random spectra; see Section~\ref{sec:experiments} for specific values used.}\\
 
-The factorisation of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
-In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages \citep[e.g.,][]{scikit-learn}.
+The factorisation of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages.
+In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log{\vec{S}}$: this is the default behaviour in many packages \citep[e.g.,][]{scikit-learn}.
 We found very similar looking basis vectors if we initialized from random values, but it took more training iterations.
@@
 This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of A- to M-type stars was also sufficiently flexible to model many O-type stars, but there was structured residuals in the wings of hydrogen lines.
-For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\
+For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.} \added{In the applications here the training is performed once using hundreds of thousands of spectra for 10,000 steps, which takes approximately 30 minutes on a single CPU core.
+At inference time, fitting a single HARPS spectrum ($\sim$3$\times$10$^6$ pixels) takes a few core-seconds.}\\
 
@@
 This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \texttt{ATLAS9} code with \project{MARCS} spherical and plane-parallel models atmospheres.
-While the grid does include spectra for hotter stars computed with ATLAS atmospheres, we restricted ourselves to MARCS model atmospheres for this work.
+\added{The native resolution of the theoretical grid should be at least comparable to the target instrument resolution.} While the grid does include spectra for hotter stars computed with ATLAS atmospheres, we restricted ourselves to MARCS model atmospheres for this work.
 The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument \citep{2003Msngr.114...20M}, and importantly it includes the theoretical continuum for each spectrum.
-We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$, truncated the convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements.
+We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$\added{; this convolution is applied to each spectrum before NMF training, ensuring the basis vectors are learned at the target resolution.
+In experiments with high rotational broadening (Section~\ref{sec:experiments}) we also apply an rotational convolution to the learned basis vectors}.
+We truncated the instrument-convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 \added{(which occur rarely due to numerical precision in the synthesis code or interpolation, not physical emission)} as they violate our NMF requirements.
 The matrix $\vec{S}$ includes all BOSZ spectra, which we randomly shuffled at every iteration during factorisation.
-We factorised this matrix into $K = 16$ basis vectors by coordinate descent for {10,000} steps.\\
+We factorised this matrix into $K = 16$ basis vectors by coordinate descent for {10,000} steps.
+\added{The choice of $K=16$ stellar basis vectors was arbitrarily determined by starting with $K=2$, testing consistency of continuum-normalization in practice, and doubling $K$ until the performance seemed reasonable.
+We discuss the choice of $K$ in Section~\ref{sec:discussion}, but in short if we care about modelling the continuum correctly then we should but as much capacity as possible into the other things \citep[e.g., stellar absorption by making $K$ very high;][]{Bickel:1998}.}\\
+
 % After training, the Frobenius loss, also known as the L2 reconstruction loss
@@
 
-\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line function $\log{f(\lambda_i;\vecalpha)}$, which follows from Equations~\ref{eq:log_y} and \ref{eq:nmf},
-\begin{align}
-    \log{f(\lambda_i;\vecalpha)} = -\vec{F}\transpose\vecalpha \label{eq:log-f}
-\end{align}
-where $\vecalpha \in [0, \infty)$ is a column vector of $K$ \emph{stellar basis weights} to be solved at inference time.
-Here, $\vecalpha$ is analogous to a column in $\vecW_\star$: it represents the $K$ weights needed to reconstruct the rectified stellar flux from the $K$ basis vectors $\vec{F}$.
+\noindent{}With the non-negative matrix $\vec{F}$ we can now define the logarithm of stellar line function $\log{f(\lambda_i;\vecalpha)}$.
+\added{Since $-\log\vec{S} \approx \vec{W}_\star\transpose\vec{F}$ from Equation~\ref{eq:nmf}, we have $\log\vec{S} \approx -\vec{W}_\star\transpose\vec{F}$, and replacing the training weights $\vec{W}_\star$ with inference-time weights $\vecalpha$ yields the log stellar absorption model to be $\vecalpha\transpose\vec{F}$.
+At inference time there is a transpose necessary to ensure the data are a column vector, leading to}
+\begin{eqnarray}
+    \log{f(\lambda_i;\vecalpha)} &=& -\vec{F}\transpose\vecalpha \label{eq:log-f} \quad .
+\end{eqnarray}
+\noindent{}\removed{where $\vecalpha \in [0, \infty)$ is a column vector of $K$ \emph{stellar basis weights} to be solved at inference time.} Here, $\vecalpha \in [0, \infty)$ is analogous to a column vector of \emph{stellar basis weights} in $\vecW_\star$: it represents the $K$ weights needed to reconstruct the rectified stellar flux from the $K$ basis vectors $\vec{F}$.
 Note that because $\vecalpha$ and $\vec{F}$ are both restricted to have non-negative elements, this severely restricts the flexibility of $f(...)$, leaving $h(...)$ to represent the smooth continuum-instrument response.
-\\
+\added{Specifically, since $\vec{F} \geq 0$ and $\vecalpha \geq 0$, we have $-\vec{F}\transpose\vecalpha \leq 0$, meaning $\log{f} \leq 0$ and thus $f \leq 1$.
+The stellar model can only produce absorption (flux $\leq 1$), never values exceeding unity.
+Any observed flux above the stellar model prediction must therefore be captured by the continuum component $h(...)$.}\\
 
@@
 The original grid of telluric transmission spectra was not available to us.
-Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, where the grid is compressed using Principal Component Analysis and an inverse hyperbolic sine transformation.
+Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, where the grid is compressed using Principal Component Analysis (PCA) and an inverse hyperbolic sine transformation.
 We reconstructed the original grid from these approximations, and factorised the negative log of theoretical telluric absorption
@@
 \end{eqnarray}
-\noindent{}into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$ using $L = 4$ basis vectors.
+\noindent{}into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$ using $L = 4$ basis vectors\added{, the dimensionality of the PCA-compressed grid}.
 We made {10,000} iterations using coordinate descent.
-With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
+With the non-negative matrix $\vec{G}$\added{, like Equation~\ref{eq:log-f},} the function $\log{g\left(\lambda_i;\vecbeta\right)}$ \added{becomes}\removed{is simply}
 \begin{eqnarray}
-    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta
+    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta \label{eq:log-g}
 \end{eqnarray}
@@
 %We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}).
-Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric transmission.\\
+Like $\vecalpha$, $\vecbeta$ is analogous to a column \added{vector} in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric transmission.\\
 
 There are many suitable choices for the logarithm of the continuum-instrument response model $\log{h(\lambda_i;\vecgamma)}$.
-Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
+Here we chose a Fourier basis of \added{$F$} sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs.
 The component $\log{h(\lambda_i;\vecgamma)}$ is expressed compactly as
 \begin{align}
-    \log{h(\lambda_i;\vecgamma)} = \vec{H}\vecgamma
+    \log{h(\lambda_i;\vecgamma)} = \vec{H}\vecgamma \label{eq:log-h}
 \end{align}
@@
 
-\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model in Equation \ref{eq:log_y} for the log-transformed data $\vec{Y}$,
+\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model in Equation \ref{eq:log_y} for the log-transformed data $\vec{Y}$ \added{using Equations~\ref{eq:log-f}, \ref{eq:log-g}, and \ref{eq:log-h}}:
 \begin{equation}
@@
 \end{equation}
-where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
+where the \removed{parameters and design matrices are stacked to construct}
+\added{design matrices are stacked to construct} $\vec{A}$ and \added{the unknown parameters are stacked into} $\vec{X}$:
 \begin{eqnarray}
@@
 For simplicity we use the truncated reflective trust region algorithm where we stop optimizing once the relative improvement has reached machine precision.
+\added{We denote the optimised parameters with hats (e.g., $\boldsymbol{\hat{\alpha}}$, $\boldsymbol{\hat{\beta}}$, $\boldsymbol{\hat{\gamma}}$) to distinguish them from the general parameters.}
 %In later sections we describe some situations where we may want to penalize values of $\vecalpha$.
@@
 No explicit cut was made to ensure a minimum S/N ratio, but in practice most exposures are of high quality.
-The B- and A-type example stars show significant rotational broadening.
-For these stars, we adopt the $v\sin{i}$ values reported by \citet{Barbieri:2023} and convolve the stellar basis vectors before inference.
+\added{Table~\ref{tab:sample} summarises the stars used and their approximate S/N ratios.} The B- and A-type example stars show significant rotational broadening.
+For these stars, we adopt the $v\sin{i}$ values reported by \citet{Barbieri:2023} and convolve the \added{already-learned} stellar basis vectors before inference.
 In Section~\ref{sec:discussion} we discuss situations when radial velocity or rotational broadening measurements are not available.\\
+
+\added{
+\begin{table}
+    \centering
+    \caption{Summary of example stars used in Experiment 1.}
+    \begin{tabular}{lccc}
+        \hline
+        Star & Spectral Type & Approx.\ S/N & Notes \\
+        \hline
+        HD 45418  & B4 & 150 & Blaze-corrected \\
+        HD 186543 & A7 & 200 & Blaze-corrected \\
+        HD 222595 & F6 & 180 & -- \\
+        HD 20794  & G8 & 250 & -- \\
+        HD 203050 & K2 & 120 & Red giant \\
+        HD 85512  & K6 & 100 & -- \\
+        HD 95735  & M1 & 80 & -- \\
+        \hline
+    \end{tabular}
+    \label{tab:sample}
+\end{table}
+}
 
@@
     \includegraphics[width=\textwidth]{example.pdf}
-    \caption{An example fit to a single \eso/\harps\ spectrum of HD~222595 (black), showing the continuum fits per echelle order (dotted), the stellar absorption model (red) and the telluric transmission model (blue), all fit simultaneously with a constrained linear absorption model.
-Near the strong Ca H and K lines it is clear that there are \emph{no} `continuum pixels' in those echelle orders, yet the model provides an excellent estimate of the true continuum.
-\label{fig:example-echelle}}
+    \caption{An example fit to a single \eso/\harps\ spectrum of HD~222595 (black), showing the continuum fits per echelle order \removed{(dotted)}, the stellar absorption model (\removed{red}\added{blue}) and the telluric transmission model (\removed{blue}\added{orange}), all fit simultaneously with a constrained linear absorption model.
+Near the strong Ca H and K lines (see inset) it is clear that there are \emph{no} `continuum pixels' in those echelle orders, yet the model provides an excellent estimate of the true continuum.
+\added{The telluric model used is imperfect (see inset), but sufficient to capture most telluric absorption.}\label{fig:example-echelle}}
 \end{sidewaysfigure}
@@
 Using the calibration blaze files mitigated much of the visible structure, providing spectra with a simpler, flatter shape.
-For this reason, for HD 186543 and HD 45418 we fit blaze-corrected echelle orders.\\
+For this reason, for HD 186543 and HD 45418 we fit blaze-corrected echelle orders.
+\added{We leave these examples in to demonstrate the applicability of our method to both blaze-corrected and blaze-uncorrected spectra.}\\
 
 When fitting each exposure we took the computed Barycentric Earth radial velocity and the radial velocity measured by the \harps\ data reduction pipeline to shift the stellar absorption basis vectors $\vec{F}$ to the observed frame.
-In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citep{Trifonov:2020}.
-Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors, 4 telluric vectors, and {9} Fourier modes per echelle order, totalling {668} model parameters per spectrum.\\
+In situations where no radial velocity was available from the \harps\ data reduction pipeline, we adopted the radial velocity reported by \citet{Trifonov:2020}.
+Each exposure has $\sim3\times10^6$ pixels across 72 echelle orders, which we model with 16 stellar basis vectors and 4 telluric vectors \added{(see earlier justification)}, and {9} Fourier modes per echelle order\added{, which seemed sufficient flexibility to capture the continuum shape}, totalling {668} model parameters per spectrum.\\
+%Similarly, the 4 telluric vectors capture the dominant modes of atmospheric variation.
+The 9 Fourier modes seemed provide sufficient flexibility to capture the blaze function and continuum shape per order without overfitting.
 
-We used the same model setup for the Experiment 2.
-In this experiment we collected many spectra taken of the same star at low and high (e.g., 10 to 200) S/N ratios.
-The lower bound on S/N ratio might be unrealistic since spectroscopists prefer measuring stellar parameters and chemical abundances from higher quality spectra, but the bound is necessary to evaluate performance at low S/N ratios.
+We used the same model setup for \removed{the} Experiment 2\added{, where we collected many spectra of the same star at many (e.g., between 10 to 200) S/N ratios}.
+\removed{In this experiment we collected many spectra taken of the same star at low and high (e.g., 10 to 200) S/N ratios.} The lower bound on S/N ratio might be unrealistic since spectroscopists prefer measuring stellar parameters and chemical abundances from higher quality spectra, but the bound is necessary to evaluate performance at low S/N ratios.
 The \eso/\harps\ catalog revealed that not many stars meet our repeated S/N sampling requirement.
 When sorted by number of exposures in descending order, $\alpha$-Centauri A was the first star that met our requirements in repeated S/N sampling.
-We constructed bins of S/N from 0 to 200 in steps of 10, and selected a random $\alpha$-Centauri A spectrum for every S/N ratio bin interval to provide an even sampling of spectra at low and high S/N ratios.
-We fit our model to each spectrum, treating it as totally indepdendent from all others.\\
+We constructed bins of S/N from 0 to 200 \removed{in steps of 10}, and selected a random $\alpha$-Centauri A spectrum for every S/N ratio bin interval to provide an even sampling of spectra at low and high S/N ratios.
+We fit our model to each spectrum, treating it as totally independent from all others.\\
 
 \begin{figure*}
-    \includegraphics[width=\textwidth, height=\textwidth]{../code/harps_examples.pdf}
+    \includegraphics[width=\textwidth, height=\textwidth]{../src/clam/harps_examples.pdf}
     \caption{Continuum-rectified \harps\ spectra of example spectral types from M to B.
@@
 This is expected; the continuum estimate provided by the model matches theoretical rectified spectra.
-It illustrates that the continuum estimate that we fit simultaneously with stellar absorption does provide a closer estimate to the true continuum rather than a pseudo-continuum.\\
+It illustrates that the continuum estimate that we fit simultaneously with stellar absorption does provide a closer estimate to the true continuum rather than a pseudo-continuum.
+\added{However, this region presents a greater challenge for continuum normalization, as the high density of absorption features makes it difficult to identify true continuum points.}\\
 
@@
 In any case, the model appears to be sufficiently accurate to describe the continuum without substantial bias, so it is not a major concern for our purposes.
-\\
+\added{If the theoretical models were perfect descriptions of stellar spectra, and our NMF approximation was also perfect, then any pixels exceeding the dashed line in Figure~\ref{fig:harps-examples} should be consistent with photon noise and emission lines.
+However, neither is true.
+This motivates that we should add flexibility to our stellar absorption model by increasing $K$ or building the stellar absorption model from observed spectral templates.}\\
 
@@
 This is apparent when closely comparing the fitted model with the data (e.g., if one could zoom in finely to Figure~\ref{fig:example-echelle}), but it is otherwise not obvious.
-The model telluric transmission predicts features that are clearly broader than the data, but the telluric basis vectors are a sufficiently good description for the data that the telluric absorption does not get captured by the continuum basis vectors and adversely affect the fit.\\
+The model telluric transmission predicts features that are clearly broader than the data, but the telluric basis vectors are a sufficiently good description for the data that the telluric absorption does not get captured by the continuum basis vectors and adversely affect the fit.
+\added{In other words, the telluric features describe narrow atmospheric features while the continuum basis vectors describe smooth functions.} \added{Higher-resolution telluric templates (e.g., ESO \texttt{SkyCalc}) could improve the fit in telluric-contaminated regions; our choice of the \texttt{PypeIt} grid was driven by convenience rather than optimality.
+Our approach can readily accommodate any telluric grid.}\\
 
@@
 This experiment forces us to select a summary statistic to evaluate the consistency of our continuum rectification.
-We adopt the 95th percentile in rectified flux as a statistic to compare to normalisation methods described in the literature, but we prefer to view how the \emph{distribution} of normalised flux values change as a function of S/N ratio.
-Both properties are shown in Figure~\ref{fig:repeatability-with-snr}, where the line indicates the 95th percentile of normalised flux, and the shaded area shows the distribution of normalised flux values.
+We adopt the 95th percentile in rectified flux as a statistic to compare to normalization methods described in the literature, but we prefer to view how the \emph{distribution} of normalized flux values change as a function of S/N ratio.
+Both properties are shown in Figure~\ref{fig:repeatability-with-snr}, where the line indicates the 95th percentile of normalized flux, and the shaded area shows the distribution of normalized flux values.
 Above S/N $\sim$ 30, both indicators are extremely stable and do not vary significantly with S/N ratio.
 This is notable, as most spectra acquired for the purpose of stellar parameters and chemical abundances will aim for S/N $>$ 30, but spectra below S/N $<$ 50 are routinely described as being problematic to rectify consistently \citep{Cretignier:2020}.
-From this figure it highlights that we can achieve consistent continuum normalisation whether the star has S/N $\sim$ 30 or 300.
+From this figure it highlights that we can achieve consistent continuum normalization whether the star has S/N $\sim$ 30 or 300.
 Notably, as stated earlier, in this experiment we did not use blaze-corrected spectra.\\% More consistent results might be achievable using blaze-corrected spectra to mitigate CCD ghosts.\\
@@
 We can make quantitative statements on the consistency of continuum rectification for $\alpha$-Centauri A.
-Taking the 95th percentile of normalised flux as a proxy statistic for the normalisation consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalised flux is 0.22\%, which is equivalent a spread in normalised flux of 0.0022.
+Taking the 95th percentile of normalized flux as a proxy statistic for the normalization consistency, we find that above S/N $> 100$, the standard deviation of the 95th percentile of normalized flux is 0.22\%, which is equivalent a spread in normalized flux of 0.0022.
 This variation is smaller than the typical flux errors, and only increases marginally when we include lower quality spectra.
-Taking all spectra with S/N $>$ 50, we find the scatter in the 95th percentile of normalised flux to be 0.35\%.
+Taking all spectra with S/N $>$ 50, we find the scatter in the 95th percentile of normalized flux to be 0.35\%.
 With S/N $>$ 30, this rises marginally to 0.46\%.
@@
     \includegraphics*[width=\textwidth, height=3in]{alpha-cen-repeat-snr.pdf}
-    \caption{Distribution (shaded) of rectified flux of $\alpha$-Centauri A as a function of S/N ratio, with every spectrum fit independently.
-The line represents the 95th percentile, where the scatter in this percentile is 0.46\% for spectra with S/N $>$ 30, and 0.22\% for everything with S/N $>$ 100.
-\label{fig:repeatability-with-snr}}
+    \caption{Distribution (shaded) of rectified flux values of $\alpha$-Centauri A as a function of S/N ratio, with every spectrum fit independently.
+The shaded region line represents distribution of flux values for each spectrum (e.g., each vertical distribution represents all $\sim3\times10^6$ pixels in a single spectrum).
+The blue line indicates the 95th percentile, which has a 1-$\sigma$ spread of 0.46\% for spectra with S/N $>$ 30, and 0.22\% for everything with S/N $>$ 100.\label{fig:repeatability-with-snr}}
 \end{figure}
@@
 This can easily model flux values much larger than 1.
-The situation does not get much better if we include basis vectors for the continuum, because much of that smooth structure can be captured by the PCA spectral model.\\
+The situation does not get much better if we include basis vectors for the continuum, because much of that smooth structure can be captured by the PCA spectral model.
+\added{One might argue that truncating PCA to a limited number of components would ameliorate this issue, but the fundamental problem remains: PCA basis vectors and weights can take any sign, so even with few components the model can produce unphysical flux values.
+While both PCA and NMF require choosing the number of components (an admittedly arbitrary choice), NMF's non-negativity provides a physical constraint that PCA lacks --- it guarantees the stellar model produces only absorption ($f \leq 1$), regardless of how many components are used.}\\
 
-Another somewhat unexpected advantage of using NMF is that the basis vectors themselves appear to be highly interpretable.
+Another \removed{somewhat unexpected} advantage of using NMF is that the basis vectors themselves appear to be \added{qualitatively} \removed{highly} interpretable.
 For example, one basis vector has only strong hydrogen lines in it.
@@
 General wisdom when working with compressed representations is that the basis vectors should generally be treated as latent (unobservable) vectors, and there is risk associated with trying to physically interpret them.
-We avoid physical interpretability here, but in subsequent work we discuss how the basis weights and vectors can be interpreted (Casey et al., in prep).
+We avoid \added{detailed} physical interpretability here, but in subsequent work we discuss how the basis weights and vectors can be interpreted (Casey et al., in prep).
 Here we simply reiterate that the stellar parameters used to construct the grid are not used by the NMF process, and the spectra are randomly shuffled during each iteration of factorisation, so there is no possible `leakage' of information about effective temperature, surface gravity, metallicity, or other properties.
@@
 
-Chosing not to interpret the basis vectors or weights has specific advantages.
-For example, if we were to build some mapping from stellar parameters to basis weights, then one could construct a forward model for the data where the model parameters are the continuum coefficients and the stellar parameters (e.g., $T_\mathrm{eff}$, $\log{g}$), instead of continuum coefficients and basis weights.
+\added{Our model occupies a middle ground between purely data-driven methods and traditional forward modelling.
+The NMF basis is derived from theoretical spectra, encoding physical priors about stellar absorption, but at inference time the basis weights are determined by the data alone without imposing a mapping to stellar parameters.} \removed{Chosing not to interpret the basis vectors or weights has specific advantages.} For example, if we were to build some mapping from stellar parameters to basis weights, then one could construct a forward model for the data where the model parameters are the continuum coefficients and the stellar parameters (e.g., $T_\mathrm{eff}$, $\log{g}$), instead of continuum coefficients and basis weights.
 That model would be far less flexible than the model we present here.
@@
 When we use existing grids with this problem (e.g., like those used in \texttt{FERRE}, as part of \texttt{ASPCAP}), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would find a biased estimate of effective temperature, or a worse fit to the data, or both.
-Instead, by not restricting ourselves to a mapping between stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.\\
+Instead, by not restricting ourselves to a mapping between stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.
+\added{We caution, however, that an excellent fit does not guarantee accurate stellar parameters --- our model is designed for continuum normalization, not parameter inference.
+The flexibility that enables good fits could also mask model deficiencies; users should not interpret small residuals as validation of the underlying physics.}\\
 
@@
 There is very little we can do about that error mis-match here.
-The second form of error arises in our factorisation of the grid.
+\added{However, one advantage from this factorization is that we learn absorption \emph{vectors}, and the weights we fit at test time do not need to represent a set of weights that exist in the training set.
+For example, if TiO absorption was under-predicted by spectral synthesis codes, we could simply infer a slightly higher basis weight for the basis vector(s) that describe TiO absorption.
+A related concern is if there are absorption features that are simply not modelled at all by spectral synthesis codes.
+Such a scenario might be unmodelled molecular features that get absorbed by the continuum component, biasing the normalization.
+The non-negativity constraint limits this effect to a degree: the continuum can only explain flux \emph{above} what the stellar model predicts, not flux deficits.} The second form of error arises in our factorisation of the grid.
 We could add more basis vectors, train for longer, even include estimated per-pixel model uncertainties, but the reconstruction will never be perfect.
@@
 If we assume that the factorisation of stellar spectra introduces a small error relative to the usual mis-match between models and data, then a constrained linear model represents a promising avenue for identifying regions where models and observations disagree the most.
-The simplest path for doing this might be to look at the total squared residuals as a function of pixel across many stars, which would immediately indicate the areas where the model does not have capacity to predict the data.
+\added{Unlike traditional forward modelling where fitting a single spectrum can take minutes to hours, our approach fits a spectrum in seconds, enabling systematic analysis of millions of stars.} The simplest path for doing this might be to look at the total squared residuals as a function of pixel across many stars, which would immediately indicate the areas where the model does not have capacity to predict the data.
 Next steps might be to improve modelling for that region (e.g., through better line lists) or might be to iteratively update the basis vectors in a data-driven way so that they better match the data.
@@
 \\
-%We have demonstrated the effectiveness of a constrained linear absorption model for continuum normalisation across a range of spectral types.
+%We have demonstrated the effectiveness of a constrained linear absorption model for continuum normalization across a range of spectral types.
 
-There are not many literature metrics to choose from when evaluating performance of continuum normalisation.
+There are not many literature metrics to choose from when evaluating performance of continuum normalization.
 We have qualitatively shown the distribution of rectified flux values for $\alpha$-Centauri A, and quantitatively estimated the precision from the scatter in the 95th percentile of rectified flux values.
 These statistics are computed using all wavelengths, but it is well-known that continuum is easier to model at redder optical wavelengths because there are fewer absorption features.
-One excellent examination of continuum normalisation consistency is that of \citet{Cretignier:2020}, where they report consistent continuum normalisation with \texttt{RASSINE} at the level of 2\% based on applications to extremely high quality ($S/N > 1000$) \eso/\harps\ spectra of $\alpha$-Centauri B stacked over consecutive nights, and 0.29\% for a comparably high quality spectrum of the Sun.
-Their statistic is defined as the 2$\sigma$ of $1-f_i$, where $f_i$ is the $i$th normalised flux value at the anchor point locations used to build the continuum.
-The anchor points are chosen by the user, which makes for a less direct comparison with our statistic, but for these purposes we will assume that the continuum anchor points are perfectly chosen and truely represent the continuum in the star.
-Given these assumptions, for comparison purposes we take that an expert user of \texttt{RASSINE} \citep{Cretignier:2020} can achieve 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.15\% consitency for comparable Solar spectra.
-If \citet{Cretignier:2020} restrict their statistic to wavelengths redder than 4500\,\AA, their precision for $\alpha$-Centauri B improves to 0.6\%.
-In comparison, for spectra we use that is 1/10th the S/N ratio, we find our method yields consistent continuum estimates at the level of 0.22\%, highlighting that our approach provides best-in-class results for consistency in continuum normalisation.\\
+One excellent examination of continuum normalization \removed{consistency} is that of \citet{Cretignier:2020}, where they report \removed{consistent} \added{\texttt{RASSINE} provides} continuum normalization \removed{with \texttt{RASSINE}} \added{accurate} at the level of 2\% based on applications to extremely high quality ($S/N > 1000$) \eso/\harps\ spectra of $\alpha$-Centauri B stacked over consecutive nights, and 0.29\% for a comparably high quality spectrum of the Sun.
+\added{We note that these metrics from \citet{Cretignier:2020} represent \emph{accuracy} (comparison to theoretical models at anchor points), not \emph{consistency} across different observations.
+Since any systematic offset between model and \texttt{RASSINE}-derived continuum would be constant across time, the reproducibility of \texttt{RASSINE} may be better than these accuracy figures suggest.} Their statistic is defined as the 2$\sigma$ of $1-f_i$, where $f_i$ is the $i$th normalized flux value at the anchor point locations used to build the continuum.
+The anchor points are chosen by the user, which makes for a less direct comparison with our statistic\added{, as we evaluate consistency across different observations rather than accuracy against models}.
+\removed{but for these purposes we will assume that the continuum anchor points are perfectly chosen and truely represent the continuum in the star.
+Given these assumptions, for comparison purposes we take that an expert user of \texttt{RASSINE} \citep{Cretignier:2020} can achieve 1\% consistency in continuum estimates for S/N $> 1000$ spectra of $\alpha$-Centauri B, and 0.15\% consitency for comparable Solar spectra.} If \citet{Cretignier:2020} restrict their statistic to wavelengths redder than 4500\,\AA, their \added{accuracy} \removed{precision} for $\alpha$-Centauri B improves to 0.6\%.
+\added{Our metric measures a different quantity --- the reproducibility of the 95th percentile of rectified flux across independent observations --- but the comparison illustrates that our method achieves sub-percent consistency even at S/N ratios an order of magnitude lower than the stacked \texttt{RASSINE} spectra.}\\
 
@@
 
-Arguments on the consistency of continuum normalisation are precision-based arguments, which says nothing about accuracy.
+Arguments on the consistency of continuum normalization are precision-based arguments, which says nothing about accuracy.
 That is unfortunate for us because any pseudo-continuum method explicitly only needs to make arguments about precision: a pseudo-continuum is not intended to be accurate!
@@
 But M-dwarf stars are not standard candles.
-Put simply, measuring accuracy of continuum is hard, and we don't try it, but we think our continuum estimate is better than a pseudo-continum.\\
+Put simply, measuring accuracy of continuum is hard, and we don't try it\added{.
+We argue our estimate is preferable to a pseudo-continuum for several reasons: (1) it is physically motivated, using theoretical spectra that encode knowledge of where absorption features occur; (2) it is reproducible across S/N ratios without parameter tuning; and (3) in regions with no true continuum pixels (e.g., blue wavelengths for cool stars), our method still provides a meaningful estimate while pseudo-continuum methods must either extrapolate or fail.
+However, we acknowledge that in regions where models are poor (e.g., molecular bands), a well-chosen pseudo-continuum from an expert practitioner might be more accurate than our model-based estimate}.\removed{, but we think our continuum estimate is better than a pseudo-continum.}\\
 
@@
 We introduce a novel approach to modelling stellar spectra as a linear combination of highly constrained absorption vectors, which we show is sufficient for simultaneously fitting stellar lines, telluric transmission, and the joint continuum-instrument response.
-A constrained linear absorption model has many advantagous properties over alternative data-driven methods or machine learning techniques used in stellar spectroscopy.
-The model structure restricts basis vectors to be strictly additive, ensuring separability between the basis vectors for absorption and continuum.
-The model requires no initial guess or prior of the stellar parameters, and the linearity ensures that inference is robust, stable, and fast.
-An extension of this work could also solve for other linear nuisances (e.g., radial velocity, rotational broadening).
-Finally, an unexpected result of this work is that the basis vectors found from factorising model rectified spectra appear to be highly interpretable, which is promising for their use for interpretable constrained linear absorption models.\\
+\added{A one-time training step factorises a grid of theoretical spectra to produce the basis vectors, with the number of components $K$ as the primary hyperparameter.} A constrained linear absorption model has many advantageous properties over alternative data-driven methods or machine learning techniques used in stellar spectroscopy.
+The model structure restricts basis vectors to be strictly additive, ensuring separability between the basis vectors for absorption and continuum\added{, with the non-negativity constraint guaranteeing that the rectified stellar flux is restricted to $f \leq 1$}.
+The model requires no initial guess or prior of the stellar parameters \added{at inference time}, and the linearity ensures that inference is robust, stable, and fast.
+\added{We demonstrate consistency in continuum normalization at the 0.2\% level for S/N $>$ 100 spectra and better than 0.5\% at S/N $\sim$ 30.} An extension of this work could also solve for other linear nuisances (e.g., radial velocity, rotational broadening).
+\removed{Finally, an unexpected result of this work is that the basis vectors found from factorising model rectified spectra appear to be highly interpretable, which is promising for their use for interpretable constrained linear absorption models.} \added{The basis vectors appear qualitatively interpretable, which we explore further in subsequent work (Casey et al., in prep).}\\
 
@@
     Guy Stringfellow (Boulder),
-    Julianne Dalcanton (flatiron Institute), and
+    Julianne Dalcanton (Flatiron Institute), and
     Michael Blanton (NYU).
